### 2017-05-12 traittemplate.Rmd
Rmarkdown interpreter of csv file to create html

---
title: "Ecological Traitdata Standard"
author: "Florian D. Schneider, Nadja Simons, Caterina Penone, Andreas Ostrowski"
date: "12 Mai 2017"
output: html_document
---

**This is the documentation of the Ecological Traitdata Standard (ETS), version 0.2**

This defined vocabulary aims at providing all essential columns for raw data of functional trait measurements for ecological research.

The column names are partly aligning with names from Darwin Core Standard and Extensions (Names compliant with DWC are referenced thus; the full Darwin Core can be found here: http://rs.tdwg.org/dwc/terms/index.htm)

The template is ordered into five different sections for column names, as well as a vocabulary for metadata information of particular importance for trait data.

1. Trait measurement data
2. Resolution of data / information on specimen
3. Data source / information on measurement
4. Measurement 
5. Location

Except for the first section, all columns are optional. Only the mandatory fields are required to produce valid trait-datasets. Some mandatory fields will be assisted by the R-Script. The columns labelled 'highly recommended' serve a better comparability of data across datasets.

This list will also be available as

- a csv/txt/xlsx table file which will be provided for download on BExIS including full documentation in the metadata (for manual use, this may be uploaded to BExIs as soon it's ready), and
- a machine readable template glossary (xml file or csv with additional flags and markers) which serves as input reference for software tools. (can we provide a simple API?)

```{r, results = 'asis', echo = FALSE}
library(data.table)
library(knitr)
library(pander)
glossary <- read.csv("template_glossary.csv")

columnNames <- glossary$columnName
sections <- glossary$section
sectionNames <- levels(glossary$section)

glossary <- as.data.table(glossary)
glossary <- setNames(split(glossary[,-c(1), with = FALSE], f = as.integer(glossary$section)), sectionNames)

glossary <- lapply(glossary, function(x) setNames(split(x, f = seq_along(x$columnName)), x$columnName))

```

# Core traitdata columns

Generally, for the essential primary data (species name, trait name, trait value), the trait standard recommends to keep the original naming and value of the data provider to ensure compatibility on the provider's side and to check for inconsistencies and misspellings. The original data provider's information is indexed by appending `_user` to the column name. The R script helps in transferring those entries into the accepted names and factor levels provided by the lookup tables.

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[1]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[as.character(i)]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

# Columns referring to Measurement or Fact

This section provides additional information about the specimen, which are not captured by the taxon information above.

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[2]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[as.character(i)]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

# Columns referring to specimen

this section of columns aims for identifying the methodology and primary source of the data and keep the reference to the actual specimen (e.g. for museum collections or related data analysis).

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[3]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[i]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

# Columns referring to sampling event or origin of record

this section of columns aims for identifying the methodology and primary source of the record and keep the reference to the actual specimen (e.g. for museum collections or related data analysis). It also provides columns for georeferencing the specimen.

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[4]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[i]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

# Columns for aggregate measures

Some data may not refer to a single specimen or measurement, but aggregate repeated measurements into an average.

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[5]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[i]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

# Additional columns for use in Biodiversity Exploratories

This section records location in the context of the exploratories. From `ExploratotriesPlotID` a detailled georeference can be inferred. Additional spatial resolution, e.g. on subplots, may be provided in `locationID` of section sampling event.

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[7]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[i]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

# Metadata vocabulary

```{r, results = 'asis', echo = FALSE}

j <- sectionNames[6]

 for(i in columnNames[sections == j]) {
     dd <- as.data.frame(t(glossary[[j]][[i]]))
     colnames(dd) <- i
     print(kable(dd, format = "markdown"))
 }

```

----

<a rel="license" href="http://creativecommons.org/licenses/by/4.0/"><img alt="Creative Commons License" style="border-width:0" src="https://i.creativecommons.org/l/by/4.0/88x31.png" /></a><br /><span xmlns:dct="http://purl.org/dc/terms/" property="dct:title">Ecological Traitdata Standard </span> by <span xmlns:cc="http://creativecommons.org/ns#" property="cc:attributionName">Florian D. Schneider, Nadja Simons, Caterina Penone, Andreas Ostrowski</span> is licensed under a <a rel="license" href="http://creativecommons.org/licenses/by/4.0/">Creative Commons Attribution 4.0 International License</a>.

### 2017-05-12 docs/traitdatastandard.Rmd
move to docs folder

--- 2017-05-12 traittemplate.Rmd
@@
 date: "12 Mai 2017"
-output: html_document
+output: 
+  html_document: 
+    keep_md: yes
 ---
@@
 library(pander)
-glossary <- read.csv("template_glossary.csv")
+glossary <- read.csv("../template_glossary.csv")
 

### 2017-06-21 docs/traitdatastandard.Rmd

--- 2017-05-12 traittemplate.Rmd
@@
 date: "12 Mai 2017"
-output: html_document
+output: 
+  html_document: 
+    keep_md: yes
+    toc: yes
 ---
@@
 **This is the documentation of the Ecological Traitdata Standard (ETS), version 0.2** 
-
 This defined vocabulary aims at providing all essential columns for raw data of functional trait measurements for ecological research.
@@
 
-The template is ordered into five different sections for column names, as well as a vocabulary for metadata information of particular importance for trait data.
+The template is ordered into six different sections, as well as a vocabulary for metadata information of particular importance for trait data.
 
 1. Trait measurement data
-2. Resolution of data / information on specimen
-3. Data source / information on measurement
-4. Measurement 
-5. Location
+2. Information about the Measurement or Fact
+3. Resolution of data / information on specimen
+4. Columns referring to sampling event or origin of record
+5. Columns for aggregate measures 
+6. Location information
 
@@
 library(knitr)
-library(pander)
-glossary <- read.csv("template_glossary.csv")
+glossary <- read.csv("../template_glossary.csv")
 
@@
  for(i in columnNames[sections == j]) {
-     dd <- as.data.frame(t(glossary[[j]][[as.character(i)]]))
+     dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
      colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
      print(kable(dd, format = "markdown"))
@@
  for(i in columnNames[sections == j]) {
-     dd <- as.data.frame(t(glossary[[j]][[as.character(i)]]))
+   dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
      colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
    print(kable(dd, format = "markdown"))
@@
  for(i in columnNames[sections == j]) {
-     dd <- as.data.frame(t(glossary[[j]][[i]]))
+    dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
      colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
  print(kable(dd, format = "markdown"))
@@
  for(i in columnNames[sections == j]) {
-     dd <- as.data.frame(t(glossary[[j]][[i]]))
+      dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
      colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
     print(kable(dd, format = "markdown"))
@@
  for(i in columnNames[sections == j]) {
-     dd <- as.data.frame(t(glossary[[j]][[i]]))
+ dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
      colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
      print(kable(dd, format = "markdown"))
@@
  for(i in columnNames[sections == j]) {
-     dd <- as.data.frame(t(glossary[[j]][[i]]))
+      dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
      colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
      print(kable(dd, format = "markdown"))

### 2017-07-06 docs/traitdatastandard.Rmd
start revising data standard description

@@
 
-**This is the documentation of the Ecological Traitdata Standard (ETS), version 0.2** 
+**This is the documentation of the Ecological Traitdata Standard (ETS), version 0.3**
+
 This defined vocabulary aims at providing all essential columns for raw data of functional trait measurements for ecological research.
 
-The column names are partly aligning with names from Darwin Core Standard and Extensions (Names compliant with DWC are referenced thus; the full Darwin Core can be found here: http://rs.tdwg.org/dwc/terms/index.htm)
+Most terms relate to terms from the Darwin Core Standard and Extensions (Names compliant with DWC are referenced thus in field 'Refines'; the full Darwin Core can be found here: http://rs.tdwg.org/dwc/terms/index.htm)
 
-The template is ordered into six different sections, as well as a vocabulary for metadata information of particular importance for trait data.
+The data standard is ordered into a **core section** with essential columns for trait data, extensions which are allowing to provide additional layers of information, as well as a vocabulary for metadata information of particular importance for trait data.
+A final section provides defined terms for lists of trait definitions, also termed a Trait Thesaurus.
 
-1. Trait measurement data
-2. Information about the Measurement or Fact
-3. Resolution of data / information on specimen
-4. Columns referring to sampling event or origin of record
-5. Columns for aggregate measures 
-6. Location information
+Extensions:
 
-Except for the first section, all columns are optional.
-Only the mandatory fields are required to produce valid trait-datasets.
-Some mandatory fields will be assisted by the R-Script.
-The columns labelled 'highly recommended' serve a better comparability of data across datasets.
+- the `Occurence` extension contains information on the level of individual specimens, such as date and location and method of sampling and preservation, or physiological specifications of the phenotype, such as sex, life stage or age.
+- the `MeasurementOrFact` extension takes information at the level of single measurements or reported values, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- The extension `BiodiversityExploratories` provides columns for localisation for trait data from the Biodiversity Exploratories sites (ww.biodiversity-exploratories.de).
 
@@
 columnNames <- glossary$columnName
-sections <- glossary$section
-sectionNames <- levels(glossary$section)
+sections <- glossary$Namespace
+sectionNames <- levels(glossary$Namespace)
 
 glossary <- as.data.table(glossary)
-glossary <- setNames(split(glossary[,-c(1), with = FALSE], f = as.integer(glossary$section)), sectionNames)
+glossary <- setNames(split(glossary[,-c(1), with = FALSE], f = as.integer(glossary$Namespace)), sectionNames)
 
@@
 
-Generally, for the essential primary data (species name, trait name, trait value), the trait standard recommends to keep the original naming and value of the data provider to ensure compatibility on the provider's side and to check for inconsistencies and misspellings.
-The original data provider's information is indexed by appending `_user` to the column name.
-The R script helps in transferring those entries into the accepted names and factor levels provided by the lookup tables.
+Generally, for the essential primary data (taxon assignment, trait name, trait value), the trait data standard recommends to keep the original naming and value of the data provider to ensure compatibility on the provider's side and to check for inconsistencies and misspellings.
+To ensure compatibility with other datasets, the original data provider's information should be transferred into a standardized columns indexed by appending `Std` to the column name.
+The R package 'traitdataform' helps in transferring those entries into the accepted taxon names and trait definitions provided by public ontologies.
 
@@
 
-j <- sectionNames[1]
+j <- "Traitdata"
 
@@
 
-# Columns referring to Measurement or Fact
+# Extension: Measurement or Fact
 
@@
 
-j <- sectionNames[2]
+j <- "MeasurementOrFact"
 
@@
 
-# Columns referring to specimen
+# Columns referring to occurence
 
@@
 
-j <- sectionNames[3]
-
- for(i in columnNames[sections == j]) {
-    dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
-     colnames(dd) <- i
-     cat("## `", i, "`  ", sep = "")
- print(kable(dd, format = "markdown"))
- }
-
-```
-
-# Columns referring to sampling event or origin of record
-
-this section of columns aims for identifying the methodology and primary source of the record and keep the reference to the actual specimen (e.g.
-for museum collections or related data analysis).
-It also provides columns for georeferencing the specimen.
-
-```{r, results = 'asis', echo = FALSE}
-
-j <- sectionNames[4]
-
- for(i in columnNames[sections == j]) {
-      dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
-     colnames(dd) <- i
-     cat("## `", i, "`  ", sep = "")
-    print(kable(dd, format = "markdown"))
- }
-
-```
-
-# Columns for aggregate measures
-
-Some data may not refer to a single specimen or measurement, but aggregate repeated measurements into an average.
-
-```{r, results = 'asis', echo = FALSE}
-
-j <- sectionNames[5]
+j <- "Occurence"
 
@@
 
-j <- sectionNames[7]
+j <- "BiodiversityExploratories"
 
@@
 
-j <- sectionNames[6]
+j <- "Metadata"
+
+ for(i in columnNames[sections == j]) {
+      dd <- as.data.frame(t(glossary[[j]][[as.character(i)]][,-1, with=FALSE]))
+     colnames(dd) <- i
+     cat("## `", i, "`  ", sep = "")
+     print(kable(dd, format = "markdown"))
+ }
+
+```
+
+# Terms for Traitlists / Trait Thesaurus
+
+```{r, results = 'asis', echo = FALSE}
+
+j <- "Traitlist"
 

### 2017-09-12 manuscript/manuscript.Rmd
manuscript: rename and major edit (no tracking), add autobibliography

---
title: "Introducing an Ecological Trait-data Standard"
author: "Florian D. Schneider, Nadja K. Simons, Andreas Ostrowski, Caterina Penone, Malte Jochum, Gaetane Le Provost"
geometry: margin=1.5in
bibliography: div05.bib
csl: amnat.csl
output:
  word_document: default
  pdf_document: default
header-includes: \usepackage{lineno}
fontsize: 12pt
---

\linenumbers

<!-- 
The aim of the methods paper is to 
- highlight the importance of data standards for functional traits
- provide standards and tools for decentralised implementation

# Introduction

1. importance of trait based ecology, current research questions, current applications of traits. 
2. Overview of existing database initiatives. Offering access management and harmonization;

(table 1 listing of open an restricted trait databases and datasets)

3. Heterogeneity of data standards and centralised databases. Open Access, publication in low-threshold data repositories becomes more important. Decentralisation of databases and importance of data brokering requires data standards. 
- lack of a universal traitdata standard.

## Common structure of trait datasets

1. minimal definition of a trait dataset
2. possible data formats: matrix, occurrence table, measurement long table 
- species--trait matrix is common for sharing taxon-level averages or facts, common for publishing species-level data and literature data. but not raw-data. The means of obtaining the average are intransparent and may be biased by local conditions. 
- occurrence table is common for empirical replicated measurements of traits, raw data. Ideal for correlation analysis of traits. Also, multivariate traits, e.g. morphometric landmarks or biochemical trais. Not ideal for merging data from multiple sources. 
- measurement table, longtable: ideal for raw data. Machine readable, simple to standardize and for sharing data, merging and combining data. 
- additional information may apply to different layers: on measurement or occurrence, co-variate analysis, environmental filters, spatial distribution.

(figure 1 species-trait matrix plus additional information, vs longtable plus additional information )

# Towards an Ecological Traitdata Standard

1. definition of structure: longtable format is most feasible to carry information at measurement level, link to external datasets via globally unique identifiers (or dataset specific)
- add occurrence level information (e.g. location, sex) and measurement level (e.g. aggregated data, statistics) information
- apply taxon name standardization
- apply trait standardization, link to definitions (trait lists, thesaurus) via globally unique identifiers, semantic network; apply unit and factor level harmonization

## A glossary of terms

- We provide: Glossary of terms for trait data and trait definitions

(table 2: most relevant terms of the traitdata standard, scheme of a term definitition)
(table 3: terms for trait definitions / trait lists)

## Computational tools for producing compliant data

- R package traitdataformat
- standardize and harmonize trait data for publication
- merge, split, apply trait data of different sources

(figure 2 process chart from original file to standardized output)

# Discussion

- With these tools, traitdata will be easier to harmonize 
- decentralised structures, not limited by infrastructural projects
- open source community development, invite contributions to traitdata standard and toolchain of R-package
- encourage development of trait definitions and semantic ontologies
- a global analysis of ecological traits will be facilitated

-->

# Introduction

*< importance of trait based ecology, current research questions, current applications of traits. >*

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]<!--^[Some frameworks differentiate effect traits and response traits (or performace and fitness traits), depending on the research question [refs, @violle07]]-->.

Early trait-based approaches have therefore been used to describe the role of a species in an ecosystem or its ability to persist under certain environmental conditions (refs).

Being assessed with a variety of research questions in mind, traits have proven useful in bridging individual level behaviour and physiology into processes at the ecosystem scale [@diaz13]. The analysis of correlations between traits and intraspecific variation informs about the physiological and evolutionary trade-offs and helps to identify economic strategies of organisms beyond taxonomy [@menezes10; @forister15].

Inferring the unknown trophic role and ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.

By looking at community weighted means and variance of functional traits, researchers aim to understand changes in functional composition or the loss of functional diversity and ecosystem services [@wood15; @bello10] in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10]  <!-- Gossner et al & other Explo papers -->. Since traits are closely related to the underlying ecological mechanisms, such patterns in functional community composition is promising versatile statistical tools to assess the current state of an ecosystem (@refs).

Traits furthermore allow comparision across trophic levels. Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs]. 
This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.

Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
A global synthesis of traits brings major insights in ecosystem processes and ecosystem services, enables projections in future climate and land use scenarios. Traits link ecological research to evolution, and management (functions and services) for biodiversity conservation (land use intensification), climate research (environmental filters). 
It is thus an interdisciplinary/cross-topic field with great potential.

In recent years, functional ecology has thus grown into a major field in ecological research!
However, a integration across trophic levels, taxa and organisational scales remains difficult because of different organismal, methodological, and functional focus. 
A harmonization of trait data standards would be of huge benefit for our understanding of ecosystems.

<!-- 
Flo: This section should probably include some more exploratories papers.

Traits can be measurements of organism morphology, life-history characteristics such as reproductive strategies, physiological traits including metabolism and photosynthetic activity, feeding traits, biochemical and isotopic compounds, behavioural traits, as well as environmental traits.

Refer to Kattge et al 2011, Violle et al 2007, Diaz et al 2007, McGill et al. 2006, Aubin et al. 2013, Lavorel et al. 2002...
## trait-based research within the Biodiversity Exploratories

Refer to successfully published papers on arthropod traits, multi-trophic trait variation etc.

Allan et al 2015
Boerschig et al 2013
Gossner et al 2015 & 2016
Simons et al 2014 & 2016
Kühsel et al 2015 & 2016
Birkhofer et al 2015
Chisté et al 2016
Gamez-Virues et al 2015

Summarize different questions, currently proposed in above- and below ground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.

Need to unify for multi-trophic integration

Trait matching integrates function across trophic levels, influence of land use and environmental filtering on trait composition, Exploratories are ideal framework for testing

-- Cat: here I would organize a bit differently to link better to the previous section:--
Traits can bridge biodiversity and functioning and that they allow comparision across trophic levels. Furthermore trait composition is influenced by land use and environmental filtering of traits. 
Exploratories are ideal framework for testing these questions because they are framed to study interactions between LUI, biodiversity and functions. 
However, multi-trophic integration is difficult because of different standards, methods, focus..
 needs unification in order to advance research and make data use possible for future generations

<!-- Nadja: I like Caterinas suggestion --> 
<!-- Flo: Yes, makes sense. -->
<!-- Malte: I agree, Caterina's suggestion makes it easier to follow for the reader. -->

*< Overview of existing database initiatives. Offering access management and harmonization >*

Depending on the organism groups and research questions, trait data have been standardised into centralised databases. Most prominently, plant traits from many different sources, including the extenisve LEDA traitbase for the Northwest-European flora [@kleyer08], for plant traits have been collated into the quickly growing TRY database [@kattge11]. TRY collates trait data from other published databases with individually owned datasets and provides an extended access management and data query services. Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] for root measurements [Fine-root Ecology Database; @iversen17]. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage.

In the animal kingdom, specialised trait-databases have been created that cover different organism groups or interaction types or ecosystems. For instance, The BETSI database collects traits on soil invertebrats (http://betsi.cesab.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades. 
However, a unified platform and harmonizing scheme for the datasets is still lacking.

Microbial ecology also makes great use of trait-based approaches to asses genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->

<!-- Flo: I would leave this open spreadsheet out for now. It would be quite a lot of effort to collect something representative. It might be an exercise for the exploratories people working on traits, for a start. In principle, any open datasets could be linked and imported into the R-package. I will highlight this possibility when talking about the package. --> 
<!-- As an attempt to track these unstandardized datasets, we initiated a living spreadsheet registry which invites submissions of any open data (Google Spreadsheet link, table 1).

(table of existing databases? name, url, focus, number of records, openness of data, maintained by) -->

<!-- Caterina: I like this table idea but should we limit it to plants and arthropods or also other organisms? Europe or whole world? I would opt for plants and arthropods in Europe but in case this becomes a publication open a collaborative file where anybody in the world can add the information for any organism in any part of the world.

Flo: I like this idea of starting a collaborative list on trait datasets. This
   1. illustrates the hetereogeneity of trait data and the need for a traitdata standard
   2. encourages a global traitdata analysis based on heterogeneous sources, which is enabled by the tools we provide here. See https://github.com/fdschneider/bexis_traits/issues/20
   -->

*< Heterogeneity of data standards and centralised databases >*

It becomes apparent, that all databases come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus. Certainly, a harmonization of trait data can be achieved by services like TRY, which take a huge effort to attract data submissions and map them into a common scheme. 
Furthermore, these databases provide access management services and tracking of data usage in studies and secondary publications. Mandatory Open Access publication still seems to prevent some data providers from uploading their data, why many databases encourage a permissive data policy but do not enforce it. 
<!-- Malte: I think this definitely needs to be backed up with a reference. -->

Centralised databases like TRY experience rapid growth but may at some point reach a resource limit in personnel and funding. Repeating this exercise for other organism groups or other trait-based research questions would be desireable. However, it has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits. [refs]

On the other hand, public research funding agencies increasinly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand. The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16]. 
At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Given this prospect, there will be no lack of data, but a lack of data standardization.

<!--Today, several general biodiversity data intitiatives are striving for a global integration of ecological and biological information. Only few so far developed universal data frameworks for trait data, such as the EOL TraitBank, which is the most general framework to date (Parr et al 2015). TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. -->
To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance. GFBio is aiming to ease data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.

Such ontologies are being developed for plant traits, for instance in the TOP Thesaurus of plant traits that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches  the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].

With this paper, we propose a roadmap towards a comprehensive trait data standard, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

## Common structure of trait datasets

As a minimum consensus, trait datasets may be defined as follows: 
A trait-dataset contains quantitative measurements or qualitative facts (i.e. trait values) <!-- Cat: the difference is not totally clear to me - is it the difference between field vs literature gathered data?  Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs. literature, but not entirely. A differentiation that has rarely been specified in the trait-litereature it seems. Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further. -->  about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an of individual or parts of an individual (i.e. an entity) that can be assigned to a biological taxon (i.e. a species or higher taxon). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon. 
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'.

*< data structure >*

Trait-datasets take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums. The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level (@laliberte10). Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (Refs).

For investigations of phenotypic variation of traits,  would be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation and sexual dimorphism. We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species. This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.

Computationally most effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) via a unique identifier which is assigned to a taxon. This allows for repeated measurements, even on a single individual. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID. <!-- Malte: I think the difference between the long-table format and the previous one is not clear enough. -->
The latter format thus provides the highest resolution and ideal for storing raw data of trait measurements in huge databases and link them via the respective unique identifiers to additional information, such as occurrenceID or locationID.

<!-- Cat: here I guess that a figure showing few rows of a species x trait matrix and a long-format one with links to other information (metadata,..) woud help quite a lot. By experience some users are not be familiar with the interest of using the long-table format.-->
<!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->

# Proposing an ecological traitdata standard

In this paper, we propose a universal scheme of defined column names and data structure that captures the different degrees of resolution and measurement detail for multiple use cases of trait data. We also provide computational tools in an R package to help users to transfer their original data into the proposed scheme.

While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we want to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources. 
In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of trait-datasets proposed above.
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits.

With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL)  TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC). We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility. The full list of terms of the traitdata standard including further explanation can be assessed at http://www.xxxxx.

(table 2: most relevant terms of the traitdata standard, scheme of a term definition)

## minimal requirement of a trait-dataset

For the minimal definition of trait-datasets posed above, the core content of a row is the reported measurement or fact for a single observation, which is composed of a value (`traitValue`) and -- for numeric values -- a standard unit (`traitUnit`).
To associate the measurement or fact to a clear trait definition, a unique identifier links each row to a descriptive trait name (`traitName`).

The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained via globally accepted identifiers. To provide an unambiguous reference which is easy to read for researchers and for software, this identifier is provided in the form of a descriptive taxon name (`scientificName`) as well as an unambiguous ID (`taxonID`).

*< global taxonomy standards >*

To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may contribute to different communities by allowing for a double record of both user-specific and standardised entries (as practiced in TRY, -@kattge11 ). This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right. We propose to add a second set of columns that contain standardized entries of the taxon name (`scientificNameStd`) that maps synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain to a globally accessible URI that maps to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') . 
Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambigueities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorization (Shan et al 2012, Schrodt et al 2015).

*< relating data to trait definitions >*

A similar approach to standardised trait names (`traitNameStd`) and harmonized reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
Similar as for taxonID, the field  `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitionsor a measurement methodology.

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
Resources for semantic ontologies exist for plant traits (e.g., TOP Plant Trait Ontology [@garnier17]), soil invertebrates (T-Sita, @ref), or biomedical traits of vertebrates (Vertebrate Trait Ontology [@park13]). However, for most organism groups and research contexts we attest a deficit of public ontologies with openly accessible URIs.

Therefore, if no published trait definition is available that can be referenced, traitdatasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte12; @garnier17]. 
This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and a unique (alphanumeric) identifier (`traitID`) as well as an unambiguous verbal definition (`traitDescription`), the legit factor levels (for categorical data, `factorLevels`) or expected standard units (for numerical data).

Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related Terms (`relatedTerm`) or synonyms (`synonym`). For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@Ref semantic database methods). This extends the trait list into a semantic web resource and facilitates the classification of traits and allows cross-taxon comparative studies at the level of broader terms. (more detail on semantic web ontologies can be found ...)

<!-- Flo: For now removing this block on the steps necessary to create web ontologies of traits. The section above is covering the essentials and implies that further work is required. We might add this following section as a supplement or within the traitdata standard. -->
<!--

## towards a distributed semantic web ontology of traits

Thus, a trait thesaurus or ontology would assign trait names with A) an unambiguous definition of the trait and B) an expected format of measured values or reported facts, and might additionally provide C) semantic relationships for deriving a hierarchical or tree-based classification of traits.

An ontology is going one step further by providing an explicit definition of a term including a formal model of the conceptual objects and relationships, or entities and qualities [@garnier17, @gruber95], ideally structured following the guidelines for a semantic web standard [ref].

Traits are not only defined in terms of their interpretation, but are ideally also standardised in terms of numerical units and, even more important, the use of factor levels. This is challenging, given the range of data types that fall within datasets of functional traits.

Numerical values represent measurements of length, volumes, ratios, rates or timespans. Integer values may apply to count data (e.g. eggs per clutch). 
Binary data (encoded as 0 or 1) or logical data (coded as TRUE or FALSE) may apply to qualitative traits such as specific behaviour during mating (e.g. are territories defended) or specialisation to a given habitat (e.g. species restricted to relicts of primeval forests). Many traits are categorical and allow for a constrained set of factor levels, such as sex differences in wing morphology (both sexes winged, both sexes unwinged, only males winged, only females winged) or unconstrained entries such as color. Categorical traits often are ordinal, i.e. they have a logical sequence as in the case of life stages or hibernation stages, or habitat preference traits such as horizontal stratum use. 
Finally there are specific formats of multivariate trait values, e.g. x-y-z coordinates of a landmark measured in a standardized 3D space [ref] or relative abundance of chain-lengths in biochemical compounds [].
To cope with this variety of data types, definitions should refer to other well-defined terms of other ontologies that describe standard units, morphological body parts, protein characteristics (Protein Ontology) or chemical terms (e.g. the ChEBI, http://www.obofoundry.org/ontology/chebi.html).

Multiple approaches have spawned around the initialisation of trait databases. Such reference definitions are currently being published in methodological handbooks (Moretti et al , ) or lists of accepted traits (databases), or an online ontology or thesaurus (TOP, T-Sita, ), most advanced certainly for plant traits compiled in the Thesaurus of Plant characteristics (TOP, @garnier17).

Online ontologies extend into (machine readable) semantic web resources by providing a hierarchical classification of traits (TOP) or a relational tree of functional traits (e.g. TOP or T-SITA). .

Each trait definition may link to a broader or narrower term. For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. (Ref semantic database methods)
This links traits of similar functional meaning and allows cross-taxon comparative studies at the level of broader terms.

These systematic approaches to traits will be very useful for comparing similar traits measured in different taxa at higher hierarchical levels.
Furthermore, trait definitions may refer to related terms that describe a similar feature but with other means, or synonyms defined in other trait ontologies. By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.

The benefit of such classifications will increase if open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. To harmonize trait data across databases, future trait standard initiatives should provide this functionality. 
Online ontologies hosted with accredited ontology servers have the advantage of providing a persistent and direct link of the term on the internet (a *Uniform Resource Identifier*, URI).
Terminology portals or registries, such as the GFBio Terminology Service or ontobee, may provide a central host for trait ontologies.

Thus, a trait thesaurus or ontology would assign trait names with A) an unambiguous definition of the trait and B) an expected format of measured values or reported facts, and might additionally provide C) semantic relationships for deriving a hierarchical or tree-based classification of traits. It additionally might relate to a more specific method of measurement.

The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards. This effort is time consuming and lengthy, but is returns synergies and great scientific benefit by enabling a global synthesis, as the example of a global plant trait database TRY shows intricately [@diaz16]. 
-->

*< other identifiers >*

<!--Additional information may apply to different layers of the core data: on measurement or occurrence, co-variate analysis, environmental filters, spatial distribution. -->

Further detail might be stored along with each measurement or fact reported in the dataset (i.e. each row of the dataset). Those information might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset. The terms for this layer of information are defined by the measurement or fact extension. The `measurementID` would also be used to link multivariate measurement data that originate from the same measurement event.

The measurement may also be referred to a single specimen or occurrence via a uniue identifier (`occurrenceID`). For most measured trait data (see above) this is a vital information that links related measurements of different traits taken on the same specimen. 
This identifier is usually dataset-specific and can be defined by the author and described with further detail in a separate data table using the columns provided in the occurrence extension detailed below. Some data-types may use global identifiers for occurrence data, e.g. a GBIF URI or a museum collection code that is publicly available.
 <!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear. It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc. -->

Additionally, information about the authorship and ownership of the data and the terms of use should be considered. These information may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).

<!-- 
(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits. However, such network data may be of interest to be combined with trait data. ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names. -->

## Extensions for additional data layers

Many terms of this section refine terms of the Darwin Core Standard and its own MeasurementOrFact and [Occurrence](http://tools.gbif.org/dwca-validator/extension.do?id=http://rs.tdwg.org/dwc/terms/Occurrence#Occurrence) extensions. The terms may be applied as columns in a supplementary data sheet or along with the core data described above.

### additional information on measurement or fact

As a high-level discrimination of the source of the measurement or fact, the Darwin Core Term `basisOfRecord` takes an entry about the type of trait data recorded: Were they taken by own measurement (distinguish "LivingSpecimen", "PreservedSpecimen", "FossilSpecimen") or taken from literature ("literatureData"), from an existing trait database ("traitDatabase"), or is it expert knowledge ("expertKnowledge"). It is highly recommended to provide further detail about the source in the column `basisOfRecordDescription`.

To keep track potential sources of noise or bias in measured data, the method of measurement (`measurementMethod`), the person conducting the measurement (`measurementDeterminedBy`), and the date at which the measurement was obtained (`measurementDeterminedDate`) are recorded. 
Authors would often report aggregate data of repeated or pooled measurements, e.g. by weighing multiple individuals simultaneously and calculating an average. In these cases, recording the number of individuals (`individualCount`) along with a dispersion measure (e.g. variance or standard deviation, `dispersion`) or range of values (e.g. min and max of values observed in the field `measurementValueMin`, `measurementValueMax`) is adviced. The field `statisticalMethod` names the method for data aggregation (e.g. mean or median) as well as the variation or range (e.g. reporting variance or standard deviation).

For data not obtained from own measurement, the field `references` provides a precise reference to the source of data (e.g. a book or existing database) or the authority of expert knowledge. 
For literature data, the original source might report trait values on the family or genus level, but the dataset author infers and reports the trait data at species level (e.g. if the entire genus reportedly shares the same trait value). To preserve this information, the column `measurementResolution` should report the taxon rank for which the reported value was originally assessed.

### additional information on observation context (occurrence)

This category of terms contains further information about the individual specimen or occurrence that has been observed and measured.

For both literature and measured data, trait values may be recorded for different sub-categories of individuals of a taxon to capture polymorphisms, for instance differentiated by sex or life stage. The template provides the fields `sex`, `lifeStage`, `age`, and `morphotype` for this distinction.

Seasonal variation of traits may be recored by assigning a date and time of sampling to the occurrence, using the fields `year`,	`month` and	`day`, depending on resolution. The field definitions of the Darwin Core Standard can be applied instead, to refer to a geological stratum, for instance. 
Sampling may be further specified using a unique identifier for the sampling event `eventID`	which references to an external dataset. The record of a `samplingProtocol` may capture bias in samling methods. Further procedures and methods of preservation should be reported in `preparations`.

To capture geographic variation of traits, a set of fields for georeferencing can put the observation into spatial and ecological context (`habitat`, 	`decimalLongitude`,	`decimalLatitude`, `elevation`,	`geodeticDatum`, `verbatimLocality`, `country`, `countryCode`). The field `locationID` may be used to reference the occurrence to a dataset-specific or global identifier. This allows the trait data to double as observation data, e.g. for upload to the GBIF database.

<!-- ## The Biodiversity Exploratories Extensions and template

Data will be uploaded to BExIS using the template to join data to the trait dataset while keeping full data authority: author information and ownership, access right management, as provided by BExIS.

If applicable, trait values can be assigned to regions, plot IDs or sampling events which will then allow users to combine trait data with abundance biodiversity, ecosystem functioning, other observational data or the management regime on the sampled plots.

Metadata:
Each trait dataset uploaded to BExIS and added to the trait database will be accompanied with a metadata file containing additional information on data origin, method, authors and owners (in redundancy with primary data, see above) and possibly more detailed definition of the traits recorded. 
-->

## information on attribution and permissions to use

Individual measurements and facts will likely already belong to a larger set of traits or a separate trait database before they are added to a combined trait database. To retain the rights of the original data contributor, the field `rightsHolder` states the person or organization who owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and `license` specifies under which terms and conditions the data can be used, re-used and/or published. 
This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`,`authorLastname` and `authorFirstname`. These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.

<!--

Finally, there is the set of information that applies to the entire trait-dataset, which classifies them as metadata. Since trait data are of great use for synthesis studies, information about how the data may be distributed, re-used and attributed to are of particular importance for trait datasets. Most researchers encourage re-use of their published datasets while making sure they are sufficiently credited. The use of permissive licenses for traitdata publications, such as Creative Commons Attribution or Creative Commons Zero/Public Domain release, has been established as the gold standard.

<!-- Where is this data record available online? Source in the Measurements or Facts extension file is for a url to be displayed alongside the data point, so EOL visitors can click through to an online resource where the data originated. This might be a dataset stored at an online repository, a taxon page on a website, or any other online location.
What shall I cite if I use this data record in a publication? BibliographicCitation in the Measurements or Facts extension file is for the suggested citation format for a user of this dataset to include in a references list or bibliography. Even if the data record or dataset have never been published elsewhere, we recommend that you craft an appropriate citation so others can provide proper credit.
If the dataset is from a literature review, and one or more references are available for individual data points, these references should be recorded in the References extension and listed by their Reference IDs in the ReferenceID field of the Measurements or Facts extension file, see below.  -->

-->

# Computational tools for producing compliant data

We provide an R package to assist in producing data compliant with the trait data standard proposed above. The package is being developed as an open source project on GitHub (https://github.com/fdschneider/traitdataform).

There are two major use cases for the package:

- preparation of own trait datasets for upload into public data bases, and
- harmonizing trait datasets from different sources by moulding them into a unified format.

The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table data into a measurement longtable format. It also maps column names into terms provided in the trait data standard.

Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy (taxonomic ontology server) by calling the function `standardize.taxonomy()`

The `standardize.traits()` function matches user provided trait names onto public trait ontologies (for now using GBIF Backbone Taxonomy) and links it to a trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonizes trait values into target units and legit factor levels.

The output of these functions can easily be merged using the `rbind()` function of R base.

In principle, transferring and harmonizing traitdata using the package is as simple as:

```r
library("traitdataform")
traitdataset1 <- standardize.traits(read.csv("path/to/data.csv"),
            thesaurus = as.thesaurus("http://url.of/thesaurus.csv"),
            taxa = "name_correct", units = "mm")
```

For a detailed description on how to use the package see the package vignette (). The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page ().

<!--More step-by-step examples here on how to create linked datasets, add a trait thesaurus, merge two datasets  -->

(figure 2 process chart from original file to standardized output)

# Conclusion

In this paper, we propose a universal scheme for ecological trait data. 
We also provide a toolbox in R to transfer data into this scheme.

We further encourage the development of trait definitions in semantic ontologies. The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards. This effort is time consuming and lengthy, but is returns synergies and great scientific benefit by enabling a global synthesis, as the example of a global plant trait database TRY shows intricately [@diaz16].

These tools help to standardize trait datasets before upload to central as well as distributed data repositories and to harmonize and combine raw data from heterogeneous sources. We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@ref].

The trait data standard as well as the R package are published for open source community development. Contributions are invited and vital for the future development and extension of the package.

<!--
## Facilitate data availability & Synthesis mainstreaming

Trait-data may serve as a pilot case for data standardisation for particular types of data assessed within the exploratories. This might be transferable to other data types, as well.

potential to expand the framework across research projects.

Comparability of traits across taxa

Gap analysis

specificity of BE trait-data is local context of exploratories, if trait data are assessed on the site, trait variation can be related to land-use intensity. In this spatial extent, this would be a novelty.

-->

# Acknowledgements

Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project.

We thank the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
infrastructure; Christiane Fischer and Jule Mangels for giving
support through the central office, Michael Owonibi for
managing the central data base, and Markus Fischer, Eduard Linsenmair,
Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
Ernst-Detlef Schulze, Wolfgang W.Weisser and the late Elisabeth Kalko
for their role in setting up the Biodiversity Exploratories project.

The work has been partly funded by the DFG Priority Program
1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). Field work
permits were issued by the responsible state environmental offices of
Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
BbgNatSchG).

# Authors' contributions

# References

\footnotesize

Allan, E., Manning, P., Alt, F., Binkenstein, J., Blaser, S., Bluthgen, N., Bohm, S., Grassein, F., Holzel, N., Klaus, V.H.,
Kleinebecker, T., Morris, E.K., Oelmann, Y., Prati, D., Renner, S.C., Rillig, M.C., Schaefer, M., Schloter, M., Schmitt, B.,
Schoning, I., Schrumpf, M., Solly, E., Sorkau, E., Steckel, J., Steffen-Dewenter, I., Stempfhuber, B., Tschapka, M.,
Weiner, C.N., Weisser, W.W., Werner, M., Westphal, C., Wilcke, W., Fischer, M., 2015. Land use intensification alters
ecosystem multifunctionality via loss of biodiversity and changes to functional composition. Ecol Lett. doi:
10.1111/ele.12469

Aubin, I., Venier, L., Pearce, J., Moretti, M., 2013. Can a trait-based multi-taxa approach improve our assessment of
forest management impact on biodiversity? 22, 2957-2975. doi: 10.1007/s10531-013-0565-6

Birkhofer, K., Diekötter, T., Meub, C., Stötzel, K., Wolters, V., 2015. Optimizing arthropod predator conservation in
permanent grasslands by considering diversity components beyond species richness. 211, 65-72. doi: 10.1016/j.agee.2015.05.014

Birkhofer, K., Smith, H.G., Weisser, W.W., Wolters, V., Gossner, M.M., 2015. Land-use effects on the functional
distinctness of arthropod communities. Ecography 38, 889-900. doi: 10.1111/ecog.01141

Börschig, C., Klein, A.M., von Wehrden, H., Krauss, J., 2013. Traits of butterfly communities change from specialist
to generalist characteristics with increasing land-use intensity. Basic Appl Ecol 14, 547-554. doi: 10.1016/j.baae.2013.09.002

Chisté, M.N., Mody, K., Gossner, M.M., Simons, N.K., Köhler, G., Weisser, W.W., Blüthgen, N., 2016. Losers, winners,
and opportunists: How grassland land-use intensity affects orthopteran communities. 7, e01545. doi: 10.1002/ecs2.1545

Entling, W., Schmidt, M.H., Bacher, S., Brandl, R., Nentwig, W., 2007. Niche properties of Central European spiders:
shading, moisture and the evolution of the habitat niche. Global Ecol Biogeogr 16, 440-448. doi: 10.1111/j.1466-8238.2006.00305.x

Gamez-Virues, S., Perovic, D.J., Gossner, M.M., Borschig, C., Bluthgen, N., de Jong, H., Simons, N.K., Klein, A.M.,
Krauss, J., Maier, G., Scherber, C., Steckel, J., Rothenwohrer, C., Steffan-Dewenter, I., Weiner, C.N., Weisser, W.,
Werner, M., Tscharntke, T., Westphal, C., 2015. Landscape simplification filters species traits and drives biotic
homogenization. 6, 8568. doi: 10.1038/ncomms9568

Gossner, M.M., Lewinsohn, T.M., Kahl, T., Grassein, F., Boch, S., Prati, D., Birkhofer, K., Renner, S.C., Sikorski, J.,
Wubet, T., Arndt, H., Baumgartner, V., Blaser, S., Bluthgen, N., Borschig, C., Buscot, F., Diekotter, T., Jorge, L.R.,
Jung, K., Keyel, A.C., Klein, A.M., Klemmer, S., Krauss, J., Lange, M., Muller, J., Overmann, J., Pasalic, E., Penone, C.,
Perovic, D.J., Purschke, O., Schall, P., Socher, S.A., Sonnemann, I., Tschapka, M., Tscharntke, T., Turke, M., Venter,
P.C., Weiner, C.N., Werner, M., Wolters, V., Wurst, S., Westphal, C., Fischer, M., Weisser, W.W., Allan, E., 2016.
Land-use intensification causes multitrophic homogenization of grassland communities. 540, 266-269. doi:
10.1038/nature20575

Gossner, M.M., Simons, N.K., Achtziger, R., Blick, T., Dorow, W.H.O., Dziock, F., Köhler, F., Rabitsch, W., Weisser,
W.W., 2015. A summary of eight traits of Coleoptera, Hemiptera, Orthoptera and Araneae, occurring in grasslands
in Germany. Scientific Data 2, 150013. doi: 10.1038/sdata.2015.13

Gossner, M.M., Simons, N.K., Höck, L., Weisser, W.W., 2015. Morphometric measures of Heteroptera sampled in
grasslands across three regions of Germany. 96, 1154. doi: 10.1890/14-2159.1

Renner, S.C., Gossner, M.M., Kahl, T., Kalko, E.K., Weisser, W.W., Fischer, M., Allan, E., 2014. Temporal changes in
randomness of bird communities across Central Europe. Plos One 9, e112347. doi: 10.1371/journal.pone.0112347

Simons, N.K., Gossner, M.M., Lewinsohn, T.M., Boch, S., Lange, M., Muller, J., Pasalic, E., Socher, S.A., Turke, M.,
Fischer, M., Weisser, W.W., 2014. Resource-mediated indirect effects of grassland management on arthropod
diversity. PLoS One 9, e107033. doi: 10.1371/journal.pone.0107033

Simons, N.K., Weisser, W.W., Gossner, M.M., 2016. Multi-taxa approach shows consistent shifts in arthropod functional traits along grassland land-use intensity gradient. Ecology 97, 754-764. doi: 10.1890/15-0616.1

Weiner, C.N., Werner, M., Linsenmair, K.E., Bluthgen, N., 2014. Land-use impacts on plant-pollinator networks:
interaction strength and specialization predict pollinator declines. Ecology 95, 466-474.

Weiner, C.N., Werner, M., Linsenmair, K.E., Blüthgen, N., 2011. Land use intensity in grasslands: Changes in
biodiversity, species composition and specialisation in flower visitor networks. 12, 292-299. doi: 10.1016
j.baae.2010.08.006

### 2017-09-15 manuscript/manuscript.Rmd
update manuscript, insert figures and external links

@@
 title: "Introducing an Ecological Trait-data Standard"
-author: "Florian D. Schneider, Nadja K. Simons, Andreas Ostrowski, Caterina Penone, Malte Jochum, Gaetane Le Provost"
-geometry: margin=1.5in
-bibliography: div05.bib
+author: "Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas Ostrowski, Caterina Penone"
 csl: amnat.csl
 output:
-  word_document: default
+  word_document:
+    keep_md: yes
   pdf_document: default
+geometry: margin=1.5in
 header-includes: \usepackage{lineno}
 fontsize: 12pt
+bibliography: div05.bib
 ---
@@
 \linenumbers
-
-<!-- 
-The aim of the methods paper is to 
-- highlight the importance of data standards for functional traits
-- provide standards and tools for decentralised implementation
-
-# Introduction
-
-1. importance of trait based ecology, current research questions, current applications of traits. 
-2. Overview of existing database initiatives. Offering access management and harmonization;
-
-(table 1 listing of open an restricted trait databases and datasets)
-
-3.
-Heterogeneity of data standards and centralised databases.
-Open Access, publication in low-threshold data repositories becomes more important.
-Decentralisation of databases and importance of data brokering requires data standards.
-- lack of a universal traitdata standard.
-
-## Common structure of trait datasets
-
-1. minimal definition of a trait dataset
-2. possible data formats: matrix, occurrence table, measurement long table 
-- species--trait matrix is common for sharing taxon-level averages or facts, common for publishing species-level data and literature data.
-but not raw-data.
-The means of obtaining the average are intransparent and may be biased by local conditions.
-- occurrence table is common for empirical replicated measurements of traits, raw data.
-Ideal for correlation analysis of traits.
-Also, multivariate traits, e.g.
-morphometric landmarks or biochemical trais.
-Not ideal for merging data from multiple sources.
-- measurement table, longtable: ideal for raw data. Machine readable, simple to standardize and for sharing data, merging and combining data. 
-- additional information may apply to different layers: on measurement or occurrence, co-variate analysis, environmental filters, spatial distribution.
-
-(figure 1 species-trait matrix plus additional information, vs longtable plus additional information )
-
-# Towards an Ecological Traitdata Standard
-
-1. definition of structure: longtable format is most feasible to carry information at measurement level, link to external datasets via globally unique identifiers (or dataset specific)
-- add occurrence level information (e.g. location, sex) and measurement level (e.g. aggregated data, statistics) information
-- apply taxon name standardization
-- apply trait standardization, link to definitions (trait lists, thesaurus) via globally unique identifiers, semantic network; apply unit and factor level harmonization
-
-## A glossary of terms
-
-- We provide: Glossary of terms for trait data and trait definitions
-
-(table 2: most relevant terms of the traitdata standard, scheme of a term definitition)
-(table 3: terms for trait definitions / trait lists)
-
-## Computational tools for producing compliant data
-
-- R package traitdataformat
-- standardize and harmonize trait data for publication
-- merge, split, apply trait data of different sources
-
-(figure 2 process chart from original file to standardized output)
-
-# Discussion
-
-- With these tools, traitdata will be easier to harmonize 
-- decentralised structures, not limited by infrastructural projects
-- open source community development, invite contributions to traitdata standard and toolchain of R-package
-- encourage development of trait definitions and semantic ontologies
-- a global analysis of ecological traits will be facilitated
-
--->
 
@@
 The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level (@laliberte10).
-Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (Refs).
+Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).
 
@@
 We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.
-The full list of terms of the traitdata standard including further explanation can be assessed at http://www.xxxxx.
+The full list of terms of the traitdata standard including further explanation can be assessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
 
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.75\textwidth, angle=90]{figures/fig2.png}
+\caption{Process chart of the functions provided within the R package.}
+
+\end{figure}
+
 Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy (taxonomic ontology server) by calling the function `standardize.taxonomy()`
@@
 
-In principle, transferring and harmonizing traitdata using the package is as simple as:
+The package functions comprise a toolchain where each function can be piped as an input into the next.
+A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
 
@@
 library("traitdataform")
-traitdataset1 <- standardize.traits(read.csv("path/to/data.csv"),
+traitdataset1 <- standardize(read.csv("path/to/data.csv"),
             thesaurus = as.thesaurus("http://url.of/thesaurus.csv"),
@@
 
-For a detailed description on how to use the package see the package vignette ().
-The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page ().
-
-<!--More step-by-step examples here on how to create linked datasets, add a trait thesaurus, merge two datasets  -->
-
-(figure 2 process chart from original file to standardized output)
+For a detailed description on how to use the package see the package vignette (https://github.com/fdschneider/traitdataform/raw/master/docs/vignette.pdf).
+The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
 

### 2017-09-21 docs/traitdatastandard.Rmd
build traitdatastandard.html

@@
 This defined vocabulary aims at providing all essential columns for raw data of functional trait measurements for ecological research.
+Most terms relate to terms from the Darwin Core Standard and it's Extensions (terms of DWC are referenced thus in field 'Refines'; the full Darwin Core Standard can be found here: http://rs.tdwg.org/dwc/terms/index.htm)
 
-Most terms relate to terms from the Darwin Core Standard and Extensions (Names compliant with DWC are referenced thus in field 'Refines'; the full Darwin Core can be found here: http://rs.tdwg.org/dwc/terms/index.htm)
-
-The data standard is ordered into a **core section** with essential columns for trait data, extensions which are allowing to provide additional layers of information, as well as a vocabulary for metadata information of particular importance for trait data.
+The glossary of terms is ordered into a **core section** with essential columns for trait data, extensions which are allowing to provide additional layers of information, as well as a vocabulary for metadata information of particular importance for trait data.
 A final section provides defined terms for lists of trait definitions, also termed a Trait Thesaurus.
 
-Extensions:
+**Extensions:**
 
@@
 
-This list will also be available as
+**Structure of trait data:**
+
+The traitdata standard requires data to be stored in a long table format containing one measurement per row described by the column terms provided in the core section.
+(see Whitepaper for further considerations)
+
+There are two ways of integrating the extensions:
+
+- **within the same data file**: additional columns are provided to describe the properties concerning the measurement or the occurence of the specimen.
+The output file may be stored as a csv or txt table or an excel spreadsheet.
+- **in separate data files**: the main file refers to additional data files in the fields `measurementID` or `occurenceID`.
+This is usually the case if the occurences are externally documented, for instance as specimens from a museum.
+This also applies if the data are stored as an Excel spreadsheet or as a Darwin Core Archive (as proposed for TraitBank).
+
+This list of terms will also be available as
 
@@
 
-```{r, results = 'asis', echo = FALSE}
+```{r, results = 'asis', echo = FALSE, warning = FALSE}
 library(data.table)
@@
 glossary <- read.csv("../template_glossary.csv")
+urlroot <- "http://fdschneider.de/bexis_traits/traitdatastandard.html"
 
@@
 
-Generally, for the essential primary data (taxon assignment, trait name, trait value), the trait data standard recommends to keep the original naming and value of the data provider to ensure compatibility on the provider's side and to check for inconsistencies and misspellings.
-To ensure compatibility with other datasets, the original data provider's information should be transferred into a standardized columns indexed by appending `Std` to the column name.
-The R package 'traitdataform' helps in transferring those entries into the accepted taxon names and trait definitions provided by public ontologies.
+For the essential primary data (trait value, taxon assignment, trait name), the trait data standard recommends to report the original naming and value scheme as used by the data provider.
+However, to ensure compatibility with other datasets, the original data provider's information should be doubled into standardized columns indexed by appending `Std` to the column name.
+This ensures compatibility on the provider's side and transparency for data users on the reported measurements and facts, and enables checking for inconsistencies and misspellings in the complete dataset provided by the author.
+If provided, the standardized fields allow merging heterogeneous data sources into a single table to perform further analyses.
+This practice of double bookkeeping of trait data has successfully established for the TRY database on plant traits, for instance (Kattge ..).
+
+The R package 'traitdataform' (https://fdschneider.github.com/traitdataform) provides tools to transfer heterogeneous datasets into a longtable format and to create standardized taxa and trait columns, based on public ontologies.
+
+By linking to (public) ontologies via the field `taxonID`, further taxonomic information can be extracted for analysis.
+Alternatively, `taxonID` may also link to an accompanying datasheet that contains information on the taxonomic resolution or specification of the observation.
+
+Similarly, linking to trait terminologies (a 'Thesaurus') via the field `traitID` allows an unambiguous interpretation of the trait measurement.
+If no online ontology is available, an accompanying dataset should specify the trait definition.
+For setting up such a Thesaurus, we propose the use of terms provided in section 'Traitlist' below.
 
@@
      cat("## `", i, "`  ", sep = "")
+     cat("\n")
+     cat("[go to top](", urlroot, ")", sep = "") 
+     cat("  |  [direct link](", urlroot, "#", tolower(i), ")", sep = "")
+
    print(kable(dd, format = "markdown"))
@@
 
-This section provides additional information about the specimen, which are not captured by the taxon information above.
+This section provides additional information about a reported measurement or fact and in most cases can easily be included as extra columns to the core dataset.
+The columns would contain detail on the methodology of measuring and reporting of aggregated data.
+
+In case of facts reported from literature or from expert knowledge, or cited from other databases, please include bibliographic citation of the original data source in field 'references'.
 
@@
 
-# Columns referring to occurence
+# Extension: Occurence
 
-this section of columns aims for identifying the methodology and primary source of the data and keep the reference to the actual specimen (e.g. for museum collections or related data analysis).
+this section of columns aims for identifying the methodology and primary source of the data and keep the reference to the actual specimen (e.g.
+for museum collections or related data analysis).
+It also may add detail on the condition of the observed specimen, its sex, morphotype or developmental stage.
+Especially for analyses of intra-specific trait variation, this composes valuable data.
+
+For trait data compiled from literature or expert knowledge, the level of information on an 'occurence' would not apply, since no specific individual has been observed.
+In this case, the field 'occurenceID' should be left blank in the core data.
 
@@
      cat("## `", i, "`  ", sep = "")
+     cat("\n")
+     cat("[go to top](", urlroot, ")", sep = "") 
+     cat("  |  [direct link](", urlroot, "#", tolower(i), ")", sep = "") 
  print(kable(dd, format = "markdown"))
@@
 
-# Additional columns for use in Biodiversity Exploratories
+# Extension: Biodiversity Exploratories
 
@@
      cat("## `", i, "`  ", sep = "")
+     cat("\n")
+     cat("[go to top](", urlroot, ")", sep = "") 
+     cat("  |  [direct link](", urlroot, "#", tolower(i), ")", sep = "") 
      print(kable(dd, format = "markdown"))
@@
 # Metadata vocabulary
+
+These data are required when combining datasets from different sources.
+We suggest integrating a `datasetID` referencing to an external data sheet or adding columns to the dataset to keep important information about authorship and terms of use at the measurement level.
 
@@
 
-# Terms for Traitlists / Trait Thesaurus
+# Terms for Traitlists (a 'Trait Thesaurus')
+
+Ontologies for functional traits are beeing developed for different organism groups, mostly centered around certain research questions or subjects of study.
+To date, the TRY database takes the most inclusive approach on functional traits for vascular plants (Kattge) and describes XX traits and put them into the hierarchical structure of an ontology.
+
+For some animal groups, similar approaches do exist, but few are available as an online ontology.
+
+As a starting point for creating an ontology for functional traits, we propose the following terms for trait lists (also termed 'lookup tables' of traits, or trait 'Thesaurus'), to describe functional traits that are in the focus of a single research project.
+Those trait lists would be uploaded along with the trait dataset.
+
+Using this standardized terminology will allow merging trait definitions from multiple sources.
+We encourage providing these lookup tables as an open ressource on public ontology and terminology servers, to enable a global referencing.
 

### 2017-09-21 manuscript/manuscript.Rmd
Minor corrections to manuscript.

@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]<!--^[Some frameworks differentiate effect traits and response traits (or performace and fitness traits), depending on the research question [refs, @violle07]]-->.
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->.
 
@@
 In recent years, functional ecology has thus grown into a major field in ecological research!
-However, a integration across trophic levels, taxa and organisational scales remains difficult because of different organismal, methodological, and functional focus. 
+However, an integration across trophic levels, taxa and organisational scales remains difficult because of different organismal, methodological, and functional focus. 
 A harmonization of trait data standards would be of huge benefit for our understanding of ecosystems.
@@
 In this paper, we propose a universal scheme of defined column names and data structure that captures the different degrees of resolution and measurement detail for multiple use cases of trait data.
-We also provide computational tools in an R package to help users to transfer their original data into the proposed scheme.
+We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
 
@@
 GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`).
-The field `taxonID` therefore should contain to a globally accessible URI that maps to the precise taxon definition at the time of the description or measurement (e.g.
+The field `taxonID` therefore should contain a globally accessible URI that maps to the precise taxon definition at the time of the description or measurement (e.g.
 for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') .
-Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambigueities with taxonomic homonyms. 
+Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambiguities with taxonomic homonyms. 
 Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
-This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorization (Shan et al 2012, Schrodt et al 2015).
+This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorization [@schrodt15](Shan et al 2012, Schrodt et al 2015).
 
@@
 A similar approach to standardised trait names (`traitNameStd`) and harmonized reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
-Similar as for taxonID, the field  `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitionsor a measurement methodology.
+Similar as for taxonID, the field  `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitions or a measurement methodology.
 
@@
 The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards.
-This effort is time consuming and lengthy, but is returns synergies and great scientific benefit by enabling a global synthesis, as the example of a global plant trait database TRY shows intricately [@diaz16].
+This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of a global plant trait database TRY shows intricately [@diaz16].
 

### 2017-09-21 manuscript/manuscript.Rmd
Manuscript: changes to Introduction

@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->.
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
+<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->
+<!--Nadja: I would not mention the distinction between effect and response traits-->
+Traits have also proven useful for bridging individual level behaviour and physiology to processes at the ecosystem scale [@diaz13], for understanding physiological and evolutionary trade-offs, and for identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15].
+Inferring an unknown trophic role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e.
+the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.
+<!--Nadja: I am not sure if I understand what you mean with "have not been observed in the ecosystem".-->
 
-Early trait-based approaches have therefore been used to describe the role of a species in an ecosystem or its ability to persist under certain environmental conditions (refs).
+On the level of communities, community weighted means or variance of functional traits are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014].
+Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].
 
-Being assessed with a variety of research questions in mind, traits have proven useful in bridging individual level behaviour and physiology into processes at the ecosystem scale [@diaz13].
-The analysis of correlations between traits and intraspecific variation informs about the physiological and evolutionary trade-offs and helps to identify economic strategies of organisms beyond taxonomy [@menezes10; @forister15].
+Traits furthermore allow comparison across trophic levels.
+Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved.
+This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.
+<!--Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
+Nadja: I am not sure how this fits here as it is very specific. Maybe phrase it more generally?-->
 
-Inferring the unknown trophic role and ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.
-
-By looking at community weighted means and variance of functional traits, researchers aim to understand changes in functional composition or the loss of functional diversity and ecosystem services [@wood15; @bello10] in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10]  <!-- Gossner et al & other Explo papers -->.
-Since traits are closely related to the underlying ecological mechanisms, such patterns in functional community composition is promising versatile statistical tools to assess the current state of an ecosystem (@refs).
-
-Traits furthermore allow comparision across trophic levels.
-Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved.
-This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].
-This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.
-
-Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
-A global synthesis of traits brings major insights in ecosystem processes and ecosystem services, enables projections in future climate and land use scenarios.
-Traits link ecological research to evolution, and management (functions and services) for biodiversity conservation (land use intensification), climate research (environmental filters).
+A global synthesis of traits can bring major insights in ecosystem processes and enable projections in future climate and land use scenarios.
+Traits link ecological research to evolution, to management (functions and services) for biodiversity conservation (land use intensification), and to climate research (environmental filters).
+<!--Nadja: Not sure if my changes make sense. I also think that the brackets make this sentence too complicated.-->
 It is thus an interdisciplinary/cross-topic field with great potential.
 
-In recent years, functional ecology has thus grown into a major field in ecological research!
-However, an integration across trophic levels, taxa and organisational scales remains difficult because of different organismal, methodological, and functional focus. 
-A harmonization of trait data standards would be of huge benefit for our understanding of ecosystems.
+<!--In recent years, functional ecology has thus grown into a major field in ecological research!
+Nadja: We don't really need this sentence.-->
+
+However, an integration of several studies across trophic levels, taxa or organisational scales remains difficult because of different organismal, methodological, and functional focus.
+A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and make data use possible for future generations.
 
@@
 <!-- Malte: I agree, Caterina's suggestion makes it easier to follow for the reader. -->
+<!-- Nadja: I think the section above is now quite close to Caterinas suggestion-->
 
@@
 
-Depending on the organism groups and research questions, trait data have been standardised into centralised databases.
-Most prominently, plant traits from many different sources, including the extenisve LEDA traitbase for the Northwest-European flora [@kleyer08], for plant traits have been collated into the quickly growing TRY database [@kattge11].
-TRY collates trait data from other published databases with individually owned datasets and provides an extended access management and data query services.
-Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] for root measurements [Fine-root Ecology Database; @iversen17].
-The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage.
+Trait data have already been standardised into centralised databases for specific organism groups and research questions.
+Most prominently, plant traits from many different sources --including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08]-- have been collated into the quickly growing TRY database [@kattge11].
+TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications.
+Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] or root measurements [Fine-root Ecology Database; @iversen17].
+The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme.
+Repeating this exercise for other organism groups or other trait-based research questions would be desireable.
+
+Microbial ecology also makes great use of trait-based approaches to asses genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->
 
@@
 For instance, The BETSI database collects traits on soil invertebrats (http://betsi.cesab.org/).
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades.
-However, a unified platform and harmonizing scheme for the datasets is still lacking.
-
-Microbial ecology also makes great use of trait-based approaches to asses genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->
 
@@
 *< Heterogeneity of data standards and centralised databases >*
+<!-- Mandatory Open Access publication still seems to prevent some data providers from uploading their data, why many databases encourage a permissive data policy but do not enforce it. -->
+<!-- Malte: I think this definitely needs to be backed up with a reference. -->
+<!-- Nadja: I don't think we need to talk about the issue of data not being published openly because we won't fix this just by providing a commont trait data standard -->
 
-It becomes apparent, that all databases come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus.
-Certainly, a harmonization of trait data can be achieved by services like TRY, which take a huge effort to attract data submissions and map them into a common scheme.
-Furthermore, these databases provide access management services and tracking of data usage in studies and secondary publications.
-Mandatory Open Access publication still seems to prevent some data providers from uploading their data, why many databases encourage a permissive data policy but do not enforce it.
-<!-- Malte: I think this definitely needs to be backed up with a reference. -->
+<!-- Centralised databases like TRY experience rapid growth but may at some point reach a resource limit in personnel and funding.
+Nadja: Not sure where this fits anymore -->
 
-Centralised databases like TRY experience rapid growth but may at some point reach a resource limit in personnel and funding.
-Repeating this exercise for other organism groups or other trait-based research questions would be desireable.
-However, it has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits.
-[refs]
+Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades.
+It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus.
+Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts produced for plant traits (@refs).
 
-On the other hand, public research funding agencies increasinly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand.
-The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].
+Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand.
+Additionally, the training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].
 At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
@@
 Such ontologies are being developed for plant traits, for instance in the TOP Thesaurus of plant traits that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
-On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches  the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].
+On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches or the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].
 

### 2017-09-21 manuscript/manuscript.Rmd
Manuscript: Changes to main text. Include first version of table 1.

@@
 
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades.
+Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g.
+@Gossner2015SData,@Gossner2015Ecology,@Renner2014].
 It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus.
@@
 A trait-dataset contains quantitative measurements or qualitative facts (i.e.
-trait values) <!-- Cat: the difference is not totally clear to me - is it the difference between field vs literature gathered data?
-Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs.
-literature, but not entirely.
-A differentiation that has rarely been specified in the trait-litereature it seems.
-Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further.
--->  about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
-traits) observed on an of individual or parts of an individual (i.e.
+trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
+traits) observed on an individual or parts of an individual (i.e.
 an entity) that can be assigned to a biological taxon (i.e.
@@
 
+<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?
+Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs.
+literature, but not entirely.
+A differentiation that has rarely been specified in the trait-litereature it seems.
+Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further.
+Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive).
+Maybe it helps if we add an example after the definition?-->
+
 *< data structure >*
 
-Trait-datasets take different formats.
+Trait-datasets usually take different formats.
 For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
@@
 on variation of means or literature source) in secondary colums.
-The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level (@laliberte10).
+The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10].
 Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).
 
-For investigations of phenotypic variation of traits,  would be recorded per observation, i.e.
+Phenotypic variation of traits within species would usually be recorded per observation, i.e.
 each individual specimen would be recorded in a row, with the respective trait values stored in columns.
@@
 
-Computationally most effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
+Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns.
+Computationally more effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
 a specimen) via a unique identifier which is assigned to a taxon.
@@
 <!-- Malte: I think the difference between the long-table format and the previous one is not clear enough.
+Nadja: Is this better now?
 -->
-The latter format thus provides the highest resolution and ideal for storing raw data of trait measurements in huge databases and link them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
+The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
 
@@
 While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we want to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
-In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of trait-datasets proposed above.
+In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of trait-datasets proposed above (Table 1).
 We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
@@
 We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.
-The full list of terms of the traitdata standard including further explanation can be assessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
+The full list of terms of the traitdata standard including further explanation can be acessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
 
-(table 2: most relevant terms of the traitdata standard, scheme of a term definition)
+Table 1: Minimal set of terms for the traitdata standard.
+valueType can be of type character, factor, boolean or numerical.
+vocabulary is only needed for factorial trait data and defines the factor levels which are allowed.
+The identifier is a unique and globally valid URI.
+
+| Term | valueType | vocabulary | Identifier | Definition | 
+|------|-----------|------------|------------|------------|
+|scientificName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificname |Original character string provided as species name by the data owner (kept for reference only)|
+|taxonID | factor | | http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonid | Verified name identifier of the species or subspecies (or higher taxon) for which this value was collected.
+Each entry should linkt to a published reference species list/taxonomy.
+Avoid synonyms.|
+|taxonRank | factor | domain; kingdom; subkingdom; superhpylum [...] |http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonrank | The taxonomic rank of the most specific name in the scientificName.
+Recommended best practice is to use a controlled vocabulary.|
+|kingdom | factor | |http://fdschneider.de/bexis_traits/traitdatastandard.html#kingdom | The full scientific name of the kingdom in which the taxon is classified.| 
+|traitName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#traitname |Name of the trait as used by the data provider (kept for compatibility on provider side).|
+|traitID|factor|see traitlist|http://fdschneider.de/bexis_traits/traitdatastandard.html#traitid ||
+|traitValue|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvalue |Measured raw value or factor level for this species trait as measured and provided by the author.|
+|traitUnit|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunit |Reports the unit that the author's raw data were measured in, if applicable (only for numeric values).|
+|scientificNameStd|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificnamestd |Full scientific name (with authorship and date information if known).|
+|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive Name of the trait reported.
+This should follow the controlled vocabulary of a thesaurus of traits available online or published along with the dataset.|
+|traitValueStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvaluestd |Standardised measured value or factor level for this species trait.
+Numerical values must use the correct unit.
+Factor levels must be compliant with the thesaurus of traits.|
+|traitUnitStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunitstd |The units associated with the measurementValue.
+The expected standard unit for this trait as defined in the traitlist/ thesaurus.|
+|measurementID|	character||	http://fdschneider.de/bexis_traits/traitdatastandard.html#measurementid |A unique, dataset-specific identifier.
+Links multi-value trait measurements, e.g.
+x-y-z coordinates of a morphometric landmark, biochemical compound quantities for different chainlengths.
+In this case, the trait names must specifiy the sub-measurement, e.g.
+"landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
+|occurenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurenceid|A character string user defined by the data author.
+It is recommended best practice to use a globally unique identifier, if available, e.g.
+a museum ID referring to the individual specimen from which the data were measured.
+usually does not apply for literature data.|
+|warnings|character|	|	|warnings from the R-script will be stored here, e.g.
+ragarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table.
+User defined warnings and flags can be added as well, e.g.
+'NOTUSE'.|
+
+<!-- Nadja: Is this the kind of table you had in mind? -->
 
@@
 
-For the minimal definition of trait-datasets posed above, the core content of a row is the reported measurement or fact for a single observation, which is composed of a value (`traitValue`) and -- for numeric values -- a standard unit (`traitUnit`).
-To associate the measurement or fact to a clear trait definition, a unique identifier links each row to a descriptive trait name (`traitName`).
+The core content of a row in the trait-dataset is the reported measurement or fact for a single observation, which is composed of a value (`traitValue`) and -- for numeric values -- a standard unit (`traitUnit`).
+To associate the measurement or fact to a clear trait definition, a unique identifier (`traitID`) links each row to a descriptive trait name (`traitName`).
 
@@
 Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right.
-We propose to add a second set of columns that contain standardized entries of the taxon name (`scientificNameStd`) that maps synonyms to an accepted species name according to a published taxonomic ontology (e.g.
+We propose to add a second set of columns that contain standardized entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
 GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`).
@@
 A similar approach to standardised trait names (`traitNameStd`) and harmonized reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
-Similar as for taxonID, the field  `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitions or a measurement methodology.
+Similar as for `taxonID`, the field  `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitions or a measurement methodology.
 
@@
 However, for most organism groups and research contexts we attest a deficit of public ontologies with openly accessible URIs.
-
-Therefore, if no published trait definition is available that can be referenced, traitdatasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
+Therefore, if no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
 In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte12; @garnier17].
@@
 -->
+<!-- Nadja: I agree that this section is not needed in the paper, but it would certainly be nice to publish it in a supplement or with the traidata standard -->
 
@@
 The measurement may also be referred to a single specimen or occurrence via a uniue identifier (`occurrenceID`).
-For most measured trait data (see above) this is a vital information that links related measurements of different traits taken on the same specimen.
+For most measured trait data this is a vital information that links related measurements of different traits taken on the same specimen.
 This identifier is usually dataset-specific and can be defined by the author and described with further detail in a separate data table using the columns provided in the occurrence extension detailed below.
@@
 -->
+ <!-- Nadja: I like the idea of having a figure explaining the different terms -->
 
@@
 
-To keep track potential sources of noise or bias in measured data, the method of measurement (`measurementMethod`), the person conducting the measurement (`measurementDeterminedBy`), and the date at which the measurement was obtained (`measurementDeterminedDate`) are recorded.
+To keep track of potential sources of noise or bias in measured data, the method of measurement (`measurementMethod`), the person conducting the measurement (`measurementDeterminedBy`), and the date at which the measurement was obtained (`measurementDeterminedDate`) are recorded.
 Authors would often report aggregate data of repeated or pooled measurements, e.g.
@@
 The field definitions of the Darwin Core Standard can be applied instead, to refer to a geological stratum, for instance.
-Sampling may be further specified using a unique identifier for the sampling event `eventID`	which references to an external dataset.
+Sampling may be further specified using a unique identifier for the sampling event (`eventID`) which references to an external dataset.
 The record of a `samplingProtocol` may capture bias in samling methods.
@@
 # References
-
-\footnotesize
-
-Allan, E., Manning, P., Alt, F., Binkenstein, J., Blaser, S., Bluthgen, N., Bohm, S., Grassein, F., Holzel, N., Klaus, V.H.,
-Kleinebecker, T., Morris, E.K., Oelmann, Y., Prati, D., Renner, S.C., Rillig, M.C., Schaefer, M., Schloter, M., Schmitt, B.,
-Schoning, I., Schrumpf, M., Solly, E., Sorkau, E., Steckel, J., Steffen-Dewenter, I., Stempfhuber, B., Tschapka, M.,
-Weiner, C.N., Weisser, W.W., Werner, M., Westphal, C., Wilcke, W., Fischer, M., 2015. Land use intensification alters
-ecosystem multifunctionality via loss of biodiversity and changes to functional composition. Ecol Lett. doi:
-10.1111/ele.12469
-
-Aubin, I., Venier, L., Pearce, J., Moretti, M., 2013. Can a trait-based multi-taxa approach improve our assessment of
-forest management impact on biodiversity? 22, 2957-2975. doi: 10.1007/s10531-013-0565-6
-
-Birkhofer, K., Diekötter, T., Meub, C., Stötzel, K., Wolters, V., 2015. Optimizing arthropod predator conservation in
-permanent grasslands by considering diversity components beyond species richness. 211, 65-72. doi: 10.1016/j.agee.2015.05.014
-
-Birkhofer, K., Smith, H.G., Weisser, W.W., Wolters, V., Gossner, M.M., 2015. Land-use effects on the functional
-distinctness of arthropod communities. Ecography 38, 889-900. doi: 10.1111/ecog.01141
-
-Börschig, C., Klein, A.M., von Wehrden, H., Krauss, J., 2013. Traits of butterfly communities change from specialist
-to generalist characteristics with increasing land-use intensity. Basic Appl Ecol 14, 547-554. doi: 10.1016/j.baae.2013.09.002
-
-Chisté, M.N., Mody, K., Gossner, M.M., Simons, N.K., Köhler, G., Weisser, W.W., Blüthgen, N., 2016. Losers, winners,
-and opportunists: How grassland land-use intensity affects orthopteran communities. 7, e01545. doi: 10.1002/ecs2.1545
-
-Entling, W., Schmidt, M.H., Bacher, S., Brandl, R., Nentwig, W., 2007. Niche properties of Central European spiders:
-shading, moisture and the evolution of the habitat niche. Global Ecol Biogeogr 16, 440-448. doi: 10.1111/j.1466-8238.2006.00305.x
-
-Gamez-Virues, S., Perovic, D.J., Gossner, M.M., Borschig, C., Bluthgen, N., de Jong, H., Simons, N.K., Klein, A.M.,
-Krauss, J., Maier, G., Scherber, C., Steckel, J., Rothenwohrer, C., Steffan-Dewenter, I., Weiner, C.N., Weisser, W.,
-Werner, M., Tscharntke, T., Westphal, C., 2015. Landscape simplification filters species traits and drives biotic
-homogenization. 6, 8568. doi: 10.1038/ncomms9568
-
-Gossner, M.M., Lewinsohn, T.M., Kahl, T., Grassein, F., Boch, S., Prati, D., Birkhofer, K., Renner, S.C., Sikorski, J.,
-Wubet, T., Arndt, H., Baumgartner, V., Blaser, S., Bluthgen, N., Borschig, C., Buscot, F., Diekotter, T., Jorge, L.R.,
-Jung, K., Keyel, A.C., Klein, A.M., Klemmer, S., Krauss, J., Lange, M., Muller, J., Overmann, J., Pasalic, E., Penone, C.,
-Perovic, D.J., Purschke, O., Schall, P., Socher, S.A., Sonnemann, I., Tschapka, M., Tscharntke, T., Turke, M., Venter,
-P.C., Weiner, C.N., Werner, M., Wolters, V., Wurst, S., Westphal, C., Fischer, M., Weisser, W.W., Allan, E., 2016.
-Land-use intensification causes multitrophic homogenization of grassland communities. 540, 266-269. doi:
-10.1038/nature20575
-
-Gossner, M.M., Simons, N.K., Achtziger, R., Blick, T., Dorow, W.H.O., Dziock, F., Köhler, F., Rabitsch, W., Weisser,
-W.W., 2015. A summary of eight traits of Coleoptera, Hemiptera, Orthoptera and Araneae, occurring in grasslands
-in Germany. Scientific Data 2, 150013. doi: 10.1038/sdata.2015.13
-
-Gossner, M.M., Simons, N.K., Höck, L., Weisser, W.W., 2015. Morphometric measures of Heteroptera sampled in
-grasslands across three regions of Germany. 96, 1154. doi: 10.1890/14-2159.1
-
-Renner, S.C., Gossner, M.M., Kahl, T., Kalko, E.K., Weisser, W.W., Fischer, M., Allan, E., 2014. Temporal changes in
-randomness of bird communities across Central Europe. Plos One 9, e112347. doi: 10.1371/journal.pone.0112347
-
-Simons, N.K., Gossner, M.M., Lewinsohn, T.M., Boch, S., Lange, M., Muller, J., Pasalic, E., Socher, S.A., Turke, M.,
-Fischer, M., Weisser, W.W., 2014. Resource-mediated indirect effects of grassland management on arthropod
-diversity. PLoS One 9, e107033. doi: 10.1371/journal.pone.0107033
-
-Simons, N.K., Weisser, W.W., Gossner, M.M., 2016.
-Multi-taxa approach shows consistent shifts in arthropod functional traits along grassland land-use intensity gradient.
-Ecology 97, 754-764.
-doi: 10.1890/15-0616.1
-
-Weiner, C.N., Werner, M., Linsenmair, K.E., Bluthgen, N., 2014. Land-use impacts on plant-pollinator networks:
-interaction strength and specialization predict pollinator declines. Ecology 95, 466-474.
-
-Weiner, C.N., Werner, M., Linsenmair, K.E., Blüthgen, N., 2011. Land use intensity in grasslands: Changes in
-biodiversity, species composition and specialisation in flower visitor networks. 12, 292-299. doi: 10.1016
-j.baae.2010.08.006

### 2017-09-22 manuscript/manuscript.Rmd
Comments and edits on the Introduction

@@
 <!--Nadja: I am not sure if I understand what you mean with "have not been observed in the ecosystem".-->
+<!--Malte: I guess, you mean their behaviour has not been observed, i.e.
+might know what they look like (what traits they have) because we caught / found individuals, but we have no idea what they do, where / how they live, with whom they interact etc.
+How about "...
+are yet undescribed and their behaviour, their interactions with other organisms and the environment and their ecological function are virtually unknown."-->
 
@@
 Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].
+<!--Malte: It is not only CWM's and variance of traits, but also functional diversity metrics that are commonly used to understand change sin functional composition. I would probably add that here.-->
 
@@
 Nadja: I am not sure how this fits here as it is very specific. Maybe phrase it more generally?-->
+<!--Malte: First sentence, 'comparison across trophic levels".
+Not sure if this is the ideal wording; is it really comparisons that they allow for here?
+What about "Functional traits furthermore are an important tool for studying interactions across trophic levels." -->
 
-A global synthesis of traits can bring major insights in ecosystem processes and enable projections in future climate and land use scenarios.
-Traits link ecological research to evolution, to management (functions and services) for biodiversity conservation (land use intensification), and to climate research (environmental filters).
+Traits link ecological research to evolution, management (functions and services) for biodiversity conservation (land use intensification), and climate research (environmental filters).
+As such, a global synthesis of traits would offer major insights in ecosystem processes and enable projections in future climate and land use scenarios.
 <!--Nadja: Not sure if my changes make sense. I also think that the brackets make this sentence too complicated.-->
-It is thus an interdisciplinary/cross-topic field with great potential.
+Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential.
 
@@
 
-However, an integration of several studies across trophic levels, taxa or organisational scales remains difficult because of different organismal, methodological, and functional focus.
-A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and make data use possible for future generations.
+Despite the obvious benefit of such global trait synthesis, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus.
+A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and enable efficient data use for future generations.
 
@@
 
-Summarize different questions, currently proposed in above- and below ground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.
+Summarize different questions, currently proposed in above- and belowground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.
 
@@
 The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme.
-Repeating this exercise for other organism groups or other trait-based research questions would be desireable.
+Repeating this exercise for other organism groups or other trait-based research questions would be desirable.
 
-Microbial ecology also makes great use of trait-based approaches to asses genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->
+Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->
 
-In the animal kingdom, specialised trait-databases have been created that cover different organism groups or interaction types or ecosystems.
-For instance, The BETSI database collects traits on soil invertebrats (http://betsi.cesab.org/).
+In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
+For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/).
 
@@
 
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g.
-@Gossner2015SData,@Gossner2015Ecology,@Renner2014].
+Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g., @Gossner2015SData,@Gossner2015Ecology,@Renner2014].
 It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus.
-Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts produced for plant traits (@refs).
+Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts made for plant trait databases (@refs).
 
@@
 At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
-Given this prospect, there will be no lack of data, but a lack of data standardization.
+Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.
 
@@
 
-Such ontologies are being developed for plant traits, for instance in the TOP Thesaurus of plant traits that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
+Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
 On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches or the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].
@@
 an entity) that can be assigned to a biological taxon (i.e.
-a species or higher taxon).
+a species).
 The entity or observation (i.e.
@@
 Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation and sexual dimorphism.
-We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species.
-This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.
+We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species.<!-- Malte: ...
+Or a single measurement, right?
+--> This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.
 
@@
 Nadja: Is this better now?
+Malte: Yes, it is.
 -->
@@
 
+M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
+
 # Authors' contributions

### 2017-09-22 manuscript/manuscript.Rmd
Edits and comments on the 2nd part of the ms

@@
 
-While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we want to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
+While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we aim to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
 In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of trait-datasets proposed above (Table 1).
@@
 We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.
-The full list of terms of the traitdata standard including further explanation can be acessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
+The full list of terms of the traitdata standard including further explanation can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
 
@@
 |taxonID | factor | | http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonid | Verified name identifier of the species or subspecies (or higher taxon) for which this value was collected.
-Each entry should linkt to a published reference species list/taxonomy.
+Each entry should link to a published reference species list/taxonomy.
 Avoid synonyms.|
@@
 |scientificNameStd|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificnamestd |Full scientific name (with authorship and date information if known).|
-|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive Name of the trait reported.
+|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive name of the trait reported.
 This should follow the controlled vocabulary of a thesaurus of traits available online or published along with the dataset.|
@@
 "landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
-|occurenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurenceid|A character string user defined by the data author.
+|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurenceid|A character string user defined by the data author.
 It is recommended best practice to use a globally unique identifier, if available, e.g.
 a museum ID referring to the individual specimen from which the data were measured.
-usually does not apply for literature data.|
-|warnings|character|	|	|warnings from the R-script will be stored here, e.g.
-ragarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table.
+Usually does not apply for literature data.|
+|warnings|character|	|	|Warnings from the R-script will be stored here, e.g.
+regarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table.
 User defined warnings and flags can be added as well, e.g.
@@
 This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
-Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right.
-We propose to add a second set of columns that contain standardized entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
+Taxonomic standardisation, with respect to ongoing renamings and refinements, is a challenging issue in its own right.
+We propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
 GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`).
@@
 Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
-This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorization [@schrodt15](Shan et al 2012, Schrodt et al 2015).
+This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15](Shan et al 2012, Schrodt et al 2015).
 
@@
 
-A similar approach to standardised trait names (`traitNameStd`) and harmonized reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
+A similar approach to standardised trait names (`traitNameStd`) and harmonised reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
 Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitions or a measurement methodology.
@@
 For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@Ref semantic database methods).
-This extends the trait list into a semantic web resource and facilitates the classification of traits and allows cross-taxon comparative studies at the level of broader terms.
+This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms.
 (more detail on semantic web ontologies can be found ...)
@@
 
-Numerical values represent measurements of length, volumes, ratios, rates or timespans. Integer values may apply to count data (e.g. eggs per clutch). 
+Numerical values represent measurements of lengths, volumes, ratios, rates or timespans. Integer values may apply to count data (e.g. eggs per clutch). 
 Binary data (encoded as 0 or 1) or logical data (coded as TRUE or FALSE) may apply to qualitative traits such as specific behaviour during mating (e.g.
@@
 they have a logical sequence as in the case of life stages or hibernation stages, or habitat preference traits such as horizontal stratum use.
-Finally there are specific formats of multivariate trait values, e.g.
-x-y-z coordinates of a landmark measured in a standardized 3D space [ref] or relative abundance of chain-lengths in biochemical compounds [].
-To cope with this variety of data types, definitions should refer to other well-defined terms of other ontologies that describe standard units, morphological body parts, protein characteristics (Protein Ontology) or chemical terms (e.g.
+Finally, there are specific formats of multivariate trait values, e.g.
+x-y-z coordinates of a landmark measured in a standardised 3D space [ref] or relative abundance of chain-lengths in biochemical compounds [].
+To cope with this variety of data types, definitions should refer to other well-defined terms of existing ontologies that describe standard units, morphological body parts, protein characteristics (Protein Ontology) or chemical terms (e.g.
 the ChEBI, http://www.obofoundry.org/ontology/chebi.html).
@@
 -->
-<!-- Nadja: I agree that this section is not needed in the paper, but it would certainly be nice to publish it in a supplement or with the traidata standard -->
+<!-- Nadja: I agree that this section is not needed in the paper, but it would certainly be nice to publish it in a supplement or with the traidata standard. Malte: Agreed. -->
 
@@
 each row of the dataset).
-Those information might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset.
+These details might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset.
 The terms for this layer of information are defined by the measurement or fact extension.
@@
 Additionally, information about the authorship and ownership of the data and the terms of use should be considered.
-These information may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).
+This information may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).
 
@@
 Individual measurements and facts will likely already belong to a larger set of traits or a separate trait database before they are added to a combined trait database.
-To retain the rights of the original data contributor, the field `rightsHolder` states the person or organization who owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and `license` specifies under which terms and conditions the data can be used, re-used and/or published.
+To retain the rights of the original data contributor, the field `rightsHolder` states the person or organisation who owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and `license` specifies under which terms and conditions the data can be used, re-used and/or published.
 This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`,`authorLastname` and `authorFirstname`.
@@
 - preparation of own trait datasets for upload into public data bases, and
-- harmonizing trait datasets from different sources by moulding them into a unified format.
+- harmonising trait datasets from different sources by moulding them into a unified format.
 
@@
 
-The `standardize.traits()` function matches user provided trait names onto public trait ontologies (for now using GBIF Backbone Taxonomy) and links it to a trait ontology or a user-provided table of trait definitions, i.e.
+The `standardize.traits()` function matches user-provided trait names onto public trait ontologies (for now using GBIF Backbone Taxonomy) and links it to a trait ontology or a user-provided table of trait definitions, i.e.
 a thesaurus of traits.
-The same function harmonizes trait values into target units and legit factor levels.
+The same function harmonises trait values into target units and legit factor levels.
 
 The output of these functions can easily be merged using the `rbind()` function of R base. 
+<!-- Malte: Shouldn't this be cbind() ? These functions do not add additional rows but additional columns, right?-->
 
 The package functions comprise a toolchain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
+A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonising traitdata as simple as:
 
@@
 We further encourage the development of trait definitions in semantic ontologies.
-The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards.
-This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of a global plant trait database TRY shows intricately [@diaz16].
+The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards.
+This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of the global plant trait database TRY shows intricately [@diaz16].
 
-These tools help to standardize trait datasets before upload to central as well as distributed data repositories and to harmonize and combine raw data from heterogeneous sources.
+These tools help to standardize trait datasets before upload to central as well as distributed data repositories and to harmonise and combine raw data from heterogeneous sources.
 We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@ref].
@@
 Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
-Ernst-Detlef Schulze, Wolfgang W.Weisser and the late Elisabeth Kalko
+Ernst-Detlef Schulze, Wolfgang W. Weisser and the late Elisabeth Kalko
 for their role in setting up the Biodiversity Exploratories project.
@@
 M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
+<!-- Malte: I just started adding personal funding sources here. -->
 

### 2017-09-25 manuscript/manuscript.Rmd
major edits on manuscript

@@
 title: "Introducing an Ecological Trait-data Standard"
-author: "Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas Ostrowski, Caterina Penone"
+author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas
+  Ostrowski, Caterina Penone
 csl: amnat.csl
 output:
+  pdf_document: default
   word_document:
     keep_md: yes
-  pdf_document: default
 geometry: margin=1.5in
@@
 
+# Abstract
+
+1.	The assessment and analysis of functional traits has been established in all domains of ecological research promising great advancements for biodiversity and climate research. 
+2.	Trait data are stored in heterogeneous formats, measured at different resolutions and using different methods, comprising literature and measured data, and focused on a variety of research questions and organism groups.
+3.	To harmonize and standardize trait data, we present a first version of an ecological trait-data standard for storing and sharing functional trait data.
+4.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
+5.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+
 # Introduction
 
-*< importance of trait based ecology, current research questions, current applications of traits. >*
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
-<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->
-<!--Nadja: I would not mention the distinction between effect and response traits-->
-Traits have also proven useful for bridging individual level behaviour and physiology to processes at the ecosystem scale [@diaz13], for understanding physiological and evolutionary trade-offs, and for identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15].
-Inferring an unknown trophic role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.
-<!--Nadja: I am not sure if I understand what you mean with "have not been observed in the ecosystem".-->
-<!--Malte: I guess, you mean their behaviour has not been observed, i.e.
-might know what they look like (what traits they have) because we caught / found individuals, but we have no idea what they do, where / how they live, with whom they interact etc.
-How about "...
-are yet undescribed and their behaviour, their interactions with other organisms and the environment and their ecological function are virtually unknown."-->
+Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13].
+Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15].
+Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e.
+the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
+For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits.
+This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].
 
-On the level of communities, community weighted means or variance of functional traits are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014].
+On the level of species assemblages, community weighted means or variance of functional traits or multivariate measures of 'functional diversity' [@petchey??] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @bello10; @hopfenmuller14; @jennings09; @moretti13;  @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014].
 Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].
-<!--Malte: It is not only CWM's and variance of traits, but also functional diversity metrics that are commonly used to understand change sin functional composition. I would probably add that here.-->
+Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.
 
-Traits furthermore allow comparison across trophic levels.
-Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved.
-This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.
-<!--Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
-Nadja: I am not sure how this fits here as it is very specific. Maybe phrase it more generally?-->
-<!--Malte: First sentence, 'comparison across trophic levels".
-Not sure if this is the ideal wording; is it really comparisons that they allow for here?
-What about "Functional traits furthermore are an important tool for studying interactions across trophic levels." -->
-
-Traits link ecological research to evolution, management (functions and services) for biodiversity conservation (land use intensification), and climate research (environmental filters).
-As such, a global synthesis of traits would offer major insights in ecosystem processes and enable projections in future climate and land use scenarios.
-<!--Nadja: Not sure if my changes make sense. I also think that the brackets make this sentence too complicated.-->
+In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.
 Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential.
 
-<!--In recent years, functional ecology has thus grown into a major field in ecological research!
-Nadja: We don't really need this sentence.-->
+## Overview of trait-data initiatives
 
-Despite the obvious benefit of such global trait synthesis, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus.
-A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and enable efficient data use for future generations.
+Despite the obvious benefit, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits heterogeneous.
+Trait data have therefore been standardised into centralised databases for specific organism groups and research questions.
+These services provide platforms for consensus building in methods, attract data submissions from a focal community of researchers and map them into a common scheme.
+Furthermore, the databases provide access management services and provide data usage policies, while simplifying data queries for synthesis researchers.
 
-<!-- 
-Flo: This section should probably include some more exploratories papers.
+Most prominently, plant traits from many different sources -- including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08] -- have been collated into the quickly growing TRY database [@kattge11].
+Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17].
+Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebaur17].
+Trait-data for lower plants are less completely assessed.
+<!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications.
+The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme.
+Repeating this exercise for other organism  -->
 
-Traits can be measurements of organism morphology, life-history characteristics such as reproductive strategies, physiological traits including metabolism and photosynthetic activity, feeding traits, biochemical and isotopic compounds, behavioural traits, as well as environmental traits.
-
-Refer to Kattge et al 2011, Violle et al 2007, Diaz et al 2007, McGill et al. 2006, Aubin et al. 2013, Lavorel et al. 2002...
-## trait-based research within the Biodiversity Exploratories
-
-Refer to successfully published papers on arthropod traits, multi-trophic trait variation etc.
-
-Allan et al 2015
-Boerschig et al 2013
-Gossner et al 2015 & 2016
-Simons et al 2014 & 2016
-Kühsel et al 2015 & 2016
-Birkhofer et al 2015
-Chisté et al 2016
-Gamez-Virues et al 2015
-
-Summarize different questions, currently proposed in above- and belowground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.
-
-Need to unify for multi-trophic integration
-
-Trait matching integrates function across trophic levels, influence of land use and environmental filtering on trait composition, Exploratories are ideal framework for testing
-
--- Cat: here I would organize a bit differently to link better to the previous section:--
-Traits can bridge biodiversity and functioning and that they allow comparision across trophic levels. Furthermore trait composition is influenced by land use and environmental filtering of traits. 
-Exploratories are ideal framework for testing these questions because they are framed to study interactions between LUI, biodiversity and functions. 
-However, multi-trophic integration is difficult because of different standards, methods, focus..
- needs unification in order to advance research and make data use possible for future generations
-
-<!-- Nadja: I like Caterinas suggestion --> 
-<!-- Flo: Yes, makes sense. -->
-<!-- Malte: I agree, Caterina's suggestion makes it easier to follow for the reader. -->
-<!-- Nadja: I think the section above is now quite close to Caterinas suggestion-->
-
-*< Overview of existing database initiatives. Offering access management and harmonization >*
-
-Trait data have already been standardised into centralised databases for specific organism groups and research questions.
-Most prominently, plant traits from many different sources --including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08]-- have been collated into the quickly growing TRY database [@kattge11].
-TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications.
-Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] or root measurements [Fine-root Ecology Database; @iversen17].
-The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme.
-Repeating this exercise for other organism groups or other trait-based research questions would be desirable.
-
-Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->
+Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14].<!-- more on microbial databases -->
 
 In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
-For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/).
+For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/), or the Carabids.org web portal (http://www.carabids.org/).
+Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades[e.g., @Gossner2015SData, @Gossner2015Ecology, @Renner2014].
+However, a unified platform and harmonizing scheme for the datasets is still lacking.
+It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits.
+[refs]
 
-<!-- Flo: I would leave this open spreadsheet out for now.
-It would be quite a lot of effort to collect something representative.
-It might be an exercise for the exploratories people working on traits, for a start.
-In principle, any open datasets could be linked and imported into the R-package.
-I will highlight this possibility when talking about the package.
--->
-<!-- As an attempt to track these unstandardized datasets, we initiated a living spreadsheet registry which invites submissions of any open data (Google Spreadsheet link, table 1).
+A global overview of existing trait data for all taxa and trait types is difficult to obtain.
+As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus, and extend of traits covered, their location on the internet and the terms of use.
+We invite data owners and users to register further datasets qualifying as trait data.
 
-(table of existing databases? name, url, focus, number of records, openness of data, maintained by) -->
+## Need for ontologies for data standardisation
 
-<!-- Caterina: I like this table idea but should we limit it to plants and arthropods or also other organisms?
-Europe or whole world?
-I would opt for plants and arthropods in Europe but in case this becomes a publication open a collaborative file where anybody in the world can add the information for any organism in any part of the world.
-
-Flo: I like this idea of starting a collaborative list on trait datasets. This
-   1. illustrates the hetereogeneity of trait data and the need for a traitdata standard
-   2. encourages a global traitdata analysis based on heterogeneous sources, which is enabled by the tools we provide here. See https://github.com/fdschneider/bexis_traits/issues/20
-   -->
-
-*< Heterogeneity of data standards and centralised databases >*
-<!-- Mandatory Open Access publication still seems to prevent some data providers from uploading their data, why many databases encourage a permissive data policy but do not enforce it. -->
-<!-- Malte: I think this definitely needs to be backed up with a reference. -->
-<!-- Nadja: I don't think we need to talk about the issue of data not being published openly because we won't fix this just by providing a commont trait data standard -->
-
-<!-- Centralised databases like TRY experience rapid growth but may at some point reach a resource limit in personnel and funding.
-Nadja: Not sure where this fits anymore -->
-
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g., @Gossner2015SData,@Gossner2015Ecology,@Renner2014].
-It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus.
-Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts made for plant trait databases (@refs).
-
-Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand.
-Additionally, the training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].
-At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
+Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15].
+However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits are compliant with this demand for open data.
+<!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].--> Open-access publications in low-threshold data repositories such as Figshare, datadryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
 Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.
@@
 GFBio is aiming to ease data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.
+Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
+On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches (T-SITA, http://t-sita.cesab.org/, @pey14).
 
-Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
-On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches or the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].
+While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored.
+Such a general data structure for quantitative trait data has been developed within the Encyclopedia of Life Project for upload of data into TraitBank [@parr16], relying on definitions of terms provided by the Darwin Core Standard (DWC, ...).
+However, the Terms of the Darwin Core Standard and available Extensions do not cover all of the aspects required in the various use-cases of trait-based research.
 
-With this paper, we propose a roadmap towards a comprehensive trait data standard, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+# Proposing an ecological traitdata standard
+
+With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+
+We propose a versatile data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data.
+We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
+
+With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL) TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC).
+We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.
 
@@
 
-*< data structure >*
-
 Trait-datasets usually take different formats.
-For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
+For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
 This format is usually reporting missing data as NA.
@@
 on variation of means or literature source) in secondary colums.
-The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10].
+The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
+Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10].
 Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).
@@
 each individual specimen would be recorded in a row, with the respective trait values stored in columns.
-Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation and sexual dimorphism.
-We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species.<!-- Malte: ...
-Or a single measurement, right?
---> This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.
+In these 'occurrence data tables', each row is centered around a single physical occurrence of a species.
+Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
+Occurence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
+This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.
 
 Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns.
-Computationally more effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
+Each trait column represents one measured value.
+
+Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
 a specimen) via a unique identifier which is assigned to a taxon.
@@
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.
-<!-- Malte: I think the difference between the long-table format and the previous one is not clear enough.
-Nadja: Is this better now?
-Malte: Yes, it is.
--->
+Longtable dataset purport multiple advantages for the data manipulation (e.g.
+filtering, subsetting and aggregating data), visualization (e.g.
+plot measured values by factor variable or taxon) and statistical modeling (e.g.
+ANOVA for testing difference of trait value by sex) [@wickham14].
+They further draw from a defined set of columns, which makes it easier to merge datasets.
+
 The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
+Each row of the dataset can be interpreted as a statement of an 'entity x having a quality y'.
+To comply with the EST, datasets must be reformatted to the longtable format.
 
@@
 
-# Proposing an ecological traitdata standard
+## common terms for trait data
 
-In this paper, we propose a universal scheme of defined column names and data structure that captures the different degrees of resolution and measurement detail for multiple use cases of trait data.
-We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
+In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset compliant with the EST, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above (Table 1).
+We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
+Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).
+The full list of terms  can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
 
-While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we aim to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
-In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of trait-datasets proposed above (Table 1).
-We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
-Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits.
-
-With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL) TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC).
-We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.
-The full list of terms of the traitdata standard including further explanation can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
+A dataset that applies these terms should relate to the EST in its metadata.
+Since the traitdata standard is a resource under open community development, the metadata also should refer to a version number.
+This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g.
+http://terminologies.gfbio.de/traitdatastandard/v1.0 !invalid URL!).
 
@@
 
-## minimal requirement of a trait-dataset
+## core elements of a trait-dataset
 
@@
 
-The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained via globally accepted identifiers.
-To provide an unambiguous reference which is easy to read for researchers and for software, this identifier is provided in the form of a descriptive taxon name (`scientificName`) as well as an unambiguous ID (`taxonID`).
+The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained in the form of a descriptive taxon name (`scientificName`).
 
-*< global taxonomy standards >*
+### relating to global taxonomy standards
 
+Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right.
 To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable.
-We consider the fact that most authors have their own schemes for standardisation and may contribute to different communities by allowing for a double record of both user-specific and standardised entries (as practiced in TRY, -@kattge11 ).
+We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11].
 This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
-Taxonomic standardisation, with respect to ongoing renamings and refinements, is a challenging issue in its own right.
-We propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
+We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
 GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`).
-The field `taxonID` therefore should contain a globally accessible URI that maps to the precise taxon definition at the time of the description or measurement (e.g.
+The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g.
 for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') .
@@
 
-*< relating data to trait definitions >*
+### relating data to trait definitions
 
@@
 -->
-Resources for semantic ontologies exist for plant traits (e.g., TOP Plant Trait Ontology [@garnier17]), soil invertebrates (T-Sita, @ref), or biomedical traits of vertebrates (Vertebrate Trait Ontology [@park13]).
-However, for most organism groups and research contexts we attest a deficit of public ontologies with openly accessible URIs.
-Therefore, if no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
+Resources for semantic ontologies exist for plant traits (e.g., TOP Plant Trait Ontology [@garnier17]), soil invertebrates (T-Sita, @pey14), or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13].
+However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.
+
+### Defining trait thesauri and ontologies
+
+If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
 In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte12; @garnier17].
-This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and a unique (alphanumeric) identifier (`traitID`) as well as an unambiguous verbal definition (`traitDescription`), the legit factor levels (for categorical data, `factorLevels`) or expected standard units (for numerical data).
+This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and define a unique dataset-specific identifier (`traitID`) as well as a short, unambiguous verbal definition (`traitDescription`) which may make use of standard terms provided in other Ontologies.
+As an Example,  the definition for 'fruit mass' in TOP reads: "the mass (PATO:mass), either fresh or dried, of a fruit (PO:fruit)", referring to Phenotypic Characeristics Ontology PATO and Planteome Plant Ontolgy, PO (http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass).
+The thesaurus should also constrain the legit factor levels (for categorical data, `factorLevels`) or expected standard units (`traitUnitStd` for numerical data).
+The type of values should be differentiated in the field `valueType` by specifying 'numerical', 'logical', 'integer', 'categorical' traits.
+
+Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements.
+The Handbook for Measurement of tratits [@moretti17] is a model example of such a methodological approach.
 
@@
 
-*< other identifiers >*
-
-<!--Additional information may apply to different layers of the core data: on measurement or occurrence, co-variate analysis, environmental filters, spatial distribution. -->
+### relating core data to other datasets
 
@@
 These details might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset.
-The terms for this layer of information are defined by the measurement or fact extension.
+The terms for this layer of information are defined by the 'Measurement or Fact' extension.
 The `measurementID` would also be used to link multivariate measurement data that originate from the same measurement event.
@@
 This identifier is usually dataset-specific and can be defined by the author and described with further detail in a separate data table using the columns provided in the occurrence extension detailed below.
-Some data-types may use global identifiers for occurrence data, e.g.
-a GBIF URI or a museum collection code that is publicly available.
+Some data-types may benefit from existing global identifiers for occurrence data, e.g.
+a GBIF URI or a museum collection code referencing the precise specimen at which the measuremetn was taken and that is publicly available.
  <!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
@@
 
+### Metadata
+
 Additionally, information about the authorship and ownership of the data and the terms of use should be considered.
-This information may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).
+
+Individual measurements and facts may already be part of a larger set of traits or a separate trait database before they are added to a combined trait dataset.
+To retain the rights of the original data contributor, the field `rightsHolder` states the person or organization that owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and license specifies under which terms and conditions the data can be used, re-used and/or published.
+This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`, `authorLastname` and `authorFirstname`.
+These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.
+These information usually may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).
 
@@
 The terms may be applied as columns in a supplementary data sheet or along with the core data described above.
+Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
+They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr15, Darwin core …).
+If the additional information provided is limited to few types, the extension columns can also be added to the core data table.
 
@@
 --> 
-
+<!-- 
 ## information on attribution and permissions to use
@@
 These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.
-
-<!--
 
@@
 If the dataset is from a literature review, and one or more references are available for individual data points, these references should be recorded in the References extension and listed by their Reference IDs in the ReferenceID field of the Measurements or Facts extension file, see below.
--->
-
 -->
@@
 
-The output of these functions can easily be merged using the `rbind()` function of R base. 
-<!-- Malte: Shouldn't this be cbind() ? These functions do not add additional rows but additional columns, right?-->
+Several datasets that have been produced by these functions can easily be merged using the `rbind()` function of R base.
 
@@
 For a detailed description on how to use the package see the package vignette (https://github.com/fdschneider/traitdataform/raw/master/docs/vignette.pdf). 
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain (@gossner15; @ricklefs17; @vanderplas).
 The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
@@
 
+While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we aim to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
+
 We further encourage the development of trait definitions in semantic ontologies.
@@
 
-These tools help to standardize trait datasets before upload to central as well as distributed data repositories and to harmonise and combine raw data from heterogeneous sources.
+These tools help to standardize trait datasets before upload to central as well as distributed data repositories.
+In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful ontologies for the standardization of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby opening the doors for new insights on global biodiversity patterns.
+In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
+Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources.
 We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@ref].
 
-The trait data standard as well as the R package are published for open source community development. Contributions are invited and vital for the future development and extension of the package.
-
-<!--
-## Facilitate data availability & Synthesis mainstreaming
-
-Trait-data may serve as a pilot case for data standardisation for particular types of data assessed within the exploratories. This might be transferable to other data types, as well.
-
-potential to expand the framework across research projects.
-
-Comparability of traits across taxa
-
-Gap analysis
-
-specificity of BE trait-data is local context of exploratories, if trait data are assessed on the site, trait variation can be related to land-use intensity.
-In this spatial extent, this would be a novelty.
-
--->
+The trait data standard as well as the R package are published for open source community development.
+Contributions are invited and vital for the future development of the data standard and an extension and improvement of the R package.
 
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project  and to Matthias Biber and Diana Bowler  for comments on the trait data standard and R-package.
 
@@
 # References
+
+\footnotesize

### 2017-09-26 manuscript/manuscript.Rmd
further edits on manuscript, fix bibliography, add figures

@@
 title: "Introducing an Ecological Trait-data Standard"
-author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas
-  Ostrowski, Caterina Penone
+author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas
+  Ostrowski, Caterina Penone, Nadja K. Simons
 csl: amnat.csl
@@
 1.	The assessment and analysis of functional traits has been established in all domains of ecological research promising great advancements for biodiversity and climate research.
-2.	Trait data are stored in heterogeneous formats, measured at different resolutions and using different methods, comprising literature and measured data, and focused on a variety of research questions and organism groups.
-3.	To harmonize and standardize trait data, we present a first version of an ecological trait-data standard for storing and sharing functional trait data.
-4.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
-5.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types.
+2.	To harmonize and standardize trait data, we present a first version of an ecological trait-data standard for storing and sharing functional trait data.
+3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
+4.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+
+# Key-words:
+
+functional ecology, traits, semantic web, ontologies, scientific data,
 
@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
 
@@
 For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits.
-This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].
+This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @weiner14] and parameterize ecological population-dynamic models [@refs].
 
-On the level of species assemblages, community weighted means or variance of functional traits or multivariate measures of 'functional diversity' [@petchey??] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @bello10; @hopfenmuller14; @jennings09; @moretti13;  @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014].
-Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].
-Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.
+On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @simons16;@hopfenmuller14; @mouillot13;  @vandewalle10; @borschig13].
+Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].
+Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
 
@@
 Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17].
-Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebaur17].
+Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
 Trait-data for lower plants are less completely assessed.
@@
 
-Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14].<!-- more on microbial databases -->
+Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level in community-aggregated traits [@krause14a; @fierer14; CAT, @fierer12] derived from metagenomic analysis.
+<!-- databases or datasets on microbial traits?
+-->
 
@@
 For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/), or the Carabids.org web portal (http://www.carabids.org/).
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades[e.g., @Gossner2015SData, @Gossner2015Ecology, @Renner2014].
+Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades[e.g., @gossner15, @gossner15c, @renner14].
 However, a unified platform and harmonizing scheme for the datasets is still lacking.
 It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits.
-[refs]
+[@refs]
 
@@
 Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
-On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches (T-SITA, http://t-sita.cesab.org/, @pey14).
+On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
 
@@
 The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
-Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10].
+Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a].
 Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).
@@
 
-Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
+Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14a], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
 a specimen) via a unique identifier which is assigned to a taxon.
@@
 plot measured values by factor variable or taxon) and statistical modeling (e.g.
-ANOVA for testing difference of trait value by sex) [@wickham14].
+ANOVA for testing difference of trait value by sex) [@wickham14a].
 They further draw from a defined set of columns, which makes it easier to merge datasets.
@@
 The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
-Each row of the dataset can be interpreted as a statement of an 'entity x having a quality y'.
+Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'.
 To comply with the EST, datasets must be reformatted to the longtable format.
@@
 <!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
+\caption{Process chart of the functions provided within the R package.}
+
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
+\caption{Process chart of the functions provided within the R package.}
+
+\end{figure}
 
@@
 
+<!--
 Table 1: Minimal set of terms for the traitdata standard.
@@
 Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
-This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15](Shan et al 2012, Schrodt et al 2015).
+This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].
 
@@
 -->
-Resources for semantic ontologies exist for plant traits (e.g., TOP Plant Trait Ontology [@garnier17]), soil invertebrates (T-Sita, @pey14), or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13].
+Resources for semantic ontologies exist for plant traits [e.g.
+TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13].
 However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.
@@
 If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte12; @garnier17].
+In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
 This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and define a unique dataset-specific identifier (`traitID`) as well as a short, unambiguous verbal definition (`traitDescription`) which may make use of standard terms provided in other Ontologies.
@@
 Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements.
-The Handbook for Measurement of tratits [@moretti17] is a model example of such a methodological approach.
+The Handbook for Measurement of tratits [@moretti17a] is a model example of such a methodological standardization approach.
 
 Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related Terms (`relatedTerm`) or synonyms (`synonym`).
-For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@Ref semantic database methods).
+For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@ref semantic database methods).
 This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms.
-(more detail on semantic web ontologies can be found ...)
+(more detail on semantic web ontologies can be found @refs...)
 
@@
 The terms may be applied as columns in a supplementary data sheet or along with the core data described above.
-Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
-They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr15, Darwin core …).
+Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14a).
+They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr16, Darwin core …).
 If the additional information provided is limited to few types, the extension columns can also be added to the core data table.
@@
 \centering
-\includegraphics[width=0.75\textwidth, angle=90]{figures/fig2.png}
+\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
 \caption{Process chart of the functions provided within the R package.}
@@
 For a detailed description on how to use the package see the package vignette (https://github.com/fdschneider/traitdataform/raw/master/docs/vignette.pdf).
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain (@gossner15; @ricklefs17; @vanderplas).
+
 The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
+
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
+To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains a set of functions to extract trait data via the open API interfaces of Birdlife, EOL TraitBank or BetyDB.
+We provide a script (see online appendix and R-package vignette) to demonstrates the harmonization of these different data sources.
 
@@
 Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources.
-We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@ref].
+We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
 

### 2017-09-26 manuscript/manuscript.Rmd
adding figure captions

@@
 
-functional ecology, traits, semantic web, ontologies, scientific data,
+functional ecology, traits, semantic web, ontologies, data standard
 
@@
 \includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Process chart of the functions provided within the R package.}
+\caption{Types of ecological trait data assume different entities or reported quantities.
+a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
+reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, refering to the entire taxon (or a subset, e.g.
+a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}
 
@@
 \includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Process chart of the functions provided within the R package.}
+\caption{Formats used for trait datasets.
+Identifiers relate the measurement to other layers of information on the specimen, the taxon or the trait measured.
+a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e.
+one physical occurence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.
+}
 

### 2017-10-27 manuscript/manuscript.Rmd
start revision of manuscript, add references, include comments of Klaus and Catrin

@@
 title: "Introducing an Ecological Trait-data Standard"
-author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas
-  Ostrowski, Caterina Penone, Nadja K. Simons
+author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
+  Caterina Penone, Nadja K. Simons
 csl: amnat.csl
 output:
-  pdf_document: default
   word_document:
     keep_md: yes
+  pdf_document: default
 geometry: margin=1.5in
@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions (@refs).
 
-Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13].
-Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15].
-Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e.
+<!-- more refs:  Reich et al., 1997; Wright et al., 2004; Chave et al., 2009; Schrodt et al., 2015-->
+
+Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13; @diaz16].
+Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09].
+Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e.
 the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
 For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits.
-This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @weiner14] and parameterize ecological population-dynamic models [@refs].
+This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13] which may be used to inform ecological networks [@munoz17; @weiner14; @miranda13; @digel11] and parameterize ecological population-dynamic models [@schneider12].
+On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
+Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel17; @lajoje15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].
+Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
+In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.
 
-On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @simons16;@hopfenmuller14; @mouillot13;  @vandewalle10; @borschig13].
-Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].
-Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
-
-In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.
 Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
+<!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->
 
@@
 
-Despite the obvious benefit, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits heterogeneous.
+Despite these obvious benefit and grand prospects, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.
 Trait data have therefore been standardised into centralised databases for specific organism groups and research questions.
@@
 
-Most prominently, plant traits from many different sources -- including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08] -- have been collated into the quickly growing TRY database [@kattge11].
+Most prominently, plant traits from many different sources  have been collated into the quickly growing TRY database [@kattge11; e.g.
+including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].
 Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17].
@@
 In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
-For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/), or the Carabids.org web portal (http://www.carabids.org/).
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades[e.g., @gossner15, @gossner15c, @renner14].
-However, a unified platform and harmonizing scheme for the datasets is still lacking.
-It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits.
-[@refs]
+For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], or the Carabids.org web portal (http://www.carabids.org/).
+Besides these databases, an unknown plethora of single datasets has been published along with article publications over the decades [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> ].
+In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g.
+@ricklefs17; @amphibio17; @iversen17; @myhrvold15; @kissling14; @jones09].
+Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking.
+It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].
 
@@
 
-## Need for ontologies for data standardisation
+## Need for ontologies to enable data standardisation
 
 Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15].
-However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits are compliant with this demand for open data.
-<!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].--> Open-access publications in low-threshold data repositories such as Figshare, datadryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
+However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits comply with this demand for open data.
+While it seems clear that factual data, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@ref], there is still no broad consensus achieved in the scientific community whether such data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
+<!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].
+legal standpoint in most countries: descriptive data are not copyright protected.
+Publicly fundet research must provide products.
+is still debated by the scientific community.
+-->
+
+Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
+Although open for manual download, data on those platforms might be stored in proprietary (e.g.
+.xlsx) or binary (e.g.
+.pdf) data formats which make a programmatical extraction tedious or impossible.
+Since the platforms are open for any type of scientific data, no data standards are enforced upon the uploaded material.
 Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.
 
-<!--Today, several general biodiversity data intitiatives are striving for a global integration of ecological and biological information.
-Only few so far developed universal data frameworks for trait data, such as the EOL TraitBank, which is the most general framework to date (Parr et al 2015).
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive.
--->
 To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance.
-GFBio is aiming to ease data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.
-Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org].
-On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
+One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.
+
+Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16].
+On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13].
+A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).
 
 While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
-Such a general data structure for quantitative trait data has been developed within the Encyclopedia of Life Project for upload of data into TraitBank [@parr16], relying on definitions of terms provided by the Darwin Core Standard (DWC, ...).
-However, the Terms of the Darwin Core Standard and available Extensions do not cover all of the aspects required in the various use-cases of trait-based research.
+The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables.
+The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12) and relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
+GBIF Taxonomy) to capture synonyms, misspellings and controversies  [@parr15].
+Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions.
+TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive.
+Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].
+
+However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities.
+The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered.
+To date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
+<!-- examples for blind spots -->
 
@@
 
-With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the considerations of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the work of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
 
-We propose a versatile data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data.
+We define a versatile, minimal data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data.
 We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
 
-With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL) TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC).
-We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility.
+With this framework, we build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12].
+We further expand the definitions of these terms for the use case of trait data and add terms which cover the special demands of trait-based research, while maintaining semantic compatibility with existing initatives.
 
@@
 Trait-datasets usually take different formats.
-For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
+For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix (a.k.a.
+wide-table format), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
 This format is usually reporting missing data as NA.
 It may store additional information (e.g.
-on variation of means or literature source) in secondary colums.
+on variation of means or literature source) in secondary colums or accompanying datasheets.
 The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
 Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a].
-Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).
+Also at this level of information, missing entries, e.g.
+about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15].
+However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].
 
@@
 each individual specimen would be recorded in a row, with the respective trait values stored in columns.
-In these 'occurrence data tables', each row is centered around a single physical occurrence of a species.
+In these occurrence data tables, each row is centered around a single physical occurrence of a species.
 Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
-Occurence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
+Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
 This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.
 
-Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns.
+Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns.
 Each trait column represents one measured value.
+The width of the table varies with the number of traits included.
+Merging data from different sources requires manual harmonization of these structures.
 
 Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14a], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
-a specimen) via a unique identifier which is assigned to a taxon.
-This allows for repeated measurements, even on a single individual.
+a specimen) which is assigned to a specific taxon.
+This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID.
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.
@@
 a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e.
-one physical occurence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.
+one physical occurrence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.
 }
@@
 
-## common terms for trait data
+## Common terms for trait data
 
-In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset compliant with the EST, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above (Table 1).
-We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire record (i.e. metadata). 
-Another set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).
-The full list of terms  can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
+In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above.
+We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata). 
+A second set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail). 
+The full glossary of terms can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
+Were possible, we relate these terms to existing terms in ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core ().
 
-A dataset that applies these terms should relate to the EST in its metadata.
-Since the traitdata standard is a resource under open community development, the metadata also should refer to a version number.
+A dataset that applies these terms should refer to the data standard in its metadata.
+Since the traitdata standard is a resource under open community development, this reference also should include the version number.
 This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g.
@@
 "landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
-|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurenceid|A character string user defined by the data author.
+|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurrenceid|A character string user defined by the data author.
 It is recommended best practice to use a globally unique identifier, if available, e.g.
@@
 
-## core elements of a trait-dataset
+### core elements of a trait-dataset
 
@@
 The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g.
-for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424') .
-Keeping an additional reference to a kingdom and class (using field `kingdom` or `class` as described by Darwin Core Standard) will avoid ambiguities with taxonomic homonyms. 
+for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424').
+Keeping a reference to a kingdom and class (using field `kingdom`, `class` as described by Darwin Core Standard and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
 Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
 This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].
+Alternatively to expanding the number of columns, these data could be contained in a separate data file which relates to the core dataset via the field `ScientificName`.
 
@@
 
-A similar approach to standardised trait names (`traitNameStd`) and harmonised reported values (`traitValueStd` and `traitUnitStd`) in compliance with published ontologies is advised. 
-Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI or DOI) of trait definitions or a measurement methodology.
+A similar approach to standardised trait names and harmonised reported values in compliance with published ontologies is advised.
+Here, column `traitNameStd` should report the accepted trait name in an online ontology, while the value and unit are reported in columns `traitValueStd` and `traitUnitStd`.
+Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI) of trait definitions or a measurement methodology.
 
@@
 -->
-Resources for semantic ontologies exist for plant traits [e.g.
-TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13].
+To date, semantic ontologies exist for plant traits [e.g.
+TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples).
 However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.
 
-### Defining trait thesauri and ontologies
+<!--### Defining own trait thesauri and ontologies-->
 
@@
 Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements.
-The Handbook for Measurement of tratits [@moretti17a] is a model example of such a methodological standardization approach.
+The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.
 
@@
 This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms.
-(more detail on semantic web ontologies can be found @refs...)
+(more detail on semantic web ontologies can be found @refs...).
+
+The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
+More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
+We encourage hosting such semantic web resources developed for own projects on public repositories like OBO Foundry, Github or the GFBio Terminology service.
+We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().
 
@@
 
-### relating core data to other datasets
-
-Further detail might be stored along with each measurement or fact reported in the dataset (i.e.
-each row of the dataset).
-These details might refer to the origin of the information, the measurement method or statistical precision of the reported value and may be given in a related data sheet which is linked to the data using a unique identifier (`measurementID`) or in additional columns to the core dataset.
-The terms for this layer of information are defined by the 'Measurement or Fact' extension.
-The `measurementID` would also be used to link multivariate measurement data that originate from the same measurement event.
-
-The measurement may also be referred to a single specimen or occurrence via a uniue identifier (`occurrenceID`).
-For most measured trait data this is a vital information that links related measurements of different traits taken on the same specimen.
-This identifier is usually dataset-specific and can be defined by the author and described with further detail in a separate data table using the columns provided in the occurrence extension detailed below.
-Some data-types may benefit from existing global identifiers for occurrence data, e.g.
-a GBIF URI or a museum collection code referencing the precise specimen at which the measuremetn was taken and that is publicly available.
- <!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
-It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc.
--->
- <!-- Nadja: I like the idea of having a figure explaining the different terms -->
-
 ### Metadata
 
-Additionally, information about the authorship and ownership of the data and the terms of use should be considered.
+Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
 
@@
 To retain the rights of the original data contributor, the field `rightsHolder` states the person or organization that owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and license specifies under which terms and conditions the data can be used, re-used and/or published.
-This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`, `authorLastname` and `authorFirstname`.
+This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`, `author` <!-- -->.
 These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.
@@
 
-Many terms of this section refine terms of the Darwin Core Standard and its own MeasurementOrFact and [Occurrence](http://tools.gbif.org/dwca-validator/extension.do?id=http://rs.tdwg.org/dwc/terms/Occurrence#Occurrence) extensions.
+Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the raw data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.
+
+In most cases those information would be kept in seperate data sheets of a spreadsheet document, like for instance the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited.
+In this case, a unique identifier might link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+
+The trait data standard  provides two extensions of the namespace that should be used to describe these data:
+
+- the `occurrence` extension contains information on the level of individual specimens, such as date and location and method of sampling and preservation, or physiological specifications of the phenotype, such as sex, life stage or age.
+- the `measurementOrFact` extension takes information at the level of single measurements or reported values, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+
+Many terms of this section refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions  and EOL Trait Bank use of those terms.
+
+Some data-types may benefit from existing global identifiers for occurrence data, e.g.
+a GBIF URI or a museum collection code referencing the precise specimen at which the measuremetn was taken and that is publicly available.
+
+ <!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
+It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc.
+-->
+ <!-- Nadja: I like the idea of having a figure explaining the different terms -->
+
 The terms may be applied as columns in a supplementary data sheet or along with the core data described above.
 Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14a).
-They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr16, Darwin core …).
-If the additional information provided is limited to few types, the extension columns can also be added to the core data table.
+They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr16, Darwin core ...).
+In that case, the traitdata table should at least contain a column with the respective identifier that directs to the covariate datasheet.
+
+However, it may be safer to add these information directly as own columns of the data table to enable an analysis of cofactors and correlations further down the road.
+This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
 
@@
 
+<!-- Add quick examples here -->
+
 Several datasets that have been produced by these functions can easily be merged using the `rbind()` function of R base.
@@
 Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
-To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains a set of functions to extract trait data via the open API interfaces of Birdlife, EOL TraitBank or BetyDB.
-We provide a script (see online appendix and R-package vignette) to demonstrates the harmonization of these different data sources.
+To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
+The package  'TR8' provides similar access to plant traits from a list of databases (including LEDA, BiolFlor and Ellenberg Values)[@bocci15] and aggregates them into a species-trait matrix.
+However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
+We provide a script (see online appendix and R-package vignette) to demonstrate the harmonization of these different data sources into our data standard.
 
@@
 In this paper, we propose a universal scheme for ecological trait data. 
-We also provide a toolbox in R to transfer data into this scheme.
+We also provide a toolbox in R to transfer and harmonize data into this scheme.
 
-While existing initiatives for standardising data are focused on a constrained organism group, ecosystem type or region and constitute centralised infrastructures, we aim to facilitate the use of a standard format of trait data by suggessting a flexible formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
+To date, existing initiatives and centralised infrastructures for standardisation and data  hosting are focused on a constrained organism group, ecosystem type or region.
+To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
 
@@
 In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
+
 Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources.
-We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
+<!-- enables quality control on aggregated trait measures, verification procedures and opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values --> We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
 
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project  and to Matthias Biber and Diana Bowler  for comments on the trait data standard and R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Catrin Westphal, Runa Boeddinghaus for comments on the manuscript, the trait data standard and the R-package.
 
@@
 M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
-<!-- Malte: I just started adding personal funding sources here. -->
 
 # Authors' contributions
+
+FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the glossary of terms of the trait data standard; FDS led the writing of the manuscript; NKS authored the list of traits; FDS developed the R package.
+All authors contributed critically to the drafts and gave final approval for publication.
 

### 2017-11-03 manuscript/manuscript.Rmd
Comments and additional references for the manuscript.

@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions (@refs).
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@Forsman2015] (@refs).
 
@@
 <!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->
+<!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level.
+The analysis presented here is only a first step to disentangle within- and between-species variability.
+[...] A more empirically based representation of functional diversity is expected to contribute to an improved prediction of biome boundary shifts in a changing environment.
+[...] Empirical inference of this kind is expected to stimulate and strengthen hypothesis-driven research (Golub, 2010; Weinberg, 2010) towards a unified ecological assessment of plant traits and their role for the functioning of the terrestrial biosphere." [@kattge11] / "Our traitbased approach to modelling competition makes generalization possible across the forest ecosystems of the world and their highly diverse species composition.
+[...] A challenge for the future is to move beyond growth to analyse all key demographic rates and life-history stages, and analyse how traits influence competitive outcomes and stable coexistence at the population level."[@Kunstler2016] / "Traits include those that in combination allow a first-order assessment of species' similarity in "ecological niches" or the overlap in their functional role.
+The standardized interpretation and adherence to source information and clear documentation aims to provide a replicable and transparent basis for research.
+[...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->
 
@@
 including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].
-Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17].
+Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and root measurements [Fine-root Ecology Database; @iversen17].
 Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
@@
 Repeating this exercise for other organism -->
+<!-- Nadja: This is of course the major achievement of TRY, but I think we don't have to mention it because this not our immediate goal with the trait standard.
+The important bit for us is that we provide tools to harmonize datasets.
+-->
 
@@
 In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
-For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], or the Carabids.org web portal (http://www.carabids.org/).
-Besides these databases, an unknown plethora of single datasets has been published along with article publications over the decades [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> ].
+For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
+Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @Sokol2011, @Price2013 <!-- Nadja: oldest relevant pub.
+in Dryad-->].
 In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g.
@@
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus, and extend of traits covered, their location on the internet and the terms of use.
+As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use.
 We invite data owners and users to register further datasets qualifying as trait data.
@@
 ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names.
--->
+--> <!--Nadja: I don't think that we need to mention this.
+If you think that this is important, it should be moved more to the beginning, e.g.
+within the section "Common terms for trait data"-->
 
@@
 Some data-types may benefit from existing global identifiers for occurrence data, e.g.
-a GBIF URI or a museum collection code referencing the precise specimen at which the measuremetn was taken and that is publicly available.
+a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
 
@@
 This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
+
+<!-- Nadja: can those last two paragraphs be merged with the second paragraph? I feel like we are repeating ourselves here. -->
 

### 2017-11-08 manuscript/manuscript.Rmd
updates on manuscript

@@
 output:
+  pdf_document: default
   word_document:
     keep_md: yes
-  pdf_document: default
 geometry: margin=1.5in
@@
 Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
-<!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->
 <!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level.
@@
 [...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->
+<!-- Prospect: what will trait analysis acheive, if global data across taxa are available,
+
+- intra-specific variation and evolutionary trade-offs
+- global variation
+- community functional pattern, comparative study across continents
+- big data challenges
+-->
 
@@
 
-Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation.
+Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation.
 Although open for manual download, data on those platforms might be stored in proprietary (e.g.
@@
 .pdf) data formats which make a programmatical extraction tedious or impossible.
-Since the platforms are open for any type of scientific data, no data standards are enforced upon the uploaded material.
+Since the platforms are open for any type of scientific data, no data standards are forced upon the uploaded material.
+Wheter open or not, the publication of original data is increasingly required by journals to warrant reproducibility of science.
 Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.
@@
 Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16].
-On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13].
+On the side of animal traits, a couple of initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13].
 A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).
@@
 The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables.
-The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12) and relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
+The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
 GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr15].
@@
 However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities.
-The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered.
-To date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
+The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data at file-publishing services.
+None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standads.
+One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
 <!-- examples for blind spots -->
+The other reason is competence: trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.
 
@@
 
-With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the work of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-
-We define a versatile, minimal data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data.
+With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research.
 We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
 
-With this framework, we build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12].
-We further expand the definitions of these terms for the use case of trait data and add terms which cover the special demands of trait-based research, while maintaining semantic compatibility with existing initatives.
+We build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12].
+We further expand the definitions of these terms  and add new key terms for the use case of trait data, particularly considering the work of @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+
+The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution.
+We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.
 
@@
 As a minimum consensus, trait datasets may be defined as follows: 
-A trait-dataset contains quantitative measurements or qualitative facts (i.e.
+A trait dataset contains quantitative measurements or qualitative facts (i.e.
 trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
-traits) observed on an individual or parts of an individual (i.e.
+traits) observed on an individual organism or parts of an individual (i.e.
 an entity) that can be assigned to a biological taxon (i.e.
-a species).
+a species or higher-level taxon) <!-- or larger organisational unit, e.g.
+a community -->.
+
+Quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig.
+1*a*), by aggregation of replicated measurements on multiple entities (Fig.
+1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g.
+databases, Fig.
+1*c*).
+Qualitative facts are assignments of an entity to a categorical level, e.g.
+of a behavioural or life-history trait (Fig.
+1.*d*).
 The entity or observation (i.e.
@@
 This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
-
 <!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?
@@
 Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive).
-Maybe it helps if we add an example after the definition?-->
-
-Trait-datasets usually take different formats.
-For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix (a.k.a.
-wide-table format), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
-This format is usually reporting missing data as NA.
-It may store additional information (e.g.
-on variation of means or literature source) in secondary colums or accompanying datasheets.
-The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
-Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a].
-Also at this level of information, missing entries, e.g.
-about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15].
-However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].
-
-Phenotypic variation of traits within species would usually be recorded per observation, i.e.
-each individual specimen would be recorded in a row, with the respective trait values stored in columns.
-In these occurrence data tables, each row is centered around a single physical occurrence of a species.
-Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
-Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
-This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.
-
-Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns.
-Each trait column represents one measured value.
-The width of the table varies with the number of traits included.
-Merging data from different sources requires manual harmonization of these structures.
-
-Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14a], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e.
-a specimen) which is assigned to a specific taxon.
-This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID.
-Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.
-Longtable dataset purport multiple advantages for the data manipulation (e.g.
-filtering, subsetting and aggregating data), visualization (e.g.
-plot measured values by factor variable or taxon) and statistical modeling (e.g.
-ANOVA for testing difference of trait value by sex) [@wickham14a].
-They further draw from a defined set of columns, which makes it easier to merge datasets.
-
-The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
-Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'.
-To comply with the EST, datasets must be reformatted to the longtable format.
-
-<!-- Cat: here I guess that a figure showing few rows of a species x trait matrix and a long-format one with links to other information (metadata,..) woud help quite a lot.
-By experience some users are not be familiar with the interest of using the long-table format.-->
-<!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->
+Maybe it helps if we add an example after the definition?
+--> <!--FDS: is this sufficiently defined now?
+the distinction between measured and literature data is spelled-out further below, when discussing the formats.
+Quantitative values can be taken from literature, though, like body sizes for example.
+-->
 
@@
 \end{figure}
+
+Trait-datasets usually take different formats.
+For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a.
+wide-table format,  Fig.
+2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
+This format is usually reporting missing data as 'NA'.
+It may store additional information (e.g.
+on variation of means or literature source) in secondary colums or accompanying datasheets.
+The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
+Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10].
+Also at this level of information, missing entries, e.g.
+about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15].
+However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].
+
+Phenotypic variation of traits within species would usually be recorded per observation, i.e.
+each individual specimen would be recorded in a row, with the respective trait values stored in columns.
+In these occurrence data tables (Fig.
+2*b*), each row is centered around a single physical occurrence of a species.
+Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
+Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
+This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.
+
+Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns.
+Each trait column represents one measured value.
+The width of the table varies with the number of traits included.
+Merging data from different sources requires manual harmonization of these structures.
+
+Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr15], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
+This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID.
+Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
+Longtable dataset purport multiple advantages for the data manipulation (e.g.
+filtering, subsetting and aggregating data), visualization (e.g.
+plot measured values by factor variable or taxon) and statistical modeling (e.g.
+ANOVA for testing difference of trait value by sex) [@wickham14].
+They further draw from a defined set of columns, which simplifies merging of datasets.
+
+The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
+Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'.
+To comply with the ETS, datasets must be reformatted to the longtable format (Fig. 2*c*).
+
+<!-- Cat: here I guess that a figure showing few rows of a species x trait matrix and a long-format one with links to other information (metadata,..) woud help quite a lot.
+By experience some users are not be familiar with the interest of using the long-table format.-->
+<!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->
 
@@
 In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above.
+To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers.
 We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).
-A second set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail). 
-The full glossary of terms can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).
-Were possible, we relate these terms to existing terms in ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core ().
 
-A dataset that applies these terms should refer to the data standard in its metadata.
-Since the traitdata standard is a resource under open community development, this reference also should include the version number.
-This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g.
-http://terminologies.gfbio.de/traitdatastandard/v1.0 !invalid URL!).
+Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).
+
+The online reference glossary of terms is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the ETS as "Schneider et al.
+2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/".
+In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.
+Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically via an API (@ref).
+Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
 
@@
 
+For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa.
+However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or identifiers instead.
+It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.
+
 ### relating to global taxonomy standards
@@
 We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11].
-This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
+This duplication of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
 We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
@@
 for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424').
-Keeping a reference to a kingdom and class (using field `kingdom`, `class` as described by Darwin Core Standard and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
+Keeping a reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
 Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
@@
 Here, column `traitNameStd` should report the accepted trait name in an online ontology, while the value and unit are reported in columns `traitValueStd` and `traitUnitStd`.
-Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI) of trait definitions or a measurement methodology.
+Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URIs) of trait definitions or a measurement methodology.
 
@@
 TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples).
-However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.
+However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs which would be used as `taxonID`.
 
@@
 Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements.
-The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.
+The Handbook for Measurement of traits [@moretti17] is a model example of such a methodological standardization approach.
 
-Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related Terms (`relatedTerm`) or synonyms (`synonym`).
+Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related terms (`relatedTerm`) or synonyms (`synonym`).
 For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@ref semantic database methods).
 This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms.
-(more detail on semantic web ontologies can be found @refs...).
+(more detail on semantic web ontologies can be found in the online reference for the ETS at @refs...).
 
@@
 More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
-We encourage hosting such semantic web resources developed for own projects on public repositories like OBO Foundry, Github or the GFBio Terminology service.
+We encourage hosting such semantic web resources developed for own projects as public resources, e.g.
+on OBO Foundry, the GFBio Terminology service, or on GitHub.
 We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().
@@
 ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names.
---> <!--Nadja: I don't think that we need to mention this.
-If you think that this is important, it should be moved more to the beginning, e.g.
-within the section "Common terms for trait data"-->
+-->
 
@@
 
-Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the raw data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.
+Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the core data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.
 
-In most cases those information would be kept in seperate data sheets of a spreadsheet document, like for instance the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited.
-In this case, a unique identifier might link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+To avoid redundancy and duplication of content, those information would usually be kept in seperate data sheets of a spreadsheet document.
+Examples are the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited.
+A unique identifier would link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
 
-The trait data standard  provides two extensions of the namespace that should be used to describe these data:
+For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').
+EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@braak11].
+These are .zip archives containing data table txt-files along with a descriptive metadata file (in .xml format).
+Detailled descriptions and tools for validation can be found on the website of EOL (http://eol.org/info/cp_archives) and GBIF (http://tools.gbif.org/dwca-assistant/).
+
+Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
+However, for standalone dataset publications on a hosting service without such file format recommendations (e.g.
+Figshare), data authors may want to avoid the expansion into multiple data tables and rather keep all columns in the same file, which is legitimate.
+It also may be safer to add such information directly as own columns of the data table to facilitate an analysis of cofactors and correlations further down the road.
+This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
+The trade-off is data-consistency and readability vs. avoidance of duplication.
+
+Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g.
+microbial functional traits assessed at the community level taken from a soil sample.
+Here, location or measurement information are in the primary focus of the investigation.
+A general definition, whether a column describes asset data or is part of the central dataset is ill advised.
+Therefore, our glossary of terms should be used to describe the scientific data according to the study context.
+
+The ETS provides two extensions of the namespace that should be used to describe these datasets:
 
@@
 
-Many terms of this section refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions  and EOL Trait Bank use of those terms.
+Many terms of this section refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions  and EOL TraitBank use of those terms (http://eol.org/info/structured_data_archives).
 
@@
  <!-- Nadja: I like the idea of having a figure explaining the different terms -->
-
-The terms may be applied as columns in a supplementary data sheet or along with the core data described above.
-Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14a).
-They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr16, Darwin core ...).
-In that case, the traitdata table should at least contain a column with the respective identifier that directs to the covariate datasheet.
-
-However, it may be safer to add these information directly as own columns of the data table to enable an analysis of cofactors and correlations further down the road.
-This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
-
-<!-- Nadja: can those last two paragraphs be merged with the second paragraph? I feel like we are repeating ourselves here. -->
 
@@
 
-Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy (taxonomic ontology server) by calling the function `standardize.taxonomy()`
+Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added as duplicate columns by calling the function `standardize.taxonomy()`
 
-The `standardize.traits()` function matches user-provided trait names onto public trait ontologies (for now using GBIF Backbone Taxonomy) and links it to a trait ontology or a user-provided table of trait definitions, i.e.
+The `standardize.traits()` function matches user-provided trait names onto public trait ontologies and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
 a thesaurus of traits.
@@
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15a; @ricklefs17; @vanderplas17].
 To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Catrin Westphal, Runa Boeddinghaus for comments on the manuscript, the trait data standard and the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal  for comments on the manuscript, the trait data standard and the R-package.
 

### 2017-11-08 manuscript/manuscript.Rmd
fixing bibliography

@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@Forsman2015] (@refs).
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@forsman15].
 
@@
 On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
-Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel17; @lajoje15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].
+Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel16; @lajoie15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].
 Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
@@
 For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @Sokol2011, @Price2013 <!-- Nadja: oldest relevant pub.
+Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @sokol11, @price13 <!-- Nadja: oldest relevant pub.
 in Dryad-->].
 In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g.
-@ricklefs17; @amphibio17; @iversen17; @myhrvold15; @kissling14; @jones09].
+@ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
 Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking.
@@
 While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
-The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables.
-The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
-GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr15].
+The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables.
+The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczorek12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
+GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16].
 Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions.
@@
 The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
-Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10].
+Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a].
 Also at this level of information, missing entries, e.g.
 about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15].
-However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].
+However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].
 
@@
 
-Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr15], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
+Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
 This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID.
@@
 Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements.
-The Handbook for Measurement of traits [@moretti17] is a model example of such a methodological standardization approach.
+The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.
 
@@
 For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').
-EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@braak11].
+EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@gbif17].
 These are .zip archives containing data table txt-files along with a descriptive metadata file (in .xml format).
@@
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15a; @ricklefs17; @vanderplas17].
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
 To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.

### 2017-11-09 manuscript/manuscript.Rmd
major revision manuscript

@@
 ---
-title: "Introducing an Ecological Trait-data Standard"
+title: "Towards an Ecological Trait-data Standard"
 author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
@@
 
-1.	The assessment and analysis of functional traits has been established in all domains of ecological research promising great advancements for biodiversity and climate research.
+1.	The assessment and analysis of functional traits has been established in all domains of ecological research, promising great advancements for biodiversity and climate research.
 A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types.
-2.	To harmonize and standardize trait data, we present a first version of an ecological trait-data standard for storing and sharing functional trait data.
+2.	To harmonize and standardize datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
 3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@forsman15].
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07].
+The scope of trait-based research spans from the evolutionary basis of individual-level properties to global patterns of biodiversity and ecosystem functioning.
 
-<!-- more refs:  Reich et al., 1997; Wright et al., 2004; Chave et al., 2009; Schrodt et al., 2015-->
-
-Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13; @diaz16].
-Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09].
+Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04].
 Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e.
 the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
-For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits.
-This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13] which may be used to inform ecological networks [@munoz17; @weiner14; @miranda13; @digel11] and parameterize ecological population-dynamic models [@schneider12].
-On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
-Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel16; @lajoie15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].
-Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
-In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.
+Mutual 'trait-matching' of co-evolved organisms enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parameterize ecological population-dynamic models [@schneider12; @kalinkat13].
+These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
+Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97].
+Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
+Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
 
-Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
-<!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level.
-The analysis presented here is only a first step to disentangle within- and between-species variability.
-[...] A more empirically based representation of functional diversity is expected to contribute to an improved prediction of biome boundary shifts in a changing environment.
-[...] Empirical inference of this kind is expected to stimulate and strengthen hypothesis-driven research (Golub, 2010; Weinberg, 2010) towards a unified ecological assessment of plant traits and their role for the functioning of the terrestrial biosphere." [@kattge11] / "Our traitbased approach to modelling competition makes generalization possible across the forest ecosystems of the world and their highly diverse species composition.
-[...] A challenge for the future is to move beyond growth to analyse all key demographic rates and life-history stages, and analyse how traits influence competitive outcomes and stable coexistence at the population level."[@Kunstler2016] / "Traits include those that in combination allow a first-order assessment of species' similarity in "ecological niches" or the overlap in their functional role.
-The standardized interpretation and adherence to source information and clear documentation aims to provide a replicable and transparent basis for research.
-[...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->
-<!-- Prospect: what will trait analysis acheive, if global data across taxa are available,
+In recent years, functional ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
+Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. 
+However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. 
+To mobilize data for this enterprise, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources.
+The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way.
+Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g.
+Figshare, DataDryad).
+Given this prospect, there will be no lack of data, but a lack of data standardisation.
 
-- intra-specific variation and evolutionary trade-offs
-- global variation
-- community functional pattern, comparative study across continents
-- big data challenges
--->
+In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users.
+We further (2.) review existing trait databases and web services, as well as initiatives for standardisation.
+We (3.) describe the common practice for structuring trait data for ecological research. 
+Based on these considerations, we (4.) propose a low-threshold struture and vocabulary for describing trait-datasets compatible with existing ontologies.
+(5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.
 
-## Overview of trait-data initiatives
+With an easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold  trait data standardisation.
 
-Despite these obvious benefit and grand prospects, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.
+# 1. Need for data standardisation
+
+Despite the obvious benefit and grand prospects of functional approaches, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.
+
 Trait data have therefore been standardised into centralised databases for specific organism groups and research questions.
-These services provide platforms for consensus building in methods, attract data submissions from a focal community of researchers and map them into a common scheme.
-Furthermore, the databases provide access management services and provide data usage policies, while simplifying data queries for synthesis researchers.
+These compiled data are either being published as static data tables or in the form of sophisticated web-services (see below for a review of current initiatives).
+Services attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, but more importantly, provide platforms for consensus building in methods.
+Furthermore, database services offer access control and data usage policies, to claim the rights of the original data providers, while simplifying data queries for synthesis researchers.
 
-Most prominently, plant traits from many different sources  have been collated into the quickly growing TRY database [@kattge11; e.g.
-including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].
+Nonetheless, open-access publications in low-threshold data repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation.
+Although open for manual download, data on low-threshold data platforms might be stored in proprietary (e.g.
+.xlsx) or binary (e.g.
+.pdf) data formats which make a programmatical extraction tedious or impossible.
+
+On the other hand, not all centralised databases for functional traits comply with the criteria of open access, which leaves them at a disadvantage compared to low-threshold repositories. 
+Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit.
+Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
+However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
+Low-threshold data repositories assume public domain and Creative Commons Attribution licenses compliant with the demands of journals and research funding.
+
+Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
+To cope with this trend towards distributed data sources, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance.
+These services acquire public data from other sources and simplify the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
+One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+
+# 2. Overview of trait-data initiatives
+
+In this section, we list two kinds of initiatives: the first aims at the harmonization of trait data into trait databases; the second develops trait ontologies and data standards.
+Often ontologies are being developed to serve a particular database project and thus are being developed in concert, as in the case of the TRY plant database [@kattge11] and the TOP Thesaurus of plant traits [@garnier17].
+
+## Databases and Datasets
+
+Plant flower, leaf and stem traits have been measured for decades, and were compiled into comprehensive databases and published as static datasets, e.g.
+the LEDA traitbase for the Northwest-European flora [@kleyer08].
+Regarding webservices, most prominently, plant traits from many different sources have been collated into the extensive TRY database [@kattge11].
 Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and root measurements [Fine-root Ecology Database; @iversen17].
@@
 Trait-data for lower plants are less completely assessed.
-<!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications.
-The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme.
-Repeating this exercise for other organism -->
-<!-- Nadja: This is of course the major achievement of TRY, but I think we don't have to mention it because this not our immediate goal with the trait standard.
-The important bit for us is that we provide tools to harmonize datasets.
--->
 
-Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level in community-aggregated traits [@krause14a; @fierer14; CAT, @fierer12] derived from metagenomic analysis.
-<!-- databases or datasets on microbial traits?
--->
+Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12; @ulrich08] derived from metagenomic analysis [@torsvik02; @langille13].
+Databases are at hand to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
+the KEGG orthology, @kanehisa12].
 
-In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
+In the animal kingdom, n unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17; @sokol11; @price13].
+In the past two decades, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g.
+@ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
+Specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
 For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @sokol11, @price13 <!-- Nadja: oldest relevant pub.
-in Dryad-->].
-In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g.
-@ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
-Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking.
-It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].
+Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is lacking.
+It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@moretti17].
+
+A general database for major physiological and life-history traits of all taxa is the Encyclopedia of Life Trait Bank project [@parr16, see below for more detail].
 
@@
 
-## Need for ontologies to enable data standardisation
+## Thesauri and Ontologies
 
-Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15].
-However, although the number of trait-based studies grows constantly, not all scientific databases for functional traits comply with this demand for open data.
-While it seems clear that factual data, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@ref], there is still no broad consensus achieved in the scientific community whether such data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
-<!--The training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16].
-legal standpoint in most countries: descriptive data are not copyright protected.
-Publicly fundet research must provide products.
-is still debated by the scientific community.
--->
+A glossary or thesaurus of terms would be primarily a listing of definitions.
+Expanding on this, ontologies are linking individual terms with each other by formally defining the relationship between terms, to the ends of enabling a computational interpretation of data.
 
-Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation.
-Although open for manual download, data on those platforms might be stored in proprietary (e.g.
-.xlsx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious or impossible.
-Since the platforms are open for any type of scientific data, no data standards are forced upon the uploaded material.
-Wheter open or not, the publication of original data is increasingly required by journals to warrant reproducibility of science.
-Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.
-
-To facilitate the standardized upload of research data to reliable and interlinked data infrastructures, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance.
-One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data.
-
-Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16].
-On the side of animal traits, a couple of initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13].
+Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16].
+On the side of animal traits, a couple of initiatives published comprehensive trait standards, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13].
 A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).
 
-While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
+Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measureing.
+For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in Thesauri.
+
+While trait thesauri define traits for focal groups of organisms, they are not specifing the  formats or structures in which trait data should be stored. 
 The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables.
-The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczorek12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
+The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
 GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16].
@@
 However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities.
-The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data at file-publishing services.
-None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standads.
+The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
+None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
 One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
@@
 
-# Proposing an ecological traitdata standard
-
-With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research.
-We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
-
-We build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12].
-We further expand the definitions of these terms  and add new key terms for the use case of trait data, particularly considering the work of @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-
-The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution.
-We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.
-
-## Common structure of trait datasets
+# 3. Common structure of trait datasets
 
@@
 The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
-Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'.
-To comply with the ETS, datasets must be reformatted to the longtable format (Fig. 2*c*).
+Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig.
+2*c*).
 
@@
 
-## Common terms for trait data
+# 4. Proposing an ecological traitdata standard
 
-In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above.
-To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers.
-We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).
+With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research.
+We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices.
 
-Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).
+Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits (see below for further detail).
 
-The online reference glossary of terms is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the ETS as "Schneider et al.
+The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the data standard as "Schneider et al.
 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/".
 In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.
-Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically via an API (@ref).
+Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e.
+via the Application Programming Interface, API).
 Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
 
-<!--
-Table 1: Minimal set of terms for the traitdata standard.
-valueType can be of type character, factor, boolean or numerical.
-vocabulary is only needed for factorial trait data and defines the factor levels which are allowed.
-The identifier is a unique and globally valid URI.
+We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
+The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution.
+We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.
 
-| Term | valueType | vocabulary | Identifier | Definition | 
-|------|-----------|------------|------------|------------|
-|scientificName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificname |Original character string provided as species name by the data owner (kept for reference only)|
-|taxonID | factor | | http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonid | Verified name identifier of the species or subspecies (or higher taxon) for which this value was collected.
-Each entry should link to a published reference species list/taxonomy.
-Avoid synonyms.|
-|taxonRank | factor | domain; kingdom; subkingdom; superhpylum [...] |http://fdschneider.de/bexis_traits/traitdatastandard.html#taxonrank | The taxonomic rank of the most specific name in the scientificName.
-Recommended best practice is to use a controlled vocabulary.|
-|kingdom | factor | |http://fdschneider.de/bexis_traits/traitdatastandard.html#kingdom | The full scientific name of the kingdom in which the taxon is classified.| 
-|traitName | character | |http://fdschneider.de/bexis_traits/traitdatastandard.html#traitname |Name of the trait as used by the data provider (kept for compatibility on provider side).|
-|traitID|factor|see traitlist|http://fdschneider.de/bexis_traits/traitdatastandard.html#traitid ||
-|traitValue|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvalue |Measured raw value or factor level for this species trait as measured and provided by the author.|
-|traitUnit|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunit |Reports the unit that the author's raw data were measured in, if applicable (only for numeric values).|
-|scientificNameStd|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#scientificnamestd |Full scientific name (with authorship and date information if known).|
-|traitNameStd|factor||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitnamestd |Descriptive name of the trait reported.
-This should follow the controlled vocabulary of a thesaurus of traits available online or published along with the dataset.|
-|traitValueStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitvaluestd |Standardised measured value or factor level for this species trait.
-Numerical values must use the correct unit.
-Factor levels must be compliant with the thesaurus of traits.|
-|traitUnitStd|character||http://fdschneider.de/bexis_traits/traitdatastandard.html#traitunitstd |The units associated with the measurementValue.
-The expected standard unit for this trait as defined in the traitlist/ thesaurus.|
-|measurementID|	character||	http://fdschneider.de/bexis_traits/traitdatastandard.html#measurementid |A unique, dataset-specific identifier.
-Links multi-value trait measurements, e.g.
-x-y-z coordinates of a morphometric landmark, biochemical compound quantities for different chainlengths.
-In this case, the trait names must specifiy the sub-measurement, e.g.
-"landmark32_x", and must be specified in a reference trait list, given in the field "measurementMethod".|
-|occurrenceID|character|	|http://fdschneider.de/bexis_traits/traitdatastandard.html#occurrenceid|A character string user defined by the data author.
-It is recommended best practice to use a globally unique identifier, if available, e.g.
-a museum ID referring to the individual specimen from which the data were measured.
-Usually does not apply for literature data.|
-|warnings|character|	|	|Warnings from the R-script will be stored here, e.g.
-regarding a lack of match between the provided taxonID and the ontology, or the trait names or values, a mismatch in the units provided and the unit expected according to the trait table.
-User defined warnings and flags can be added as well, e.g.
-'NOTUSE'.|
+## core structure for trait data
 
-<!-- Nadja: Is this the kind of table you had in mind? -->
+As described above, longtable formats are the most predictable and compatible format for complex datasets.
+
+If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
+
+If taxon information and metadata need to be included, datasets may apply the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core Archives [DwCA;].
+
+<!--- -TO DO ---->
+
+## vocabulary for trait data
+
+In the subsequent sections we propose a set of terms for composing a trait dataset, starting with the columns required for our minimal definition of 'core' trait-datasets proposed above.
+To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
+We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).
 
@@
 
-Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements.
-The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.
-
 Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related terms (`relatedTerm`) or synonyms (`synonym`).
-For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@ref semantic database methods).
-This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms.
-(more detail on semantic web ontologies can be found in the online reference for the ETS at @refs...).
+For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
+This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms [@garnier17].
 
@@
 
-The ETS provides two extensions of the namespace that should be used to describe these datasets:
+The data standard provides two extensions of the namespace that should be used to describe these datasets:
 
@@
 
-For a detailed description on how to use the package see the package vignette (https://github.com/fdschneider/traitdataform/raw/master/docs/vignette.pdf).
-
+For a detailed description on how to use the package see the package documentation. 
 The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
@@
 To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases (including LEDA, BiolFlor and Ellenberg Values)[@bocci15] and aggregates them into a species-trait matrix.
+The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
 However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
-We provide a script (see online appendix and R-package vignette) to demonstrate the harmonization of these different data sources into our data standard.
+We provide a script (see online appendix and R-package vignette <!-- add URL-->) to demonstrate the harmonization of very different data sources into our data standard.
 
@@
 
-In this paper, we propose a universal scheme for ecological trait data. 
+In this paper, we propose a general, low-threshold structure and vocabulary for ecological trait data. 
 We also provide a toolbox in R to transfer and harmonize data into this scheme.
 
-To date, existing initiatives and centralised infrastructures for standardisation and data  hosting are focused on a constrained organism group, ecosystem type or region.
+To date, existing initiatives and centralized infrastructures for standardization and data  hosting are focused on a constrained organism group, ecosystem type or region.
 To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
 
-We further encourage the development of trait definitions in semantic ontologies.
-The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards.
+We further encourage the development of trait thesauri in semantic ontologies.
+The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
 This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of the global plant trait database TRY shows intricately [@diaz16].
 
-These tools help to standardize trait datasets before upload to central as well as distributed data repositories.
+The tools proposed here help to standardize trait datasets before upload to central as well as distributed data repositories.
 In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful ontologies for the standardization of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby opening the doors for new insights on global biodiversity patterns.

### 2017-11-10 manuscript/manuscript.Rmd
start shortening manuscript: remove definitions of terms into website

@@
 title: "Towards an Ecological Trait-data Standard"
-author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
-  Caterina Penone, Nadja K. Simons
+author:
+- name: Florian D. Schneider
+- name: Malte Jochum
+- name: Gaetane LeProvost
+- name: Andreas Ostrowski
+- name: Caterina Penone
+- name: Nadja K. Simons
 csl: amnat.csl
 output:
-  pdf_document: default
+  pdf_document:
+    fig_caption: yes
+    fig_crop: no
+    template: pretty.template
+  html_document:
+    fig_caption: yes
+    fig_width: 4
+    highlight: null
+    keep_md: yes
+    theme: united
   word_document:
-    keep_md: yes
-geometry: margin=1.5in
-header-includes: \usepackage{lineno}
-fontsize: 12pt
+    fig_caption: yes
+    fig_width: 6
 bibliography: div05.bib
@@
 Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
-To cope with this trend towards distributed data sources, data brokering services such as the German Federation on Biological Data (gfbio.org) are likely to gain importance.
-These services acquire public data from other sources and simplify the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
+To cope with this trend towards distributed data sources, data brokering services (e.g.
+the German Federation on Biological Data; gfbio.org)  or database management systems for scientific projects (e.g.
+BExIS, XX ) are likely to gain importance.
+These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
 One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+
+After all, to mobilise the broad wealth of data into structured databases, it requires a certain minimum of information from the data providers.
+Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.
 
@@
 
-# 3. Common structure of trait datasets
+# 3. The structure of trait datasets
 
@@
 
-With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research.
+With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research.
 We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices.
+We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices and existing standards for biodiversity data management.
 
-Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits (see below for further detail).
+The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/.
 
-The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the data standard as "Schneider et al.
-2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/".
-In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.
-Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e.
-via the Application Programming Interface, API).
-Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
-
-We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.
 The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution.
@@
 
-## core structure for trait data
+## structure for trait data
 
-As described above, longtable formats are the most predictable and compatible format for complex datasets.
+As described above, tidy longtable formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.
 
+To enable a better standardization of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
+In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
 If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
+For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').
 
-If taxon information and metadata need to be included, datasets may apply the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core Archives [DwCA;].
+The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al.
+2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/".
+In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
 
-<!--- -TO DO ---->
+In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
+Therefore we recommend use of  Darwin Core Archives [DwCA; @robertson09], the structure proposed for GBIF [@gbif17; http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr15; http://eol.org/info/cp_archives].
+This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.
 
@@
 
-In the subsequent sections we propose a set of terms for composing a trait dataset, starting with the columns required for our minimal definition of 'core' trait-datasets proposed above.
-To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
-We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).
+Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/TraitDataStandard/. 
+Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e.
+via the Application Programming Interface, API).
+Whereever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
 
-### core elements of a trait-dataset
+### Core columns
 
-The core content of a row in the trait-dataset is the reported measurement or fact for a single observation, which is composed of a value (`traitValue`) and -- for numeric values -- a standard unit (`traitUnit`).
-To associate the measurement or fact to a clear trait definition, a unique identifier (`traitID`) links each row to a descriptive trait name (`traitName`).
-
-The core data are also keeping a record of the scientific taxon for which the measurement or fact was obtained in the form of a descriptive taxon name (`scientificName`).
-
+To describe the minimum core content, where each row is the reported measurement or fact for a single observation, the following columns are required: a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); a descriptive trait name (`traitName`);  the scientific taxon for which the measurement or fact was obtained (`scientificName`).
 For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
 However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or identifiers instead.
+
+### Relate to ontologies
+
 It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.
+This is acheived by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g.
+GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').
+By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practiced in TRY; @kattge11].
+This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
+Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms.
+It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
+This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].
+
+If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
+In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
+See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].
+We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
+
+<!--
 
@@
 
-<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.
--->
 To date, semantic ontologies exist for plant traits [e.g.
@@
 
-<!--### Defining own trait thesauri and ontologies-->
+### Defining own trait thesauri and ontologies
 
@@
 We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().
+-->
 
@@
 
-Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the core data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.
-
+Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
+<!--
 To avoid redundancy and duplication of content, those information would usually be kept in seperate data sheets of a spreadsheet document.
@@
 Figshare), data authors may want to avoid the expansion into multiple data tables and rather keep all columns in the same file, which is legitimate.
+-->
+
+<!--- WORK HERE!!! -->
+
 It also may be safer to add such information directly as own columns of the data table to facilitate an analysis of cofactors and correlations further down the road.
@@
 
-# Computational tools for producing compliant data
+# 5. Computational tools for producing compliant data
 

### 2017-11-11 manuscript/manuscript.Rmd
massive shortening of manuscript; move definitions to online resource

@@
 
-### Relate to ontologies
+### Relate to global taxon and trait definitions
 
@@
 
+The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
+More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
+
 If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
+A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
+Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
+units or legit factor levels) of measured values or reported facts.
+It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
+We encourage publishing trait thesauri developed for the own project context as public resources, e.g. on OBO Foundry, the GFBio Terminology service, or on other stable web services.
+
 See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].
 We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
-
-<!--
-
-### relating to global taxonomy standards
-
-Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right.
-To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable.
-We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11].
-This duplication of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
-We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g.
-GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`).
-The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g.
-for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424').
-Keeping a reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
-Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
-This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].
-Alternatively to expanding the number of columns, these data could be contained in a separate data file which relates to the core dataset via the field `ScientificName`.
-
-### relating data to trait definitions
-
-A similar approach to standardised trait names and harmonised reported values in compliance with published ontologies is advised.
-Here, column `traitNameStd` should report the accepted trait name in an online ontology, while the value and unit are reported in columns `traitValueStd` and `traitUnitStd`.
-Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URIs) of trait definitions or a measurement methodology.
-
-To date, semantic ontologies exist for plant traits [e.g.
-TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples).
-However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs which would be used as `taxonID`.
-
-### Defining own trait thesauri and ontologies
-
-If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-In its simplest form, a trait thesaurus would provide a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
-This listing or table should at minimum provide a human readable, informative trait name (`traitName`) and define a unique dataset-specific identifier (`traitID`) as well as a short, unambiguous verbal definition (`traitDescription`) which may make use of standard terms provided in other Ontologies.
-As an Example,  the definition for 'fruit mass' in TOP reads: "the mass (PATO:mass), either fresh or dried, of a fruit (PO:fruit)", referring to Phenotypic Characeristics Ontology PATO and Planteome Plant Ontolgy, PO (http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass).
-The thesaurus should also constrain the legit factor levels (for categorical data, `factorLevels`) or expected standard units (`traitUnitStd` for numerical data).
-The type of values should be differentiated in the field `valueType` by specifying 'numerical', 'logical', 'integer', 'categorical' traits.
-
-Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related terms (`relatedTerm`) or synonyms (`synonym`).
-For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms [@garnier17].
-
-The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
-More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
-We encourage hosting such semantic web resources developed for own projects as public resources, e.g.
-on OBO Foundry, the GFBio Terminology service, or on GitHub.
-We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().
--->
-
-<!-- Flo: For now removing this block on the steps necessary to create web ontologies of traits.
-The section above is covering the essentials and implies that further work is required.
-We might add this following section as a supplement or within the traitdata standard.
--->
-<!--
-
-## towards a distributed semantic web ontology of traits
-
-Thus, a trait thesaurus or ontology would assign trait names with A) an unambiguous definition of the trait and B) an expected format of measured values or reported facts, and might additionally provide C) semantic relationships for deriving a hierarchical or tree-based classification of traits.
-
-An ontology is going one step further by providing an explicit definition of a term including a formal model of the conceptual objects and relationships, or entities and qualities [@garnier17, @gruber95], ideally structured following the guidelines for a semantic web standard [ref].
-
-Traits are not only defined in terms of their interpretation, but are ideally also standardised in terms of numerical units and, even more important, the use of factor levels.
-This is challenging, given the range of data types that fall within datasets of functional traits.
-
-Numerical values represent measurements of lengths, volumes, ratios, rates or timespans. Integer values may apply to count data (e.g. eggs per clutch). 
-Binary data (encoded as 0 or 1) or logical data (coded as TRUE or FALSE) may apply to qualitative traits such as specific behaviour during mating (e.g.
-are territories defended) or specialisation to a given habitat (e.g.
-species restricted to relicts of primeval forests).
-Many traits are categorical and allow for a constrained set of factor levels, such as sex differences in wing morphology (both sexes winged, both sexes unwinged, only males winged, only females winged) or unconstrained entries such as color.
-Categorical traits often are ordinal, i.e.
-they have a logical sequence as in the case of life stages or hibernation stages, or habitat preference traits such as horizontal stratum use.
-Finally, there are specific formats of multivariate trait values, e.g.
-x-y-z coordinates of a landmark measured in a standardised 3D space [ref] or relative abundance of chain-lengths in biochemical compounds [].
-To cope with this variety of data types, definitions should refer to other well-defined terms of existing ontologies that describe standard units, morphological body parts, protein characteristics (Protein Ontology) or chemical terms (e.g.
-the ChEBI, http://www.obofoundry.org/ontology/chebi.html).
-
-Multiple approaches have spawned around the initialisation of trait databases.
-Such reference definitions are currently being published in methodological handbooks (Moretti et al , ) or lists of accepted traits (databases), or an online ontology or thesaurus (TOP, T-Sita, ), most advanced certainly for plant traits compiled in the Thesaurus of Plant characteristics (TOP, @garnier17).
-
-Online ontologies extend into (machine readable) semantic web resources by providing a hierarchical classification of traits (TOP) or a relational tree of functional traits (e.g. TOP or T-SITA). .
-
-Each trait definition may link to a broader or narrower term.
-For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-(Ref semantic database methods)
-This links traits of similar functional meaning and allows cross-taxon comparative studies at the level of broader terms.
-
-These systematic approaches to traits will be very useful for comparing similar traits measured in different taxa at higher hierarchical levels.
-Furthermore, trait definitions may refer to related terms that describe a similar feature but with other means, or synonyms defined in other trait ontologies.
-By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.
-
-The benefit of such classifications will increase if open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
-To harmonize trait data across databases, future trait standard initiatives should provide this functionality.
-Online ontologies hosted with accredited ontology servers have the advantage of providing a persistent and direct link of the term on the internet (a *Uniform Resource Identifier*, URI).
-Terminology portals or registries, such as the GFBio Terminology Service or ontobee, may provide a central host for trait ontologies.
-
-Thus, a trait thesaurus or ontology would assign trait names with A) an unambiguous definition of the trait and B) an expected format of measured values or reported facts, and might additionally provide C) semantic relationships for deriving a hierarchical or tree-based classification of traits.
-It additionally might relate to a more specific method of measurement.
-
-The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the standards.
-This effort is time consuming and lengthy, but is returns synergies and great scientific benefit by enabling a global synthesis, as the example of a global plant trait database TRY shows intricately [@diaz16].
--->
-<!-- Nadja: I agree that this section is not needed in the paper, but it would certainly be nice to publish it in a supplement or with the traidata standard. Malte: Agreed. -->
-
 ### Metadata
@@
 Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
+We define a set of terms that allows to relate trait data to authorship and define a bibliographic reference and license model.
 
-Individual measurements and facts may already be part of a larger set of traits or a separate trait database before they are added to a combined trait dataset.
-To retain the rights of the original data contributor, the field `rightsHolder` states the person or organization that owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and license specifies under which terms and conditions the data can be used, re-used and/or published.
-This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`, `author` <!-- -->.
-These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.
-These information usually may be kept in the metadata of the dataset, but if datasets from different sources are merged, those should be referred to by a unique identifier (`datasetID`) or be reported as additional columns in the merged dataset (`author`, `license`, ...; see Dublin Core Metadata standards, Ref).
+In the case of raw measurement data, these information apply to the entire trait dataset, which classifies them as metadata stored in the information structure where the data are published. 
+In cases, where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be linked to the measurement level, via an unambiguous `datasetID` and a descriptive `datasetName`.
 
@@
 
-## Extensions for additional data layers
+### Extensions for additional data layers
 
@@
 A unique identifier would link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+This is usually the case if the occurrences are externally documented, for instance as specimens from a museum. This also applies if the data are stored in additional data tables.
 
@@
 - the `occurrence` extension contains information on the level of individual specimens, such as date and location and method of sampling and preservation, or physiological specifications of the phenotype, such as sex, life stage or age.
-- the `measurementOrFact` extension takes information at the level of single measurements or reported values, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- the `measurementOrFact` extension takes information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
 
-Many terms of this section refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions  and EOL TraitBank use of those terms (http://eol.org/info/structured_data_archives).
+Many terms of these extensions refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
 
-Some data-types may benefit from existing global identifiers for occurrence data, e.g.
+Some data-types may directly refer to existing global identifiers for occurrence data, e.g.
 a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
@@
 
+<!--
 ### additional information on measurement or fact
@@
 for upload to the GBIF database.
-
-<!-- ## The Biodiversity Exploratories Extensions and template
-
-Data will be uploaded to BExIS using the template to join data to the trait dataset while keeping full data authority: author information and ownership, access right management, as provided by BExIS.
-
-If applicable, trait values can be assigned to regions, plot IDs or sampling events which will then allow users to combine trait data with abundance biodiversity, ecosystem functioning, other observational data or the management regime on the sampled plots.
-
-Metadata:
-Each trait dataset uploaded to BExIS and added to the trait database will be accompanied with a metadata file containing additional information on data origin, method, authors and owners (in redundancy with primary data, see above) and possibly more detailed definition of the traits recorded.
---> 
-<!-- 
-## information on attribution and permissions to use
-
-Individual measurements and facts will likely already belong to a larger set of traits or a separate trait database before they are added to a combined trait database.
-To retain the rights of the original data contributor, the field `rightsHolder` states the person or organisation who owns or manages the rights to the data; `bibliographicCitation` states a bibliographic reference which should be cited when the data is used; and `license` specifies under which terms and conditions the data can be used, re-used and/or published.
-This information always applies to one single fact or measurement, further information on the larger dataset which originally contained this entry can be stored in `datasetID`, `datasetName`,`authorLastname` and `authorFirstname`.
-These columns should hence give credit to the person who compiled the original dataset and signs responsible for the correct identification and reporting of the rights holder.
-
-Finally, there is the set of information that applies to the entire trait-dataset, which classifies them as metadata.
-Since trait data are of great use for synthesis studies, information about how the data may be distributed, re-used and attributed to are of particular importance for trait datasets.
-Most researchers encourage re-use of their published datasets while making sure they are sufficiently credited.
-The use of permissive licenses for traitdata publications, such as Creative Commons Attribution or Creative Commons Zero/Public Domain release, has been established as the gold standard.
-
-<!-- Where is this data record available online?
-Source in the Measurements or Facts extension file is for a url to be displayed alongside the data point, so EOL visitors can click through to an online resource where the data originated.
-This might be a dataset stored at an online repository, a taxon page on a website, or any other online location.
-What shall I cite if I use this data record in a publication?
-BibliographicCitation in the Measurements or Facts extension file is for the suggested citation format for a user of this dataset to include in a references list or bibliography.
-Even if the data record or dataset have never been published elsewhere, we recommend that you craft an appropriate citation so others can provide proper credit.
-If the dataset is from a literature review, and one or more references are available for individual data points, these references should be recorded in the References extension and listed by their Reference IDs in the ReferenceID field of the Measurements or Facts extension file, see below.
 -->
@@
 
-We provide an R package to assist in producing data compliant with the trait data standard proposed above.
-The package is being developed as an open source project on GitHub (https://github.com/fdschneider/traitdataform).
+To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
+The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
+However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
 
-There are two major use cases for the package:
+We provide the R package 'traitdataform' to assist in producing data compliant with the trait data standard proposed here. There are two major use cases for the package:
 
-- preparation of own trait datasets for upload into public data bases, and
+- preparing of own trait datasets for publication on low-threshold fileservers and project databases, and
 - harmonising trait datasets from different sources by moulding them into a unified format.
 
-The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table data into a measurement longtable format.
+The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format.
 It also maps column names into terms provided in the trait data standard.
@@
 
-Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added as duplicate columns by calling the function `standardize.taxonomy()`
+Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added to the core data by calling the function `standardize.taxonomy()`.
 
@@
 <!-- Add quick examples here -->
-
-Several datasets that have been produced by these functions can easily be merged using the `rbind()` function of R base.
 
@@
 
-For a detailed description on how to use the package see the package documentation. 
+Several datasets that have been produced by these functions can easily be appended using the `rbind()` function of R base.
+
+To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identfier using the function `merge()`.
+This enables an easy handling of data sources that have been stored in a Darwin Core Archive database format.
+
+For a detailed description on how to use the package see the package documentation website. 
 The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
-To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
-However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
-We provide a script (see online appendix and R-package vignette <!-- add URL-->) to demonstrate the harmonization of very different data sources into our data standard.
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to datasets listed in our online spreadsheet that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
 
@@
 
-To date, existing initiatives and centralized infrastructures for standardization and data  hosting are focused on a constrained organism group, ecosystem type or region.
+To date, existing initiatives and centralised infrastructures for standardisation and data  hosting are focused on a constrained organism group, ecosystem type or region.
 To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
 
-We further encourage the development of trait thesauri in semantic ontologies.
+The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories.
+In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby opening the doors for new insights on global biodiversity patterns.
+In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
+
+We therefore encourage the development of trait thesauri in semantic ontologies.
 The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
@@
 
-The tools proposed here help to standardize trait datasets before upload to central as well as distributed data repositories.
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful ontologies for the standardization of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby opening the doors for new insights on global biodiversity patterns.
-In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
-
 Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources.
-<!-- enables quality control on aggregated trait measures, verification procedures and opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values --> We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
+Such tools to handle the extensive raw data resources on traits are a prerequisite for research on trait variation along environmental gradients or intra-specific variation.
+It also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
+We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
 

### 2017-11-13 manuscript/manuscript.Rmd
finalize manuscript, spell checked (british), added examples to R package, shorten text on extensions

@@
 author:
-- name: Florian D. Schneider
-- name: Malte Jochum
-- name: Gaetane LeProvost
-- name: Andreas Ostrowski
-- name: Caterina Penone
-- name: Nadja K. Simons
+- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
+- name: Malte Jochum \footnotemark[4] 
+- name: Gaetane LeProvost \footnotemark[2] 
+- name: Andreas Ostrowski \footnotemark[5] 
+- name: Caterina Penone \footnotemark[4] 
+- name: Nadja K. Simons \footnotemark[6] 
 csl: amnat.csl
@@
     template: pretty.template
+    keep_tex: yes
   html_document:
@@
 
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
+\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
+\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
+\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[5]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
+\footnotetext[6]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+
 \linenumbers
@@
 A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types.
-2.	To harmonize and standardize datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
+2.	To harmonize and standardise datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
 3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
-4.	With these decentralized tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+4.	With these decentralised tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.
 
@@
 the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
-Mutual 'trait-matching' of co-evolved organisms enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parameterize ecological population-dynamic models [@schneider12; @kalinkat13].
+Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13].
 These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
@@
 Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g.
-Figshare, DataDryad).
+Figshare, Dryad).
 Given this prospect, there will be no lack of data, but a lack of data standardisation.
@@
 We (3.) describe the common practice for structuring trait data for ecological research. 
-Based on these considerations, we (4.) propose a low-threshold struture and vocabulary for describing trait-datasets compatible with existing ontologies.
+Based on these considerations, we (4.) propose a low-threshold structure and vocabulary for describing trait-datasets compatible with existing ontologies.
 (5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.
@@
 
-Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12; @ulrich08] derived from metagenomic analysis [@torsvik02; @langille13].
+Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12] derived from metagenomic analysis [@torsvik02; @langille13].
 Databases are at hand to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
@@
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use.
+As an attempt to track  databases and unstandardised datasets from independent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use.
 We invite data owners and users to register further datasets qualifying as trait data.
@@
 
-Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16].
-On the side of animal traits, a couple of initiatives published comprehensive trait standards, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13].
-A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).
+Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16].
+On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13].
+Hosting services for  biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).
 
-Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measureing.
+Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measuring.
 For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in Thesauri.
 
-While trait thesauri define traits for focal groups of organisms, they are not specifing the  formats or structures in which trait data should be stored. 
+While trait thesauri define traits for focal groups of organisms, they are not specifying the  formats or structures in which trait data should be stored. 
 The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables.
-The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
+The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
 GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16].
@@
 The entity or observation (i.e.
-the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon.
+the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.
 This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
@@
 a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, refering to the entire taxon (or a subset, e.g.
+reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
 a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}
@@
 It may store additional information (e.g.
-on variation of means or literature source) in secondary colums or accompanying datasheets.
+on variation of means or literature source) in secondary columns or accompanying datasheets.
 The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
@@
 Also at this level of information, missing entries, e.g.
-about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15].
+about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using Bayesian approaches [@schrodt15].
 However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].
@@
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
-Longtable dataset purport multiple advantages for the data manipulation (e.g.
-filtering, subsetting and aggregating data), visualization (e.g.
-plot measured values by factor variable or taxon) and statistical modeling (e.g.
+Long-table dataset purport multiple advantages for the data manipulation (e.g.
+filtering, sub-setting and aggregating data), visualization (e.g.
+plot measured values by factor variable or taxon) and statistical modelling (e.g.
 ANOVA for testing difference of trait value by sex) [@wickham14].
@@
 a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e.
-one physical occurrence of a taxon; c) a measurement longtable is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.
+one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.
 }
@@
 We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major webservices and existing standards for biodiversity data management.
+We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.
 
@@
 
-As described above, tidy longtable formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.
+As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.
 
-To enable a better standardization of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
+To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
 In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
-If data contain multilayered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
+If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
 For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').
@@
 In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
-Therefore we recommend use of  Darwin Core Archives [DwCA; @robertson09], the structure proposed for GBIF [@gbif17; http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr15; http://eol.org/info/cp_archives].
+Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09], the structure proposed for GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr16,  http://eol.org/info/cp_archives].
 This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.
+
+In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road.
+This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
+The trade-off is data-consistency and readability *vs.* avoidance of content duplication.
+
+Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g.
+microbial functional traits assessed at the community level taken from a soil sample.
+Here, location or measurement information are in the primary focus of the investigation (see vocabulary extensions below).
+A general definition, whether a column describes asset data or is part of the central dataset is ill advised.
+Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
 
@@
 via the Application Programming Interface, API).
-Whereever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
+Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
 
@@
 
-It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.
-This is acheived by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g.
+It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.
+This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g.
 GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').
-By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practiced in TRY; @kattge11].
+By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11].
 This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
@@
 
-The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
-More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
+The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
+The accessibility by software differs widely.
+For biodiversity data, the OBO Foundry (http://www.obofoundry.org/) or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer API services for computational access.
+More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
 
@@
 It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
-We encourage publishing trait thesauri developed for the own project context as public resources, e.g. on OBO Foundry, the GFBio Terminology service, or on other stable web services.
+By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset.
+The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.
 
@@
 We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
+
 ### Metadata
@@
 Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a set of terms that allows to relate trait data to authorship and define a bibliographic reference and license model.
+We define a set of terms (https://ecologicaltraitdata.github.io/TraitDataStandard/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.
 
@@
 Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-<!--
-To avoid redundancy and duplication of content, those information would usually be kept in seperate data sheets of a spreadsheet document.
-Examples are the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited.
-A unique identifier would link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
-This is usually the case if the occurrences are externally documented, for instance as specimens from a museum. This also applies if the data are stored in additional data tables.
 
-For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').
-EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@gbif17].
-These are .zip archives containing data table txt-files along with a descriptive metadata file (in .xml format).
-Detailled descriptions and tools for validation can be found on the website of EOL (http://eol.org/info/cp_archives) and GBIF (http://tools.gbif.org/dwca-assistant/).
+The data standard provides two extensions of the vocabulary that should be used to describe these information:
 
-Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
-However, for standalone dataset publications on a hosting service without such file format recommendations (e.g.
-Figshare), data authors may want to avoid the expansion into multiple data tables and rather keep all columns in the same file, which is legitimate.
--->
-
-<!--- WORK HERE!!! -->
-
-It also may be safer to add such information directly as own columns of the data table to facilitate an analysis of cofactors and correlations further down the road.
-This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
-The trade-off is data-consistency and readability vs. avoidance of duplication.
-
-Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample.
-Here, location or measurement information are in the primary focus of the investigation.
-A general definition, whether a column describes asset data or is part of the central dataset is ill advised.
-Therefore, our glossary of terms should be used to describe the scientific data according to the study context.
-
-The data standard provides two extensions of the namespace that should be used to describe these datasets:
-
-- the `occurrence` extension contains information on the level of individual specimens, such as date and location and method of sampling and preservation, or physiological specifications of the phenotype, such as sex, life stage or age.
-- the `measurementOrFact` extension takes information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- the `occurrence` extension (https://ecologicaltraitdata.github.io/TraitDataStandard/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age.
+This also includes the date and geographical location, as well as the method of sampling and preservation, which provides an important resource to analyse trait variation due to differences in space and time.
+- the `measurementOrFact` extension (https://ecologicaltraitdata.github.io/TraitDataStandard/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+It aims at important information to track potential sources of noise or bias in measured data (e.g.
+variation in measurement method) or aggregated values (e.g.
+statistical method applied), as well as the source of reported facts (e.g.
+literature source or expert reference).
 
@@
 
-Some data-types may directly refer to existing global identifiers for occurrence data, e.g.
+Same as for taxon or trait information, these additional layers of information can either be added as extra columns to the core dataset or, to  avoid redundancy and duplication of content, be kept in separate data sheets of a Darwin Core Archive (see above).
+A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
 a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
@@
  <!-- Nadja: I like the idea of having a figure explaining the different terms -->
-
-<!--
-### additional information on measurement or fact
-
-As a high-level discrimination of the source of the measurement or fact, the Darwin Core Term `basisOfRecord` takes an entry about the type of trait data recorded: Were they taken by own measurement (distinguish "LivingSpecimen", "PreservedSpecimen", "FossilSpecimen") or taken from literature ("literatureData"), from an existing trait database ("traitDatabase"), or is it expert knowledge ("expertKnowledge").
-It is highly recommended to provide further detail about the source in the column `basisOfRecordDescription`.
-
-To keep track of potential sources of noise or bias in measured data, the method of measurement (`measurementMethod`), the person conducting the measurement (`measurementDeterminedBy`), and the date at which the measurement was obtained (`measurementDeterminedDate`) are recorded.
-Authors would often report aggregate data of repeated or pooled measurements, e.g.
-by weighing multiple individuals simultaneously and calculating an average.
-In these cases, recording the number of individuals (`individualCount`) along with a dispersion measure (e.g.
-variance or standard deviation, `dispersion`) or range of values (e.g.
-min and max of values observed in the field `measurementValueMin`, `measurementValueMax`) is adviced.
-The field `statisticalMethod` names the method for data aggregation (e.g.
-mean or median) as well as the variation or range (e.g.
-reporting variance or standard deviation).
-
-For data not obtained from own measurement, the field `references` provides a precise reference to the source of data (e.g. a book or existing database) or the authority of expert knowledge. 
-For literature data, the original source might report trait values on the family or genus level, but the dataset author infers and reports the trait data at species level (e.g.
-if the entire genus reportedly shares the same trait value).
-To preserve this information, the column `measurementResolution` should report the taxon rank for which the reported value was originally assessed.
-
-### additional information on observation context (occurrence)
-
-This category of terms contains further information about the individual specimen or occurrence that has been observed and measured.
-
-For both literature and measured data, trait values may be recorded for different sub-categories of individuals of a taxon to capture polymorphisms, for instance differentiated by sex or life stage.
-The template provides the fields `sex`, `lifeStage`, `age`, and `morphotype` for this distinction.
-
-Seasonal variation of traits may be recored by assigning a date and time of sampling to the occurrence, using the fields `year`,	`month` and	`day`, depending on resolution.
-The field definitions of the Darwin Core Standard can be applied instead, to refer to a geological stratum, for instance.
-Sampling may be further specified using a unique identifier for the sampling event (`eventID`) which references to an external dataset.
-The record of a `samplingProtocol` may capture bias in samling methods.
-Further procedures and methods of preservation should be reported in `preparations`.
-
-To capture geographic variation of traits, a set of fields for georeferencing can put the observation into spatial and ecological context (`habitat`, 	`decimalLongitude`,	`decimalLatitude`, `elevation`,	`geodeticDatum`, `verbatimLocality`, `country`, `countryCode`).
-The field `locationID` may be used to reference the occurrence to a dataset-specific or global identifier.
-This allows the trait data to double as observation data, e.g.
-for upload to the GBIF database.
--->
 
@@
 
-- preparing of own trait datasets for publication on low-threshold fileservers and project databases, and
+- preparing of own trait datasets for publication on low-threshold file-servers and project databases, and
 - harmonising trait datasets from different sources by moulding them into a unified format.
 
+A comprehensive documentation of the package is found in the package vignette, and on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
+
+The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
+
 The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format.
-It also maps column names into terms provided in the trait data standard.
+It also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
+This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:
+
+```
+library("traitdataform")
+dataset <- as.traitdata(read.csv("path/to/data.csv"), 
+             traits = c("body_length", "antenna_length", 
+                        "metafemur_length"),
+             units = "mm",
+             taxa = "name_correct", 
+             keep = c(locationID = "location")
+             )
+
+```
+
+The column containing taxon names is given in parameter 'taxa'.
+Note that the parameter 'keep' specifies and renames any data that should be maintained in the output.
+The parameter 'units' is used to specify the units of measurement.
 
@@
 Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added to the core data by calling the function `standardize.taxonomy()`.
-
-The `standardize.traits()` function matches user-provided trait names onto public trait ontologies and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
+Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
 a thesaurus of traits.
 The same function harmonises trait values into target units and legit factor levels.
+The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
 
-<!-- Add quick examples here -->
+```
+traitlist <- as.thesaurus(
+       body_length = as.trait("body_length", traitUnit = "mm", traitUnitStd = "mm", 
+         traitID = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length"), 
+       antenna_length = as.trait("antenna_length", traitUnitStd = "mm", 
+         traitID = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Antenna_length"),
+       metafemur_length = as.trait("metafemur_length", traitUnit = "mm",
+         traitID = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Femur_length")
+)
 
-The package functions comprise a toolchain where each function can be piped as an input into the next.
+datasetStd <- standardize.traits(dataset, thesaurus = traitlist)
+```
+
+The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`.
+Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`).
+Future iterations of the R package will aim at automatising the generation of trait thesaurus objects from globally available Ontologies.
+
+The package functions form a toolchain where each function can be piped as an input into the next.
 A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonising traitdata as simple as:
 
-```r
-library("traitdataform")
-traitdataset1 <- standardize(read.csv("path/to/data.csv"),
-            thesaurus = as.thesaurus("http://url.of/thesaurus.csv"),
-            taxa = "name_correct", units = "mm")
+```
+datasetStd <- standardize(read.csv("path/to/data.csv"),
+            thesaurus = traitlist,
+            taxa = "name_correct", 
+            units = "mm"
+            )
 ```
 
-Several datasets that have been produced by these functions can easily be appended using the `rbind()` function of R base.
+Datasets that have been produced by these functions can easily be appended using the `rbind()` function of R base, while maintaining any available metadata information as separate column entries.
 
-To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identfier using the function `merge()`.
-This enables an easy handling of data sources that have been stored in a Darwin Core Archive database format.
+To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identifier using the function `merge()`.
+This enables an easy handling of data sources that have been stored in a Darwin Core Archive format.
 
-For a detailed description on how to use the package see the package documentation website. 
-The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
-
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to datasets listed in our online spreadsheet that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @ricklefs17; @vanderplas17].
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @ricklefs17; @vanderplas17; @myhrvold15].
+See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
+We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
 

### 2017-11-13 manuscript/manuscript.Rmd
update figure 2

@@
 output:
+  word_document:
+    fig_caption: yes
+    fig_width: 6
   pdf_document:
@@
     fig_crop: no
+    keep_tex: yes
     template: pretty.template
-    keep_tex: yes
   html_document:
@@
     theme: united
-  word_document:
-    fig_caption: yes
-    fig_width: 6
 bibliography: div05.bib
@@
 \caption{Formats used for trait datasets.
-Identifiers relate the measurement to other layers of information on the specimen, the taxon or the trait measured.
-a) taxon-level traits are often published as a compiled species $\times$ traits matrix; b) measured data report measurements on the specimen-level, i.e.
-one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to an occurrence and taxon via unambiguous identifiers.
-}
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) wide-tables report measurements on the specimen-level, i.e.
+one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to a trait definition and taxon via unambiguous identifiers.
+Additional identifiers relate each row to other layers of information on d) the specimen (occurrence) or e) the origin or confidence on the reported measurement or fact.}
 

### 2017-11-14 manuscript/manuscript.Rmd
manuscript: minor note
fix URL and title

@@
 ---
-title: "Towards an Ecological Trait-data Standard"
+title: "Introducing an Ecological Trait-data Standard"
 author:
@@
 Hosting services for  biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).
+<!--General classes of traits are defined within the GeoBON Essential Biodiversity Variables [@pereira13].-->
 
@@
 
-The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/.
+The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/ETS/.
 
@@
 The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al.
-2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/".
+2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v1.0/".
 In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
@@
 
-Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/TraitDataStandard/. 
+Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/ETS/. 
 Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e.
@@
 It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
-By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset.
+By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset.
 The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.
 
-See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].
+See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus].
 We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
@@
 Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a set of terms (https://ecologicaltraitdata.github.io/TraitDataStandard/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.
+We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.
 
@@
 
-- the `occurrence` extension (https://ecologicaltraitdata.github.io/TraitDataStandard/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age.
+- the `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age.
 This also includes the date and geographical location, as well as the method of sampling and preservation, which provides an important resource to analyse trait variation due to differences in space and time.
-- the `measurementOrFact` extension (https://ecologicaltraitdata.github.io/TraitDataStandard/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- the `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
 It aims at important information to track potential sources of noise or bias in measured data (e.g.
@@
 
-The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
+The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
 

### 2017-11-28 manuscript/manuscript.Rmd
start major revision after comments: work on abstract, section 2&3 (turn into review)

@@
 ---
-title: "Introducing an Ecological Trait-data Standard"
+title: "Towards an Ecological Trait-data Standard"
 author:
@@
 - name: Caterina Penone \footnotemark[4]
+- name: Markus Fischer
+- name: Pete Manning 
+- name: Birgitta König-Ries
 - name: Nadja K. Simons \footnotemark[6]
@@
 output:
-  word_document:
-    fig_caption: yes
-    fig_width: 6
   pdf_document:
@@
     theme: united
+  word_document:
+    fig_caption: yes
+    fig_width: 6
 bibliography: div05.bib
@@
 
-1.	The assessment and analysis of functional traits has been established in all domains of ecological research, promising great advancements for biodiversity and climate research.
-A wealth of data is available from multiple sources and for many organism groups but is lacking standardisation due to heterogeneous data formats and data types.
-2.	To harmonize and standardise datasets, we present a general vocabulary and structure for an ecological trait-data standard for storing and sharing functional trait data.
-3.	We further provide an R-package that assists in transferring own data structures into the data standard and to harmonize and merge trait datasets from heterogeneous sources. 
-4.	With these decentralised tools for traitdata management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+1.	The use of functional trait data is now widespread in all domains of ecological research, thanks to the great potential for such data to deliver general conclusions regarding the nature of biodiversity.
+Accordingly, a wealth of trait data is now available for many organism groups but due to lacking standardisation these data come in heterogeneous data formats and data types.
+2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
+3. To aid the the standardisation and harmonization of distributed data sources, we present a general and simple vocabulary and structure for storing and sharing ecological trait data. 
+4.
+We provide an R-package that enables the transformation of any data into the proposed form.
+This also allows to harmonize and easily merge trait datasets from heterogeneous sources, which greatly facilitates the compilation of databses of particular research focus.
+5. With these decentralised tools for trait-data management, we intend to facilitate the exchange and analysis of trait data for ecological research.
 
@@
 
-functional ecology, traits, semantic web, ontologies, data standard
+functional ecology, functional traits, semantic web, ontologies, data standard
 
@@
 Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07].
-The scope of trait-based research spans from the evolutionary basis of individual-level properties to global patterns of biodiversity and ecosystem functioning.
+Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
+For example, trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity and ecosystem functioning (refs). 
+Using traits is a promising venue to bypass taxonomic impediment, i.e.
+the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring both the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
+With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].
 
-Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04].
+<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04].
 Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e.
@@
 Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
+--> 
+<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
+Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->
 
-In recent years, functional ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
-Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. 
 However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. 
@@
 Nonetheless, open-access publications in low-threshold data repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation.
-Although open for manual download, data on low-threshold data platforms might be stored in proprietary (e.g.
-.xlsx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious or impossible.
-
-On the other hand, not all centralised databases for functional traits comply with the criteria of open access, which leaves them at a disadvantage compared to low-threshold repositories. 
-Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit.
-Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
-However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
-Low-threshold data repositories assume public domain and Creative Commons Attribution licenses compliant with the demands of journals and research funding.
 
@@
 
-# 2. Overview of trait-data initiatives
+# 2. A review of trait-data initiatives
 
-In this section, we list two kinds of initiatives: the first aims at the harmonization of trait data into trait databases; the second develops trait ontologies and data standards.
-Often ontologies are being developed to serve a particular database project and thus are being developed in concert, as in the case of the TRY plant database [@kattge11] and the TOP Thesaurus of plant traits [@garnier17].
+In this section, we review four kinds of initiatives of relevance for trait data standardisation: 
+the first are projects that work with trait datasets that have been assembled out of a primary interest in the project context; 
+the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
+the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and Thesauri; 
+the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'semantic ontologies'.
 
-## Databases and Datasets
+We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11] and the TOP Thesaurus of plant traits [@garnier17].
+The objective of this review is to show how the degree of trait-data standardisation in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources.
 
-Plant flower, leaf and stem traits have been measured for decades, and were compiled into comprehensive databases and published as static datasets, e.g.
-the LEDA traitbase for the Northwest-European flora [@kleyer08].
-Regarding webservices, most prominently, plant traits from many different sources have been collated into the extensive TRY database [@kattge11].
-Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and root measurements [Fine-root Ecology Database; @iversen17].
-Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
-Trait-data for lower plants are less completely assessed.
+## Trait datasets
 
-Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12] derived from metagenomic analysis [@torsvik02; @langille13].
-Databases are at hand to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
-the KEGG orthology, @kanehisa12].
+In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books.
+With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
+A plethora of such static datasets has been published along with scientific articles [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13].
 
-In the animal kingdom, n unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17; @sokol11; @price13].
-In the past two decades, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g.
-@ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
-Specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems.
-For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is lacking.
-It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@moretti17].
-
-A general database for major physiological and life-history traits of all taxa is the Encyclopedia of Life Trait Bank project [@parr16, see below for more detail].
+Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org.
+These  platforms come with very low thresholds in terms of metadata documentation and standardisation.
+Furthermore, they warrant long-term accessibility (offering DOIs for referencing), and assume Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
 
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-As an attempt to track  databases and unstandardised datasets from independent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use.
-We invite data owners and users to register further datasets qualifying as trait data.
+Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list).
+We invite data owners and users to register further datasets qualifying as trait data to this spreadsheet.
 
-## Thesauri and Ontologies
+However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g.
+.xlsx, .docx) or binary (e.g.
+.pdf) data formats which make a programmatical extraction tedious and depending on commercial software.
+The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work.
+With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information.
 
-A glossary or thesaurus of terms would be primarily a listing of definitions.
-Expanding on this, ontologies are linking individual terms with each other by formally defining the relationship between terms, to the ends of enabling a computational interpretation of data.
+## Trait databases
 
-Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16].
+<!-- community database backend, but static publication --> 
+In the past two decades, many distributed datasets have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
+@klotz02; @iversen17; @kleyer08; @ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
+[@wright04; ]
+Most of these data are being published as static datasets and are being updated irregularly.
+Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative.
+Few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.
+
+<!-- database backend, selective download, user access management --> 
+A more dynmically growing, advanced online portal,  offer 
+selective data download 
+user access and rights management 
+webservices (API)
+
+most prominently, plant traits from many different sources have been collated into the extensive TRY database [@kattge11].
+Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
+
+Specialised trait-databases have been created that cover different taxa, interaction types (e.g.
+pollination traits or feeding relationships), or ecosystem types.
+For example, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
+
+<!--Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12].
+Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13].
+Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
+the KEGG orthology, @kanehisa12].
+-->
+
+Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is still lacking.
+The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
+
+On the other hand, not all centralized databases for functional traits comply with the criteria of open access, which leaves them at a disadvantage compared to low-threshold repositories.
+
+## Thesauri and Ontologies for traits
+
+In database management, a glossary or thesaurus of terms is primarily a listing of definitions.
+Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data.
+Relating the trait definition to methodological definitions on trait measurements is key for harmonizing trait data already at the physical process of measuring.
+For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in thesauri and ontologies.
+
+In the context of trait-based research, several thesauri of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16].
 On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13].
-Hosting services for  biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).
-<!--General classes of traits are defined within the GeoBON Essential Biodiversity Variables [@pereira13].-->
+More general classes of traits are defined within the GeoBON Essential Biodiversity Variables [@pereira13].
+Platforms hosting biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).
 
-Relating the trait definition to methodological definitions on trait measurements are key for harmonizing trait data already at the physical process of measuring.
-For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in Thesauri.
+Most of these ontologies are rather simple in structure, and mainly concerned with defining terms, rather than defining relationships of terms.
+The degree to which they define body-parts, life-history properties, and measurement procedures, or how they define expected standard units, differs widely.
 
-While trait thesauri define traits for focal groups of organisms, they are not specifying the  formats or structures in which trait data should be stored. 
-The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables.
-The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
+## Ontologies for the semantic web
+
+A general database for major physiological and life-history traits of all kingdoms is the Encyclopedia of Life Trait Bank project [@parr16, see below for more detail].
+
+While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored. 
+Recently, the Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables.
+The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
 GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16].
@@
 
-However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities.
+However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in biodiversity data management for structured input from the research communities.
 The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
@@
 <!-- examples for blind spots -->
-The other reason is competence: trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.
+The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.
 
@@
 As a minimum consensus, trait datasets may be defined as follows: 
-A trait dataset contains quantitative measurements or qualitative facts (i.e.
+a trait dataset contains quantitative measurements or qualitative facts (i.e.
 trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
@@
 
+<!--Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit.
+Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
+However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
+-->
+
 <!-- 
@@
 a thesaurus of traits.
-The same function harmonises trait values into target units and legit factor levels.
+The same function harmonises trait values by converting values into target units and legit factor levels.
+
 The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
@@
 traitlist <- as.thesaurus(
-       body_length = as.trait("body_length", traitUnit = "mm", traitUnitStd = "mm", 
-         traitID = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length"), 
-       antenna_length = as.trait("antenna_length", traitUnitStd = "mm", 
-         traitID = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Antenna_length"),
-       metafemur_length = as.trait("metafemur_length", traitUnit = "mm",
-         traitID = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Femur_length")
+   body_length = as.trait("body_length", 
+     expectedUnit = "mm", valueType = "numeric", 
+     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length"), 
+   antenna_length = as.trait("antenna_length", 
+     expectedUnit = "mm", valueType = "numeric", 
+     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Antenna_length"),
+   metafemur_length = as.trait("metafemur_length", 
+     expectedUnit = "mm", valueType = "numeric", 
+     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Femur_length")
 )
@@
 
+# Online Resources
+
+The online reference for the Ecological Traitdata Standard described in this paper is https://ecologicaltraitdata.github.io/ETS/
+
+The package website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform
+
+The living spreadsheet on existing trait datasets and databases is found at
+
+Development of the Standard and the R-package is coordinated via https://github.com/EcologicalTraitData/ .
+
 # References

### 2017-11-28 manuscript/manuscript.Rmd
update review section: merge ontology concepts from later sections, remove data table formats (for now)

@@
 
-# 1. Need for data standardisation
-
 Despite the obvious benefit and grand prospects of functional approaches, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.
@@
 
-# 2. A review of trait-data initiatives
+# 1. A review of trait-data initiatives
 
@@
 With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13].
+A plethora of such static datasets has been published along with scientific articles [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].
 
@@
 
-## Trait databases
+## Database initiatives
 
@@
 In the past two decades, many distributed datasets have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
-[@wright04; ]
-Most of these data are being published as static datasets and are being updated irregularly.
-Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative.
+@klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
+<!--[@wright04; ] -->
+Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16]. 
+Most of these databases are published as static datasets and are updated irregularly. Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
 Few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.
@@
 <!-- database backend, selective download, user access management --> 
-A more dynmically growing, advanced online portal,  offer 
-selective data download 
-user access and rights management 
-webservices (API)
+More specialised trait-databases have been created that cover different taxa, interaction types (e.g.
+pollination traits or feeding relationships), or ecosystem types.
+These centralized database initiatives invite data submissions from  a defined research field and offer to take care of the harmonization process and greatly facilitate data synthesis.
+For example, traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
+The BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
 
-most prominently, plant traits from many different sources have been collated into the extensive TRY database [@kattge11].
-Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
-
-Specialised trait-databases have been created that cover different taxa, interaction types (e.g.
-pollination traits or feeding relationships), or ecosystem types.
-For example, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
+By aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other platform.
+The online platform enables selective data download and user permission and rights management.
+Being a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is still lacking.
+The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
 
@@
 
-Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is still lacking.
-The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
-
-On the other hand, not all centralized databases for functional traits comply with the criteria of open access, which leaves them at a disadvantage compared to low-threshold repositories.
+Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.
 
@@
 
-In database management, a glossary or thesaurus of terms is primarily a listing of definitions.
+In database management, a glossary or thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
 Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data.
-Relating the trait definition to methodological definitions on trait measurements is key for harmonizing trait data already at the physical process of measuring.
+Providing relational ontologies in a machine readable form makes them compliant with semantic web standards, thus supporting the goal of widely interlinked data structures for the world wide web [@berners-lee01; @page08].
+
+<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
+units or legit factor levels) of measured values or reported facts.
+It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
+-->
+
+Standard trait definitions have been provided in all shades along this continuum, ranging from listings of definitions in glossaries, over well-defined thesauri to relational ontologies.
+The initiatives behind these are first and foremost concerned with consensus building on trait definitions.
+Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring.
 For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in thesauri and ontologies.
+Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
+Thesauri of trait definitions have been developed for plants [e.g.
+the Planteome ontology for crop plants, @walls12; @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
+the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13].
+Comprehensive trait ontologies, including hierarchies of parent terms, have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and on the side of animal traits, in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
 
-In the context of trait-based research, several thesauri of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16].
-On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13].
-More general classes of traits are defined within the GeoBON Essential Biodiversity Variables [@pereira13].
-Platforms hosting biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).
+These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control [@]. 
+For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
+These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets.
+For example, a measurement of seed size would be linked to Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
 
-Most of these ontologies are rather simple in structure, and mainly concerned with defining terms, rather than defining relationships of terms.
-The degree to which they define body-parts, life-history properties, and measurement procedures, or how they define expected standard units, differs widely.
+To be unambiguous, any thesaurus should be defining terms based on other well-defined terms from semantic ontologies.
+The definition provided here contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
+This way, each trait definition may link to a broader or narrower term.
+For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
+Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies.
+By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.
 
-## Ontologies for the semantic web
+The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+To harmonize trait data across databases, future trait standard initiatives should provide this functionality.
+More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.
 
-A general database for major physiological and life-history traits of all kingdoms is the Encyclopedia of Life Trait Bank project [@parr16, see below for more detail].
+## Trait-data structures for the semantic web
 
-While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored. 
-Recently, the Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables.
-The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g.
-GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16].
-Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions.
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive.
-Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].
+While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g.
+for taxa.
+To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.
 
-However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in biodiversity data management for structured input from the research communities.
-The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
-None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
-One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
-<!-- examples for blind spots -->
-The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.
-
-# 3. The structure of trait datasets
-
-As a minimum consensus, trait datasets may be defined as follows: 
-a trait dataset contains quantitative measurements or qualitative facts (i.e.
+Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
+More specifically, a trait dataset contains quantitative measurements or qualitative facts (i.e.
 trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
-traits) observed on an individual organism or parts of an individual (i.e.
-an entity) that can be assigned to a biological taxon (i.e.
-a species or higher-level taxon) <!-- or larger organisational unit, e.g.
-a community -->.
-
-Quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig.
+traits) observed on an biological entity (i.e.
+an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e.
+a species or higher-level taxon).
+Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig.
 1*a*), by aggregation of replicated measurements on multiple entities (Fig.
@@
 the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.
-This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
-<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?
-Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs.
-literature, but not entirely.
-A differentiation that has rarely been specified in the trait-litereature it seems.
-Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further.
-Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive).
-Maybe it helps if we add an example after the definition?
---> <!--FDS: is this sufficiently defined now?
-the distinction between measured and literature data is spelled-out further below, when discussing the formats.
-Quantitative values can be taken from literature, though, like body sizes for example.
--->
 
@@
 
-Trait-datasets usually take different formats.
-For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a.
-wide-table format,  Fig.
-2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available.
-This format is usually reporting missing data as 'NA'.
+These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain taxon names, sex or life stage, time and date of observation, as well as measurement values and methodological detail.
+
+Recently, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
+The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
+For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology.
+Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
+TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
+The archive also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.
+
+However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand in biodiversity data management for structured input from the research communities.
+
+The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
+None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
+One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
+<!-- examples for blind spots -->
+The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.
+
+<!--
+Depending on the research objective, trait-data are stored in a range of different table formats.
+In the subsequent paragraphs, we discuss the most common formats  in relation to the use cases in trait-based ecology.
+
+### a) Species wide-table
+
+If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (Fig.
+2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available.
+This format needs to clearly differentiate missing data and traits not applicable to a given taxon.
 It may store additional information (e.g.
 on variation of means or literature source) in secondary columns or accompanying datasheets.
-The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon).
-Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a].
+The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
+Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a].
 Also at this level of information, missing entries, e.g.
-about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using Bayesian approaches [@schrodt15].
-However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].
+about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14] and using Bayesian approaches [@schrodt15].
+<!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].
+
+### b) Occurrence wide-table
 
@@
 This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.
+
+### c.) measurement long-table
 
@@
 
-# 4. Proposing an ecological traitdata standard
+# 2. Proposing an ecological traitdata standard
 
-With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research.
+With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
 We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
@@
 In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
-Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09], the structure proposed for GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr16,  http://eol.org/info/cp_archives].
-This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.
+Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09].
 
@@
 It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.
-This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g.
-GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').
+This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g.
+GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.
 By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11].
@@
 
-The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
+<!--The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
 The accessibility by software differs widely.
@@
 More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
+-->
 
 If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
+<!-- A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
 Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
@@
 It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
+-->
 By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset.
@@
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @ricklefs17; @vanderplas17; @myhrvold15].
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @vanderplas17; @myhrvold15].
 See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.

### 2017-11-30 manuscript/manuscript.Rmd
revise manuscript: minor phrasing, add notes of DF and AG

@@
 
-# 1. A review of trait-data initiatives
+# 1. A review of trait-standard initiatives
 
@@
 With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].
+A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].
 
@@
 Furthermore, they warrant long-term accessibility (offering DOIs for referencing), and assume Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
+<!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ?
+-->
 
@@
 The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work.
+<!--To cope with this trend towards distributed data sources, data brokering services (e.g.
+the German Federation on for Biological Data; gfbio.org) or database management systems for scientific projects (e.g.
+BExIS, XX ) are likely to gain importance.
+These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
+One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.--><!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
 With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information.
@@
 The online platform enables selective data download and user permission and rights management.
-Being a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
-Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is still lacking.
+As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
 The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
@@
 
-In database management, a glossary or thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
+A major problem in trait-data standardisation is the reference to accepted and unambiguous trait definitions. This is acheived via thesauri and ontologies. 
+In database management, a thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
 Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data.
@@
 For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
+<!-- Mention also NCBO BioPortal?
+https://bioportal.bioontology.org/ -->
 These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets.
@@
 of a behavioural or life-history trait (Fig.
-1.*d*).
+1*d*).
 The entity or observation (i.e.
@@
 \end{figure}
+<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”.
+Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->
 
@@
 These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain taxon names, sex or life stage, time and date of observation, as well as measurement values and methodological detail.
+The terminologies provided by these standards are quite universal and cover most use cases of trait data.
+<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
+The Trait Data Standard is a standard on how to store trait data measurements.
+In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard.
+For instance the ABCD standard has a huge section for MeasurementsAndFacts.
+And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible).
+Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD.
+While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.
+-->
 
@@
 TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-The archive also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.
+The archive also integrates a defined framework for metadata in the Ecological Metadata Language [EML, @knb11], which should be used for describing and linking the data tables.
 
@@
 Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
-This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.
+This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data but rarely found for reported qualitative facts obtained from the literature.
 
@@
 <!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->
-
+<!--
 \begin{figure}[t]
@@
 \end{figure}
+-->
 
@@
 
-With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
+With this paper, we propose a versatile vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
+
 We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
@@
 
-## structure for trait data
+## Standard structure for trait data
 
@@
 In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
+<!-- AG: This was a little bit confusing for me.
+In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon.
+On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name.
+In any case I would suggest to recommend to keep both the URI and the literal.-->
 If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
@@
 Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09].
+<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
 
@@
 This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
+<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
 The trade-off is data-consistency and readability *vs.* avoidance of content duplication.
@@
 
-## vocabulary for trait data
+## Standard vocabulary for trait data
 
 Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/ETS/. 
-Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e.
+Individual terms <!-- DF:Please explain the difference between the “individual terms” and the “field entries” above-->  should be referenced via their global identifier, which is hosted by the GFBio Terminology Service <!-- DF: Please add citation to the TS Paper in Datenbanken-Spektrum, see https://terminologies.gfbio.org/about/#citation  --> and also can be accessed programmatically (i.e.
 via the Application Programming Interface, API).
@@
 
-# 5. Computational tools for producing compliant data
+# 3. Computational tools for producing compliant data
 

### 2017-12-12 manuscript/manuscript.Rmd
add comments of BKR and AO

@@
 To cope with this trend towards distributed data sources, data brokering services (e.g.
-the German Federation on Biological Data; gfbio.org)  or database management systems for scientific projects (e.g.
-BExIS, XX ) are likely to gain importance.
+the German Federation on Biological Data; gfbio.org)  or database <!-- AO: Just “data” or “research data”--> management systems for scientific projects (e.g.
+BEXIS, XX <!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance.
 These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
@@
 
-After all, to mobilise the broad wealth of data into structured databases, it requires a certain minimum of information from the data providers.
+After all, to mobilise the broad wealth of data into structured databases <!-- AO: Maybe better: formats-->, it requires a certain minimum of information from the original data providers.
 Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.
+<!-- AO: The last paragraphs may need more sources/proofings?-->
 
@@
 .xlsx, .docx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious and depending on commercial software.
+.pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.
+<!-- BKR: This might also result in data becoming no longer accessible (because the last computer on which the excel version required to read a file ran died -->
 The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work. 
@@
 Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.
+<!-- AO: In a sense of data availability to the public : yes.
+But do we discuss it here?
+We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs.
+non-open data, or?
+bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
+FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
+ -->
 
@@
 It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
+--><!-- BKR: Isn’t that the same?
+Or can you have multiple parents for one node?
+FDS: agree.
+simplify phrasing.
 -->
@@
 trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
-traits) observed on an biological entity (i.e.
+traits) observed on a biological entity (i.e.
 an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e.
@@
 
-If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (Fig.
+<!-- AO: You argue somehow that matrix format I a result of lit.
+review and so on.
+I would not agree with it.
+Your example also “just” points to aggregated data which menas one row is on species (aggregated) value.
+Your example is (for me) not really a matrix.
+In a matrix you have a relation between two variables.
+In the first column and and on the upper row normally the values of one variable are written.
+An example would be species x body length.
+Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted.
+E.g.
+spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on.
+Your wexample hat already different variables (=traits) with different units as columns.
+That is a meaningfull representation of variables and values in a normal data table.
+Maybe it is good her to show the enormous number of different investigated traits and point to the resulting big (many columns = wide) tables.
+--><!--
+If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait wide-table format (Fig.
 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available.
-This format needs to clearly differentiate missing data and traits not applicable to a given taxon.
+This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g.
+'999').
 It may store additional information (e.g.
 on variation of means or literature source) in secondary columns or accompanying datasheets.
-The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
+The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
+<!-- AO: You say it already 2 sentences before --><!--
 Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a].
@@
 <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].
+--><!--AO: Why not?
+I can filter for a specific species and get all respective rows.
+But this is not a result a using a data format, it is a result of the data itself, FDS: no, since only one row per species exists; true that data itself don't cover this variation.
+-->
+<!--
 
@@
 Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns.
-Each trait column represents one measured value.
+Each trait column represents one value that applies to the entity in question (i.e.
+to a species in the first case or to an occurrence in the latter case).
 The width of the table varies with the number of traits included.
-Merging data from different sources requires manual harmonization of these structures.
+Merging data from different sources requires user-defined mapping and manual harmonization of these structures.
 
 Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
+<!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes.
+But not necessarily computational more effective at all.
+E.g.
+I start to mix up numeric values with descriptive values.
+I start to mix up different units.
+Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g.
+quality stuff.
+Ergo less effective from that perspective.
+But it has some advantages which may overrule dis.
+See below
+AO: For the whole proofing why to use such a long format I think that GBIF and DwC has already given some sentences. I strongly think their work should be considered and can be used as reference!
+We can simply argue that a well established routine should now also executed on trait level!
+Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.
+
+--><!--
 This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID.
@@
 They further draw from a defined set of columns, which simplifies merging of datasets.
+<!-- BKR: Is that really true?
+Of course, you have less problems on the schema or structure level.
+But, on the other hand, the differences (e.g.
+different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows).
+Thus, at least with respect to integration, you just move the problem from one level to the other.
+To ease integration, one would need to agree on the entries in the features column
+AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
+We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
+Again: I think we find here papers regarding DWC and GBIF (and OBIS) as source of proofing the facts.
+-->
 
@@
 We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.
+<!-- AO: Webservices!!
+;  Really to be considered as standards?
+-->
 
-The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/ETS/.
+The online reference glossary of terms and description of the proposed structure can be found at https://ecologicaltraitdata.github.io/ETS/.
 
-The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution.
+The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ...
+-->.
 We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.
@@
 As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.
+<!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
+-->
 
 To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
-In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
+In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
+<!-- BKR: There is some grammatical error here; not sure what you try to say-->
 <!-- AG: This was a little bit confusing for me.
@@
 If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
-For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').
+For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt') with a unified character encoding (e.g.
+UTF-8).
+<!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->
 
@@
 <!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
-
 In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road.
@@
 Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
+<!-- FDS: shorten this and refer to existing standards for database use. -->
 
@@
 Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms.
+<!-- BKR: Why do you need this, if taxonIDs are used.
+They should be unambiguous
+FDS: right, rephrase to: either use taxonID or provide this.-->
 It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
@@
 Many terms of these extensions refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
+<!-- BKR: Looking at your definitions, in most cases, I am not sure how you refine the terms. Is it really refinement or is it just using them with the same meaning they have in DwC?
+-->
 
@@
 
-To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
+To access data from public databases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
 The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
@@
 Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added to the core data by calling the function `standardize.taxonomy()`.
+<!-- BKR: In the example below, this looks like a manual matching.
+What is done automatically here and which information does the user have to provide?
+FDS: taxon is matched automatically, traits not.
+lack of trait APIs and webservices.
+-->
 Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
@@
 The same function harmonises trait values by converting values into target units and legit factor levels.
+<!-- BKR: Have you thought about linking to unit ontologies here?
+AO: This paper is a good reference and gives an idea of which ontology could be used;
+http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
+FDS: might be referred to in the section about thesauri -->
 
@@
 
-We therefore encourage the development of trait thesauri in semantic ontologies.
+We therefore encourage the development of trait thesauri in ontologies.
 The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.

### 2017-12-12 manuscript/manuscript.Rmd
add comments of Pete and Caterina (#26)

@@
 --> 
+<!-- CP: I like very much this whole paragraph (both writing and content), but I feel that it’s a bit out of the topic.
+It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
+I give some suggestions on how to cut it and rather develop more the following one, which people are surely less aware of and which sets more the aim of the paper.
+I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
+
+That’s the structure I try to follow:
+1. Traits are important in ecology, evolution and conservation (i.e. this paragraph)
+
+2.
+To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources.
+When we do this we can answer new questions and find generalities.
+So there is a growing interest in assembling trait data from different sources and taxa (e.g.
+give the number of papers using TRY).
+
+3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.
+
+4.
+One solution are databases.
+But these are costly and depend on the will of few persons/funding.
+They also cover few taxonomic groups.
+Matching between different databases is still difficult because of different definitions and structures.
+Most data are still uploaded separately.
+We need a solution to make their use easy.
+
+But of course it’s just a suggestion, feel free to happily ignore it!
+
+I would choose 2 or 3 (also for the previous), anyway it’s impossible to be exhaustive. Maybe put (e.g. ref1, ref2) so make clear that it’s just examples
+
+-->
+<!-- PM: Thuis is quite complex- e.g.
+‘taxonomic impediment’etc.
+I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples.
+I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.
+
+Huge number of references here- I think you can strip it down to a few key ones, and reviews.
+Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site.
+-->
 <!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
@@
 However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa.
-To mobilize data for this enterprise, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources.
+<!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific?
+-->
+To mobilize data for this enterprise <!-- PM: Please define what the exercise is- too vague at present
+-->, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources.
+<!-- PM: I think this is a key justification- perhaps say what happens if we don’t do this – non repeatable work, differing definitions of traits that may lead to scientific flaws, repetition of the painstaking task of processing trait data e.g.
+to compute species level means.
+-->
 The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way.
-Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g.
+<!-- CP: I felt that this needed a definition, not sure everybody know what an ontology is.
+Maybe we should guide more the reader that might not be familiar with the problems.
+-->
+
+Nonetheless, individual researchers often neglect the benefit of centralised databases <!-- PM: Explain what this benefit is.
+-->.
+Furthermore, many journals and research funding agencies require  data to be published without access restrictions [@swan12; @allison15].
+However, this primary data is often posted on file hosting services, where no data standards are forced upon the uploaded material (e.g.
 Figshare, Dryad).
-Given this prospect, there will be no lack of data, but a lack of data standardisation.
+The long-term outcome of this situation is an abundance of freely available data, but a lack of data standardisation.
 
@@
 We further (2.) review existing trait databases and web services, as well as initiatives for standardisation.
-We (3.) describe the common practice for structuring trait data for ecological research. 
-Based on these considerations, we (4.) propose a low-threshold structure and vocabulary for describing trait-datasets compatible with existing ontologies.
-(5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.
+We (3.) describe the common practice <!--PM: Current or your own new definition?--> for structuring trait data for ecological research. 
+Based on these considerations, we (4.) propose a low-threshold <!-- PM: What’s this? Please define
+-->  structure and vocabulary for describing trait-datasets compatible with existing ontologies <!-- PM & CAT: define--> .
+(5.) Finally, we provide an R-package that assists in aply this structure to own data and for the harmonisation of data from distributed sources.
 
-With an easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold  trait data standardisation.
+With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold <!-- PM:Again, I’m not sure what this is a low threshold of?
+--> trait data standardisation.
 
-Despite the obvious benefit and grand prospects of functional approaches, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.
+Despite the obvious benefit and great potential of functional trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
+All of these factors render trait data extremely heterogeneous.
 
-Trait data have therefore been standardised into centralised databases for specific organism groups and research questions.
-These compiled data are either being published as static data tables or in the form of sophisticated web-services (see below for a review of current initiatives).
-Services attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, but more importantly, provide platforms for consensus building in methods.
-Furthermore, database services offer access control and data usage policies, to claim the rights of the original data providers, while simplifying data queries for synthesis researchers.
+As a response to these issues, trait data have been standardized in centralized databases for specific organism groups and research questions.
+<!-- PM: (refs, also give an e.g. or two for some big ones like TRY- saw there was a new one for tropical mammals a couple of days ago too and a huge Chinese plants one- in )--> 
+These compiled data are either being published as static data tables or in the form of sophisticated online portals like the plant trait database TRY (see below for a review of current initiatives).
+These portals attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, and importantly also, provide networks for consensus building in measurement methods.
+Furthermore, database services offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.
 
-Nonetheless, open-access publications in low-threshold data repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation.
+In contrast to standardized databases, open-access data publication in low-threshold repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org is becoming increasingly frequent.
+Such repositories foster decentralised data hosting with low thresholds regarding data standardisation and documentation.
 
@@
 
-After all, to mobilise the broad wealth of data into structured databases <!-- AO: Maybe better: formats-->, it requires a certain minimum of information from the original data providers.
-Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.
-<!-- AO: The last paragraphs may need more sources/proofings?-->
+However, to mobilize the broad wealth of distributed data into structured databases <!-- AO: Maybe better: formats-->,  a certain minimum of information from original data providers is requrired, e.g.
+on the methodology and definitions of trait values or the covariates describing the ecological context of the trait observation.
+In order to get this accurate and well-defined input ecological researchers need to be aware of and understand basic principles of data standards and ontologies.
+<!-- AO: The last paragraphs may need more sources/proofings?--><!-- PM: Literacy is quite a strong word- as illiterate is quite an insult.
+I think awareness is a problem too.
+-->
+
+ More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
 
@@
 the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and Thesauri; 
-the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'semantic ontologies'.
+the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.
 
@@
 The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work.
-<!--To cope with this trend towards distributed data sources, data brokering services (e.g.
-the German Federation on for Biological Data; gfbio.org) or database management systems for scientific projects (e.g.
-BExIS, XX ) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.--><!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
-With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information.
+
+As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited.
+<!--With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information. -->
 
@@
 Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.
+<!-- FDS: not an option for primary publication ...!-->
 <!-- AO: In a sense of data availability to the public : yes.
@@
 Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
-More specifically, a trait dataset contains quantitative measurements or qualitative facts (i.e.
-trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e.
+More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
+trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
 traits) observed on a biological entity (i.e.
@@
 Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig.
-1*a*), by aggregation of replicated measurements on multiple entities (Fig.
-1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g.
+1*a*), by aggregating replicated measurements on multiple entities (Fig.
+1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g.
 databases, Fig.
@@
 The entity or observation (i.e.
-the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.
+the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
+a genus).
 
@@
 
+For the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that intends to cover many potential use cases.
+This data structure is centered around a central data table that contains a single observation of a trait per row, linking it to a taxon and possibly also a single occurrence of the plant species (i.e.
+an individual specimen or 'entity').
+<!-- FDS: enter comments of Jens Kattge here!-->
+
 Recently, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
@@
 
-However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand in biodiversity data management for structured input from the research communities.
+However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand from biodiversity data management for more structured input from the research community.
 
 The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
-None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
-One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
-<!-- examples for blind spots -->
-The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.
+<!-- PM: There is also little incentive for researchers to do this- why bother?
+Incentives are also needed.
+-->
+None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
+One reason for this is most certainly compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
+<!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific -->
+The other reason is awareness and literaci: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.
+
+As it stands, this lack of data standardization will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited. 
+To cope with the trend towards distributed data sources, data brokering services (e.g.
+the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g.
+BExIS, NEON ) are likely to gain importance.
+These services simplify and direct the standardized upload of research data and descriptive metadata into consistent and interlinked data infrastructures.
+One goal of such initiatives is to facilitate the publication and standardization of data for researchers, e.g.
+by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
 
@@
 to a species in the first case or to an occurrence in the latter case).
-The width of the table varies with the number of traits included.
-Merging data from different sources requires user-defined mapping and manual harmonization of these structures.
+As the comlumn names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.
 
-Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
+A computationally more effective method that allows for high flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16].
+In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
+<!-- FDS: also refer to kattge 11 -->
 <!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes.
@@
 --><!--
-This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID.
+This allows repeated measurements on a single individual to be stored by linking the data from separate rows via a unique specimen ID.
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
-Long-table dataset purport multiple advantages for the data manipulation (e.g.
+Long-table datasets purport multiple advantages for data manipulation (e.g.
 filtering, sub-setting and aggregating data), visualization (e.g.
@@
 ANOVA for testing difference of trait value by sex) [@wickham14].
-They further draw from a defined set of columns, which simplifies merging of datasets.
+As long-table fromats draw from a defined set of columns,  merging of datasets is much easier.
 <!-- BKR: Is that really true?
@@
 -->
-
-The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.
-Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig.
+<!--
+The long-table format therefore is ideal for storing raw data of trait measurements in huge databases, and allows the entries to be linked via the respective unique identifiers to additional information.
+Each row of the dataset can therefore be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig.
 2*c*).
@@
 
+<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
+
+FDS: see Garnier et al. but may contain:
+
+Term | Definition 
+----- | -----
+Terminology | 
+Thesaurus | 
+Ontology | 
+semantic web |
+
+-->
+
 # 2. Proposing an ecological traitdata standard
@@
 With this paper, we propose a versatile vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
+<!-- PM: It took a long time to get to this point- I suggest brininging in a section early on that describes what the paper will do as it’s not always clear where it is leading. --> 
+<!-- FDS: point out focus on distributed low-threshold data, not full-fledged database initiatives. -->
 
-We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.
+In designing this structure, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+We also paid particularl consideration the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
 <!-- AO: Webservices!!
@@
 
-The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ...
+We acknowledge that the vocabulary proposed will not capture the full complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ...
 -->.
-We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.
+Therefore, we invite researchers to contribute to the next iterations of this standard terminology and develop own applications and ontologies.
+
+<!-- PM: It seems that the object identifier for an individual is the key distinction here- this could be clearer.
+
+Maybe you address this later but does your tool assign one of these- how to you make sure they are all unique if there are multiple users- e.g.
+if you pooled together two previously complied datasets which might contain some of the same individuals?
+-->
 
@@
 
-As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.
+As described above, observation-level long-table formats are the most predictable and compatible format for the core content of trait datasets and should therefore also be used for standalone uploads on low-threshold repositories.
 <!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
 -->
+<!-- CP: Not sure this needs to be repeated -->
 
-To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
+To enable a better standardisation of nomenclatures and values, the entries of the core dataset <!-- CP: Does this need to be defined --> should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
 In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
@@
 In any case I would suggest to recommend to keep both the URI and the literal.-->
-If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
-For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt') with a unified character encoding (e.g.
+<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
+If data contain additional information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
+<!-- PM: Not clear here -->
+For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
 UTF-8).
@@
 
-The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al.
-2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v1.0/".
-In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
+In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider et al.
+2017 Ecological Traitdata Standard v0.7, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
+In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
+<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
+FDS: to consider!! -->
 
-In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
+In very complex datasets, the repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
 Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09].
+<!-- PM: Why do these avid this? -->
 <!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
-In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road.
-This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
+
+In some contexts, it may be safer to add such information directly as columns of the data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
+This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
 <!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
@@
 microbial functional traits assessed at the community level taken from a soil sample.
-Here, location or measurement information are in the primary focus of the investigation (see vocabulary extensions below).
-A general definition, whether a column describes asset data or is part of the central dataset is ill advised.
+Here, location or measurement information are the primary focus of the investigation (see vocabulary extensions below).
+<!-- PM: I get lost here-->
+A general recommendation, whether a column describes asset data or is part of the central dataset is ill advised.
 Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
-<!-- FDS: shorten this and refer to existing standards for database use. -->
+<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
 
 ## Standard vocabulary for trait data
+
+<!-- PM: Can you say a little about what this vocab is and how it was chosen/defined? -->
 
@@
 
-To describe the minimum core content, where each row is the reported measurement or fact for a single observation, the following columns are required: a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); a descriptive trait name (`traitName`);  the scientific taxon for which the measurement or fact was obtained (`scientificName`).
+To describe the minimum core content, where each row is the reported measurement or fact for a single observation, the following columns are required: 1.
+a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2.
+a descriptive trait name (`traitName`);  3.
+the scientific taxon for which the measurement or fact was obtained (`scientificName`).
 For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
-However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or identifiers instead.
+However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.
 
-### Relate to global taxon and trait definitions
+<!--### Relate to global taxon and trait definitions -->
 
-It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.
+In this case, it is essential to relate the user-defined names to a consensus standard of taxon names or or look-up tables of traits. 
 This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g.
 GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.
-By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11].
-This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side.
-Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms.
+By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
+This redundancy of data allows for continuity on the authors and data owners side while also ensuring quality checks and comparability for the data user.
+
+If no URIs are referenced for taxonID, it is highly recommended to keep an additional reference to a kingdom and order (using columns `kingdom`, `order` as described by DWC and suggested for EOL Traitbank) to avoid ambiguities with taxonomic homonyms.
 <!-- BKR: Why do you need this, if taxonIDs are used.
@@
 FDS: right, rephrase to: either use taxonID or provide this.-->
-It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level.
-This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].
+Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
+This information can then be used for trait inference at higher taxonomic resolution.
+<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
+-->
 
@@
 The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.
+<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
 
 See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus].
-We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
+We also provide an example thesaurus, that has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
 
@@
 Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.
+We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authorship, while also stating a bibliographic reference and license model.
 
-In the case of raw measurement data, these information apply to the entire trait dataset, which classifies them as metadata stored in the information structure where the data are published. 
-In cases, where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be linked to the measurement level, via an unambiguous `datasetID` and a descriptive `datasetName`.
+In the case of raw measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata. 
+In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be linked to the measurement level, via an unambiguous `datasetID` and a descriptive `datasetName`.
 
@@
 
-- the `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age.
+- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age.
 This also includes the date and geographical location, as well as the method of sampling and preservation, which provides an important resource to analyse trait variation due to differences in space and time.
-- the `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
-It aims at important information to track potential sources of noise or bias in measured data (e.g.
+- The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
 variation in measurement method) or aggregated values (e.g.
@@
 -->
+<!-- PM: All of this sounds like quite a lot of work- so will it be automated, or incentivized somehow?
+-->
 
-Same as for taxon or trait information, these additional layers of information can either be added as extra columns to the core dataset or, to  avoid redundancy and duplication of content, be kept in separate data sheets of a Darwin Core Archive (see above).
+As for taxon or trait information, these additional layers of information can either be added as extra columns to the core dataset or be kept in separate data sheets of a Darwin Core Archive (see above), this avoiding redundancy and duplication of content.
 A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
@@
 
-We provide the R package 'traitdataform' to assist in producing data compliant with the trait data standard proposed here. There are two major use cases for the package:
+Accordingly, we provide the R package 'traitdataform' to assist the production of data compliant with the trait data standard proposed here. There are two major use cases for the package:
 
-- preparing of own trait datasets for publication on low-threshold file-servers and project databases, and
-- harmonising trait datasets from different sources by moulding them into a unified format.
+1. preparing of own trait datasets for publication on low-threshold file-servers and project databases, and
+2. automating the harmonisation of trait datasets from different sources by moulding them into a unified format.
 
@@
 
-The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format.
-It also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
+The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
+2).
+This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
 This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:
@@
 Note that the parameter 'keep' specifies and renames any data that should be maintained in the output.
-The parameter 'units' is used to specify the units of measurement.
+The parameter 'units' is used to specify the input units of measurement.
 
@@
 
-Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added to the core data by calling the function `standardize.taxonomy()`.
+In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data.
 <!-- BKR: In the example below, this looks like a manual matching.
@@
 a thesaurus of traits.
-The same function harmonises trait values by converting values into target units and legit factor levels.
+The same function harmonises trait values by converting values into target units and unified factor levels.
 <!-- BKR: Have you thought about linking to unit ontologies here?
@@
 Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`).
-Future iterations of the R package will aim at automatising the generation of trait thesaurus objects from globally available Ontologies.
+Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies.
 
 The package functions form a toolchain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonising traitdata as simple as:
+A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
 
@@
 
-Datasets that have been produced by these functions can easily be appended using the `rbind()` function of R base, while maintaining any available metadata information as separate column entries.
+Datasets that have been produced by these functions can easily be appended using the function `rbind()` of R base, while maintaining any available metadata information as separate column entries.
 
@@
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @vanderplas17; @myhrvold15].
+Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
 See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
@@
 We therefore encourage the development of trait thesauri in ontologies.
-The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
-This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of the global plant trait database TRY shows intricately [@diaz16].
+The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
+This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].
 
@@
 
-The trait data standard as well as the R package are published for open source community development.
+<!--The trait data standard as well as the R package are published for open source community development.
 Contributions are invited and vital for the future development of the data standard and an extension and improvement of the R package.
+-->
 
@@
 
-We thank the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
-infrastructure; Christiane Fischer and Jule Mangels for giving
-support through the central office, Michael Owonibi for
-managing the central data base, and Markus Fischer, Eduard Linsenmair,
+We thank <!--the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
+infrastructure;--> Christiane Fischer and Jule Mangels for giving
+support through the central office, <!--Michael Owonibi for
+managing the central data base,--> and <!--Markus Fischer,--> Eduard Linsenmair,
 Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
@@
 The work has been partly funded by the DFG Priority Program
-1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). Field work
+1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). <!--Field work
 permits were issued by the responsible state environmental offices of
 Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
-BbgNatSchG).
+BbgNatSchG).-->
 
@@
 
-FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the glossary of terms of the trait data standard; FDS led the writing of the manuscript; NKS authored the list of traits; FDS developed the R package.
+FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package.
 All authors contributed critically to the drafts and gave final approval for publication.

### 2017-12-12 manuscript/manuscript.Rmd
add comments of Malte, Nadja, Gaetane (#26)

@@
 - name: Malte Jochum \footnotemark[4]
-- name: Gaetane LeProvost \footnotemark[2]
+- name: Ga\"etane Le Provost \footnotemark[2]
 - name: Andreas Ostrowski \footnotemark[5]
@@
 
-1.	The use of functional trait data is now widespread in all domains of ecological research, thanks to the great potential for such data to deliver general conclusions regarding the nature of biodiversity.
-Accordingly, a wealth of trait data is now available for many organism groups but due to lacking standardisation these data come in heterogeneous data formats and data types.
+1.	The use of trait-based approaches is now widespread in all domains of ecological research, thanks to the great potential for such data to deliver general conclusions regarding the nature of biodiversity.
+Accordingly, a wealth of trait data is now available for many organism groups but due to lacking standardisation these data come in heterogeneous data formats.
 2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
-3. To aid the the standardisation and harmonization of distributed data sources, we present a general and simple vocabulary and structure for storing and sharing ecological trait data. 
+3. In order to aid the the standardisation and harmonization of distributed data sources, we present a general and simple vocabulary and data structure for storing and sharing ecological trait data. 
 4.
@@
 This also allows to harmonize and easily merge trait datasets from heterogeneous sources, which greatly facilitates the compilation of databses of particular research focus.
-5. With these decentralised tools for trait-data management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+5.
+With these decentralised tools for trait-data harmonization, we intend to facilitate the exchange and analysis of trait data for ecological research and enable a global synthesis of traits across a wide range of organisms and ecosystems.
 
@@
 These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
-Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97].
-Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
-Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
+<!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ?
+Put instead:
+Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
+And Coux et al. 2016 Linking species functional roles to their network roles
+?
+-->
+<!--
+Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g.
+@reich97].
+<!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ?
+I’m not sure that the literature cited is always appropriate here ?
+-->
+<!--
+Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17].
 --> 
@@
 -->
+<!-- GLP: 
+The second paragraph is a bit confusing for me.
+
+I wonder also if the introduction could be restructures like this :
+
+Trait based approaches and their use
+From the individual to population
+Community level (assemblage, interactions, effects on ecosystem)
+Macroecological question
+
+ Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g.
+TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals).
+Say also something on the development of general protocols for measuring traits (e.g.
+Moretti et al.
+2016; Freschet et al.
+2017) ?
+
+Importance of data standardization.
+This is the main idea here and so I wonder if some idea of the first point (Need for data standardization) should not be merge here with the last paragraph of the introduction ?
+As there is also some repetitions between those 2 parts.
+
+There is just some copy-past to do.
+I didn’t change the structure of the introduction but if you like the structure that I proposed, we can discuss next week or I can help you with that.
+I like to write introduction 
+
+I wonder if you should speak about ‘Functional traits’here or just Traits ?
+As you are not only considered Functional traits in this paper (e.g.
+life history traits) ?
+Maybe remove the definition no ?
+-->
 <!-- PM: Thuis is quite complex- e.g.
@@
 -->
-<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
+<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions.
 Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->
 
-However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa.
+However, fully exploiting this portential relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa.
 <!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific?
@@
 
-In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users.
+In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users.
 We further (2.) review existing trait databases and web services, as well as initiatives for standardisation.
@@
 
-Despite the obvious benefit and great potential of functional trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
+Despite the obvious benefit and great potential of trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
 All of these factors render trait data extremely heterogeneous.
@@
 
-In this section, we review four kinds of initiatives of relevance for trait data standardisation: 
-the first are projects that work with trait datasets that have been assembled out of a primary interest in the project context; 
+In this section, we review four kinds of initiatives that are of relevance for trait data standardisation: 
+the first are projects that work with trait datasets that have been assembled out of a particular research interest, either by measurement or colated from literature; 
 the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
@@
 These  platforms come with very low thresholds in terms of metadata documentation and standardisation.
-Furthermore, they warrant long-term accessibility (offering DOIs for referencing), and assume Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
+Furthermore, the platforms warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
 <!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ?
@@
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list).
+Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy; Appendix A) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list).
 We invite data owners and users to register further datasets qualifying as trait data to this spreadsheet.
 
-However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g.
+However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
 .xlsx, .docx) or binary (e.g.
@@
 
-By aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other platform.
+By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform.
 The online platform enables selective data download and user permission and rights management.
@@
 traits) observed on a biological entity (i.e.
-an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e.
+an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e.
 a species or higher-level taxon).
-Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig.
+Expaning on this definition, quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
 1*a*), by aggregating replicated measurements on multiple entities (Fig.
@@
 
-Phenotypic variation of traits within species would usually be recorded per observation, i.e.
-each individual specimen would be recorded in a row, with the respective trait values stored in columns.
-In these occurrence data tables (Fig.
-2*b*), each row is centered around a single physical occurrence of a species.
+In these occurrence wide-tables (Fig.
+2*b*), each row is centered around a single physical occurrence of a species, e.g.
+phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns.
 Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
@@
 
-Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns.
+Both wide-table formats report the different trait values stored in multiple columns.
 Each trait column represents one value that applies to the entity in question (i.e.
-to a species in the first case or to an occurrence in the latter case).
-As the comlumn names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.
+to a species  or an occurrence).
+As the column names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.
 
 A computationally more effective method that allows for high flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16].
-In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
+In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  (i.e., one occurrence)  which is assigned to a specific taxon.
 <!-- FDS: also refer to kattge 11 -->
@@
 --><!--
-This allows repeated measurements on a single individual to be stored by linking the data from separate rows via a unique specimen ID.
+This allows repeated measurements on a single individual to be stored by linking the data from separate rows via a unique specimen identifier.
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
@@
 
-In designing this structure, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We also paid particularl consideration the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+In designing this terminology, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+We paid particular consideration to the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
 <!-- AO: Webservices!!
@@
 
-Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample.
+When traits are assessed on a higher level of organisation, e.g.
+microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
 Here, location or measurement information are the primary focus of the investigation (see vocabulary extensions below).
@@
 For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
-However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.
+However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.
 
@@
 -->
-By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset.
-The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.
+By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for a specific project context which should always be referenced in the core trait dataset.
+The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable internet address (via URIs).
 <!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
@@
 See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus].
-We also provide an example thesaurus, that has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
+We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
 
@@
 
-- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age.
-This also includes the date and geographical location, as well as the method of sampling and preservation, which provides an important resource to analyse trait variation due to differences in space and time.
+- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
+This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
 - The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
@@
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
+Since the intention of the package is to facilitate upload of own data to hosting services but also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
 See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
@@
 
-To date, existing initiatives and centralised infrastructures for standardisation and data  hosting are focused on a constrained organism group, ecosystem type or region.
+To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions.
 To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
 
-The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories.
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby opening the doors for new insights on global biodiversity patterns.
+The tools proposed here help to standardize trait datasets before upload to central as well as distributed data repositories.
+In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns.
 In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
@@
 Such tools to handle the extensive raw data resources on traits are a prerequisite for research on trait variation along environmental gradients or intra-specific variation.
-It also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
+This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
 We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].

### 2017-12-13 manuscript/manuscript.Rmd
revision of review; start revision of proposed structure

@@
 2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
-3. In order to aid the the standardisation and harmonization of distributed data sources, we present a general and simple vocabulary and data structure for storing and sharing ecological trait data. 
+3.
+In order to facilitate the standardisation and harmonization of distributed trait datasets, we propose a general and simple vocabulary and a two-dimensional data structure for storing and sharing ecological trait data.
 4.
 We provide an R-package that enables the transformation of any data into the proposed form.
-This also allows to harmonize and easily merge trait datasets from heterogeneous sources, which greatly facilitates the compilation of databses of particular research focus.
+This also allows to harmonize and merge trait datasets from heterogeneous sources, which facilitates the compilation of databses of particular research focus.
 5.
-With these decentralised tools for trait-data harmonization, we intend to facilitate the exchange and analysis of trait data for ecological research and enable a global synthesis of traits across a wide range of organisms and ecosystems.
+With these decentralized tools for trait-data harmonization, we intend to facilitate the exchange and analysis of trait data for ecological research and enable a global synthesis of traits across a wide range of organisms and ecosystems.
 
@@
 the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and Thesauri; 
+the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and thesauri; 
 the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.
 
-We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11] and the TOP Thesaurus of plant traits [@garnier17].
-The objective of this review is to show how the degree of trait-data standardisation in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources.
+We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the TOP Thesaurus of plant traits [@garnier17].
+We show how the degree of trait-data standardization in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources.
+The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in the appropriate form.
 
@@
 Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org. 
-These  platforms come with very low thresholds in terms of metadata documentation and standardisation.
-Furthermore, the platforms warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
+Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
 <!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ?
 -->
+However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
+.xlsx, .docx) or binary (e.g.
+.pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.
+
+Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management.
+In most of these datasets, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
+Datasets vary in the detail provided in the metadata, e.g.
+for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
+
+As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compilation of data into larger collections will be inefficient and that many published datasets will not be properly exploited.
 
@@
 
-However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
-.xlsx, .docx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.
-<!-- BKR: This might also result in data becoming no longer accessible (because the last computer on which the excel version required to read a file ran died -->
-The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work.
-
-As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited.
-<!--With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information. -->
-
 ## Database initiatives
@@
 <!-- community database backend, but static publication --> 
-In the past two decades, many distributed datasets have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
+In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
 @klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
 <!--[@wright04; ] -->
-Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16]. 
-Most of these databases are published as static datasets and are updated irregularly. Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
-Few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.
+While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
+-->
+Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
+Just as the individual datasets described above, many of these databases are published as static datasets on low-threshold file hosting platforms and are updated irregularly.
+
+Dealing with much larger amounts of data, initiatives that form around museal collection data are more concerned with data standardization.
+Aiming for a mobilization of measurements from collections, the  VertNet database compiles and harmonized large quantities of vertebrate traitdata [@guralnick16].
+The resulting data are published as static datasets which are updated as new data sources are integrated.
 
 <!-- database backend, selective download, user access management --> 
-More specialised trait-databases have been created that cover different taxa, interaction types (e.g.
-pollination traits or feeding relationships), or ecosystem types.
-These centralized database initiatives invite data submissions from  a defined research field and offer to take care of the harmonization process and greatly facilitate data synthesis.
-For example, traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
-The BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-
+More specialised trait-database platforms have been created that cover certain trait types (e.g.
+floral traits, seed traits or wood density traits), interaction types (e.g.
+pollination traits or feeding relationships), or environmental and experimental context of the trait observation (e.g.
+location or climatic data).
+These centralized database initiatives attract data submissions from a defined research field and take care of the harmonization process and greatly facilitate data synthesis. 
+<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
 By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform.
@@
 As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
 Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
@@
 
-Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.
-<!-- FDS: not an option for primary publication ...!-->
+Regarding open access, hardly any of these centralized databases for functional traits comply with the criteria demanded by journals and funding agencies for primary data publication.
+They incentivize data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
+Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
 <!-- AO: In a sense of data availability to the public : yes.
@@
 Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring.
-For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in thesauri and ontologies.
+Efforts to develop handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but are of limited effect when harmonizing trait data that predate or ignore this standard [@kattge11a].
 Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
@@
 
-These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control [@]. 
+These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. 
 For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
@@
 The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
-To harmonize trait data across databases, future trait standard initiatives should provide this functionality.
+To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
 More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.
@@
 These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain taxon names, sex or life stage, time and date of observation, as well as measurement values and methodological detail.
+These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain trait values, units, taxon names, variables such as sex or life stage, as well as ancillary information of time and date of observation, as well as methodological details.
 The terminologies provided by these standards are quite universal and cover most use cases of trait data.
+<!-- TODO FDS: dimensional and star-shema databases, relational databases Kattge et al 2011, Kimball & Ross 2002-->
 <!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
@@
 
-For the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that intends to cover many potential use cases.
-This data structure is centered around a central data table that contains a single observation of a trait per row, linking it to a taxon and possibly also a single occurrence of the plant species (i.e.
-an individual specimen or 'entity').
+Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that covers most potential use cases.
+This data structure is build around a central data table that contains observations, i.e.
+a single event of measurements on the same individual plant specimen at the same point in time.
+Identifiers link multiple measurements to the same observation, each measurement being well defined by additional standard tables.
+The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
+This structure can be implemented in any software or programming language for relational databases.
 <!-- FDS: enter comments of Jens Kattge here!-->
 
-Recently, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
+In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
 The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
 For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology.
+The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review.
 Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
 TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-The archive also integrates a defined framework for metadata in the Ecological Metadata Language [EML, @knb11], which should be used for describing and linking the data tables.
+The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].
 
-However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand from biodiversity data management for more structured input from the research community.
+However, these proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
+None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
+One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
+If the traits are assembled for a specific research question with little ancillary information, it may be safer to add them as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
+This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
+<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
+The trade-off is data-consistency and readability *vs.* avoidance of content duplication.
 
-The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
-<!-- PM: There is also little incentive for researchers to do this- why bother?
-Incentives are also needed.
+<!--When traits are assessed on a higher level of organisation, e.g.
+microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
+Here, location or measurement information are the primary focus of the investigation and must be contained in the .
+<!-- PM: I get lost here-->
+<!--A general recommendation, whether a column describes ancillary data or is part of the central dataset is ill advised.
+Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
 -->
-None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
-One reason for this is most certainly compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
+<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
+
+<!--while the existing trait data standards may cover all of the aspects required in the various use-cases of trait-based research.
 <!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific -->
-The other reason is awareness and literaci: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.
-
-As it stands, this lack of data standardization will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited. 
-To cope with the trend towards distributed data sources, data brokering services (e.g.
+The other reason is lack of awareness and literacy: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, which is not affordable for small and intermediate size research projects.
+By filling this gap, data brokering services (e.g.
 the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g.
@@
 by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
+
+The practical importance of these frameworks for data providers and data users is often not obvious and not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
+<!-- PM: There is also little incentive for researchers to do this- why bother? Incentives are also needed. -->
+The pay-off is only playing out in the long run. Citations of trait datasets are higher when they are easier to access and relate to other data due to standard structures and vocabularies. 
+Along with the push towards open access publication of primary data, the scientific community must come up with ways to incentivize and require data standardization.
 
@@
 Also at this level of information, missing entries, e.g.
-about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14] and using Bayesian approaches [@schrodt15].
+about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14; @taugourdeau14] and using Bayesian approaches [@schrodt15].
 <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].
@@
 
-A computationally more effective method that allows for high flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16].
+A computationally more effective method that allows for high flexibility is the storage of traitdata in long-table formats [@wickham14; @parr16].
 In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  (i.e., one occurrence)  which is assigned to a specific taxon.
@@
 
-With this paper, we propose a versatile vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
+As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees measurement detail for multiple use cases of trait-based ecological research.
 <!-- PM: It took a long time to get to this point- I suggest brininging in a section early on that describes what the paper will do as it’s not always clear where it is leading. --> 
@@
 
-In designing this terminology, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We paid particular consideration to the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+- we propose Vocabulary for two-dimensional tables of traits. 
+- long-table format.
+has advantage of easy merging and filtering, computing, modelling.
+But disadvantage of multi-type columns (e.g.
+for values) and repetitive content.
+For complex dimensional databases, use Kattge, DwC or ABCD instead, packaged as ZIP files with annotations in EML Metadata format rather than multi-spreadsheet files.
+- linking to standard taxonomy and trait terminology.
+Allowing for but minimizing duplicate information.
+by providing standardised duplicates, align user-provided nomenclature and trait methodology with taxonomic nomenclature and trait ontologies.
+transparency for data user.
+- Offering compatibility, input for dimensional databases by using global identifiers and internal identifiers that provide hierarchical layers (measurement, occurrence, sampling event, data source, bibliographic reference)
+- using these identifiers enables the vocabulary for for observation level data (linked via an occurrence identifier) raw measurement data (each row an own measurement identifier), as well as for aggregated data on the species level (each row an own measurement identifier plus aggregation information, like taxon rank or statistical method).
+allows for originary data (reporting own measuremetns) or compiled data sheets (reporting multiple dataset identifiers).
+In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
 <!-- AO: Webservices!!
@@
 To enable a better standardisation of nomenclatures and values, the entries of the core dataset <!-- CP: Does this need to be defined --> should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
-In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
+In these duplicate columns, values should be reported as 'harmonized' entries [@kattge11].
+
+, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
 <!-- BKR: There is some grammatical error here; not sure what you try to say-->
@@
 For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
-UTF-8).
+UTF-8 or ASCII).
 <!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->
@@
 
-In very complex datasets, the repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
-Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09].
-<!-- PM: Why do these avid this? -->
+However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. 
+In more complex datasets, the repetition of verbose entries and long character strings in a single data table should be avoided.
+Therefore we recommend use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09].
 <!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
-
-In some contexts, it may be safer to add such information directly as columns of the data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
-This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
-<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
-The trade-off is data-consistency and readability *vs.* avoidance of content duplication.
-
-When traits are assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
-Here, location or measurement information are the primary focus of the investigation (see vocabulary extensions below).
-<!-- PM: I get lost here-->
-A general recommendation, whether a column describes asset data or is part of the central dataset is ill advised.
-Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
-<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
 
@@
 
-# Conclusion
+# Discussion
 
-In this paper, we propose a general, low-threshold structure and vocabulary for ecological trait data. 
+To serve the demand for simple ways to standardise and harmonize ecological trait data, we propose a versatile vocabulary for two-dimensional datasets.  
 We also provide a toolbox in R to transfer and harmonize data into this scheme.
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal  for comments on the manuscript, the trait data standard and the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal and Jens Kattge for comments on the manuscript, the trait data standard and the R-package.
 
+The work has been partly funded by the DFG Priority Program
+1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). <!--Field work
+permits were issued by the responsible state environmental offices of
+Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
+BbgNatSchG).-->
 We thank <!--the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
@@
 for their role in setting up the Biodiversity Exploratories project.
-
-The work has been partly funded by the DFG Priority Program
-1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). <!--Field work
-permits were issued by the responsible state environmental offices of
-Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
-BbgNatSchG).-->
 

### 2018-02-13 bexis_handbook/bexis_handbook_traits.Rmd
add BExIS handbook draft

--- 2017-12-12 manuscript/manuscript.Rmd
@@
 ---
-title: "Towards an Ecological Trait-data Standard"
-author:
-- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
-- name: Malte Jochum \footnotemark[4]
-- name: Gaetane LeProvost \footnotemark[2]
-- name: Andreas Ostrowski \footnotemark[5]
-- name: Caterina Penone \footnotemark[4]
-- name: Markus Fischer
-- name: Pete Manning 
-- name: Birgitta König-Ries
-- name: Nadja K. Simons \footnotemark[6]
-csl: amnat.csl
-output:
-  pdf_document:
-    fig_caption: yes
-    fig_crop: no
-    keep_tex: yes
-    template: pretty.template
-  html_document:
-    fig_caption: yes
-    fig_width: 4
-    highlight: null
-    keep_md: yes
-    theme: united
-  word_document:
-    fig_caption: yes
-    fig_width: 6
-bibliography: div05.bib
+title: "BExIS Manual for Traitdata"
+author: "Florian D. Schneider"
+date: "13 February 2018"
+output:  pdf_document
+bibliography: ../manuscript/div05.bib
+csl: ../manuscript/amnat.csl
 ---
 
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+Functional traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.
 
-\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
-\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
-\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
-\footnotetext[5]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[6]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+Sharing and harmonizing trait data into databases is a key task of biodiversity data management, but it relies on an unambiguous assignment of trait definitions and taxa, as well as clearly defined datasets.
+To facilitate the data handling by BExIS and other downstream services (e.g.
+GFBio or TRY) we defined a standard vocabulary for publishing trait datasets.
 
-\linenumbers
+This document describes the main features of the standard vocabulary and how to prepare traitdata for upload into BExIS.
 
-# Abstract
+# Traitdata vocabulary
 
-1.	The use of functional trait data is now widespread in all domains of ecological research, thanks to the great potential for such data to deliver general conclusions regarding the nature of biodiversity.
-Accordingly, a wealth of trait data is now available for many organism groups but due to lacking standardisation these data come in heterogeneous data formats and data types.
-2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
-3. To aid the the standardisation and harmonization of distributed data sources, we present a general and simple vocabulary and structure for storing and sharing ecological trait data. 
-4.
-We provide an R-package that enables the transformation of any data into the proposed form.
-This also allows to harmonize and easily merge trait datasets from heterogeneous sources, which greatly facilitates the compilation of databses of particular research focus.
-5. With these decentralised tools for trait-data management, we intend to facilitate the exchange and analysis of trait data for ecological research.
+The full set of terms can be accessed at https://ecologicaltraitdata.github.io/ETS/.
 
-# Key-words:
+The vocabulary proposed here is meant for two-dimensional spreadsheet data.
+These simple table formats are what a typical research project on a particular organism group and locality produces.
+The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported measurement or fact (e.g.
+measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context.
+The vocabulary also contains terms for describing metadata acompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
+For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09].
+]
 
-functional ecology, functional traits, semantic web, ontologies, data standard
-
-# Introduction
-
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07].
-Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
-For example, trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity and ecosystem functioning (refs). 
-Using traits is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring both the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
-With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].
-
-<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04].
-Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
-Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13].
-These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
-Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97].
-Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
-Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17].
---> 
-<!-- CP: I like very much this whole paragraph (both writing and content), but I feel that it’s a bit out of the topic.
-It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
-I give some suggestions on how to cut it and rather develop more the following one, which people are surely less aware of and which sets more the aim of the paper.
-I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
-
-That’s the structure I try to follow:
-1. Traits are important in ecology, evolution and conservation (i.e. this paragraph)
-
-2.
-To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources.
-When we do this we can answer new questions and find generalities.
-So there is a growing interest in assembling trait data from different sources and taxa (e.g.
-give the number of papers using TRY).
-
-3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.
-
-4.
-One solution are databases.
-But these are costly and depend on the will of few persons/funding.
-They also cover few taxonomic groups.
-Matching between different databases is still difficult because of different definitions and structures.
-Most data are still uploaded separately.
-We need a solution to make their use easy.
-
-But of course it’s just a suggestion, feel free to happily ignore it!
-
-I would choose 2 or 3 (also for the previous), anyway it’s impossible to be exhaustive. Maybe put (e.g. ref1, ref2) so make clear that it’s just examples
-
--->
-<!-- PM: Thuis is quite complex- e.g.
-‘taxonomic impediment’etc.
-I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples.
-I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.
-
-Huge number of references here- I think you can strip it down to a few key ones, and reviews.
-Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site.
--->
-<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
-Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->
-
-However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa.
-<!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific?
--->
-To mobilize data for this enterprise <!-- PM: Please define what the exercise is- too vague at present
--->, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources.
-<!-- PM: I think this is a key justification- perhaps say what happens if we don’t do this – non repeatable work, differing definitions of traits that may lead to scientific flaws, repetition of the painstaking task of processing trait data e.g.
-to compute species level means.
--->
-The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way.
-<!-- CP: I felt that this needed a definition, not sure everybody know what an ontology is.
-Maybe we should guide more the reader that might not be familiar with the problems.
--->
-
-Nonetheless, individual researchers often neglect the benefit of centralised databases <!-- PM: Explain what this benefit is.
--->.
-Furthermore, many journals and research funding agencies require  data to be published without access restrictions [@swan12; @allison15].
-However, this primary data is often posted on file hosting services, where no data standards are forced upon the uploaded material (e.g.
-Figshare, Dryad).
-The long-term outcome of this situation is an abundance of freely available data, but a lack of data standardisation.
-
-In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users.
-We further (2.) review existing trait databases and web services, as well as initiatives for standardisation.
-We (3.) describe the common practice <!--PM: Current or your own new definition?--> for structuring trait data for ecological research. 
-Based on these considerations, we (4.) propose a low-threshold <!-- PM: What’s this? Please define
--->  structure and vocabulary for describing trait-datasets compatible with existing ontologies <!-- PM & CAT: define--> .
-(5.) Finally, we provide an R-package that assists in aply this structure to own data and for the harmonisation of data from distributed sources.
-
-With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold <!-- PM:Again, I’m not sure what this is a low threshold of?
---> trait data standardisation.
-
-Despite the obvious benefit and great potential of functional trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
-All of these factors render trait data extremely heterogeneous.
-
-As a response to these issues, trait data have been standardized in centralized databases for specific organism groups and research questions.
-<!-- PM: (refs, also give an e.g. or two for some big ones like TRY- saw there was a new one for tropical mammals a couple of days ago too and a huge Chinese plants one- in )--> 
-These compiled data are either being published as static data tables or in the form of sophisticated online portals like the plant trait database TRY (see below for a review of current initiatives).
-These portals attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, and importantly also, provide networks for consensus building in measurement methods.
-Furthermore, database services offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.
-
-In contrast to standardized databases, open-access data publication in low-threshold repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org is becoming increasingly frequent.
-Such repositories foster decentralised data hosting with low thresholds regarding data standardisation and documentation.
-
-Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.
-To cope with this trend towards distributed data sources, data brokering services (e.g.
-the German Federation on Biological Data; gfbio.org)  or database <!-- AO: Just “data” or “research data”--> management systems for scientific projects (e.g.
-BEXIS, XX <!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
-
-However, to mobilize the broad wealth of distributed data into structured databases <!-- AO: Maybe better: formats-->,  a certain minimum of information from original data providers is requrired, e.g.
-on the methodology and definitions of trait values or the covariates describing the ecological context of the trait observation.
-In order to get this accurate and well-defined input ecological researchers need to be aware of and understand basic principles of data standards and ontologies.
-<!-- AO: The last paragraphs may need more sources/proofings?--><!-- PM: Literacy is quite a strong word- as illiterate is quite an insult.
-I think awareness is a problem too.
--->
-
- More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
-
-# 1. A review of trait-standard initiatives
-
-In this section, we review four kinds of initiatives of relevance for trait data standardisation: 
-the first are projects that work with trait datasets that have been assembled out of a primary interest in the project context; 
-the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and Thesauri; 
-the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.
-
-We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11] and the TOP Thesaurus of plant traits [@garnier17].
-The objective of this review is to show how the degree of trait-data standardisation in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources.
-
-## Trait datasets
-
-In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books.
-With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].
-
-Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org.
-These  platforms come with very low thresholds in terms of metadata documentation and standardisation.
-Furthermore, they warrant long-term accessibility (offering DOIs for referencing), and assume Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
-<!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ?
--->
-
-A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list).
-We invite data owners and users to register further datasets qualifying as trait data to this spreadsheet.
-
-However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g.
-.xlsx, .docx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.
-<!-- BKR: This might also result in data becoming no longer accessible (because the last computer on which the excel version required to read a file ran died -->
-The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work.
-
-As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited.
-<!--With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information. -->
-
-## Database initiatives
-
-<!-- community database backend, but static publication --> 
-In the past two decades, many distributed datasets have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
-<!--[@wright04; ] -->
-Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16]. 
-Most of these databases are published as static datasets and are updated irregularly. Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
-Few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.
-
-<!-- database backend, selective download, user access management --> 
-More specialised trait-databases have been created that cover different taxa, interaction types (e.g.
-pollination traits or feeding relationships), or ecosystem types.
-These centralized database initiatives invite data submissions from  a defined research field and offer to take care of the harmonization process and greatly facilitate data synthesis.
-For example, traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
-The BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-
-By aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other platform.
-The online platform enables selective data download and user permission and rights management.
-As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
-Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
-The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
-
-<!--Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12].
-Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13].
-Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
-the KEGG orthology, @kanehisa12].
--->
-
-Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.
-<!-- FDS: not an option for primary publication ...!-->
-<!-- AO: In a sense of data availability to the public : yes.
-But do we discuss it here?
-We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs.
-non-open data, or?
-bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
-FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
- -->
-
-## Thesauri and Ontologies for traits
-
-A major problem in trait-data standardisation is the reference to accepted and unambiguous trait definitions. This is acheived via thesauri and ontologies. 
-In database management, a thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
-Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data.
-Providing relational ontologies in a machine readable form makes them compliant with semantic web standards, thus supporting the goal of widely interlinked data structures for the world wide web [@berners-lee01; @page08].
-
-<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
-units or legit factor levels) of measured values or reported facts.
-It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
---><!-- BKR: Isn’t that the same?
-Or can you have multiple parents for one node?
-FDS: agree.
-simplify phrasing.
--->
-
-Standard trait definitions have been provided in all shades along this continuum, ranging from listings of definitions in glossaries, over well-defined thesauri to relational ontologies.
-The initiatives behind these are first and foremost concerned with consensus building on trait definitions.
-Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring.
-For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in thesauri and ontologies.
-Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
-Thesauri of trait definitions have been developed for plants [e.g.
-the Planteome ontology for crop plants, @walls12; @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
-the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13].
-Comprehensive trait ontologies, including hierarchies of parent terms, have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and on the side of animal traits, in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
-
-These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control [@]. 
-For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
-<!-- Mention also NCBO BioPortal?
-https://bioportal.bioontology.org/ -->
-These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets.
-For example, a measurement of seed size would be linked to Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
-
-To be unambiguous, any thesaurus should be defining terms based on other well-defined terms from semantic ontologies.
-The definition provided here contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
-This way, each trait definition may link to a broader or narrower term.
-For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies.
-By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.
-
-The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
-To harmonize trait data across databases, future trait standard initiatives should provide this functionality.
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.
-
-## Trait-data structures for the semantic web
-
-While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g.
-for taxa.
-To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.
-
-Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
-More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
-trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
-traits) observed on a biological entity (i.e.
-an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e.
-a species or higher-level taxon).
-Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig.
-1*a*), by aggregating replicated measurements on multiple entities (Fig.
-1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g.
-databases, Fig.
-1*c*).
-Qualitative facts are assignments of an entity to a categorical level, e.g.
-of a behavioural or life-history trait (Fig.
-1*d*).
-The entity or observation (i.e.
-the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
-a genus).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Types of ecological trait data assume different entities or reported quantities.
-a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}
-
-\end{figure}
-<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”.
-Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->
-
-These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain taxon names, sex or life stage, time and date of observation, as well as measurement values and methodological detail.
-The terminologies provided by these standards are quite universal and cover most use cases of trait data.
-<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
-The Trait Data Standard is a standard on how to store trait data measurements.
-In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard.
-For instance the ABCD standard has a huge section for MeasurementsAndFacts.
-And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible).
-Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD.
-While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.
--->
-
-For the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that intends to cover many potential use cases.
-This data structure is centered around a central data table that contains a single observation of a trait per row, linking it to a taxon and possibly also a single occurrence of the plant species (i.e.
-an individual specimen or 'entity').
-<!-- FDS: enter comments of Jens Kattge here!-->
-
-Recently, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
-The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
-For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology.
-Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-The archive also integrates a defined framework for metadata in the Ecological Metadata Language [EML, @knb11], which should be used for describing and linking the data tables.
-
-However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand from biodiversity data management for more structured input from the research community.
-
-The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
-<!-- PM: There is also little incentive for researchers to do this- why bother?
-Incentives are also needed.
--->
-None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards.
-One reason for this is most certainly compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research.
-<!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific -->
-The other reason is awareness and literaci: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.
-
-As it stands, this lack of data standardization will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited. 
-To cope with the trend towards distributed data sources, data brokering services (e.g.
-the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g.
-BExIS, NEON ) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into consistent and interlinked data infrastructures.
-One goal of such initiatives is to facilitate the publication and standardization of data for researchers, e.g.
-by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
-
-<!--
-Depending on the research objective, trait-data are stored in a range of different table formats.
-In the subsequent paragraphs, we discuss the most common formats  in relation to the use cases in trait-based ecology.
-
-### a) Species wide-table
-
-<!-- AO: You argue somehow that matrix format I a result of lit.
-review and so on.
-I would not agree with it.
-Your example also “just” points to aggregated data which menas one row is on species (aggregated) value.
-Your example is (for me) not really a matrix.
-In a matrix you have a relation between two variables.
-In the first column and and on the upper row normally the values of one variable are written.
-An example would be species x body length.
-Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted.
-E.g.
-spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on.
-Your wexample hat already different variables (=traits) with different units as columns.
-That is a meaningfull representation of variables and values in a normal data table.
-Maybe it is good her to show the enormous number of different investigated traits and point to the resulting big (many columns = wide) tables.
---><!--
-If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait wide-table format (Fig.
-2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available.
-This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g.
-'999').
-It may store additional information (e.g.
-on variation of means or literature source) in secondary columns or accompanying datasheets.
-The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
-<!-- AO: You say it already 2 sentences before --><!--
-Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a].
-Also at this level of information, missing entries, e.g.
-about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14] and using Bayesian approaches [@schrodt15].
-<!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].
---><!--AO: Why not?
-I can filter for a specific species and get all respective rows.
-But this is not a result a using a data format, it is a result of the data itself, FDS: no, since only one row per species exists; true that data itself don't cover this variation.
--->
-<!--
-
-### b) Occurrence wide-table
-
-Phenotypic variation of traits within species would usually be recorded per observation, i.e.
-each individual specimen would be recorded in a row, with the respective trait values stored in columns.
-In these occurrence data tables (Fig.
-2*b*), each row is centered around a single physical occurrence of a species.
-Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
-Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
-This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data but rarely found for reported qualitative facts obtained from the literature.
-
-### c.) measurement long-table
-
-Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns.
-Each trait column represents one value that applies to the entity in question (i.e.
-to a species in the first case or to an occurrence in the latter case).
-As the comlumn names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.
-
-A computationally more effective method that allows for high flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16].
-In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence).
-<!-- FDS: also refer to kattge 11 -->
-<!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes.
-But not necessarily computational more effective at all.
-E.g.
-I start to mix up numeric values with descriptive values.
-I start to mix up different units.
-Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g.
-quality stuff.
-Ergo less effective from that perspective.
-But it has some advantages which may overrule dis.
-See below
-AO: For the whole proofing why to use such a long format I think that GBIF and DwC has already given some sentences. I strongly think their work should be considered and can be used as reference!
-We can simply argue that a well established routine should now also executed on trait level!
-Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.
-
---><!--
-This allows repeated measurements on a single individual to be stored by linking the data from separate rows via a unique specimen ID.
-Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
-Long-table datasets purport multiple advantages for data manipulation (e.g.
-filtering, sub-setting and aggregating data), visualization (e.g.
-plot measured values by factor variable or taxon) and statistical modelling (e.g.
-ANOVA for testing difference of trait value by sex) [@wickham14].
-As long-table fromats draw from a defined set of columns,  merging of datasets is much easier.
-<!-- BKR: Is that really true?
-Of course, you have less problems on the schema or structure level.
-But, on the other hand, the differences (e.g.
-different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows).
-Thus, at least with respect to integration, you just move the problem from one level to the other.
-To ease integration, one would need to agree on the entries in the features column
-AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
-We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
-Again: I think we find here papers regarding DWC and GBIF (and OBIS) as source of proofing the facts.
--->
-<!--
-The long-table format therefore is ideal for storing raw data of trait measurements in huge databases, and allows the entries to be linked via the respective unique identifiers to additional information.
-Each row of the dataset can therefore be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig.
-2*c*).
-
-<!-- Cat: here I guess that a figure showing few rows of a species x trait matrix and a long-format one with links to other information (metadata,..) woud help quite a lot.
-By experience some users are not be familiar with the interest of using the long-table format.-->
-<!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->
-<!--
-\begin{figure}[t]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) wide-tables report measurements on the specimen-level, i.e.
-one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to a trait definition and taxon via unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the specimen (occurrence) or e) the origin or confidence on the reported measurement or fact.}
-
-\end{figure}
--->
-
-<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
-
-FDS: see Garnier et al. but may contain:
-
-Term | Definition 
------ | -----
-Terminology | 
-Thesaurus | 
-Ontology | 
-semantic web |
-
--->
-
-# 2. Proposing an ecological traitdata standard
-
-With this paper, we propose a versatile vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
-<!-- PM: It took a long time to get to this point- I suggest brininging in a section early on that describes what the paper will do as it’s not always clear where it is leading. --> 
-<!-- FDS: point out focus on distributed low-threshold data, not full-fledged database initiatives. -->
-
-In designing this structure, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We also paid particularl consideration the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
-<!-- AO: Webservices!!
-;  Really to be considered as standards?
--->
-
-The online reference glossary of terms and description of the proposed structure can be found at https://ecologicaltraitdata.github.io/ETS/.
-
-We acknowledge that the vocabulary proposed will not capture the full complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ...
--->.
-Therefore, we invite researchers to contribute to the next iterations of this standard terminology and develop own applications and ontologies.
-
-<!-- PM: It seems that the object identifier for an individual is the key distinction here- this could be clearer.
-
-Maybe you address this later but does your tool assign one of these- how to you make sure they are all unique if there are multiple users- e.g.
-if you pooled together two previously complied datasets which might contain some of the same individuals?
--->
-
-## Standard structure for trait data
-
-As described above, observation-level long-table formats are the most predictable and compatible format for the core content of trait datasets and should therefore also be used for standalone uploads on low-threshold repositories.
-<!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
--->
-<!-- CP: Not sure this needs to be repeated -->
-
-To enable a better standardisation of nomenclatures and values, the entries of the core dataset <!-- CP: Does this need to be defined --> should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
-In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
-<!-- BKR: There is some grammatical error here; not sure what you try to say-->
-<!-- AG: This was a little bit confusing for me.
-In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon.
-On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name.
-In any case I would suggest to recommend to keep both the URI and the literal.-->
-<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
-If data contain additional information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
-<!-- PM: Not clear here -->
-For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
-UTF-8).
-<!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->
-
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider et al.
-2017 Ecological Traitdata Standard v0.7, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
-In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
-FDS: to consider!! -->
-
-In very complex datasets, the repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14].
-Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09].
-<!-- PM: Why do these avid this? -->
-<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
-
-In some contexts, it may be safer to add such information directly as columns of the data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
-This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
-<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
-The trade-off is data-consistency and readability *vs.* avoidance of content duplication.
-
-Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample.
-Here, location or measurement information are the primary focus of the investigation (see vocabulary extensions below).
-<!-- PM: I get lost here-->
-A general recommendation, whether a column describes asset data or is part of the central dataset is ill advised.
-Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
-<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
-
-## Standard vocabulary for trait data
-
-<!-- PM: Can you say a little about what this vocab is and how it was chosen/defined? -->
-
-Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/ETS/. 
-Individual terms <!-- DF:Please explain the difference between the “individual terms” and the “field entries” above-->  should be referenced via their global identifier, which is hosted by the GFBio Terminology Service <!-- DF: Please add citation to the TS Paper in Datenbanken-Spektrum, see https://terminologies.gfbio.org/about/#citation  --> and also can be accessed programmatically (i.e.
-via the Application Programming Interface, API).
-Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
-
-### Core columns
-
-To describe the minimum core content, where each row is the reported measurement or fact for a single observation, the following columns are required: 1.
+We assume that data are stored in a observation table, where each row is the reported measurement or fact for a single observation.
+At minimum, the following columns are required: 1.
 a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2.
@@
 the scientific taxon for which the measurement or fact was obtained (`scientificName`).
-For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
-However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.
+<!--For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa.
+However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers.
+-->
 
-<!--### Relate to global taxon and trait definitions -->
+## Identifiers
 
-In this case, it is essential to relate the user-defined names to a consensus standard of taxon names or or look-up tables of traits. 
-This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g.
-GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.
+To make data compatible with a consensus standard of taxon names or or look-up tables of traits, it is highly recommended to map them to globally valid terminologies by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
+GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').
+
 By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
@@
 
-If no URIs are referenced for taxonID, it is highly recommended to keep an additional reference to a kingdom and order (using columns `kingdom`, `order` as described by DWC and suggested for EOL Traitbank) to avoid ambiguities with taxonomic homonyms.
-<!-- BKR: Why do you need this, if taxonIDs are used.
-They should be unambiguous
-FDS: right, rephrase to: either use taxonID or provide this.-->
-Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
-This information can then be used for trait inference at higher taxonomic resolution.
-<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
--->
-
-<!--The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
-The accessibility by software differs widely.
-For biodiversity data, the OBO Foundry (http://www.obofoundry.org/) or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer API services for computational access.
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
--->
-
-If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-<!-- A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
-Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
-units or legit factor levels) of measured values or reported facts.
-It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
--->
-By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset.
-The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.
-<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
-
-See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus].
-We also provide an example thesaurus, that has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
-
-### Metadata
-
-Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authorship, while also stating a bibliographic reference and license model.
-
-In the case of raw measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata. 
-In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be linked to the measurement level, via an unambiguous `datasetID` and a descriptive `datasetName`.
-
-<!--Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit.
-Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
-However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
--->
-
-<!-- 
-(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits.
-However, such network data may be of interest to be combined with trait data.
-) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names.
--->
-
-### Extensions for additional data layers
+## Extensions
 
 Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
+Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context.
 
@@
 
-Many terms of these extensions refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
-<!-- BKR: Looking at your definitions, in most cases, I am not sure how you refine the terms. Is it really refinement or is it just using them with the same meaning they have in DwC?
--->
-<!-- PM: All of this sounds like quite a lot of work- so will it be automated, or incentivized somehow?
--->
+# R-package 'traitdataform'
 
-As for taxon or trait information, these additional layers of information can either be added as extra columns to the core dataset or be kept in separate data sheets of a Darwin Core Archive (see above), this avoiding redundancy and duplication of content.
-A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
-Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
-a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
-
-<!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
-It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc.
--->
- <!-- Nadja: I like the idea of having a figure explaining the different terms -->
-
-# 3. Computational tools for producing compliant data
-
-To access data from public databases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
-However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
-
-Accordingly, we provide the R package 'traitdataform' to assist the production of data compliant with the trait data standard proposed here. There are two major use cases for the package:
+We provide the R package 'traitdataform' to assist the production of data compliant with the trait data standard proposed here. There are two major use cases for the package:
 
@@
 A comprehensive documentation of the package is found in the package vignette, and on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
-
-The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
-\caption{Process chart of the functions provided within the R package.}
-
-\end{figure}
-
 In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data.
@@
 The same function harmonises trait values by converting values into target units and unified factor levels.
-<!-- BKR: Have you thought about linking to unit ontologies here?
-AO: This paper is a good reference and gives an idea of which ontology could be used;
-http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
-FDS: might be referred to in the section about thesauri -->
 
-The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
+# Upload to BExIS
 
-```
-traitlist <- as.thesaurus(
-   body_length = as.trait("body_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length"), 
-   antenna_length = as.trait("antenna_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Antenna_length"),
-   metafemur_length = as.trait("metafemur_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Femur_length")
-)
-
-datasetStd <- standardize.traits(dataset, thesaurus = traitlist)
-```
-
-The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`.
-Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`).
-Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies.
-
-The package functions form a toolchain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
-
-```
-datasetStd <- standardize(read.csv("path/to/data.csv"),
-            thesaurus = traitlist,
-            taxa = "name_correct", 
-            units = "mm"
-            )
-```
-
-Datasets that have been produced by these functions can easily be appended using the function `rbind()` of R base, while maintaining any available metadata information as separate column entries.
-
-To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identifier using the function `merge()`.
-This enables an easy handling of data sources that have been stored in a Darwin Core Archive format.
-
-Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
-See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
-We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
-
-# Conclusion
-
-In this paper, we propose a general, low-threshold structure and vocabulary for ecological trait data. 
-We also provide a toolbox in R to transfer and harmonize data into this scheme.
-
-To date, existing initiatives and centralised infrastructures for standardisation and data  hosting are focused on a constrained organism group, ecosystem type or region.
-To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
-
-The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories.
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby opening the doors for new insights on global biodiversity patterns.
-In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
-
-We therefore encourage the development of trait thesauri in ontologies.
-The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
-This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].
-
-Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources.
-Such tools to handle the extensive raw data resources on traits are a prerequisite for research on trait variation along environmental gradients or intra-specific variation.
-It also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
-We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
-
-<!--The trait data standard as well as the R package are published for open source community development.
-Contributions are invited and vital for the future development of the data standard and an extension and improvement of the R package.
--->
-
-# Acknowledgements
-
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal  for comments on the manuscript, the trait data standard and the R-package.
-
-We thank <!--the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
-infrastructure;--> Christiane Fischer and Jule Mangels for giving
-support through the central office, <!--Michael Owonibi for
-managing the central data base,--> and <!--Markus Fischer,--> Eduard Linsenmair,
-Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
-Ernst-Detlef Schulze, Wolfgang W. Weisser and the late Elisabeth Kalko
-for their role in setting up the Biodiversity Exploratories project.
-
-The work has been partly funded by the DFG Priority Program
-1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). <!--Field work
-permits were issued by the responsible state environmental offices of
-Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
-BbgNatSchG).-->
-
-M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
-
-# Authors' contributions
-
-FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package.
-All authors contributed critically to the drafts and gave final approval for publication.
-
-# Online Resources
-
-The online reference for the Ecological Traitdata Standard described in this paper is https://ecologicaltraitdata.github.io/ETS/
-
-The package website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform
-
-The living spreadsheet on existing trait datasets and databases is found at
-
-Development of the Standard and the R-package is coordinated via https://github.com/EcologicalTraitData/ .
+When uploading the dataset to BExIS,
 
 # References
-
-\footnotesize

### 2018-02-14 bexis_handbook/bexis_handbook_traits.Rmd
update bexis handbook

@@
 The vocabulary also contains terms for describing metadata acompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
-For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09].
+For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
 ]
@@
 a descriptive trait name (`traitName`);  3.
-the scientific taxon for which the measurement or fact was obtained (`scientificName`).
+the scientific taxon for which the measurement or fact was obtained (`scientificName`).^[Another common format for trait data are species--trait matrices, where each row is a species and multiple trait values are stored in columns.
+These have to be converted into observation tables when applying the trait vocabulary.]
 <!--For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa.
@@
 
-To make data compatible with a consensus standard of taxon names or or look-up tables of traits, it is highly recommended to map them to globally valid terminologies by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
-GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').
+To make data compatible with a consensus standard of taxon names or a look-up tables of traits, it is highly recommended to map them to globally valid terminologies by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
+GBIF Backbone Terminology)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'].
 
@@
 literature source or expert reference).
+- The `BiodiversityExploratories` extension provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`
 
@@
 1. preparing of own trait datasets for publication on low-threshold file-servers and project databases, and
-2. automating the harmonisation of trait datasets from different sources by moulding them into a unified format.
+2.
+automating the harmonisation of trait datasets from different sources by moulding them into a unified format^[The package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
+See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
+We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
+].
 
-A comprehensive documentation of the package is found in the package vignette, and on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
+A comprehensive documentation of the package is found in the package website (http://EcologicalTraitData.github.io/traitdataform/), and on its Github repository (https://github.com/EcologicalTraitData/traitdataform).
 
-The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
-2).
-This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
+Installation of the package is possible via the devtools package directly from Github:
+
+```
+install.packages('devtools')
+devtools::install_github('EcologicalTraitData/traitdataform')
+```
+
+The key function of the package is `as.traitdata()`.
+This function maps column names into terms provided in the standard trait vocabulary and adds metadata as attributes to the output object.
+The function also will mould a species-trait-matrix into a observation table format.
+If data are measured on a single individual the function adds an occurrenceID to keep the reference context of individual measurements.
 This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:
@@
 Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
-a thesaurus of traits.
+a thesaurus of traits^[This requires some manual mapping.
+Please see the package website for instructions.
+Also note the trait thesaurus provided for the Exploratories (https://ecologicaltraitdata.github.io/TraitDataList/), which will be included in the package in due time.].
 The same function harmonises trait values by converting values into target units and unified factor levels.
+
+The package functions form a toolchain where each function can be piped as an input into the next.
+A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
+
+```
+datasetStd <- standardize(read.csv("path/to/data.csv"),
+            thesaurus = traitlist,
+            taxa = "name_correct", 
+            units = "mm"
+            )
+```
+
+When saving your data, it is recommended to use
 
@@
 
-When uploading the dataset to BExIS,
+When uploading the dataset to BExIS, you can use the predefined template "traitTemplate", which already contains a list of column definitions.
+
+On page 2: Most likely, the number of plots is irrelevant to your trait data, but assign whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system.
+The taxon field is of particular importance here.
+The field 'bioticDataTypes'  should be set to "Trait Data".
+
+On page 3: describe the methodology underlying your trait values thoroughly.
+If data were extracted from literature or other database sources, declare that data are "literature data" and - if applicable - point to the column "basisOfRecord" where you should distinguish the source of your data.
+
+On page 4: When providing keywords, you should specify which kind of traits are reported (e.g. "morphological", "behaviour", "life-history", "functional").
+
+On page 5: Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd").
+
+On page 6: Set 'exploratory' and 'keyPlots' to 'none' if your trait data are not assigned to a single plot and site.
+If you created a .txt or .csv file from the R package, chose fileFormat 'ASCII' and pay attention to the field delimiter and decimal Character.
+Missing values created by the R package are empty fields, thus enter "".
+Known missing values (i.e.
+values have not been measured for any reason) should not exist in the observation table format (the R package eliminates them by default).
+If you chose to keep rows for lacking trait measurements in the table, they should be marked as "NA".
+The section on 'matrix data' should be irrelevant if the data have been correctly transferred into observation tables.
+The variables (i.e.
+column names) are already provided by the template.
+You may add non-standard columns if necessary.
+
+<!--Eliminate columns that are not present in your dataset. 
+-->
+
+On page 7: if the trait data are extracted from one or multiple existing databases, it is necessary to refer to those here (by name and by original paper reference).
+If the data have been uploaded before in a non-standard format, please provide the dataset ID.
 

### 2018-02-14 bexis_handbook/bexis_handbook_traits.Rmd
complete bexis handbook, first draft

@@
 date: "13 February 2018"
-output:  pdf_document
+output:  
+  pdf_document:
+    includes:
+      in_header: header.tex
+    toc: yes
+fontsize: 10pt
+geometry: margin=1.5in
 bibliography: ../manuscript/div05.bib
@@
 
-Functional traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.
+# Ecological Traits
+
+Ecological traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.
 
@@
 literature source or expert reference).
-- The `BiodiversityExploratories` extension provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`
+- The `BiodiversityExploratories` extension  (http://ecologicaltraitdata.github.io/ETS/#extension-biodiversity-exploratories)  provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`
 
@@
 
-When saving your data, it is recommended to use
+For saving your data for upload in BExIS, you would use
+
+```
+write.csv(datasetStd, file = "filename.csv", row.names = FALSE)
+```
+
+which and omits the row names when produces a comma separated file with points as decimal delimiter.
 

### 2018-02-28 bexis_handbook/bexis_handbook_traits.Rmd
update bexis handbook for traits, include comments of Andreas, Nadja, Malte, Caterina.

--- 2018-02-14 bexis_handbook/bexis_handbook_traits.Rmd
@@
 title: "BExIS Manual for Traitdata"
-author: "Florian D. Schneider"
-date: "13 February 2018"
-output:  pdf_document
+author: "Florian D. Schneider (florian.dirk.schneider@gmail.com)"
+date: "28 February 2018 (v0.2)"
+output:  
+  pdf_document:
+    includes:
+      in_header: header.tex
+    toc: yes
+    fig_caption: true
+fontsize: 10pt
+geometry: margin=1.5in
 bibliography: ../manuscript/div05.bib
@@
 
-Functional traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.
+# Ecological traits
+
+Ecological traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.
 
@@
 
-The vocabulary proposed here is meant for two-dimensional spreadsheet data.
+The vocabulary proposed here is designed for two-dimensional spreadsheet data.
 These simple table formats are what a typical research project on a particular organism group and locality produces.
@@
 measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context.
-The vocabulary also contains terms for describing metadata acompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
-For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
+The vocabulary also contains terms for describing metadata accompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
+For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend to use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
 ]
 
-We assume that data are stored in a observation table, where each row is the reported measurement or fact for a single observation.
-At minimum, the following columns are required: 1.
+We assume that data are stored in an observation table, where each row is the reported measurement or fact for a single observation.
+At minimum, the following columns are required (Fig.
+1): 1.
 a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2.
@@
 
+![Elements of the original data set (species $\times$ trait matrix) are transferred into a observation table format with additional columns for trait and taxon standardization.
+Further information on the level of measurements, occurrences (i.e.
+individual specimens) and sampling locations can be added via the extended vocabulary.](figures.PNG)
+
 ## Identifiers
 
-To make data compatible with a consensus standard of taxon names or a look-up tables of traits, it is highly recommended to map them to globally valid terminologies by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
-GBIF Backbone Terminology)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'].
+To make data compatible with a consensus standard of taxon names or a look-up tables of traits, it is highly recommended to map the dataset-specific names provided for taxa (`scientificName`) and traits (`traitName`) onto globally valid terminologies by adding uniform ressource identifiers ([URIs](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier)) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by human-readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
+GBIF Backbone Terminology; see Fig.
+1)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'].
 
@@
 Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context.
+Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context (Fig.
+1).
 
-The data standard provides two extensions of the vocabulary that should be used to describe these information:
+The data standard provides two extensions of the vocabulary that should be used to describe this information:
 
-- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
-This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
 - The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
@@
 literature source or expert reference).
-- The `BiodiversityExploratories` extension provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`
+- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
+This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
+- The `BiodiversityExploratories` extension  (http://ecologicaltraitdata.github.io/ETS/#extension-biodiversity-exploratories)  provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`.
 
@@
 
-A comprehensive documentation of the package is found in the package website (http://EcologicalTraitData.github.io/traitdataform/), and on its Github repository (https://github.com/EcologicalTraitData/traitdataform).
+A comprehensive documentation of the package is found on the package website (http://EcologicalTraitData.github.io/traitdataform/), and on its Github repository (https://github.com/EcologicalTraitData/traitdataform).
 
@@
 This function maps column names into terms provided in the standard trait vocabulary and adds metadata as attributes to the output object.
-The function also will mould a species-trait-matrix into a observation table format.
+The function also will mould a species-trait-matrix into an observation table format.
 If data are measured on a single individual the function adds an occurrenceID to keep the reference context of individual measurements.
@@
             taxa = "name_correct", 
-            units = "mm"
+            units = "mm",
+            bexis = TRUE
             )
+
+write.csv(datasetStd, file = "filename.csv", row.names = FALSE)
+
 ```
 
-When saving your data, it is recommended to use
+This creates a clean text file (i.e. it omits the row names when producing a comma separated file with points as decimal delimiter) ready for upload on BExIS.
 
@@
 
-On page 2: Most likely, the number of plots is irrelevant to your trait data, but assign whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system.
+**On page 2:** Please assign the number of plots the data are related to (leave empty for data not assessed from exploratory specimens), and whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system.
 The taxon field is of particular importance here.
@@
 
-On page 3: describe the methodology underlying your trait values thoroughly.
+**On page 3:** describe the methodology underlying your trait values thoroughly.
 If data were extracted from literature or other database sources, declare that data are "literature data" and - if applicable - point to the column "basisOfRecord" where you should distinguish the source of your data.
 
-On page 4: When providing keywords, you should specify which kind of traits are reported (e.g. "morphological", "behaviour", "life-history", "functional").
+**On page 4:** When providing keywords, you should specify which kind of traits are reported (e.g. "morphological", "behaviour", "life-history", "functional").
 
-On page 5: Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd").
+**On page 5:** Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd").
 
-On page 6: Set 'exploratory' and 'keyPlots' to 'none' if your trait data are not assigned to a single plot and site.
-If you created a .txt or .csv file from the R package, chose fileFormat 'ASCII' and pay attention to the field delimiter and decimal Character.
-Missing values created by the R package are empty fields, thus enter "".
+**On page 6:** Set 'exploratory' and 'keyPlots' to 'none' if your trait data has no column with specific exploratories information, otherwise enter the column that contains the plot ID (by default "ExploratoriesPlotID").
+If you created a .txt or .csv file from the R package, choose fileFormat 'ASCII' and double check for the correct field delimiter (usually ",") and decimal separator (usually ".").
+Structural missing values for empty columns created by the R package are empty fields, thus enter "".
 Known missing values (i.e.
-values have not been measured for any reason) should not exist in the observation table format (the R package eliminates them by default).
-If you chose to keep rows for lacking trait measurements in the table, they should be marked as "NA".
-The section on 'matrix data' should be irrelevant if the data have been correctly transferred into observation tables.
+trait values which have not been measured for any reason) should not exist in the observation table format (the R package eliminates them by default, if they were correctly marked as NA in the original data).
+If you choose to keep rows for lacking trait measurements in the table, they should be marked as "NA".
+The section on 'matrix data' can be ignored since the data have been transferred into observation table format.
 The variables (i.e.
@@
 
-<!--Eliminate columns that are not present in your dataset. 
--->
-
-On page 7: if the trait data are extracted from one or multiple existing databases, it is necessary to refer to those here (by name and by original paper reference).
+**On page 7:** if the trait data are extracted from one or multiple existing databases, it is necessary to refer to those here (by name and by original paper reference).
 If the data have been uploaded before in a non-standard format, please provide the dataset ID.
 
+# Assistance and Feedback
+
+This project is work in progress. If something doesn't work as expected, please get in touch and report problems:
+
+- For troubleshooting during file upload, please contact BExIS Helpdesk (mail@?.de).
+For issues with the definitions of the vocabulary, please get involved in the discussion on Github (https://github.com/EcologicalTraitData/ETS/issues).
+- For comments on this manual, please contact us directly (florian.dirk.schneider@gmail.com). 
+- For bug reports or feature requests of the R-package, please file an issue on Github (https://github.com/EcologicalTraitData/traitdataform/issues).
+
 # References

### 2018-04-23 bexis_handbook/bexis_handbook_traits.Rmd
finalize BExIS manual, plus example

@@
 author: "Florian D. Schneider (florian.dirk.schneider@gmail.com)"
-date: "28 February 2018 (v0.2)"
+date: "23 April 2018 (v0.3)"
 output:  
@@
       in_header: header.tex
-    toc: yes
+    toc: no
     fig_caption: true
 fontsize: 10pt
-geometry: margin=1.5in
+geometry: margin=1.3in
 bibliography: ../manuscript/div05.bib
@@
 ---
+
+## Contributors
+
+Malte Jochum, Gaetane LeProvost, Caterina Penone, Andreas Ostrowski, Nadja K. Simons
 
@@
 These simple table formats are what a typical research project on a particular organism group and locality produces.
-The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported measurement or fact (e.g.
-measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context.
+The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported taxon, the measurement or fact (e.g.
+measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context of the occurrence.
 The vocabulary also contains terms for describing metadata accompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
@@
 
-We assume that data are stored in an observation table, where each row is the reported measurement or fact for a single observation.
+We assume that data are stored in an observation table, where each row is the reported measurement or fact for a single observation^[Another common format for trait data are species--trait matrices, where each row is a species and multiple trait values are stored in columns.
+These have to be converted into observation tables when applying the trait vocabulary.
+The R-package 'traitdataform' (see below) facilitates this conversion.].
 At minimum, the following columns are required (Fig.
@@
 a descriptive trait name (`traitName`);  3.
-the scientific taxon for which the measurement or fact was obtained (`scientificName`).^[Another common format for trait data are species--trait matrices, where each row is a species and multiple trait values are stored in columns.
-These have to be converted into observation tables when applying the trait vocabulary.]
+the scientific taxon for which the measurement or fact was obtained (`scientificName`).
 <!--For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa.
@@
 ![Elements of the original data set (species $\times$ trait matrix) are transferred into a observation table format with additional columns for trait and taxon standardization.
-Further information on the level of measurements, occurrences (i.e.
-individual specimens) and sampling locations can be added via the extended vocabulary.](figures.PNG)
+Further information on the level of taxa, ameasurements, occurrences (i.e.
+individual specimens) and sampling locations can be added via the extended vocabulary.](fig1.png)
 
@@
 
-Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context (Fig.
+Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the taxon, to the individual or specimen, or to the reported fact, measurement or sampling event.
+Here, the assignment of a unique identifier per single taxon (`taxonID`), measurement event (`measurementID`), or individual occurrence (`occurrenceID`), defines the structure that these data would take in a relational database context (Fig.
 1).
 
-The data standard provides two extensions of the vocabulary that should be used to describe this information:
+The data standard provides four extensions of the vocabulary that should be used to describe this information:
 
+- The `taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides terms to describe the hierarchical taxonomy information and taxon rank of the measurement. 
 - The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
@@
 
-1. preparing of own trait datasets for publication on low-threshold file-servers and project databases, and
+1. preparing of own trait datasets for publication on low-threshold file-servers (e.g. Figshare or Researchgate) and project databases (like BExIS), and
 2.
-automating the harmonisation of trait datasets from different sources by moulding them into a unified format^[The package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
+automatizing the harmonisation of trait datasets from different sources by moulding them into a unified format^[The package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
 See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
@@
 
-Installation of the package is possible via the devtools package directly from Github:
+Installation of the package is done via the 'devtools' package directly from Github^[The package is in beta stage, that is it is not yet complete in its desired functionality and undergoes heavy revisions.
+If you have difficulties with the package or the provided documentation, please report these on https://github.com/EcologicalTraitData/traitdataform/issues]:
 
@@
 
-In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data.
-<!-- BKR: In the example below, this looks like a manual matching.
-What is done automatically here and which information does the user have to provide?
-FDS: taxon is matched automatically, traits not.
-lack of trait APIs and webservices.
--->
+In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` as well as hierarchical taxonomy information to the core data.
+
 Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
-a thesaurus of traits^[This requires some manual mapping.
+a thesaurus of traits^[This requires some manual trait name mapping.
 Please see the package website for instructions.
@@
 
-write.csv(datasetStd, file = "filename.csv", row.names = FALSE)
-
+write.table(datasetStd, file = "filename.txt", 
+            row.names = FALSE, 
+            sep = "\t", 
+            na = "NA", 
+            dec = "."
+            )
 ```
 
-This creates a clean text file (i.e. it omits the row names when producing a comma separated file with points as decimal delimiter) ready for upload on BExIS.
+This creates a clean text file (i.e. it omits the row names when producing a tab-separated file with points as decimal delimiter) ready for upload on a data repository (like Figshare).
+
+An extended working example (with inline comments) can be found appended to this manual.
+
+**Always double-check the content of your output file before uploading it to an online repository!**
+
+As a particular use case, the package includes a function `standardize.exploratories()` for data assembled within the Biodiversity Exploratories project.
+The function maps plot IDs to universal geolocation and habitat information and adds these as columns to the core dataset.
+This is only relevant for traits that have been measured from specimens collected on the sites of the Exploratories.
 
@@
 
-When uploading the dataset to BExIS, you can use the predefined template "traitTemplate", which already contains a list of column definitions.
+You should use the predefined template "traitTemplate" when creating a new trait dataset on BExIS. Please pay special attention to the following points when providing the Metadata information:
 
-**On page 2:** Please assign the number of plots the data are related to (leave empty for data not assessed from exploratory specimens), and whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system.
+**On page 2:** Please assign the number of plots the data are related to (leave empty for data not assessed from exploratory specimens; this will be verified during data upload against the data provided in `BEPlotID`), and whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system.
 The taxon field is of particular importance here.
-The field 'bioticDataTypes'  should be set to "Trait Data".
+The field 'bioticDataTypes'  should be set to "Trait Data" and, if applicable, also to "Morphological".
 
@@
 
-**On page 5:** Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd").
+**On page 5:** Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd" or just "yyyy").
 
-**On page 6:** Set 'exploratory' and 'keyPlots' to 'none' if your trait data has no column with specific exploratories information, otherwise enter the column that contains the plot ID (by default "ExploratoriesPlotID").
-If you created a .txt or .csv file from the R package, choose fileFormat 'ASCII' and double check for the correct field delimiter (usually ",") and decimal separator (usually ".").
-Structural missing values for empty columns created by the R package are empty fields, thus enter "".
+**On page 6:** Set 'exploratory' and 'keyPlots' to 'none' if your trait data has no column with specific exploratories information, otherwise enter the column that contains the plot ID (by default "BEPlotID").
+If you created a .txt file from the R package, choose fileFormat 'ASCII' and double check for the correct field delimiter (ideally tab, i.e.
+"\t") and decimal separator (usually ".").
+Structural missing values for empty columns created by the R package are filled with "NA", thus enter "NA" for missing data.
 Known missing values (i.e.
-trait values which have not been measured for any reason) should not exist in the observation table format (the R package eliminates them by default, if they were correctly marked as NA in the original data).
-If you choose to keep rows for lacking trait measurements in the table, they should be marked as "NA".
+trait values which have not been measured for some reason although they could have) should not exist in the observation table format.
+The R package eliminates NAs in your original trait matrix by default when converting it to the long-table format.
+If you chose to keep these data, they should be differentiated as "n.d." (for "not determined") as values in the original data file.
+Reflect this annotation in the BExIS Metadata.
 The section on 'matrix data' can be ignored since the data have been transferred into observation table format.
@@
 
-# Assistance and Feedback
+After creating a Metadata object, you can append/upload a dataset to your BExIS entry. Please make sure that the order of columns is correctly aligned with the definitions of the template.
+
+By default, the function `standardize.exploratories()` produces an output containing *all* columns offered by the trait vocabulary (except the ones reserved for trait thesauri, i.e.
+lists of trait definitions) in the order expected by the BExIS template.
+(but please double-check!)
+
+If you create your trait data manually, i.e.
+without the assistance of the R-package, you can download an empty template as an xlsx file from BExIS after you created the Metadata object.
+This template contains all columns in the correct order.
+Please fill those row-wise according to the definitions provided by the vocabulary.
+At minimum, the columns `traitName`, `traitValue`, `scientificName` must be provided, but we highly recommend to relate taxon name with a uniform resource identifier provided in `taxonID`, and `occurrenceID` if multiple trait measurements have been taken from the same specimen.
+Provide as much information as is available to you!
+
+# Getting Help and providing feedback
 
@@
 
-- For troubleshooting during file upload, please contact BExIS Helpdesk (mail@?.de).
-For issues with the definitions of the vocabulary, please get involved in the discussion on Github (https://github.com/EcologicalTraitData/ETS/issues).
+- For troubleshooting during file upload, please contact BExIS Helpdesk (bexis@listserv.uni-jena.de). 
 - For comments on this manual, please contact us directly (florian.dirk.schneider@gmail.com). 
+- For issues with the definitions of the vocabulary, please get involved in the discussion on Github (https://github.com/EcologicalTraitData/ETS/issues). 
 - For bug reports or feature requests of the R-package, please file an issue on Github (https://github.com/EcologicalTraitData/traitdataform/issues).

### 2018-04-23 bexis_handbook/example01.Rmd
finalize BExIS manual, plus example

---
title: "Example 1"
output:  
  pdf_document:
    includes:
      in_header: header.tex
fontsize: 10pt
geometry: margin=1.5in
---

```{r, eval = FALSE}

# load library
library(traitdataform)

# load non-standardized dataset from BExIS 
# (requires access credentials and access rights)

moths <- read.service(21247, dec = ",")

# transform to observation long-table:

dataset1 <- as.traitdata(moths, 
                  taxa = "species",  # which column contains taxa?
                  traits = c("weight",   # which columns contain traits?
                             "wing_length", "wing_width", 
                             "wing_area", "wing_loading"
                             ),
                  units = c("mg", "mm", # which units do apply
                            "mm", "mm2", "mg/mm2"), 
                  keep = c(locationID = "plot") # which other cols to keep
                  )

# standardize taxonomy:
# applies a matching of taxonomic names to GBIF backbone taxonomy

dataset1Std <- standardize.taxonomy(dataset1)

# define a user-provided trait thesaurus:

traitlist <- as.thesaurus(data.frame(
  trait = c(
    "weight", 
    "wing_length",
    "wing_width",
    "wing_area",
    "wing_loading"
  ),
  traitDescription = c("dry weight of the individual", 
                       "the distance between base and apex (xi)", 
                       "the distance between tornus and costa (yi)", 
                       "wing_width x wing_length", 
                       "ratio of wing_area to body_mass"
                       ),
  valueType = c("numeric"),
  expectedUnit = c("mg", "mm", "mm", "mm2", "mg/mm2")
)
)

# standardize traits:

dataset1Std <- standardize.traits(dataset1Std, traitlist)

# standardize occurrence context by Exploratories Plot ID:

dataset1Std <- standardize.exploratories(dataset1Std)

# save to txt file, ready for BExIS upload: 
# omitting row names, use tab as field delimiter, 
# point as decimal delimiter, NA as missing value

write.table(dataset1Std, file = "mothtraitsStd.txt", 
            row.names = FALSE, 
            sep = "\t", 
            na = "NA", 
            dec = "."
            )

```

### 2018-05-01 manuscript/manuscript.Rmd
major revision of manuscript

@@
 - name: Malte Jochum \footnotemark[4]
-- name: Ga\"etane Le Provost \footnotemark[2]
+- name: Gaetane Le Provost \footnotemark[2]
 - name: Andreas Ostrowski \footnotemark[5]
 - name: Caterina Penone \footnotemark[4]
-- name: Markus Fischer
-- name: Pete Manning 
-- name: Birgitta König-Ries
+- name: Markus Fischer \footnotemark[4]\footnotemark[2]
+- name: Pete Manning \footnotemark[2]
+- name: Birgitta König-Ries\footnotemark[5]
 - name: Nadja K. Simons \footnotemark[6]
 csl: amnat.csl
+fontsize: 10pt
+geometry: margin=1in
 output:
@@
 
+<!-- CP:  I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
+
+That’s the structure I try to follow:
+1.
+Traits are important in ecology, evolution and conservation (i.e.
+It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
+
+2.
+To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources.
+When we do this we can answer new questions and find generalities.
+So there is a growing interest in assembling trait data from different sources and taxa (e.g.
+give the number of papers using TRY).
+
+3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.
+
+4.
+One solution are databases.
+But these are costly and depend on the will of few persons/funding.
+They also cover few taxonomic groups.
+Matching between different databases is still difficult because of different definitions and structures.
+Most data are still uploaded separately.
+We need a solution to make their use easy. 
+-->
+<!-- GLP: 
+ Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g.
+TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals).
+Say also something on the development of general protocols for measuring traits (e.g.
+Moretti et al.
+2016; Freschet et al.
+2017) ?
+
+Importance of data standardization.
+This is the main idea here and so I wonder if some idea of the first point (Need for data standardization) should not be merge here with the last paragraph of the introduction ?
+As there is also some repetitions between those 2 parts.
+
+I wonder if you should speak about ‘Functional traits’here or just Traits ?
+As you are not only considered Functional traits in this paper (e.g.
+life history traits) ?
+Maybe remove the definition no ?
+-->
+<!-- PM: Thuis is quite complex- e.g.
+‘taxonomic impediment’etc.
+I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples.
+I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.
+
+Huge number of references here- I think you can strip it down to a few key ones, and reviews.
+Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site.
+-->
+
+<!-- 1. Traits are important in ecology, evolution and conservation -->
+
 Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07].
 Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
-For example, trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity and ecosystem functioning (refs). 
-Using traits is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring both the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
+Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
+For example, using traits is a promising venue to bypass taxonomic impediment, i.e.
+the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
+
 With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].
@@
 --> 
-<!-- CP: I like very much this whole paragraph (both writing and content), but I feel that it’s a bit out of the topic.
-It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
-I give some suggestions on how to cut it and rather develop more the following one, which people are surely less aware of and which sets more the aim of the paper.
-I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
-
-That’s the structure I try to follow:
-1. Traits are important in ecology, evolution and conservation (i.e. this paragraph)
-
-2.
-To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources.
-When we do this we can answer new questions and find generalities.
-So there is a growing interest in assembling trait data from different sources and taxa (e.g.
-give the number of papers using TRY).
-
-3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.
-
-4.
-One solution are databases.
-But these are costly and depend on the will of few persons/funding.
-They also cover few taxonomic groups.
-Matching between different databases is still difficult because of different definitions and structures.
-Most data are still uploaded separately.
-We need a solution to make their use easy.
-
-But of course it’s just a suggestion, feel free to happily ignore it!
-
-I would choose 2 or 3 (also for the previous), anyway it’s impossible to be exhaustive. Maybe put (e.g. ref1, ref2) so make clear that it’s just examples
-
--->
-<!-- GLP: 
-The second paragraph is a bit confusing for me.
-
-I wonder also if the introduction could be restructures like this :
-
-Trait based approaches and their use
-From the individual to population
-Community level (assemblage, interactions, effects on ecosystem)
-Macroecological question
-
- Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g.
-TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals).
-Say also something on the development of general protocols for measuring traits (e.g.
-Moretti et al.
-2016; Freschet et al.
-2017) ?
-
-Importance of data standardization.
-This is the main idea here and so I wonder if some idea of the first point (Need for data standardization) should not be merge here with the last paragraph of the introduction ?
-As there is also some repetitions between those 2 parts.
-
-There is just some copy-past to do.
-I didn’t change the structure of the introduction but if you like the structure that I proposed, we can discuss next week or I can help you with that.
-I like to write introduction 
-
-I wonder if you should speak about ‘Functional traits’here or just Traits ?
-As you are not only considered Functional traits in this paper (e.g.
-life history traits) ?
-Maybe remove the definition no ?
--->
-<!-- PM: Thuis is quite complex- e.g.
-‘taxonomic impediment’etc.
-I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples.
-I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.
-
-Huge number of references here- I think you can strip it down to a few key ones, and reviews.
-Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site.
--->
 <!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions.
@@
 
-However, fully exploiting this portential relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa.
-<!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific?
--->
-To mobilize data for this enterprise <!-- PM: Please define what the exercise is- too vague at present
--->, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources.
-<!-- PM: I think this is a key justification- perhaps say what happens if we don’t do this – non repeatable work, differing definitions of traits that may lead to scientific flaws, repetition of the painstaking task of processing trait data e.g.
-to compute species level means.
--->
-The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way.
-<!-- CP: I felt that this needed a definition, not sure everybody know what an ontology is.
-Maybe we should guide more the reader that might not be familiar with the problems.
--->
+<!-- 2. assemble trait data into databases -->
 
-Nonetheless, individual researchers often neglect the benefit of centralised databases <!-- PM: Explain what this benefit is.
--->.
-Furthermore, many journals and research funding agencies require  data to be published without access restrictions [@swan12; @allison15].
-However, this primary data is often posted on file hosting services, where no data standards are forced upon the uploaded material (e.g.
-Figshare, Dryad).
-The long-term outcome of this situation is an abundance of freely available data, but a lack of data standardisation.
-
-In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users.
-We further (2.) review existing trait databases and web services, as well as initiatives for standardisation.
-We (3.) describe the common practice <!--PM: Current or your own new definition?--> for structuring trait data for ecological research. 
-Based on these considerations, we (4.) propose a low-threshold <!-- PM: What’s this? Please define
--->  structure and vocabulary for describing trait-datasets compatible with existing ontologies <!-- PM & CAT: define--> .
-(5.) Finally, we provide an R-package that assists in aply this structure to own data and for the harmonisation of data from distributed sources.
-
-With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold <!-- PM:Again, I’m not sure what this is a low threshold of?
---> trait data standardisation.
-
-Despite the obvious benefit and great potential of trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
+However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and taxonomic and regional coverage of trait data, both of present-day taxa ase well as in evolutionary deep-time.
+Researchers have a growing interest in assembling trait data from different sources since it pays off by getting answers to general questions.
+Integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
 All of these factors render trait data extremely heterogeneous.
 
-As a response to these issues, trait data have been standardized in centralized databases for specific organism groups and research questions.
-<!-- PM: (refs, also give an e.g. or two for some big ones like TRY- saw there was a new one for tropical mammals a couple of days ago too and a huge Chinese plants one- in )--> 
-These compiled data are either being published as static data tables or in the form of sophisticated online portals like the plant trait database TRY (see below for a review of current initiatives).
-These portals attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, and importantly also, provide networks for consensus building in measurement methods.
-Furthermore, database services offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.
+<!-- 3. trait databases have been developed, with standard structure and terms --> 
+To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions (ref to TRY, panthera, etc.).
+These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (ref Amphibio, panthera, mammals), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g.
+@kattge11].
+Furthermore, these portals map heterogeneous data into a common scheme, and importantly also, offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.
+To enable the compatibility of data across data-bases, semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
+Solving all these issues of data assemblage and harmonization in centralised infrastructures requires a huge effort in costs and personell (see below for a review of current initiatives).
 
-In contrast to standardized databases, open-access data publication in low-threshold repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org is becoming increasingly frequent.
-Such repositories foster decentralised data hosting with low thresholds regarding data standardisation and documentation.
+<!-- 4. distributed trait data must be integrated --> 
+Many journals and research funding agencies today require data to be published without access restrictions [@swan12; @allison15].
+An increasing number of individual projects comply  with open data standards and publish their primary data on file hosting services like Figshare, Dryad or Researchgate, where no data standards are forced upon the uploaded material.
+The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.
+Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.
 
-Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.
-To cope with this trend towards distributed data sources, data brokering services (e.g.
-the German Federation on Biological Data; gfbio.org)  or database <!-- AO: Just “data” or “research data”--> management systems for scientific projects (e.g.
-BEXIS, XX <!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+In this paper, we (1.) review existing trait databases and online portals, as well as initiatives for standardisation.
+In this review, we discuss the current practice and the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users.
+Based on these considerations, we (2.) propose a minimal structure and vocabulary for describing trait-datasets, compatible with existing terminology standards for biodiversity data.
+Finally, we (3.)  propose the development of software that assists in aplying a harmonisation of data from distributed sources.
 
-However, to mobilize the broad wealth of distributed data into structured databases <!-- AO: Maybe better: formats-->,  a certain minimum of information from original data providers is requrired, e.g.
-on the methodology and definitions of trait values or the covariates describing the ecological context of the trait observation.
-In order to get this accurate and well-defined input ecological researchers need to be aware of and understand basic principles of data standards and ontologies.
-<!-- AO: The last paragraphs may need more sources/proofings?--><!-- PM: Literacy is quite a strong word- as illiterate is quite an insult.
-I think awareness is a problem too.
--->
+With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards an accessible trait data standardisation.
 
- More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
+<!-- FDS: Why 'ecological' trait data?
+what is special about this subject?
+Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->
 
@@
 In this section, we review four kinds of initiatives that are of relevance for trait data standardisation: 
-the first are projects that work with trait datasets that have been assembled out of a particular research interest, either by measurement or colated from literature; 
-the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and thesauri; 
-the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.
+- The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
+- the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
+- the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies; 
+- the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct relational 'ontologies' for an application in a 'semantic web' of information.
 
-We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the TOP Thesaurus of plant traits [@garnier17].
+We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
 We show how the degree of trait-data standardization in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources.
@@
 Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management.
-In most of these datasets, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
+As a result of this, datasets also greatly vary in structure. 
+The column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
 Datasets vary in the detail provided in the metadata, e.g.
 for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
+If factual trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated quantitative measurements averaged from raw data, they usually are reported in a species$\times$trait wide-table format.
+Here, each row contains a species (or taxon) for which multiple traits are reported in columns.
+Similarly, when reporting raw data, researchers place observations of individual specimens in rows with multiple trait measurements applied to the same specimen placed in columns. 
+The variability of the number and meaning of columns in these tables requires manual adjustments when merging multiple datasets.
+
+<!-- Fig 1 here? -->
 
@@
 
-Regarding open access, hardly any of these centralized databases for functional traits comply with the criteria demanded by journals and funding agencies for primary data publication.
-They incentivize data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
+Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication.
+They incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
 Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
@@
 
+To conclude, there is a shortage of globally valid trait thesauri or ontologies for traits that emerged from a standardisation effort of methodologies and broad consensus process within scientific field of ecology *and* that are sufficiently interlinked and accessible to fulfil semantic web standards.
 The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
 To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.
+Such thesauri would advance the possibilities of aggregated datasets and reduce noise and uncertainty. 
+More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from distributed sources.
 
@@
 
-Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that covers most potential use cases.
+Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for databases that covers most potential use cases of trait-based ecology.
 This data structure is build around a central data table that contains observations, i.e.
 a single event of measurements on the same individual plant specimen at the same point in time.
-Identifiers link multiple measurements to the same observation, each measurement being well defined by additional standard tables.
+This structure puts emphasis on the fact that trait  data are often measured for an analysis of correlations of multiple traits.
+Identifiers link multiple measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables.
 The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
@@
 
-However, these proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
-None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
+These proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
+Hovewer, hardly any of the aforementioned aggregated and harmonized datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards.
+
 One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
-If the traits are assembled for a specific research question with little ancillary information, it may be safer to add them as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
+If the traits are assembled for a specific research question with few columns of ancillary information, it may be safer to add this information as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
 This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
-<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
-The trade-off is data-consistency and readability *vs.* avoidance of content duplication.
 
-<!--When traits are assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
-Here, location or measurement information are the primary focus of the investigation and must be contained in the .
-<!-- PM: I get lost here-->
-<!--A general recommendation, whether a column describes ancillary data or is part of the central dataset is ill advised.
-Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
--->
-<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
+The other reason is lack of awareness of data providers for trait data standardization: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.
+Investing time and energy into research for the best and most sustainable data structure is not affordable for small and intermediate-size research projects.
 
-<!--while the existing trait data standards may cover all of the aspects required in the various use-cases of trait-based research.
-<!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific -->
-The other reason is lack of awareness and literacy: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, which is not affordable for small and intermediate size research projects.
 By filling this gap, data brokering services (e.g.
-the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g.
-BExIS, NEON ) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into consistent and interlinked data infrastructures.
-One goal of such initiatives is to facilitate the publication and standardization of data for researchers, e.g.
-by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
-
-The practical importance of these frameworks for data providers and data users is often not obvious and not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
-<!-- PM: There is also little incentive for researchers to do this- why bother? Incentives are also needed. -->
-The pay-off is only playing out in the long run. Citations of trait datasets are higher when they are easier to access and relate to other data due to standard structures and vocabularies. 
-Along with the push towards open access publication of primary data, the scientific community must come up with ways to incentivize and require data standardization.
+the German Federation on for Biological Data; gfbio.org) or data management systems for scientific projects (e.g.
+BExIS, NEON<!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance.
+These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
+One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
 
@@
 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available.
+The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
+<!-- AO: You say it already 2 sentences before --><!--
 This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g.
@@
 on variation of means or literature source) in secondary columns or accompanying datasheets.
-The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
-<!-- AO: You say it already 2 sentences before --><!--
+
 Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a].
@@
 
+<!--The trade-off is data-consistency and readability *vs.* avoidance of content duplication. -->
+
+<!--When traits are assessed on a higher level of organisation, e.g.
+microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
+Here, location or measurement information are the primary focus of the investigation and must be contained in the .
+<!-- PM: I get lost here-->
+<!--A general recommendation, whether a column describes ancillary data or is part of the central dataset is ill advised.
+Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
+-->
+<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
+<!--while the existing trait data standards may cover all of the aspects required in the various use-cases of trait-based research.
+<!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific -->
+
 <!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
@@
 semantic web | 
+relational database | 
+dataset | 
+Web service | 
+Online Portal |
+Repository | 
+Darwin Core Terms |
+Darwin Core Archive |
 
@@
 
-# 2. Proposing an ecological traitdata standard
+# 2. Proposing an ecological traitdata standard vocabulary
 
-As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees measurement detail for multiple use cases of trait-based ecological research.
-<!-- PM: It took a long time to get to this point- I suggest brininging in a section early on that describes what the paper will do as it’s not always clear where it is leading. --> 
-<!-- FDS: point out focus on distributed low-threshold data, not full-fledged database initiatives. -->
+As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research.
+Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in a semantic web.
 
-- we propose Vocabulary for two-dimensional tables of traits. 
-- long-table format.
-has advantage of easy merging and filtering, computing, modelling.
-But disadvantage of multi-type columns (e.g.
-for values) and repetitive content.
-For complex dimensional databases, use Kattge, DwC or ABCD instead, packaged as ZIP files with annotations in EML Metadata format rather than multi-spreadsheet files.
-- linking to standard taxonomy and trait terminology.
-Allowing for but minimizing duplicate information.
-by providing standardised duplicates, align user-provided nomenclature and trait methodology with taxonomic nomenclature and trait ontologies.
-transparency for data user.
+By using this standard vocabulary, authors can ensure that the description of tables of trait measurements that are uploaded to distributed data repositories like Figshare, Zenodo etc.
+will be unambiguous and generally applicable.
+It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure accessibility in the long term.
+
+The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.
+
+In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
+We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
+Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
+
+## Why use the vocabulary
+
 - Offering compatibility, input for dimensional databases by using global identifiers and internal identifiers that provide hierarchical layers (measurement, occurrence, sampling event, data source, bibliographic reference)
@@
 allows for originary data (reporting own measuremetns) or compiled data sheets (reporting multiple dataset identifiers).
-In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
-<!-- AO: Webservices!!
-;  Really to be considered as standards?
--->
-
-The online reference glossary of terms and description of the proposed structure can be found at https://ecologicaltraitdata.github.io/ETS/.
-
-We acknowledge that the vocabulary proposed will not capture the full complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ...
--->.
-Therefore, we invite researchers to contribute to the next iterations of this standard terminology and develop own applications and ontologies.
-
+- wherever appropriate, use URIs for linking to external ressources. 
 <!-- PM: It seems that the object identifier for an individual is the key distinction here- this could be clearer.
-
 Maybe you address this later but does your tool assign one of these- how to you make sure they are all unique if there are multiple users- e.g.
@@
 
-## Standard structure for trait data
+- long-table format. has advantage of easy merging and filtering, computing, modelling. But disadvantage of multi-type columns (e.g. for values) and repetitive content.
 
-As described above, observation-level long-table formats are the most predictable and compatible format for the core content of trait datasets and should therefore also be used for standalone uploads on low-threshold repositories.
-<!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
--->
-<!-- CP: Not sure this needs to be repeated -->
+- linking to standard taxonomy and trait terminology. semantic web standards, URI identifiers of external resources.
 
-To enable a better standardisation of nomenclatures and values, the entries of the core dataset <!-- CP: Does this need to be defined --> should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names.
-In these duplicate columns, values should be reported as 'harmonized' entries [@kattge11].
+- Allowing for but minimizing duplicate information.
+by providing standardised duplicates, align user-provided nomenclature and trait methodology with taxonomic nomenclature and trait ontologies.
+transparency for data user.
 
-, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08].
-<!-- BKR: There is some grammatical error here; not sure what you try to say-->
-<!-- AG: This was a little bit confusing for me.
-In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon.
-On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name.
-In any case I would suggest to recommend to keep both the URI and the literal.-->
-<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
-If data contain additional information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below).
-<!-- PM: Not clear here -->
-For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
+- vocabulary is structured into sections: core, measurement, occurrence, metadata, traitlists.
+Short definitions.
+These sections are meant to guide through the vocabulary.
+their structural meaning depends on context provided by hierarchical IDs.
+
+Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
+For datasets containing multi-layered information on observations, traits, taxa and environmental context, we highly recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09].
+<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
+
+## How to apply the standard vocabulary
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
+\caption{Formats used for trait datasets.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
+
+\end{figure}
+
+Column names and, where possible, field entries of a trait datasets published on online repositories should draw from the defined vocabulary.
+
+The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
+trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
+traits) observed on a biological entity (i.e.
+an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e.
+a species or higher-level taxon).
+
+Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below).
+
+Beyond the core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information).
+These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
+For this case, we offer three extensions of the core vocabulary ("taxon", "measurement or fact", and "occurrence") that expand and refine terms of the Darwin Core Extensions (see below).
+This information may simply be added as extra columns to the core dataset.
+
+For reasons of long-term accessibility, data should not be uploaded in spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
 UTF-8 or ASCII).
-<!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->
 
 In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider et al.
-2017 Ecological Traitdata Standard v0.7, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
+2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
 In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
-FDS: to consider!! -->
 
-However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. 
-In more complex datasets, the repetition of verbose entries and long character strings in a single data table should be avoided.
-Therefore we recommend use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09].
-<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
-
-## Standard vocabulary for trait data
-
-<!-- PM: Can you say a little about what this vocab is and how it was chosen/defined? -->
-
-Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/ETS/. 
-Individual terms <!-- DF:Please explain the difference between the “individual terms” and the “field entries” above-->  should be referenced via their global identifier, which is hosted by the GFBio Terminology Service <!-- DF: Please add citation to the TS Paper in Datenbanken-Spektrum, see https://terminologies.gfbio.org/about/#citation  --> and also can be accessed programmatically (i.e.
+Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam14; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e.
 via the Application Programming Interface, API).
@@
 
-### Core columns
+Besides this use in published datasets, the standard vocabulary could be imposed in webservices, e.g.
+APIs that provide direct access to online databases or download tools that pull subsets of data from larger databases.
 
-To describe the minimum core content, where each row is the reported measurement or fact for a single observation, the following columns are required: 1.
-a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2.
-a descriptive trait name (`traitName`);  3.
-the scientific taxon for which the measurement or fact was obtained (`scientificName`).
+<!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
+It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc.
+-->
+ <!-- Nadja: I like the idea of having a figure explaining the different terms -->
+
+<!-- 
+(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model.
+However, such network data may be of interest to be combined with trait data.
+) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names.
+-->
+
+### Specifications of core terms
+
+To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum: 
+1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
+2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
+3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
 For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
@@
 
-In this case, it is essential to relate the user-defined names to a consensus standard of taxon names or or look-up tables of traits. 
-This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g.
-GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.
+In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up tables of traits. 
+This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively).
+For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.
 By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
@@
 
-If no URIs are referenced for taxonID, it is highly recommended to keep an additional reference to a kingdom and order (using columns `kingdom`, `order` as described by DWC and suggested for EOL Traitbank) to avoid ambiguities with taxonomic homonyms.
-<!-- BKR: Why do you need this, if taxonIDs are used.
-They should be unambiguous
-FDS: right, rephrase to: either use taxonID or provide this.-->
-Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
-This information can then be used for trait inference at higher taxonomic resolution.
-<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
--->
+<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
 
-<!--The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control.
-The accessibility by software differs widely.
-For biodiversity data, the OBO Foundry (http://www.obofoundry.org/) or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer API services for computational access.
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.
--->
+# Specification of traits
 
-If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-<!-- A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
-Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
-units or legit factor levels) of measured values or reported facts.
-It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
--->
-By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for a specific project context which should always be referenced in the core trait dataset.
-The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable internet address (via URIs).
+If no globally available trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
+Such a 'controlled vocabulary' would, in its simplest form, assign trait names with A) an unambiguous definition of the trait and B) an expected format  of measured values or reported facts (e.g.
+units or legit factor levels).
+
+By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context which should always be referenced in the core trait dataset.
+The thesaurus might accompany the core data in a separate file (e.g.
+within a Darwin Core Archive), or be published on any other stable internet address (via URIs).
 <!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
-
-See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus].
+See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].
 We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
 
-### Metadata
+### Specification of Metadata terms
 
@@
 
-In the case of raw measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata. 
-In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be linked to the measurement level, via an unambiguous `datasetID` and a descriptive `datasetName`.
+In the case of raw measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g.
+in a separate file or on the download website).
+In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be provided at the measurement level.
+This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
 
@@
 
-<!-- 
-(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits.
-However, such network data may be of interest to be combined with trait data.
-) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names.
--->
+<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
+FDS: to consider!! -->
 
@@
 
-- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
+- the `Taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides further terms for specifying the taxonomic resolution of the observation and to ensure correct reference in case of synonyms and homonyms.
+- The `Occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
 This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
-- The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- The `MeasurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
 It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
@@
 
-Many terms of these extensions refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
-<!-- BKR: Looking at your definitions, in most cases, I am not sure how you refine the terms. Is it really refinement or is it just using them with the same meaning they have in DwC?
--->
-<!-- PM: All of this sounds like quite a lot of work- so will it be automated, or incentivized somehow?
+Many terms of these extensions refine or copy terms of the Darwin Core Standard and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
+
+<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).
+Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
+This information can then be used for trait inference at higher taxonomic resolution.
+<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
 -->
 
-As for taxon or trait information, these additional layers of information can either be added as extra columns to the core dataset or be kept in separate data sheets of a Darwin Core Archive (see above), this avoiding redundancy and duplication of content.
+As for metadata information, these additional layers of information can either be added as extra columns to the core dataset or be kept in separate data sheets (published separately or as part of a Darwin Core Archive), this avoiding redundancy and duplication of content.
 A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
@@
 a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
-
-<!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
-It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc.
--->
- <!-- Nadja: I like the idea of having a figure explaining the different terms -->
 
@@
 
-Accordingly, we provide the R package 'traitdataform' to assist the production of data compliant with the trait data standard proposed here. There are two major use cases for the package:
+Accordingly, we started to develop the R package 'traitdataform', which assists the production of data compliant with the trait data standard proposed here.
+There are two major use cases for the package:
 
-1. preparing of own trait datasets for publication on low-threshold file-servers and project databases, and
+1. preparing own trait datasets for publication on public file-servers and project databases, and
 2. automating the harmonisation of trait datasets from different sources by moulding them into a unified format.
 
-A comprehensive documentation of the package is found in the package vignette, and on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
-
+A comprehensive documentation of the package is found on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
 The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
 
-The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
+<!-- The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
 2).
@@
 -->
-Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
+<!-- Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
 a thesaurus of traits.
@@
 FDS: might be referred to in the section about thesauri -->
-
+<!--
 The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
@@
 
+-->
+
 Since the intention of the package is to facilitate upload of own data to hosting services but also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
-See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
-We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
+<!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
 
@@
 
-To serve the demand for simple ways to standardise and harmonize ecological trait data, we propose a versatile vocabulary for two-dimensional datasets.  
-We also provide a toolbox in R to transfer and harmonize data into this scheme.
+<!--The practical importance of these frameworks for data providers and data users is often not obvious and not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
+
+The investment is only paying out in the long run. Citations of trait datasets are higher when they are easier to access and relate to other data due to standard structures and vocabularies. 
+Along with the push towards open access publication of primary data, the scientific community must come up with ways to incentivize and require data standardization.
+
+More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.-->
+To serve the demand for simple ways to standardise and harmonize ecological trait data, we propose a versatile vocabulary for simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
+We also present a toolbox in R to transfer and harmonize data into this scheme.
 
@@
 
-<!--The trait data standard as well as the R package are published for open source community development.
-Contributions are invited and vital for the future development of the data standard and an extension and improvement of the R package.
--->
-
 # Acknowledgements
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal and Jens Kattge for comments on the manuscript, the trait data standard and the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal and Jens Kattge for comments on the manuscript, the trait data standard vocabulary and the pre-releases of the R-package.
 
@@
 
-The online reference for the Ecological Traitdata Standard described in this paper is https://ecologicaltraitdata.github.io/ETS/
+The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/
 
-The package website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform
+The development website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform.
 
@@
 
-Development of the Standard and the R-package is coordinated via https://github.com/EcologicalTraitData/ .
+Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/ .
 

### 2018-05-02 manuscript/manuscript.Rmd
update manuscript, argument for use, author list, bibliography

@@
 - name: Markus Fischer \footnotemark[4]\footnotemark[2]
+- name: David Fichtmüller \footnotemark[5]
+- name: Anton Güntsch \footnotemark[5]
 - name: Pete Manning \footnotemark[2]
-- name: Birgitta König-Ries\footnotemark[5]
-- name: Nadja K. Simons \footnotemark[6]
+- name: Birgitta König-Ries\footnotemark[6]
+- name: Nadja K. Simons \footnotemark[7]
 csl: amnat.csl
-fontsize: 10pt
+fontsize: 12pt
 geometry: margin=1in
@@
 \footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
-\footnotetext[5]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[6]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
+\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
 
@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07].
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
 Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
@@
 With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].
+A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15;  @renner14; @vanderplas17; @price13;  @ricklefs17].
 
-Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org. 
-Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, as increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
-<!-- AG: Perhaps mention the Bouchout declaration here: http://www.bouchoutdeclaration.org/ ?
--->
+Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net, Zenodo.org. 
+Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, which supports the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
 However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
@@
 As a result of this, datasets also greatly vary in structure. 
-The column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
+Data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/), offer means to meet the high scientific criteria of data management.
+
+For trait-data, there are specific issues arising from the variability of data structure. 
+For instance, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
 Datasets vary in the detail provided in the metadata, e.g.
@@
 As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
-On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
-The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
-
-<!--Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14a; @fierer14;  @fierer12].
+Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12].
 Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13].
@@
 the KEGG orthology, @kanehisa12].
--->
+On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
+Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
+The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans, make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
 
 Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication.
-They incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
-Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
+The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
+With the pro-active turn towards open access data (as declared inthe Bouchot Declaration; http://www.bouchoutdeclaration.org/) it may be necessary to find other incentives for data submission.
+
 <!-- AO: In a sense of data availability to the public : yes.
@@
 The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
-For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology.
 The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review.
 Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
+TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
 The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].
@@
 
-By filling this gap, data brokering services (e.g.
-the German Federation on for Biological Data; gfbio.org) or data management systems for scientific projects (e.g.
-BExIS, NEON<!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance.
+By filling this gap, data brokering services [e.g.
+the German Federation on for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
+BExIS, NEON Dataportal; Diversity Workbench, diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance.
 These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
 One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
-<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
+<!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
+
+## Conclusion of review
+
+Initiatives for standardization (e.g.
+ontologies and data standards) and platforms for data management (e.g.
+database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.
+With low-threshold offers for the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, the concern is with data harmonization of these distributed data sources.
+
+The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardization on the side of data providers. 
+We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in data visibility and interconnectedness of the published data.
 
@@
 
+-->
+
 Term | Definition 
@@
 Terminology |  
+Term |
 Thesaurus | 
@@
 relational database | 
+two-dimensional data | 
 dataset | 
@@
 Repository | 
-Darwin Core Terms |
+Darwin Core Standard |
 Darwin Core Archive |
-
--->
+Application Programming Interface (API) | 
+Data standard | 
+Uniform Resource Identifier (URI) |
 
@@
 As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research.
-Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in a semantic web.
+Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.
 
@@
 We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.
+
 However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
+Also, at present we conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model.
+However, such network data may easily be combined with trait data via the taxon identifiers or by using other Extensions of the Darwin Core (DwC).
 Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
@@
 
-- Offering compatibility, input for dimensional databases by using global identifiers and internal identifiers that provide hierarchical layers (measurement, occurrence, sampling event, data source, bibliographic reference)
-- using these identifiers enables the vocabulary for for observation level data (linked via an occurrence identifier) raw measurement data (each row an own measurement identifier), as well as for aggregated data on the species level (each row an own measurement identifier plus aggregation information, like taxon rank or statistical method).
-allows for originary data (reporting own measuremetns) or compiled data sheets (reporting multiple dataset identifiers).
-- wherever appropriate, use URIs for linking to external ressources. 
-<!-- PM: It seems that the object identifier for an individual is the key distinction here- this could be clearer.
-Maybe you address this later but does your tool assign one of these- how to you make sure they are all unique if there are multiple users- e.g.
-if you pooled together two previously complied datasets which might contain some of the same individuals?
--->
+The incentives for using a unified vocabulary for trait data might not be obvious for data providers, who are mainly concerned with a particular study site, organism group and research question.
+We won't discuss here in depth how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis.
+It appears to be broad consensus that an open biodiversity science is crucial for an evidence based decision making and conservation policy on regional and global scales.
 
-- long-table format. has advantage of easy merging and filtering, computing, modelling. But disadvantage of multi-type columns (e.g. for values) and repetitive content.
+Instead, this section points on the technical aspects and advantages of readying own trait data to link with the 'semantic web' of biodiversity data and knowledge. 
+By using a constrained vocabulary with globally accessible definitions of terms, data can be accessed more easily by other researchers and datasets from different sources can be combined into aggregated datasets.
+Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary.
 
-- linking to standard taxonomy and trait terminology. semantic web standards, URI identifiers of external resources.
+It is implicit to this vocabulary to use a two-dimensional observation long-table format for the data, rather than a species-trait matrix (where some of the columns would be trait names which are not included in the vocabulary, but are part of trait thesauri).
+As this format draws from a defined set of columns,  merging of datasets is  easier.
+Long-table datasets also purport multiple advantages for data manipulation [e.g.
+filtering, sub-setting and aggregating data; @wickham14].
 
-- Allowing for but minimizing duplicate information.
-by providing standardised duplicates, align user-provided nomenclature and trait methodology with taxonomic nomenclature and trait ontologies.
-transparency for data user.
+Using well-defined identifiers ('IDs') to structure the dataset offers compatibility with the input formats for larger databases that may offer better visibility and impact of the data contribution. 
+Identifier fields are key elements to structure the two-dimensional dataset. 
+For instance, for occurrence level data, were multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link these measurements across the rows of the dataset.
+Similarly, multivariate measurements, like for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
+If literature data are provided, data are probably not structured below the taxon level and a 'taxonID' is the key identifier.
+A 'datasetID' allows to report datasets that have been compiled from multiple sources.  
+Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external ontology services (see above for resources).
 
-- vocabulary is structured into sections: core, measurement, occurrence, metadata, traitlists.
-Short definitions.
-These sections are meant to guide through the vocabulary.
-their structural meaning depends on context provided by hierarchical IDs.
-
-Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
-For datasets containing multi-layered information on observations, traits, taxa and environmental context, we highly recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr15, ideally compiled into Darwin Core Archives [DwC-A, @robertson09].
-<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
+However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
+For datasets containing multi-layered information on observations, traits, taxa and environmental context, we  recommend to consider relational datatabase structures composed of multiple spreadsheets linked via identifier columns, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
+The trade-off is data-consistency and readability in a single table *vs.* avoidance of content duplication and redundancy in a relational database.
+The standard vocabulary proposed here may still be applied within these structured databases.
 
@@
 
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider et al.
-2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
+In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A.
+and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
 In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
@@
 Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam14; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e.
-via the Application Programming Interface, API).
+via the Application Programming Interface, API **in preparation!**).
 Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
@@
 APIs that provide direct access to online databases or download tools that pull subsets of data from larger databases.
-
-<!-- Malte: For readers not that familiar with trait data, the differences between measurement, occurrence and trait ID might not be that clear.
-It might be worth adding either a sentence explaining in an example what is what OR a tiny figure that shows how a specimen is sampled, processed, traits are measured etc.
--->
- <!-- Nadja: I like the idea of having a figure explaining the different terms -->
-
-<!-- 
-(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model.
-However, such network data may be of interest to be combined with trait data.
-) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names.
--->
 
@@
 
-# Specification of traits
+### Specification of traits
 
@@
 
-Many terms of these extensions refine or copy terms of the Darwin Core Standard and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
+Many terms of these extensions refine or copy terms of the Darwin Core Standard (Dwc) and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
 

### 2018-05-02 manuscript/manuscript.Rmd
finalize ms v0.7, shared with coauthors

@@
 author:
-- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
+- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
 - name: Malte Jochum \footnotemark[4]
-- name: Gaetane Le Provost \footnotemark[2]
+- name: Gaetane Le Provost \footnotemark[3]
 - name: Andreas Ostrowski \footnotemark[5]
 - name: Caterina Penone \footnotemark[4]
-- name: Markus Fischer \footnotemark[4]\footnotemark[2]
 - name: David Fichtmüller \footnotemark[5]
 - name: Anton Güntsch \footnotemark[5]
-- name: Pete Manning \footnotemark[2]
+- name: Pete Manning \footnotemark[3]
 - name: Birgitta König-Ries\footnotemark[6]
@@
 fontsize: 12pt
-geometry: margin=1in
+geometry: margin=1.3in
 output:
-  pdf_document:
+  word_document:
     fig_caption: yes
-    fig_crop: no
-    keep_tex: yes
-    template: pretty.template
+    fig_width: 6
   html_document:
@@
     theme: united
-  word_document:
+  pdf_document:
     fig_caption: yes
-    fig_width: 6
+    fig_crop: no
+    keep_tex: yes
+    template: pretty.template
 bibliography: div05.bib
@@
 \footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
-\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
+\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
+\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
 \footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
@@
 
-functional ecology, functional traits, semantic web, ontologies, data standard
+functional ecology, species traits, semantic web, ontologies, data standard
 
@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
+Species traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
 Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
@@
 <!-- 3. trait databases have been developed, with standard structure and terms --> 
-To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions (ref to TRY, panthera, etc.).
-These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (ref Amphibio, panthera, mammals), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g.
-@kattge11].
+To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions [e.g.
+TRY, @kattge11; PanTHERIA, @jones09; AmphiBio, @oliveira17].
+These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (in the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g.
+TRY, @kattge11].
 Furthermore, these portals map heterogeneous data into a common scheme, and importantly also, offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.
@@
 
-# 1. A review of trait-standard initiatives
+<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
+
+FDS: see Garnier et al. but may contain:
+
+-->
+
+Term | Definition 
+----- | -----
+Terminology |  
+Term |
+Thesaurus | 
+Ontology | 
+semantic web | 
+relational database | 
+two-dimensional data | 
+dataset | 
+Web service | 
+Online Portal |
+Repository | 
+Darwin Core Standard |
+Darwin Core Archive |
+Application Programming Interface (API) | 
+Data standard | 
+Uniform Resource Identifier (URI) |
+
+# 1. A review of initiatives for trait-data standardisation
 
 In this section, we review four kinds of initiatives that are of relevance for trait data standardisation:
-- The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
-- the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-- the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies; 
-- the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct relational 'ontologies' for an application in a 'semantic web' of information.
+
+* The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
+* the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
+* the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies and 'ontologies'; 
+* the fourth kind of initiative feed the needs of biodiversity informatics and develop relational database structures and provide platforms to relate trait data to a wider 'semantic web' of biodiversity data.
 
@@
 
-<!-- Fig 1 here? -->
-
 As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compilation of data into larger collections will be inefficient and that many published datasets will not be properly exploited.
@@
 In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09].
+@klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09].
 <!--[@wright04; ] -->
@@
 
-<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
-
-FDS: see Garnier et al. but may contain:
-
--->
-
-Term | Definition 
------ | -----
-Terminology |  
-Term |
-Thesaurus | 
-Ontology | 
-semantic web | 
-relational database | 
-two-dimensional data | 
-dataset | 
-Web service | 
-Online Portal |
-Repository | 
-Darwin Core Standard |
-Darwin Core Archive |
-Application Programming Interface (API) | 
-Data standard | 
-Uniform Resource Identifier (URI) |
-
 # 2. Proposing an ecological traitdata standard vocabulary
@@
 
-Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam14; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e.
+Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam16; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e.
 via the Application Programming Interface, API **in preparation!**).
@@
 If no globally available trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-Such a 'controlled vocabulary' would, in its simplest form, assign trait names with A) an unambiguous definition of the trait and B) an expected format  of measured values or reported facts (e.g.
+Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g.
 units or legit factor levels).
 
-By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context which should always be referenced in the core trait dataset.
+By providing a minimal vocabulary for trait thesauri (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset.
 The thesaurus might accompany the core data in a separate file (e.g.
-within a Darwin Core Archive), or be published on any other stable internet address (via URIs).
+within a Darwin Core Archive), or be published on any other stable internet address (ideally providing stable URIs).
 <!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
-See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].
+See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].
 We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
@@
 Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authorship, while also stating a bibliographic reference and license model.
+We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model.
 
-In the case of raw measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g.
-in a separate file or on the download website).
+In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g.
+in a separate metadata file, possibly consider applying the ecological metadata language, EML).
 In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be provided at the measurement level.
@@
 
-The data standard provides two extensions of the vocabulary that should be used to describe these information:
+The data standard provides three extensions of the vocabulary that should be used to describe these information:
 
@@
 
-Since the intention of the package is to facilitate upload of own data to hosting services but also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
+Since the intention of the package is also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been released in the public domain or under creative commons licenses.
 <!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
@@
 To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions.
-To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
+To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
+It also can be applied in the communication of data content between databases.
+In addition, we also offer a software tool to handle the extensive raw data resources on traits.
 
-The tools proposed here help to standardize trait datasets before upload to central as well as distributed data repositories.
+These tools are to facilitate research on trait variation along environmental gradients or intra-specific variation based on distributed data sources.
+This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
+
+The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories.
 In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns.
 In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
+We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis.
 
-We therefore encourage the development of trait thesauri in ontologies.
+We also encourage the development of trait thesauri in more advanced and interrelated ontologies.
 The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
 This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].
-
-Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources.
-Such tools to handle the extensive raw data resources on traits are a prerequisite for research on trait variation along environmental gradients or intra-specific variation.
-This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
-We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].
 
@@
 
-FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package.
+FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package 'traitdataform'.
 All authors contributed critically to the drafts and gave final approval for publication.
@@
 
-The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/
+The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+
+Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.
 
@@
 
-The living spreadsheet on existing trait datasets and databases is found at
-
-Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/ .
+The living spreadsheet on existing trait datasets and databases is found at https://goo.gl/QxzfHy.
 

### 2018-05-16 manuscript/manuscript.Rmd
revision of introduction and glossary

@@
 output:
-  word_document:
+  pdf_document:
     fig_caption: yes
-    fig_width: 6
+    fig_crop: no
+    keep_tex: yes
+    template: pretty.template
   html_document:
@@
     theme: united
-  pdf_document:
+  word_document:
     fig_caption: yes
-    fig_crop: no
-    keep_tex: yes
-    template: pretty.template
+    fig_width: 6
 bibliography: div05.bib
@@
 
-1.	The use of trait-based approaches is now widespread in all domains of ecological research, thanks to the great potential for such data to deliver general conclusions regarding the nature of biodiversity.
-Accordingly, a wealth of trait data is now available for many organism groups but due to lacking standardisation these data come in heterogeneous data formats.
+1.	Trait-based approaches are widespread throughout ecological research, due to the great potential for trait data to deliver general and mechanistic conclusions.
+Accordingly, a wealth of trait data is available for many organism groups but, due to a lack of standardisation, these data come in heterogeneous formats.
 2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
 3.
-In order to facilitate the standardisation and harmonization of distributed trait datasets, we propose a general and simple vocabulary and a two-dimensional data structure for storing and sharing ecological trait data.
+In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a general and simple vocabulary as well as a simple data structure for storing and sharing ecological trait data.
 4.
-We provide an R-package that enables the transformation of any data into the proposed form.
-This also allows to harmonize and merge trait datasets from heterogeneous sources, which facilitates the compilation of databses of particular research focus.
+Additionally, we provide an R-package that enables the transformation of any dataset into the proposed format.
+This also allows trait datasets from heterogeneous sources to be harmonized and merged, thus facilitating data compilation for any particular research focus.
 5.
-With these decentralized tools for trait-data harmonization, we intend to facilitate the exchange and analysis of trait data for ecological research and enable a global synthesis of traits across a wide range of organisms and ecosystems.
+With these decentralised tools for trait-data harmonisation, we intend to facilitate the exchange and analysis of trait data within ecological research and enable global syntheses of traits across a wide range of taxa and ecosystems.
 
@@
 
-Species traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
-Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
-Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
-For example, using traits is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
-
-With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].
+Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
+Because trait-based approaches to ecological and evolutionary research questions allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
+Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10].
+With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
+<!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
+Using traits is also a promising means of bypassing taxonomic impediment, i.e.
+the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
+This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].
 
@@
 the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
-Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13].
+Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [ @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13].
 These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
@@
 
-However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and taxonomic and regional coverage of trait data, both of present-day taxa ase well as in evolutionary deep-time.
-Researchers have a growing interest in assembling trait data from different sources since it pays off by getting answers to general questions.
-Integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used.
-All of these factors render trait data extremely heterogeneous.
+Many issues in trait-based research arise when compiling datasets from several sources.
+Data may differ in taxonmic nomenclature and resolution (e.g.
+reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the methodology applied in measurements.
+All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
+However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to acheive sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
 
 <!-- 3. trait databases have been developed, with standard structure and terms --> 
-To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions [e.g.
-TRY, @kattge11; PanTHERIA, @jones09; AmphiBio, @oliveira17].
-These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (in the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g.
-TRY, @kattge11].
-Furthermore, these portals map heterogeneous data into a common scheme, and importantly also, offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.
-To enable the compatibility of data across data-bases, semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
-Solving all these issues of data assemblage and harmonization in centralised infrastructures requires a huge effort in costs and personell (see below for a review of current initiatives).
+To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
+PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
+These databases are then published as static data tables (i.e.
+product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g.
+TRY, www.try-db.org).
+Furthermore, these portals map heterogeneous data into a common scheme, and importantly, also offer access control and data usage policies.
+As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers.
+Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
+These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
+<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
 
 <!-- 4. distributed trait data must be integrated --> 
-Many journals and research funding agencies today require data to be published without access restrictions [@swan12; @allison15].
-An increasing number of individual projects comply  with open data standards and publish their primary data on file hosting services like Figshare, Dryad or Researchgate, where no data standards are forced upon the uploaded material.
-The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.
-Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.
+While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly.
+This is due to the fact that open-science reaches the mainstream and a growing number of journals and research funding agencies require data to be published without access restrictions [@swan12; @allison15].
+As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
+<!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
+It is likely that data will become increasingly available, but a lack of data standardisation will hamper the efficient re-use and synthesis of published datasets.
 
-In this paper, we (1.) review existing trait databases and online portals, as well as initiatives for standardisation.
-In this review, we discuss the current practice and the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users.
-Based on these considerations, we (2.) propose a minimal structure and vocabulary for describing trait-datasets, compatible with existing terminology standards for biodiversity data.
-Finally, we (3.)  propose the development of software that assists in aplying a harmonisation of data from distributed sources.
+In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation.
+We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users.
+Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that is compatible with existing terminology standards for biodiversity data.
+Finally, we present an R package that assists the harmonisation of trait data from distributed sources.
 
-With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards an accessible trait data standardisation.
+With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.
 
@@
 Term | Definition 
------ | -----
-Terminology |  
-Term |
-Thesaurus | 
-Ontology | 
-semantic web | 
-relational database | 
-two-dimensional data | 
-dataset | 
-Web service | 
-Online Portal |
-Repository | 
-Darwin Core Standard |
-Darwin Core Archive |
+----- | ---------
+Data standard | 
+Terminology |  The  body of terms used with a particular technical application in a subject of study. In this case terms related to traits in ecology and evolution?  
+Thesaurus | Controlled  vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline  [@garnier17]
+Trait ontology | Controlled  vocabulary to describe each trait as a distinguishable feature, characteristic, quality or phenotypic feature of a developing or mature individual.
+Ontologies also link the terms to each other and when possible define their hierarchy.
+Semantic web | The concept of connecting data on the internet via well-defined machine-readable standards and exchange protocolls  [@berners-lee02]
+Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being harmonized across all data entries;  
+Two-dimensional data tables | a spread-sheet containing data organised in rows and columns; in most cases these datasets are considered 'static', i.e. they are not altered or filtered across time
+Database | a set of data compiled from multiple *datasets*; i.e.
+from multiple study contexts; may take the form of a two-dimensional spreadsheet, but mostly is organised in software tools into *relational databases*
+Relational database | Two or more *two-dimensional data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
+Identifier  (ID) | 
+Uniform Resource Identifier (URI) |
+Web service | An exchange protocoll to access online databases directly and programmatically, i.e. by calls from a software tool; 
 Application Programming Interface (API) | 
-Data standard | 
-Uniform Resource Identifier (URI) |
+Online Portal | A website designed as a platform for the exchange of information, e.g.
+trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
+Repository | Archive for data files; online repositories make data available for public access, provide metadata and (not always) facilitate citations via DOIs (Digital Object Identifiers)
+Metadata | 
+Darwin Core Standard (Dwc) | Body  of standards containing a glossary of terms intended to facilitate the sharing of information about biological diversity by providing reference definitions, examples, and commentaries.
+(From http://rs.tdwg.org/dwc/)
+Darwin Core Archive (DwC-A) | A file archive (.zip) that contains structured *metadata* and primary data in the form of *two-dimensional data tables* combined into a *relational database* via *identifier* columns.
 
-# 1. A review of initiatives for trait-data standardisation
+# A review of initiatives for trait-data standardisation
 
-In this section, we review four kinds of initiatives that are of relevance for trait data standardisation:
+In this section, we review four types of initiatives that are of relevance for trait data standardisation:
 
-* The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
-* the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-* the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies and 'ontologies'; 
-* the fourth kind of initiative feed the needs of biodiversity informatics and develop relational database structures and provide platforms to relate trait data to a wider 'semantic web' of biodiversity data.
+1. Projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
+2. Initiatives that aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
+3. Initiatives that aim at the standardization and development of consensus methods and definitions for traits and provides standard terminologies and 'ontologies'; 
+4. Initiative that feed the needs of biodiversity informatics and develop relational database structures and provide platforms to relate trait data to a wider 'semantic web' of biodiversity data.
 
-We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
-We show how the degree of trait-data standardization in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources.
-The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in the appropriate form.
+We discuss these initiatives separately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
+We show how the degree of trait-data standardisation in existing datasets spans  this entire spectrum and which tools and standards do exist to achieve harmonisation of data from multiple, distributed sources.
+The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in an appropriate form.
 

### 2018-05-17 manuscript/manuscript.Rmd
update MS review parts 1-3; update glossary

@@
 While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly.
-This is due to the fact that open-science reaches the mainstream and a growing number of journals and research funding agencies require data to be published without access restrictions [@swan12; @allison15].
+This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]., which supports.
 As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
@@
 -->
+\scriptsize
 
@@
 ----- | ---------
-Data standard | 
-Terminology |  The  body of terms used with a particular technical application in a subject of study. In this case terms related to traits in ecology and evolution?  
-Thesaurus | Controlled  vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline  [@garnier17]
-Trait ontology | Controlled  vocabulary to describe each trait as a distinguishable feature, characteristic, quality or phenotypic feature of a developing or mature individual.
-Ontologies also link the terms to each other and when possible define their hierarchy.
-Semantic web | The concept of connecting data on the internet via well-defined machine-readable standards and exchange protocolls  [@berners-lee02]
+Term | a word that describes a particular *concept* as part of the specialised vocabulary of a field
+Concept | An idea, notion or object that is made explicit in a communication context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
+Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
+trait data), that is recognized by a large proportion of members of the application context
+Terminology | The  body of *terms* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
+Thesaurus | Controlled vocabulary that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte12] 
+Ontology | Controlled vocabulary that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
+defines a hierarchy of terms; thus a formal model of a the objects and their relationships in a domain of interest [@gruber95]
+Semantic web | The concept of connecting data on the internet via well-defined standards and exchange protocols [@berners-lee02]; the World Wide Web Consortium (W3C) defines standards, i.e.
+specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
 Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being harmonized across all data entries;  
-Two-dimensional data tables | a spread-sheet containing data organised in rows and columns; in most cases these datasets are considered 'static', i.e. they are not altered or filtered across time
-Database | a set of data compiled from multiple *datasets*; i.e.
-from multiple study contexts; may take the form of a two-dimensional spreadsheet, but mostly is organised in software tools into *relational databases*
-Relational database | Two or more *two-dimensional data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Identifier  (ID) | 
+Data table | a two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
+Database | a suite of data compiled from multiple *datasets*, i.e.
+from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software tools;
+Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
+Online database | A relational database that is made accessible on the internet offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
+Identifier (ID) | A unique label that relates entries within and across datasets; is used to link data tables into a relational database; can be user-specific or as a *URI* be part of a globally valid ontology or thesaurus
 Uniform Resource Identifier (URI) |
@@
 Repository | Archive for data files; online repositories make data available for public access, provide metadata and (not always) facilitate citations via DOIs (Digital Object Identifiers)
-Metadata | 
+Metadata | Data documentation representing the higher level information or instructions that describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
 Darwin Core Standard (Dwc) | Body  of standards containing a glossary of terms intended to facilitate the sharing of information about biological diversity by providing reference definitions, examples, and commentaries.
@@
 Darwin Core Archive (DwC-A) | A file archive (.zip) that contains structured *metadata* and primary data in the form of *two-dimensional data tables* combined into a *relational database* via *identifier* columns.
+Occurrence | 
+Measurement | 
+Aggregated Value |
+Value | 
+Fact | 
+Method handbook | 
+File hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; popular examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org;
+
+Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier16)
+
+\normalsize
 
@@
 
-1. Projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
-2. Initiatives that aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-3. Initiatives that aim at the standardization and development of consensus methods and definitions for traits and provides standard terminologies and 'ontologies'; 
-4. Initiative that feed the needs of biodiversity informatics and develop relational database structures and provide platforms to relate trait data to a wider 'semantic web' of biodiversity data.
+1. Initiatives that provide *trait datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature. 
+2. Initiatives that aim to harmonise trait data from the literature or from direct measurement into *trait databases* and make those data widely available. 
+3.
+Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits, and provides standard *terminologies* in the form of *thesauri* and *ontologies*.
+4. Initiatives that aim to leverage *relational database* structures and *semantic web* technology to link trait data to a wider set of biodiversity data.
 
 We discuss these initiatives separately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
-We show how the degree of trait-data standardisation in existing datasets spans  this entire spectrum and which tools and standards do exist to achieve harmonisation of data from multiple, distributed sources.
-The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in an appropriate form.
+We show how the degree of trait-data standardisation in existing datasets spans this entire spectrum and which tools and standards are applied to achieve harmonisation of data from multiple, distributed sources.
+The objective of this review is to raise awareness for the generic structure of trait data and aid researchers to share and publish own datasets in an appropriate form.
 
@@
 
-In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books.
-With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15;  @renner14; @vanderplas17; @price13;  @ricklefs17].
+In the field of comparative biology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been published in regular journal articles or books.
+With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
+A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
 
-Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net, Zenodo.org. 
-Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, which supports the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
+Today, the online publication of such data is greatly facilitated by file hosting services (e.g. Figshare.com). 
+Furthermore, these services warrant long-term accessibility, offer DOIs for referencing, and support Public Domain dedication or Creative Commons licenses.
 However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
 .xlsx, .docx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.
+.pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
 
-Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management.
-As a result of this, datasets also greatly vary in structure. 
-Data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/), offer means to meet the high scientific criteria of data management.
+Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds for *metadata* documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management.
+As a result, datasets greatly vary in structure and terminology.
 
-For trait-data, there are specific issues arising from the variability of data structure. 
-For instance, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
-Datasets vary in the detail provided in the metadata, e.g.
+For trait-data, there are typical issues arising from the variability of data structures. 
+For instance, the column descriptions andterminology applied to taxa and traits are project specific, and rarely chosen to allow translation into larger database initiatives.
+Furthermore, metadata varies in its detail, e.g.
 for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
-If factual trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated quantitative measurements averaged from raw data, they usually are reported in a species$\times$trait wide-table format.
-Here, each row contains a species (or taxon) for which multiple traits are reported in columns.
-Similarly, when reporting raw data, researchers place observations of individual specimens in rows with multiple trait measurements applied to the same specimen placed in columns. 
-The variability of the number and meaning of columns in these tables requires manual adjustments when merging multiple datasets.
+In terms of structure, trait data usually are reported in a species$\times$trait wide-table format.
+In this format, each row contains a species (or taxon) for which multiple traits are reported in columns.
+Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
+Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].
 
-As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compilation of data into larger collections will be inefficient and that many published datasets will not be properly exploited.
+<!--To remedy these problems, data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/) or GFBio [the German Federation on for Biological Data; gfbio.org; @diepenbroek14], offer means for small projects to meet the high scientific criteria of data management.-->
 
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy; Appendix A) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list).
-We invite data owners and users to register further datasets qualifying as trait data to this spreadsheet.
+Therefore, in an attempt to collate a list of existing, distributed and unstandardised datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list).
+We invite data owners and users to add further trait datasets to this spreadsheet.
+
+As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.
 
@@
 <!-- community database backend, but static publication --> 
-In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09].
-<!--[@wright04; ] -->
-While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
+In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g.
+@klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09].<!--[@wright04; ] -->
+While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
 -->
-Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
-Just as the individual datasets described above, many of these databases are published as static datasets on low-threshold file hosting platforms and are updated irregularly.
+Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
+Just as the individual datasets described above, many of these databases are published as static data tables on low-threshold file hosting platforms and are updated irregularly.
 
-Dealing with much larger amounts of data, initiatives that form around museal collection data are more concerned with data standardization.
-Aiming for a mobilization of measurements from collections, the  VertNet database compiles and harmonized large quantities of vertebrate traitdata [@guralnick16].
-The resulting data are published as static datasets which are updated as new data sources are integrated.
+As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
+With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowdsourcing (e.g.
+www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
+For example, the VertNet database compiled and harmonized large quantities of vertebrate traitdata with the aim of mobilising measurements from collections [@guralnick16].
+The resulting data are published as versioned data tables which are updated as new data sources are integrated.
 
 <!-- database backend, selective download, user access management --> 
-More specialised trait-database platforms have been created that cover certain trait types (e.g.
-floral traits, seed traits or wood density traits), interaction types (e.g.
-pollination traits or feeding relationships), or environmental and experimental context of the trait observation (e.g.
+More specialised trait-database platforms have been created to cover certain trait types (e.g.
+floral traits, seed traits, root traits or wood density traits), interaction types (e.g.
+pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g.
 location or climatic data).
-These centralized database initiatives attract data submissions from a defined research field and take care of the harmonization process and greatly facilitate data synthesis. 
+Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
 <!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
-By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform.
-The online platform enables selective data download and user permission and rights management.
+For example, by aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other trait data platform.
+The *online database* enables selective data download and user permission and rights management.
 As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
@@
 the KEGG orthology, @kanehisa12].
-On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
-The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans, make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
+For animals, a single unified platform and harmonizing scheme for animal trait data is still lacking.
+The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17].
+Nonetheless, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
 
-Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication.
-The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
-With the pro-active turn towards open access data (as declared inthe Bouchot Declaration; http://www.bouchoutdeclaration.org/) it may be necessary to find other incentives for data submission.
+Regarding open access, few of these centralised databases comply with the criteria demanded by journals and funding agencies for primary data publication.
+The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship.
+With the proactive turn towards open access data (as stated in the Bouchot Declaration; http://www.bouchoutdeclaration.org/), it may be necessary to find other incentives for data submission.
 
@@
 
-A major problem in trait-data standardisation is the reference to accepted and unambiguous trait definitions. This is acheived via thesauri and ontologies. 
-In database management, a thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
-Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data.
-Providing relational ontologies in a machine readable form makes them compliant with semantic web standards, thus supporting the goal of widely interlinked data structures for the world wide web [@berners-lee01; @page08].
+A major challenge in trait-data standardisation is the lack of widely accepted and unambiguous trait definitions.
+Previous standard definitions of trait *concepts* range from listings of selected definitions in *glossaries*, over well-defined methodological *handbooks* and and comprehensive *thesauri*, to relational definitions of trait concepts in *ontologies*.
+While glossaries may be seen as specific for a study context, the initiatives behind method handbooks, thesauri and ontologies are primarily concerned with consensus building on trait definitions in a wider community.
+
+Assigning an unambiguous methodological protocol to a trait is key for standardising the physical process of measuring.
+Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].
+Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
+
+A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
+Expanding on this, ontologies link the defined terms to each other by formally defining the relationships between them, with the objective of enabling a computational interpretation of data. 
+Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)* within own datasets.
+For example, a measurement of seed size could be linked to the Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
+Ontologies define terms based on other well-defined terms from published ontologies.
+The Planteome Trait Ontology definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
+This way, each trait definition may link to a broader or narrower term.
+For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
+Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies.
+By providing this interlinkage of trait ontologies, a web of definitions is spun across the Internet which allows researchers and search engines to relate independent trait measurements with each other and connect it to the wider *semantic web* of online data [@berners-lee01; @page08].
+The distinction of thesauri and ontologies is not truely binary.
+Rather they mark idealised ends of a spectrum.
+While thesauri may contain defined relations between terms within the standard, ontologies relate most terms to other defined concepts, and also link those to other standards.
+
+Comprehensive trait thesauri have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
+Ontologies of trait definitions have been developed for plants [e.g.
+the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
+the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13].
+The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control.
 
@@
 
-Standard trait definitions have been provided in all shades along this continuum, ranging from listings of definitions in glossaries, over well-defined thesauri to relational ontologies.
-The initiatives behind these are first and foremost concerned with consensus building on trait definitions.
-Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring.
-Efforts to develop handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but are of limited effect when harmonizing trait data that predate or ignore this standard [@kattge11a].
-Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
-Thesauri of trait definitions have been developed for plants [e.g.
-the Planteome ontology for crop plants, @walls12; @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
-the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13].
-Comprehensive trait ontologies, including hierarchies of parent terms, have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and on the side of animal traits, in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
-
-These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. 
-For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
+Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for the collaborative development [@walls12]. 
+For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
 <!-- Mention also NCBO BioPortal?
 https://bioportal.bioontology.org/ -->
-These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets.
-For example, a measurement of seed size would be linked to Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
 
-To be unambiguous, any thesaurus should be defining terms based on other well-defined terms from semantic ontologies.
-The definition provided here contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
-This way, each trait definition may link to a broader or narrower term.
-For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies.
-By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.
-
-To conclude, there is a shortage of globally valid trait thesauri or ontologies for traits that emerged from a standardisation effort of methodologies and broad consensus process within scientific field of ecology *and* that are sufficiently interlinked and accessible to fulfil semantic web standards.
-The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
-To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
-Such thesauri would advance the possibilities of aggregated datasets and reduce noise and uncertainty. 
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from distributed sources.
+To conclude, there is already a suite of globally available thesauri and ontologies for traits that emerged from standardisation efforts of methodologies and community consensus processes.
+Albeit definitions in some domains are better covered than others.
+Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
+Most importantly, while these defined vocabularies are widely used in biodiversity data management, distributed data repositories of smaller project contexts hardly make use of them. 
+A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
+To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
+For example the accessibility of ontologies will increase if advanced webservices, and particularly open Application Programming Interfaces (APIs), are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+Software then can assist researchers in linking own data to globally defined concepts.
 
@@
 
-One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
+One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for static datasets for which a two-dimensional data table might suffice.
 If the traits are assembled for a specific research question with few columns of ancillary information, it may be safer to add this information as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.

### 2018-05-18 manuscript/manuscript.Rmd
finalize ms

@@
 \footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
-\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+\footnotetext[7]{Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
 
@@
 Concept | An idea, notion or object that is made explicit in a communication context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
+Controlled vocabulary | 
 Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
@@
 www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
-For example, the VertNet database compiled and harmonized large quantities of vertebrate traitdata with the aim of mobilising measurements from collections [@guralnick16].
+For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16].
 The resulting data are published as versioned data tables which are updated as new data sources are integrated.
@@
 
-While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g.
+While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g.
 for taxa.
-To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.
+To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.
 
-Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
-More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
+Trait data have been defined by @garnier17 to follow an entity-quality model (EQ), where a trait observation is 'an entity having a quality'. 
+More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e.
 trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
@@
 a species or higher-level taxon).
-Expaning on this definition, quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
+We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
 1*a*), by aggregating replicated measurements on multiple entities (Fig.
-1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g.
+1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g.
 databases, Fig.
@@
 The entity or observation (i.e.
-the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
+the *occurrence*) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
 a genus).
@@
 
-These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain trait values, units, taxon names, variables such as sex or life stage, as well as ancillary information of time and date of observation, as well as methodological details.
-The terminologies provided by these standards are quite universal and cover most use cases of trait data.
-<!-- TODO FDS: dimensional and star-shema databases, relational databases Kattge et al 2011, Kimball & Ross 2002-->
+These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies these terms.
+These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
+The terminologies provided by these standards are quite universal and even cover most use cases of trait data.
+An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12].
 <!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
@@
 
-Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for databases that covers most potential use cases of trait-based ecology.
-This data structure is build around a central data table that contains observations, i.e.
-a single event of measurements on the same individual plant specimen at the same point in time.
-This structure puts emphasis on the fact that trait  data are often measured for an analysis of correlations of multiple traits.
-Identifiers link multiple measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables.
+Specifically designed for plant traits, @kattge11 proposed a generic database structure that covers most potential use cases of trait-based ecology.
+This data structure is built around a central data table that contains observations, i.e.
+a single event of measurement on the same individual plant specimen at the same point in time.
+This structure emphasises the fact that multiple trait data are measured on the same individual organisms and used to analyse correlations between these multiple traits.
+Identifiers link the measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables.
 The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
-This structure can be implemented in any software or programming language for relational databases.
-<!-- FDS: enter comments of Jens Kattge here!-->
+This structure can be implemented in any relational database management system.
 
-In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
-The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
-The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review.
+In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach to providing an integrated structure for trait data.
+The framework employs established terms provided by the DwC, relates trait definitions to trait ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
 Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
@@
 The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11]. 
+The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review [@franz16].
+Initiatives that aim at providing a stable reference for taxa are for instance the EOL Catalogue of Life [http://www.catalogueoflife.org/, @roskov18], the GBIF Backbone Taxonomy [@gbifsecretariat17], or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).
 
-These proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
-Hovewer, hardly any of the aforementioned aggregated and harmonized datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards.
+These proposed standards are responses to a demand from biodiversity data managers for more structured input from the research community. 
+Hovewer, hardly any of the aforementioned trait datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
+One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
+In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
+<!--If the traits are assembled for a specific research question with just few ancillary co-factors, it may be safer to append these as columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
+This way, if datasets of different sources are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
+-->
+The other reason is lack of awareness for the need for trait data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
 
-One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for static datasets for which a two-dimensional data table might suffice.
-If the traits are assembled for a specific research question with few columns of ancillary information, it may be safer to add this information as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
-This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
-
-The other reason is lack of awareness of data providers for trait data standardization: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.
-Investing time and energy into research for the best and most sustainable data structure is not affordable for small and intermediate-size research projects.
-
-By filling this gap, data brokering services [e.g.
-the German Federation on for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
-BExIS, NEON Dataportal; Diversity Workbench, diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
-<!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
+By filling this gap, data-brokering services [e.g.
+the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
+BExIS, http://bexis2.uni-jena.de/; Diversity Workbench, www.diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance.
+These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
+One goal of such initiatives is to facilitate data publications and standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+<!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->
 
@@
 
-Initiatives for standardization (e.g.
+Initiatives for standardisation (e.g.
 ontologies and data standards) and platforms for data management (e.g.
 database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.
-With low-threshold offers for the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, the concern is with data harmonization of these distributed data sources.
-
-The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardization on the side of data providers. 
-We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in data visibility and interconnectedness of the published data.
+Low-threshold repositories offer the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, but data harmonisation of these distributed data sets is currently laborious.
+The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
+We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in the visibility and interconnectedness of published data.
 
@@
 
-# 2. Proposing an ecological traitdata standard vocabulary
+# An ecological trait-data standard vocabulary
 
-As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research.
-Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.
-
-By using this standard vocabulary, authors can ensure that the description of tables of trait measurements that are uploaded to distributed data repositories like Figshare, Zenodo etc.
-will be unambiguous and generally applicable.
-It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure accessibility in the long term.
-
-The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.
+As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
+The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
+Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.
+By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable.
+It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.
 
@@
 We paid particular consideration to the work of @kattge11, @kattge11a, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
-Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.
-
-However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
-Also, at present we conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model.
-However, such network data may easily be combined with trait data via the taxon identifiers or by using other Extensions of the Darwin Core (DwC).
-Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
-
+Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.
+Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. 
+The vocabulary proposed is intended to form the foundations of a standard nomenclature that can be expanded and corrected by the wider community of researchers using trait-based approaches in ecology.
+<!--
 ## Why use the vocabulary
@@
 The incentives for using a unified vocabulary for trait data might not be obvious for data providers, who are mainly concerned with a particular study site, organism group and research question.
-We won't discuss here in depth how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis.
-It appears to be broad consensus that an open biodiversity science is crucial for an evidence based decision making and conservation policy on regional and global scales.
+We won't discuss here how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis.
+However, it appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.
 
@@
 By using a constrained vocabulary with globally accessible definitions of terms, data can be accessed more easily by other researchers and datasets from different sources can be combined into aggregated datasets.
-Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary.
-
-It is implicit to this vocabulary to use a two-dimensional observation long-table format for the data, rather than a species-trait matrix (where some of the columns would be trait names which are not included in the vocabulary, but are part of trait thesauri).
-As this format draws from a defined set of columns,  merging of datasets is  easier.
-Long-table datasets also purport multiple advantages for data manipulation [e.g.
-filtering, sub-setting and aggregating data; @wickham14].
-
-Using well-defined identifiers ('IDs') to structure the dataset offers compatibility with the input formats for larger databases that may offer better visibility and impact of the data contribution. 
-Identifier fields are key elements to structure the two-dimensional dataset. 
-For instance, for occurrence level data, were multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link these measurements across the rows of the dataset.
-Similarly, multivariate measurements, like for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
-If literature data are provided, data are probably not structured below the taxon level and a 'taxonID' is the key identifier.
-A 'datasetID' allows to report datasets that have been compiled from multiple sources.  
-Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external ontology services (see above for resources).
-
-However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
-For datasets containing multi-layered information on observations, traits, taxa and environmental context, we  recommend to consider relational datatabase structures composed of multiple spreadsheets linked via identifier columns, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
-The trade-off is data-consistency and readability in a single table *vs.* avoidance of content duplication and redundancy in a relational database.
-The standard vocabulary proposed here may still be applied within these structured databases.
+Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary. -->
 
 ## How to apply the standard vocabulary
+
+We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible. 
+The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17] each entry describes a trait value (i.e.
+quality) observed on an an individual or population (i.e.
+entity), of a biological taxon.
+It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data, rather than a species$\times$trait matrix.
+As the long-table format draws from a defined set of columns, the merging of datasets is  easier.
+Long-table datasets also purport multiple advantages for data manipulation [e.g.
+filtering, sub-setting and aggregating data, @wickham14; Fig.
+2].
+
+Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary.
+For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset.
+Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
+In literature data, summarized traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
+In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
+Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external terminology services (see above for resources).
+Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
+As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16.
+The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
+The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.
 
@@
 
-Column names and, where possible, field entries of a trait datasets published on online repositories should draw from the defined vocabulary.
-
-The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
-trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
-traits) observed on a biological entity (i.e.
-an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e.
-a species or higher-level taxon).
-
-Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below).
-
-Beyond the core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information).
-These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-For this case, we offer three extensions of the core vocabulary ("taxon", "measurement or fact", and "occurrence") that expand and refine terms of the Darwin Core Extensions (see below).
-This information may simply be added as extra columns to the core dataset.
-
-For reasons of long-term accessibility, data should not be uploaded in spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
+For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
 UTF-8 or ASCII).
@@
 In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-
-Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam16; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e.
-via the Application Programming Interface, API **in preparation!**).
+Wherever referring to individual terms of the vocabulary in publications or metadata,this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
+via the API; **in preparation!**).
 Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
 
-Besides this use in published datasets, the standard vocabulary could be imposed in webservices, e.g.
-APIs that provide direct access to online databases or download tools that pull subsets of data from larger databases.
+## Terms of the standard vocabulary
 
-### Specifications of core terms
+The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.
+The core terms describe minimal trait data according to the Entity-Quality model. 
+Beyond these core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information).
+These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
+For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence") that expand and refine terms of the Darwin Core Extensions (see below).
+This information may simply be added as extra columns to the core dataset.
+Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below). 
+The scope of the vocabulary may not yet cover all aspects morphological and evolutionary perspectives. 
+Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other Extensions of the Darwin Core.
+Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
+
+### Specification of core terms
 
@@
 3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
-For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
-However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.
-
+For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
+However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
 <!--### Relate to global taxon and trait definitions -->
-
-In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up tables of traits. 
-This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively).
+In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
+This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively).
 For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.
 By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
-This redundancy of data allows for continuity on the authors and data owners side while also ensuring quality checks and comparability for the data user.
+This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.
 
-<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
+### Extensions for additional data layers
+
+Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
+The data standard provides three extensions of the vocabulary that should be used to describe this information:
+
+- The `Taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.
+- The `MeasurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
+variation in measurement method) or aggregated values (e.g.
+statistical method applied), as well as the source of reported facts (e.g.
+literature source or expert reference).
+- The `Occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
+This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
+
+Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
+<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).
+Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
+This information can then be used for trait inference at higher taxonomic resolution.
+<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
+-->
+These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets (published separately or as part of a Darwin Core Archive), thus avoiding redundancy and duplication of content.
+A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken.
 
@@
 
-If no globally available trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
+Wherever possible, the colum `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
+If no globally available trait definition exists as a reference, trait-datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits.
 Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g.
 units or legit factor levels).
-
-By providing a minimal vocabulary for trait thesauri (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset.
+By providing a minimal vocabulary for trait thesauri (see https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset.
 The thesaurus might accompany the core data in a separate file (e.g.
@@
 <!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
-See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].
-We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
+See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards (https://ecologicaltraitdata.github.io/ETS/thesauri.html).
+We also provide an example thesaurus on Github (http://ecologicaltraitdata.github.io/TraitDataList/), which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project [@fischer10].
 
@@
 We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model.
-
 In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g.
-in a separate metadata file, possibly consider applying the ecological metadata language, EML).
-In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be provided at the measurement level.
+in a separate metadata file, possibly applying the ecological metadata language, EML).
+In cases where individual data from different sources are compiled into a trait database, these information must be provided at the measurement level.
 This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
@@
 -->
-
 <!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
@@
 
-### Extensions for additional data layers
-
-Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-
-The data standard provides three extensions of the vocabulary that should be used to describe these information:
-
-- the `Taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides further terms for specifying the taxonomic resolution of the observation and to ensure correct reference in case of synonyms and homonyms.
-- The `Occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
-This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
-- The `MeasurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
-It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
-variation in measurement method) or aggregated values (e.g.
-statistical method applied), as well as the source of reported facts (e.g.
-literature source or expert reference).
-
-Many terms of these extensions refine or copy terms of the Darwin Core Standard (Dwc) and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
-
-<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).
-Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
-This information can then be used for trait inference at higher taxonomic resolution.
-<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
--->
-
-As for metadata information, these additional layers of information can either be added as extra columns to the core dataset or be kept in separate data sheets (published separately or as part of a Darwin Core Archive), this avoiding redundancy and duplication of content.
-A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
-Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
-a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
-
-# 3. Computational tools for producing compliant data
+# Computational tools for producing compliant data
 
@@
 The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
-However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
-
-Accordingly, we started to develop the R package 'traitdataform', which assists the production of data compliant with the trait data standard proposed here.
+However, none of these packages provide the option to harmonization trait data into a unified scheme.
+To close this gap, we developed the R package 'traitdataform', which assists the production of data compliant with the trait data standard proposed above. 
 There are two major use cases for the package:
 
-1. preparing own trait datasets for publication on public file-servers and project databases, and
+1. preparing trait datasets for publication on public hosting services and project databases, and
 2. automating the harmonisation of trait datasets from different sources by moulding them into a unified format.
 
-A comprehensive documentation of the package is found on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
+A comprehensive documentation of the package can be found on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
 The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
 
-<!-- The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
-2).
+The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
+3).
 This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
+\caption{Process chart of the functions provided within the R package 'traitdataform' to apply the standard vocabulary to any trait-data table.}
+
+\end{figure}
+
+The parameter 'traits' lists column names that contain trait values.
 The column containing taxon names is given in parameter 'taxa'.
@@
 The parameter 'units' is used to specify the input units of measurement.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
-\caption{Process chart of the functions provided within the R package.}
-
-\end{figure}
-
 In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data.
@@
 FDS: might be referred to in the section about thesauri -->
-<!--
+
 The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
@@
 Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies.
-
 The package functions form a toolchain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
+A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing trait data as simple as:
 
@@
 Datasets that have been produced by these functions can easily be appended using the function `rbind()` of R base, while maintaining any available metadata information as separate column entries. 
+To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identifier using the function `merge()`.
+This enables an easy handling of data sources that originate in a relational database format.
 
-To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identifier using the function `merge()`.
-This enables an easy handling of data sources that have been stored in a Darwin Core Archive format.
-
--->
-
-Since the intention of the package is also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been released in the public domain or under creative commons licenses.
+Since the intention of the package is also to simplify the harmonisation of published trait data, the package offers direct access to trait datasets that have been released in the Public Domain or under Creative Commons licenses.
 <!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
 
-# Discussion
+# Conclusion
 
@@
 More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.-->
-To serve the demand for simple ways to standardise and harmonize ecological trait data, we propose a versatile vocabulary for simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
-We also present a toolbox in R to transfer and harmonize data into this scheme.
+To serve the demand for simple ways to standardise and harmonise ecological trait data, we propose a versatile vocabulary for simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
+With the R-package 'traitdataform', we also present a toolbox in R to transfer and harmonise data into this scheme.
+
+It appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.
+In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns.
+In terms of data science, machine-readable, ontology-based data ease the application of big-data mining and machine-learning techniques.
 
 To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions. 
-To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
-It also can be applied in the communication of data content between databases.
-In addition, we also offer a software tool to handle the extensive raw data resources on traits.
-
-These tools are to facilitate research on trait variation along environmental gradients or intra-specific variation based on distributed data sources.
-This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
-
+However, these distributed data are heterogeneous in form and description and initiatives to harmonize and compile these data require significant amounts of funding and personel. 
+To support the long-term rewards of standardization efforts, incentives should be sought to mitigate the cost of readying trait data for the 'semantic web' of biodiversity data and knowledge.
+This can be software tools or supporting infrastructures.
 The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories. 
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns.
-In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
-We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis.
-
-We also encourage the development of trait thesauri in more advanced and interrelated ontologies.
+By using a constrained vocabulary with globally accessible definitions of terms,  distributed trait data can be accessed more easily by other researchers and harmonised into aggregated datasets.
+Also, it will ease the exchange of data between databases and facilitate the development computational methods and software tools that access and handle the data, based on the standard vocabulary. 
+We also encourage the advancement of trait thesauri into more interrelated and complete ontologies. 
 The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
-This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].
+This requires significant effort, but it returns  great scientific benefit by enabling synthesis on our general understanding of biodiversity and ecosystem function.
 
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal and Jens Kattge for comments on the manuscript, the trait data standard vocabulary and the pre-releases of the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts, the trait data standard vocabulary and the pre-releases of the R-package.
 
@@
 infrastructure;--> Christiane Fischer and Jule Mangels for giving
-support through the central office, <!--Michael Owonibi for
-managing the central data base,--> and <!--Markus Fischer,--> Eduard Linsenmair,
+support through the central office, Jens Niesschulze for establishing the central data base, and Markus Fischer, Eduard Linsenmair,
 Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
@@
 
-FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package 'traitdataform'.
-All authors contributed critically to the drafts and gave final approval for publication.
+FDS, MJ, GP, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard; NKS authored the example thesaurus of traits; FDS developed the R package 'traitdataform'.
+All authors contributed critically to the structure and content of the manuscript and gave final approval for publication.
 
@@
 
-The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
 

### 2018-05-18 manuscript/appendix_A.Rmd
add Appendix A: Rmarkdown file transfer of google spreadsheet to pdf

--- 2018-05-16 manuscript/manuscript.Rmd
@@
 ---
-title: "Towards an Ecological Trait-data Standard"
+title: "Towards an Ecological Trait-data Standard - Appendix A"
 author:
@@
 - name: Nadja K. Simons \footnotemark[7]
-csl: amnat.csl
 fontsize: 12pt
-geometry: margin=1.3in
+geometry: margin=0.5in
+classoption: landscape
 output:
@@
     fig_width: 6
-bibliography: div05.bib
 ---
@@
 \footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
-\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+\footnotetext[7]{Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
 
-\linenumbers
+\newpage
 
-# Abstract
+```{r, echo = FALSE, warning = FALSE}
+options(knitr.table.format = "latex")
+library(kableExtra)
+gg <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTfQOIHjJv5x_YHgq14DR7VLTvycfsl2Njj2Ip4p2a5kca8AX5t5cmaial6Bg-MX9pBDOJaJfnlR7Qj/pub?gid=0&single=true&output=csv", encoding = "UTF-8")
+dt <- subset(gg, select = c(datasetName, type, DOI, NumberTaxa, NumberTraits, Responsible, CiteAs, TermsOfUse))
 
-1.	Trait-based approaches are widespread throughout ecological research, due to the great potential for trait data to deliver general and mechanistic conclusions.
-Accordingly, a wealth of trait data is available for many organism groups but, due to a lack of standardisation, these data come in heterogeneous formats.
-2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
-3.
-In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a general and simple vocabulary as well as a simple data structure for storing and sharing ecological trait data.
-4.
-Additionally, we provide an R-package that enables the transformation of any dataset into the proposed format.
-This also allows trait datasets from heterogeneous sources to be harmonized and merged, thus facilitating data compilation for any particular research focus.
-5.
-With these decentralised tools for trait-data harmonisation, we intend to facilitate the exchange and analysis of trait data within ecological research and enable global syntheses of traits across a wide range of taxa and ecosystems.
-
-# Key-words:
-
-functional ecology, species traits, semantic web, ontologies, data standard
-
-# Introduction
-
-<!-- CP:  I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
-
-That’s the structure I try to follow:
-1.
-Traits are important in ecology, evolution and conservation (i.e.
-It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
-
-2.
-To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources.
-When we do this we can answer new questions and find generalities.
-So there is a growing interest in assembling trait data from different sources and taxa (e.g.
-give the number of papers using TRY).
-
-3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.
-
-4.
-One solution are databases.
-But these are costly and depend on the will of few persons/funding.
-They also cover few taxonomic groups.
-Matching between different databases is still difficult because of different definitions and structures.
-Most data are still uploaded separately.
-We need a solution to make their use easy. 
--->
-<!-- GLP: 
- Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g.
-TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals).
-Say also something on the development of general protocols for measuring traits (e.g.
-Moretti et al.
-2016; Freschet et al.
-2017) ?
-
-Importance of data standardization.
-This is the main idea here and so I wonder if some idea of the first point (Need for data standardization) should not be merge here with the last paragraph of the introduction ?
-As there is also some repetitions between those 2 parts.
-
-I wonder if you should speak about ‘Functional traits’here or just Traits ?
-As you are not only considered Functional traits in this paper (e.g.
-life history traits) ?
-Maybe remove the definition no ?
--->
-<!-- PM: Thuis is quite complex- e.g.
-‘taxonomic impediment’etc.
-I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples.
-I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.
-
-Huge number of references here- I think you can strip it down to a few key ones, and reviews.
-Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site.
--->
-
-<!-- 1. Traits are important in ecology, evolution and conservation -->
-
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
-Because trait-based approaches to ecological and evolutionary research questions allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
-Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10].
-With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
-<!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
-Using traits is also a promising means of bypassing taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
-This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].
-
-<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04].
-Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
-Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [ @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13].
-These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].
-<!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ?
-Put instead:
-Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
-And Coux et al. 2016 Linking species functional roles to their network roles
-?
--->
-<!--
-Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g.
-@reich97].
-<!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ?
-I’m not sure that the literature cited is always appropriate here ?
--->
-<!--
-Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17].
---> 
-<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions.
-Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->
-
-<!-- 2. assemble trait data into databases -->
-
-Many issues in trait-based research arise when compiling datasets from several sources.
-Data may differ in taxonmic nomenclature and resolution (e.g.
-reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the methodology applied in measurements.
-All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
-However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to acheive sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
-
-<!-- 3. trait databases have been developed, with standard structure and terms --> 
-To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
-PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
-These databases are then published as static data tables (i.e.
-product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g.
-TRY, www.try-db.org).
-Furthermore, these portals map heterogeneous data into a common scheme, and importantly, also offer access control and data usage policies.
-As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers.
-Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
-These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
-<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
-
-<!-- 4. distributed trait data must be integrated --> 
-While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly.
-This is due to the fact that open-science reaches the mainstream and a growing number of journals and research funding agencies require data to be published without access restrictions [@swan12; @allison15].
-As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
-<!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
-It is likely that data will become increasingly available, but a lack of data standardisation will hamper the efficient re-use and synthesis of published datasets.
-
-In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation.
-We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users.
-Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that is compatible with existing terminology standards for biodiversity data.
-Finally, we present an R package that assists the harmonisation of trait data from distributed sources.
-
-With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.
-
-<!-- FDS: Why 'ecological' trait data?
-what is special about this subject?
-Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->
-
-<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
-
-FDS: see Garnier et al. but may contain:
-
--->
-
-Term | Definition 
------ | ---------
-Data standard | 
-Terminology |  The  body of terms used with a particular technical application in a subject of study. In this case terms related to traits in ecology and evolution?  
-Thesaurus | Controlled  vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline  [@garnier17]
-Trait ontology | Controlled  vocabulary to describe each trait as a distinguishable feature, characteristic, quality or phenotypic feature of a developing or mature individual.
-Ontologies also link the terms to each other and when possible define their hierarchy.
-Semantic web | The concept of connecting data on the internet via well-defined machine-readable standards and exchange protocolls  [@berners-lee02]
-Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being harmonized across all data entries;  
-Two-dimensional data tables | a spread-sheet containing data organised in rows and columns; in most cases these datasets are considered 'static', i.e. they are not altered or filtered across time
-Database | a set of data compiled from multiple *datasets*; i.e.
-from multiple study contexts; may take the form of a two-dimensional spreadsheet, but mostly is organised in software tools into *relational databases*
-Relational database | Two or more *two-dimensional data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Identifier  (ID) | 
-Uniform Resource Identifier (URI) |
-Web service | An exchange protocoll to access online databases directly and programmatically, i.e. by calls from a software tool; 
-Application Programming Interface (API) | 
-Online Portal | A website designed as a platform for the exchange of information, e.g.
-trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-Repository | Archive for data files; online repositories make data available for public access, provide metadata and (not always) facilitate citations via DOIs (Digital Object Identifiers)
-Metadata | 
-Darwin Core Standard (Dwc) | Body  of standards containing a glossary of terms intended to facilitate the sharing of information about biological diversity by providing reference definitions, examples, and commentaries.
-(From http://rs.tdwg.org/dwc/)
-Darwin Core Archive (DwC-A) | A file archive (.zip) that contains structured *metadata* and primary data in the form of *two-dimensional data tables* combined into a *relational database* via *identifier* columns.
-
-# A review of initiatives for trait-data standardisation
-
-In this section, we review four types of initiatives that are of relevance for trait data standardisation:
-
-1. Projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
-2. Initiatives that aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
-3. Initiatives that aim at the standardization and development of consensus methods and definitions for traits and provides standard terminologies and 'ontologies'; 
-4. Initiative that feed the needs of biodiversity informatics and develop relational database structures and provide platforms to relate trait data to a wider 'semantic web' of biodiversity data.
-
-We discuss these initiatives separately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
-We show how the degree of trait-data standardisation in existing datasets spans  this entire spectrum and which tools and standards do exist to achieve harmonisation of data from multiple, distributed sources.
-The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in an appropriate form.
-
-## Trait datasets
-
-In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books.
-With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15;  @renner14; @vanderplas17; @price13;  @ricklefs17].
-
-Today, the online publication of such data is greatly facilitated by hosting services like  Figshare.com, Dryad (datadryad.org), Researchgate.net, Zenodo.org. 
-Furthermore, these services warrant long-term accessibility (offering DOIs for referencing), and support Public Domain dedication or Creative Commons Attribution licenses, which supports the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12].
-However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
-.xlsx, .docx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk.
-
-Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management.
-As a result of this, datasets also greatly vary in structure. 
-Data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/), offer means to meet the high scientific criteria of data management.
-
-For trait-data, there are specific issues arising from the variability of data structure. 
-For instance, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives.
-Datasets vary in the detail provided in the metadata, e.g.
-for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
-If factual trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated quantitative measurements averaged from raw data, they usually are reported in a species$\times$trait wide-table format.
-Here, each row contains a species (or taxon) for which multiple traits are reported in columns.
-Similarly, when reporting raw data, researchers place observations of individual specimens in rows with multiple trait measurements applied to the same specimen placed in columns. 
-The variability of the number and meaning of columns in these tables requires manual adjustments when merging multiple datasets.
-
-As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compilation of data into larger collections will be inefficient and that many published datasets will not be properly exploited.
-
-A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-Therefore, in an attempt to collate a list of existing, distributed and unstandardized datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy; Appendix A) which lists published trait datasets, their regional and taxonomic focus, the traits covered, their location on the internet and the terms of their use (see Appendix 1 for a current excerpt of this list).
-We invite data owners and users to register further datasets qualifying as trait data to this spreadsheet.
-
-## Database initiatives
-
-<!-- community database backend, but static publication --> 
-In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09].
-<!--[@wright04; ] -->
-While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
--->
-Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
-Just as the individual datasets described above, many of these databases are published as static datasets on low-threshold file hosting platforms and are updated irregularly.
-
-Dealing with much larger amounts of data, initiatives that form around museal collection data are more concerned with data standardization.
-Aiming for a mobilization of measurements from collections, the  VertNet database compiles and harmonized large quantities of vertebrate traitdata [@guralnick16].
-The resulting data are published as static datasets which are updated as new data sources are integrated.
-
-<!-- database backend, selective download, user access management --> 
-More specialised trait-database platforms have been created that cover certain trait types (e.g.
-floral traits, seed traits or wood density traits), interaction types (e.g.
-pollination traits or feeding relationships), or environmental and experimental context of the trait observation (e.g.
-location or climatic data).
-These centralized database initiatives attract data submissions from a defined research field and take care of the harmonization process and greatly facilitate data synthesis. 
-<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
-By aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other platform.
-The online platform enables selective data download and user permission and rights management.
-As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
-Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12].
-Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13].
-Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
-the KEGG orthology, @kanehisa12].
-On the side of animals, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
-Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is still lacking.
-The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans, make harmonizing animal trait data a more complex task than for plant traits [@moretti17].
-
-Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication.
-The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.
-With the pro-active turn towards open access data (as declared inthe Bouchot Declaration; http://www.bouchoutdeclaration.org/) it may be necessary to find other incentives for data submission.
-
-<!-- AO: In a sense of data availability to the public : yes.
-But do we discuss it here?
-We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs.
-non-open data, or?
-bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
-FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
- -->
-
-## Thesauri and Ontologies for traits
-
-A major problem in trait-data standardisation is the reference to accepted and unambiguous trait definitions. This is acheived via thesauri and ontologies. 
-In database management, a thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].
-Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data.
-Providing relational ontologies in a machine readable form makes them compliant with semantic web standards, thus supporting the goal of widely interlinked data structures for the world wide web [@berners-lee01; @page08].
-
-<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
-units or legit factor levels) of measured values or reported facts.
-It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
---><!-- BKR: Isn’t that the same?
-Or can you have multiple parents for one node?
-FDS: agree.
-simplify phrasing.
--->
-
-Standard trait definitions have been provided in all shades along this continuum, ranging from listings of definitions in glossaries, over well-defined thesauri to relational ontologies.
-The initiatives behind these are first and foremost concerned with consensus building on trait definitions.
-Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring.
-Efforts to develop handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but are of limited effect when harmonizing trait data that predate or ignore this standard [@kattge11a].
-Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
-Thesauri of trait definitions have been developed for plants [e.g.
-the Planteome ontology for crop plants, @walls12; @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
-the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13].
-Comprehensive trait ontologies, including hierarchies of parent terms, have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and on the side of animal traits, in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
-
-These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. 
-For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
-<!-- Mention also NCBO BioPortal?
-https://bioportal.bioontology.org/ -->
-These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets.
-For example, a measurement of seed size would be linked to Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
-
-To be unambiguous, any thesaurus should be defining terms based on other well-defined terms from semantic ontologies.
-The definition provided here contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
-This way, each trait definition may link to a broader or narrower term.
-For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies.
-By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.
-
-To conclude, there is a shortage of globally valid trait thesauri or ontologies for traits that emerged from a standardisation effort of methodologies and broad consensus process within scientific field of ecology *and* that are sufficiently interlinked and accessible to fulfil semantic web standards.
-The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
-To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
-Such thesauri would advance the possibilities of aggregated datasets and reduce noise and uncertainty. 
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from distributed sources.
-
-## Trait-data structures for the semantic web
-
-While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g.
-for taxa.
-To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.
-
-Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
-More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
-trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
-traits) observed on a biological entity (i.e.
-an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e.
-a species or higher-level taxon).
-Expaning on this definition, quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
-1*a*), by aggregating replicated measurements on multiple entities (Fig.
-1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g.
-databases, Fig.
-1*c*).
-Qualitative facts are assignments of an entity to a categorical level, e.g.
-of a behavioural or life-history trait (Fig.
-1*d*).
-The entity or observation (i.e.
-the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
-a genus).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Types of ecological trait data assume different entities or reported quantities.
-a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}
-
-\end{figure}
-<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”.
-Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->
-
-These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain trait values, units, taxon names, variables such as sex or life stage, as well as ancillary information of time and date of observation, as well as methodological details.
-The terminologies provided by these standards are quite universal and cover most use cases of trait data.
-<!-- TODO FDS: dimensional and star-shema databases, relational databases Kattge et al 2011, Kimball & Ross 2002-->
-<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
-The Trait Data Standard is a standard on how to store trait data measurements.
-In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard.
-For instance the ABCD standard has a huge section for MeasurementsAndFacts.
-And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible).
-Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD.
-While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.
--->
-
-Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for databases that covers most potential use cases of trait-based ecology.
-This data structure is build around a central data table that contains observations, i.e.
-a single event of measurements on the same individual plant specimen at the same point in time.
-This structure puts emphasis on the fact that trait  data are often measured for an analysis of correlations of multiple traits.
-Identifiers link multiple measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables.
-The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
-This structure can be implemented in any software or programming language for relational databases.
-<!-- FDS: enter comments of Jens Kattge here!-->
-
-In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
-The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
-The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review.
-Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].
-
-These proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
-Hovewer, hardly any of the aforementioned aggregated and harmonized datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards.
-
-One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
-If the traits are assembled for a specific research question with few columns of ancillary information, it may be safer to add this information as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
-This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
-
-The other reason is lack of awareness of data providers for trait data standardization: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.
-Investing time and energy into research for the best and most sustainable data structure is not affordable for small and intermediate-size research projects.
-
-By filling this gap, data brokering services [e.g.
-the German Federation on for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
-BExIS, NEON Dataportal; Diversity Workbench, diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance.
-These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
-<!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
-
-## Conclusion of review
-
-Initiatives for standardization (e.g.
-ontologies and data standards) and platforms for data management (e.g.
-database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.
-With low-threshold offers for the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, the concern is with data harmonization of these distributed data sources.
-
-The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardization on the side of data providers. 
-We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in data visibility and interconnectedness of the published data.
-
-<!--
-Depending on the research objective, trait-data are stored in a range of different table formats.
-In the subsequent paragraphs, we discuss the most common formats  in relation to the use cases in trait-based ecology.
-
-### a) Species wide-table
-
-<!-- AO: You argue somehow that matrix format I a result of lit.
-review and so on.
-I would not agree with it.
-Your example also “just” points to aggregated data which menas one row is on species (aggregated) value.
-Your example is (for me) not really a matrix.
-In a matrix you have a relation between two variables.
-In the first column and and on the upper row normally the values of one variable are written.
-An example would be species x body length.
-Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted.
-E.g.
-spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on.
-Your wexample hat already different variables (=traits) with different units as columns.
-That is a meaningfull representation of variables and values in a normal data table.
-Maybe it is good her to show the enormous number of different investigated traits and point to the resulting big (many columns = wide) tables.
---><!--
-If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait wide-table format (Fig.
-2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available.
-The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
-<!-- AO: You say it already 2 sentences before --><!--
-This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g.
-'999').
-It may store additional information (e.g.
-on variation of means or literature source) in secondary columns or accompanying datasheets.
-
-Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a].
-Also at this level of information, missing entries, e.g.
-about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14; @taugourdeau14] and using Bayesian approaches [@schrodt15].
-<!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].
---><!--AO: Why not?
-I can filter for a specific species and get all respective rows.
-But this is not a result a using a data format, it is a result of the data itself, FDS: no, since only one row per species exists; true that data itself don't cover this variation.
--->
-<!--
-
-### b) Occurrence wide-table
-
-In these occurrence wide-tables (Fig.
-2*b*), each row is centered around a single physical occurrence of a species, e.g.
-phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns.
-Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
-Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism.
-This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data but rarely found for reported qualitative facts obtained from the literature.
-
-### c.) measurement long-table
-
-Both wide-table formats report the different trait values stored in multiple columns.
-Each trait column represents one value that applies to the entity in question (i.e.
-to a species  or an occurrence).
-As the column names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.
-
-A computationally more effective method that allows for high flexibility is the storage of traitdata in long-table formats [@wickham14; @parr16].
-In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  (i.e., one occurrence)  which is assigned to a specific taxon.
-<!-- FDS: also refer to kattge 11 -->
-<!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes.
-But not necessarily computational more effective at all.
-E.g.
-I start to mix up numeric values with descriptive values.
-I start to mix up different units.
-Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g.
-quality stuff.
-Ergo less effective from that perspective.
-But it has some advantages which may overrule dis.
-See below
-AO: For the whole proofing why to use such a long format I think that GBIF and DwC has already given some sentences. I strongly think their work should be considered and can be used as reference!
-We can simply argue that a well established routine should now also executed on trait level!
-Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.
-
---><!--
-This allows repeated measurements on a single individual to be stored by linking the data from separate rows via a unique specimen identifier.
-Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
-Long-table datasets purport multiple advantages for data manipulation (e.g.
-filtering, sub-setting and aggregating data), visualization (e.g.
-plot measured values by factor variable or taxon) and statistical modelling (e.g.
-ANOVA for testing difference of trait value by sex) [@wickham14].
-As long-table fromats draw from a defined set of columns,  merging of datasets is much easier.
-<!-- BKR: Is that really true?
-Of course, you have less problems on the schema or structure level.
-But, on the other hand, the differences (e.g.
-different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows).
-Thus, at least with respect to integration, you just move the problem from one level to the other.
-To ease integration, one would need to agree on the entries in the features column
-AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
-We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
-Again: I think we find here papers regarding DWC and GBIF (and OBIS) as source of proofing the facts.
--->
-<!--
-The long-table format therefore is ideal for storing raw data of trait measurements in huge databases, and allows the entries to be linked via the respective unique identifiers to additional information.
-Each row of the dataset can therefore be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig.
-2*c*).
-
-<!-- Cat: here I guess that a figure showing few rows of a species x trait matrix and a long-format one with links to other information (metadata,..) woud help quite a lot.
-By experience some users are not be familiar with the interest of using the long-table format.-->
-<!-- Malte: I agree (see my comment above), such a figure / would definitely help convince people (or at least not to lose them). -->
-<!--
-\begin{figure}[t]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) wide-tables report measurements on the specimen-level, i.e.
-one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to a trait definition and taxon via unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the specimen (occurrence) or e) the origin or confidence on the reported measurement or fact.}
-
-\end{figure}
--->
-
-<!--The trade-off is data-consistency and readability *vs.* avoidance of content duplication. -->
-
-<!--When traits are assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
-Here, location or measurement information are the primary focus of the investigation and must be contained in the .
-<!-- PM: I get lost here-->
-<!--A general recommendation, whether a column describes ancillary data or is part of the central dataset is ill advised.
-Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.
--->
-<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->
-<!--while the existing trait data standards may cover all of the aspects required in the various use-cases of trait-based research.
-<!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific -->
-
-# 2. Proposing an ecological traitdata standard vocabulary
-
-As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research.
-Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.
-
-By using this standard vocabulary, authors can ensure that the description of tables of trait measurements that are uploaded to distributed data repositories like Figshare, Zenodo etc.
-will be unambiguous and generally applicable.
-It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure accessibility in the long term.
-
-The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.
-
-In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
-Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.
-
-However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
-Also, at present we conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model.
-However, such network data may easily be combined with trait data via the taxon identifiers or by using other Extensions of the Darwin Core (DwC).
-Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
-
-## Why use the vocabulary
-
-The incentives for using a unified vocabulary for trait data might not be obvious for data providers, who are mainly concerned with a particular study site, organism group and research question.
-We won't discuss here in depth how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis.
-It appears to be broad consensus that an open biodiversity science is crucial for an evidence based decision making and conservation policy on regional and global scales.
-
-Instead, this section points on the technical aspects and advantages of readying own trait data to link with the 'semantic web' of biodiversity data and knowledge. 
-By using a constrained vocabulary with globally accessible definitions of terms, data can be accessed more easily by other researchers and datasets from different sources can be combined into aggregated datasets.
-Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary.
-
-It is implicit to this vocabulary to use a two-dimensional observation long-table format for the data, rather than a species-trait matrix (where some of the columns would be trait names which are not included in the vocabulary, but are part of trait thesauri).
-As this format draws from a defined set of columns,  merging of datasets is  easier.
-Long-table datasets also purport multiple advantages for data manipulation [e.g.
-filtering, sub-setting and aggregating data; @wickham14].
-
-Using well-defined identifiers ('IDs') to structure the dataset offers compatibility with the input formats for larger databases that may offer better visibility and impact of the data contribution. 
-Identifier fields are key elements to structure the two-dimensional dataset. 
-For instance, for occurrence level data, were multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link these measurements across the rows of the dataset.
-Similarly, multivariate measurements, like for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
-If literature data are provided, data are probably not structured below the taxon level and a 'taxonID' is the key identifier.
-A 'datasetID' allows to report datasets that have been compiled from multiple sources.  
-Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external ontology services (see above for resources).
-
-However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
-For datasets containing multi-layered information on observations, traits, taxa and environmental context, we  recommend to consider relational datatabase structures composed of multiple spreadsheets linked via identifier columns, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
-The trade-off is data-consistency and readability in a single table *vs.* avoidance of content duplication and redundancy in a relational database.
-The standard vocabulary proposed here may still be applied within these structured databases.
-
-## How to apply the standard vocabulary
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
-
-\end{figure}
-
-Column names and, where possible, field entries of a trait datasets published on online repositories should draw from the defined vocabulary.
-
-The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: a trait dataset contains information on quantitative measurements or qualitative facts (i.e.
-trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
-traits) observed on a biological entity (i.e.
-an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e.
-a species or higher-level taxon).
-
-Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below).
-
-Beyond the core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information).
-These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-For this case, we offer three extensions of the core vocabulary ("taxon", "measurement or fact", and "occurrence") that expand and refine terms of the Darwin Core Extensions (see below).
-This information may simply be added as extra columns to the core dataset.
-
-For reasons of long-term accessibility, data should not be uploaded in spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
-UTF-8 or ASCII).
-
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
-In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-
-Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam16; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e.
-via the Application Programming Interface, API **in preparation!**).
-Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
-
-Besides this use in published datasets, the standard vocabulary could be imposed in webservices, e.g.
-APIs that provide direct access to online databases or download tools that pull subsets of data from larger databases.
-
-### Specifications of core terms
-
-To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum: 
-1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
-2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
-3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
-For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
-However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.
-
-<!--### Relate to global taxon and trait definitions -->
-
-In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up tables of traits. 
-This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively).
-For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.
-By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
-This redundancy of data allows for continuity on the authors and data owners side while also ensuring quality checks and comparability for the data user.
-
-<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
-
-### Specification of traits
-
-If no globally available trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus.
-Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g.
-units or legit factor levels).
-
-By providing a minimal vocabulary for trait thesauri (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset.
-The thesaurus might accompany the core data in a separate file (e.g.
-within a Darwin Core Archive), or be published on any other stable internet address (ideally providing stable URIs).
-<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
-See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].
-We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).
-
-### Specification of Metadata terms
-
-Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model.
-
-In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g.
-in a separate metadata file, possibly consider applying the ecological metadata language, EML).
-In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be provided at the measurement level.
-This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
-
-<!--Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit.
-Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
-However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship.
--->
-
-<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
-FDS: to consider!! -->
-
-### Extensions for additional data layers
-
-Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-
-The data standard provides three extensions of the vocabulary that should be used to describe these information:
-
-- the `Taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides further terms for specifying the taxonomic resolution of the observation and to ensure correct reference in case of synonyms and homonyms.
-- The `Occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
-This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
-- The `MeasurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
-It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
-variation in measurement method) or aggregated values (e.g.
-statistical method applied), as well as the source of reported facts (e.g.
-literature source or expert reference).
-
-Many terms of these extensions refine or copy terms of the Darwin Core Standard (Dwc) and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
-
-<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).
-Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
-This information can then be used for trait inference at higher taxonomic resolution.
-<!-- using hierarchical probabilistic matrix factorisation [@schrodt15].
--->
-
-As for metadata information, these additional layers of information can either be added as extra columns to the core dataset or be kept in separate data sheets (published separately or as part of a Darwin Core Archive), this avoiding redundancy and duplication of content.
-A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
-Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
-a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.
-
-# 3. Computational tools for producing compliant data
-
-To access data from public databases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
-However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.
-
-Accordingly, we started to develop the R package 'traitdataform', which assists the production of data compliant with the trait data standard proposed here.
-There are two major use cases for the package:
-
-1. preparing own trait datasets for publication on public file-servers and project databases, and
-2. automating the harmonisation of trait datasets from different sources by moulding them into a unified format.
-
-A comprehensive documentation of the package is found on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
-The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
-
-<!-- The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
-2).
-This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
-This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:
-
+kable(dt, longtable=TRUE, 
+      caption = "List of published trait datasets and databases", 
+      booktabs = T) %>%
+kable_styling(latex_options = c("scale_down", "repeat_header"), font_size = 7) %>%
+  column_spec(1, width = "15em") %>%
+  column_spec(2, width = "5em") %>%
+  column_spec(3, width = "10em") %>%
+  column_spec(6, width = "10em") %>%
+  column_spec(7, width = "25em") %>%
+  column_spec(8, width = "11em") 
 ```
-library("traitdataform")
-dataset <- as.traitdata(read.csv("path/to/data.csv"), 
-             traits = c("body_length", "antenna_length", 
-                        "metafemur_length"),
-             units = "mm",
-             taxa = "name_correct", 
-             keep = c(locationID = "location")
-             )
-
-```
-
-The column containing taxon names is given in parameter 'taxa'.
-Note that the parameter 'keep' specifies and renames any data that should be maintained in the output.
-The parameter 'units' is used to specify the input units of measurement.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
-\caption{Process chart of the functions provided within the R package.}
-
-\end{figure}
-
-In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data.
-<!-- BKR: In the example below, this looks like a manual matching.
-What is done automatically here and which information does the user have to provide?
-FDS: taxon is matched automatically, traits not.
-lack of trait APIs and webservices.
--->
-<!-- Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
-a thesaurus of traits.
-The same function harmonises trait values by converting values into target units and unified factor levels.
-<!-- BKR: Have you thought about linking to unit ontologies here?
-AO: This paper is a good reference and gives an idea of which ontology could be used;
-http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
-FDS: might be referred to in the section about thesauri -->
-<!--
-The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
-
-```
-traitlist <- as.thesaurus(
-   body_length = as.trait("body_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length"), 
-   antenna_length = as.trait("antenna_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Antenna_length"),
-   metafemur_length = as.trait("metafemur_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Femur_length")
-)
-
-datasetStd <- standardize.traits(dataset, thesaurus = traitlist)
-```
-
-The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`.
-Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`).
-Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies.
-
-The package functions form a toolchain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:
-
-```
-datasetStd <- standardize(read.csv("path/to/data.csv"),
-            thesaurus = traitlist,
-            taxa = "name_correct", 
-            units = "mm"
-            )
-```
-
-Datasets that have been produced by these functions can easily be appended using the function `rbind()` of R base, while maintaining any available metadata information as separate column entries.
-
-To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identifier using the function `merge()`.
-This enables an easy handling of data sources that have been stored in a Darwin Core Archive format.
-
--->
-
-Since the intention of the package is also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been released in the public domain or under creative commons licenses.
-<!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
-
-# Discussion
-
-<!--The practical importance of these frameworks for data providers and data users is often not obvious and not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
-
-The investment is only paying out in the long run. Citations of trait datasets are higher when they are easier to access and relate to other data due to standard structures and vocabularies. 
-Along with the push towards open access publication of primary data, the scientific community must come up with ways to incentivize and require data standardization.
-
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.-->
-To serve the demand for simple ways to standardise and harmonize ecological trait data, we propose a versatile vocabulary for simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
-We also present a toolbox in R to transfer and harmonize data into this scheme.
-
-To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions.
-To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.
-It also can be applied in the communication of data content between databases.
-In addition, we also offer a software tool to handle the extensive raw data resources on traits.
-
-These tools are to facilitate research on trait variation along environmental gradients or intra-specific variation based on distributed data sources.
-This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.
-
-The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories.
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns.
-In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
-We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis.
-
-We also encourage the development of trait thesauri in more advanced and interrelated ontologies.
-The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
-This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].
-
-# Acknowledgements
-
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal and Jens Kattge for comments on the manuscript, the trait data standard vocabulary and the pre-releases of the R-package.
-
-The work has been partly funded by the DFG Priority Program
-1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). <!--Field work
-permits were issued by the responsible state environmental offices of
-Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
-BbgNatSchG).-->
-We thank <!--the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
-infrastructure;--> Christiane Fischer and Jule Mangels for giving
-support through the central office, <!--Michael Owonibi for
-managing the central data base,--> and <!--Markus Fischer,--> Eduard Linsenmair,
-Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
-Ernst-Detlef Schulze, Wolfgang W. Weisser and the late Elisabeth Kalko
-for their role in setting up the Biodiversity Exploratories project.
-
-M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
-
-# Authors' contributions
-
-FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package 'traitdataform'.
-All authors contributed critically to the drafts and gave final approval for publication.
-
-# Online Resources
-
-The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
-
-Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.
-
-The development website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform.
-
-The living spreadsheet on existing trait datasets and databases is found at https://goo.gl/QxzfHy.
-
-# References
-
-\footnotesize

### 2018-05-22 manuscript/appendix_A.Rmd
update appendix A script

@@
 gg <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTfQOIHjJv5x_YHgq14DR7VLTvycfsl2Njj2Ip4p2a5kca8AX5t5cmaial6Bg-MX9pBDOJaJfnlR7Qj/pub?gid=0&single=true&output=csv", encoding = "UTF-8")
-dt <- subset(gg, select = c(datasetName, type, DOI, NumberTaxa, NumberTraits, Responsible, CiteAs, TermsOfUse))
+dt <- subset(gg, select = c(datasetName, type, DOI, organismFocus, regionalFocus, NumberTaxa, NumberTraits, Responsible, CiteAs, TermsOfUse))
 
 kable(dt, longtable=TRUE, 
-      caption = "List of published trait datasets and databases", 
+      caption = paste0("List of published trait datasets and databases; static copy of the living spreadsheet at https://goo.gl/QxzfHy (access date ",
+        format(Sys.time(), '%Y-%m-%d'), ")"), 
       booktabs = T) %>%
-kable_styling(latex_options = c("scale_down", "repeat_header"), font_size = 7) %>%
+kable_styling(latex_options = c("scale_down", "repeat_header"), font_size = 6) %>%
   column_spec(1, width = "15em") %>%
   column_spec(2, width = "5em") %>%
-  column_spec(3, width = "10em") %>%
-  column_spec(6, width = "10em") %>%
-  column_spec(7, width = "25em") %>%
-  column_spec(8, width = "11em") 
+  column_spec(3, width = "12em") %>%
+  column_spec(4, width = "6em") %>%
+  column_spec(5, width = "6em") %>%
+  column_spec(6, width = "6em") %>%
+  column_spec(8, width = "10em") %>%
+  column_spec(9, width = "25em") %>%
+  column_spec(10, width = "11em") 
 ```

### 2018-05-22 manuscript/manuscript.Rmd
finalize manuscript v0.8; complete glossary

@@
 csl: amnat.csl
-fontsize: 12pt
+fontsize: 11pt
 geometry: margin=1.3in
@@
 <!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
-
 <!-- 4. distributed trait data must be integrated --> 
 While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly.
-This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]., which supports.
+This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15].
 As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
@@
 Finally, we present an R package that assists the harmonisation of trait data from distributed sources.
-
 With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.
@@
 -->
-\scriptsize
+\newpage
+
+\tiny
 
@@
 Term | a word that describes a particular *concept* as part of the specialised vocabulary of a field
-Concept | An idea, notion or object that is made explicit in a communication context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
-Controlled vocabulary | 
+Concept | An idea, notion or object that is made explicit in an information context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
+Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a praticular context, while no unlisted entries are accepted
+Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
 Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
-trait data), that is recognized by a large proportion of members of the application context
-Terminology | The  body of *terms* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
-Thesaurus | Controlled vocabulary that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte12] 
-Ontology | Controlled vocabulary that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
+trait data), that is recognised by a large proportion of members of the application context
+Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte13] 
+Ontology | *Controlled vocabulary* that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
 defines a hierarchy of terms; thus a formal model of a the objects and their relationships in a domain of interest [@gruber95]
-Semantic web | The concept of connecting data on the internet via well-defined standards and exchange protocols [@berners-lee02]; the World Wide Web Consortium (W3C) defines standards, i.e.
+Semantic web | The concept of connecting data on the internet via well-defined *data standards* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
 specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
-Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being harmonized across all data entries;  
+Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being homogeneous across all data entries
 Data table | a two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
 Database | a suite of data compiled from multiple *datasets*, i.e.
-from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software tools;
+from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software;
 Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Online database | A relational database that is made accessible on the internet offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
-Identifier (ID) | A unique label that relates entries within and across datasets; is used to link data tables into a relational database; can be user-specific or as a *URI* be part of a globally valid ontology or thesaurus
-Uniform Resource Identifier (URI) |
-Web service | An exchange protocoll to access online databases directly and programmatically, i.e. by calls from a software tool; 
-Application Programming Interface (API) | 
+Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
+Identifier (ID) | A unique label that relates entries within and across *datasets*; is used to connect *data tables* into a *relational database*; can be user-specific or, as a *URI*, point to a globally valid *ontology* or *thesaurus*
+Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; an example is 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
+Webservice | An exchange protocoll to access *online databases* directly and programmatically, i.e. by calls from a software tool
+Application Programming Interface (API) |  a set of clearly defined methods of communication between  software components, e.g.
+client software and the *webservice* of an online databases; APIs are usually documented on the website of a database provider
 Online Portal | A website designed as a platform for the exchange of information, e.g.
 trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-Repository | Archive for data files; online repositories make data available for public access, provide metadata and (not always) facilitate citations via DOIs (Digital Object Identifiers)
-Metadata | Data documentation representing the higher level information or instructions that describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
-Darwin Core Standard (Dwc) | Body  of standards containing a glossary of terms intended to facilitate the sharing of information about biological diversity by providing reference definitions, examples, and commentaries.
-(From http://rs.tdwg.org/dwc/)
-Darwin Core Archive (DwC-A) | A file archive (.zip) that contains structured *metadata* and primary data in the form of *two-dimensional data tables* combined into a *relational database* via *identifier* columns.
-Occurrence | 
+File repository | a structure for data files, for archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
+File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
+Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
+Darwin Core Standard (Dwc) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
+Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* and primary data combined into a *relational database* via *identifier* columns. 
+Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures. 
+Occurrence | A single observation instance of a taxon, i.e. an organism at a particular place at a particular time (http://rs.tdwg.org/dwc/terms/Organism)
+
+Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier17)
+
+<!--
 Measurement |  
@@
 Fact | 
-Method handbook | 
-File hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; popular examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org;
-
-Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier16)
+-->
 
@@
 
+\newpage
+
 # A review of initiatives for trait-data standardisation
 
-In this section, we review four types of initiatives that are of relevance for trait data standardisation:
+In this section, we review four types of initiatives that are of relevance for trait data standardisation (see Glossary in Table 1 for italicised terms):
 
@@
 A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
-
-Today, the online publication of such data is greatly facilitated by file hosting services (e.g. Figshare.com). 
+Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g. Figshare.com). 
 Furthermore, these services warrant long-term accessibility, offer DOIs for referencing, and support Public Domain dedication or Creative Commons licenses.
@@
 .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
-
 Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds for *metadata* documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management.
@@
 We invite data owners and users to add further trait datasets to this spreadsheet.
-
 As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.
@@
 As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
-With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowdsourcing (e.g.
+With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g.
 www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
@@
 To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
-For example the accessibility of ontologies will increase if advanced webservices, and particularly open Application Programming Interfaces (APIs), are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+For example the accessibility of ontologies will increase if advanced webservices, and particularly open *Application Programming Interfaces (APIs)*, are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
 Software then can assist researchers in linking own data to globally defined concepts.
@@
 entity), of a biological taxon.
-It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data, rather than a species$\times$trait matrix.
-As the long-table format draws from a defined set of columns, the merging of datasets is  easier.
+It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data (Fig.
+2 *b*), rather than a species$\times$trait matrix (Fig.
+2 *a*).
+As the long-table format draws from a defined set of columns, merging datasets is easier.
 Long-table datasets also purport multiple advantages for data manipulation [e.g.
-filtering, sub-setting and aggregating data, @wickham14; Fig.
-2].
+filtering, sub-setting and aggregating data, @wickham14].
 
-Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary.
+Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary (Fig. 2 *c* & *d*).
 For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset.
@@
 
-To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum: 
+To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum  (Fig.
+2 *b*):
 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
@@
 Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-The data standard provides three extensions of the vocabulary that should be used to describe this information:
+The data standard provides three extensions of the vocabulary that should be used to describe this information (Fig. 2 *d*):
 
@@
 
-Wherever possible, the colum `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
+Wherever possible, the column `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
 If no globally available trait definition exists as a reference, trait-datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits.
@@
 
-The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig.
+The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement long-table format (Fig.
 3).
@@
 The parameter 'units' is used to specify the input units of measurement.
-In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data.
+In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data  (Fig.
+3).
 <!-- BKR: In the example below, this looks like a manual matching.
@@
 
-The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
+The following example harmonises traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
 
@@
 
-The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`.
+The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`  (Fig.
+3).
 Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`).
 Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies.
-The package functions form a toolchain where each function can be piped as an input into the next.
+The package functions form a tool-chain where each function can be piped as an input into the next.
 A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing trait data as simple as:
@@
 
-Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.
+Any future development of the vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.
 
@@
 
-The living spreadsheet on existing trait datasets and databases is found at https://goo.gl/QxzfHy.
+An example of a published trait thesaurus for the Biodiversity Exploratories (www.biodiversity-exploratories.de) can be found at https://ecologicaltraitdata.github.io/TraitDataList/.
+
+The living spreadsheet on existing trait datasets and databases can be found at https://goo.gl/QxzfHy.
 

### 2018-05-24 manuscript/appendix_A.Rmd
update Appendix script, sort table, add URL and DOI hyperref

--- 2018-05-18 manuscript/appendix_A.Rmd
@@
 - name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
+- name: David Fichtmüller \footnotemark[5]
+- name: Anton Güntsch \footnotemark[5]
 - name: Malte Jochum \footnotemark[4]
+- name: Birgitta König-Ries\footnotemark[6]
 - name: Gaetane Le Provost \footnotemark[3]
+- name: Pete Manning \footnotemark[3]
 - name: Andreas Ostrowski \footnotemark[5]
 - name: Caterina Penone \footnotemark[4]
-- name: David Fichtmüller \footnotemark[5]
-- name: Anton Güntsch \footnotemark[5]
-- name: Pete Manning \footnotemark[3]
-- name: Birgitta König-Ries\footnotemark[6]
 - name: Nadja K. Simons \footnotemark[7]
@@
 library(kableExtra)
-gg <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTfQOIHjJv5x_YHgq14DR7VLTvycfsl2Njj2Ip4p2a5kca8AX5t5cmaial6Bg-MX9pBDOJaJfnlR7Qj/pub?gid=0&single=true&output=csv", encoding = "UTF-8")
-dt <- subset(gg, select = c(datasetName, type, DOI, NumberTaxa, NumberTraits, Responsible, CiteAs, TermsOfUse))
+gg <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTfQOIHjJv5x_YHgq14DR7VLTvycfsl2Njj2Ip4p2a5kca8AX5t5cmaial6Bg-MX9pBDOJaJfnlR7Qj/pub?output=csv", encoding = "UTF-8")
+dt <- subset(gg, select = c(datasetName, type, DOI, organismFocus, regionalFocus, NumberTaxa, NumberTraits, Responsible, CiteAs, TermsOfUse))
 
-kable(dt, longtable=TRUE, 
-      caption = "List of published trait datasets and databases", 
+#dt$DOI <- paste0("[http://dx.doi.org/", gg$DOI, "](", gg$DOI, ")")
+URL <- sapply(as.character(gg$URL), function(x) unlist(strsplit(x, split = "\n"))[1] )
+URL_resolved <- paste0("\\href{",URL,"}{", sapply(URL, substr, start = 1, stop = 30), "... \\ExternalLink} ") #
+dt$DOI <- paste0("\\href{http://dx.doi.org/", gg$DOI, "}{", gg$DOI, " \\ExternalLink} ") 
+empty <- which(dt$DOI %in% c("\\href{http://dx.doi.org/NA}{NA \\ExternalLink} ", "\\href{http://dx.doi.org/}{ \\ExternalLink} "))
+dt$DOI[empty] <- URL_resolved[empty]
+
+#dt$DOI <-  cell_spec(gg$DOI, "latex", link = paste0("http://dx.doi.org/", gg$DOI ) )
+caption = paste0("Collection of trait datasets and databases; excerpt of the living spreadsheet at https://goo.gl/QxzfHy (access date ", format(Sys.time(), '%Y-%m-%d'), ")")
+
+order <- c(14, 15,30,  # all organisms
+           1,4,5,13,18,20,21, # plants
+           11,16,25,33, # all animals
+           3,17,31, ## birds
+           7,34,12, # mammals
+           26,8,27, # other vertebrates
+           32,2, 6,9,23,10,22,29,# invertebrates
+           24,28, # fungi
+           19 # other
+           )
+
+kable(dt[order,], longtable=TRUE, escape = FALSE, row.names = FALSE,
+      caption = caption,
       booktabs = T) %>%
-kable_styling(latex_options = c("scale_down", "repeat_header"), font_size = 7) %>%
+kable_styling(latex_options = c("scale_down", "repeat_header"), font_size = 6) %>%
   column_spec(1, width = "15em") %>%
   column_spec(2, width = "5em") %>%
-  column_spec(3, width = "10em") %>%
-  column_spec(6, width = "10em") %>%
-  column_spec(7, width = "25em") %>%
-  column_spec(8, width = "11em") 
+  column_spec(3, width = "16em") %>%
+  column_spec(4, width = "6em") %>%
+  column_spec(5, width = "6em") %>%
+  column_spec(6, width = "5em") %>%
+  column_spec(7, width = "5em") %>%
+  column_spec(8, width = "8em") %>%
+  column_spec(9, width = "25em") %>%
+  column_spec(10, width = "11em") 
 ```

### 2018-05-28 manuscript/manuscript.Rmd
revise ms: comments of AG and BKR

@@
 - name: Gaetane Le Provost \footnotemark[3]
-- name: Andreas Ostrowski \footnotemark[5]
+- name: Andreas Ostrowski \footnotemark[6]
 - name: Caterina Penone \footnotemark[4]
 - name: David Fichtmüller \footnotemark[5]
+- name: Martin Gossner \footnotemark[8]
 - name: Anton Güntsch \footnotemark[5]
@@
 \footnotetext[7]{Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
 
@@
 
-1.	Trait-based approaches are widespread throughout ecological research, due to the great potential for trait data to deliver general and mechanistic conclusions.
-Accordingly, a wealth of trait data is available for many organism groups but, due to a lack of standardisation, these data come in heterogeneous formats.
+1.	Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions.
+Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
 2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
@@
 4.
-Additionally, we provide an R-package that enables the transformation of any dataset into the proposed format.
+Additionally, we provide an R-package that enables the transformation of any tabular dataset into the proposed format.
 This also allows trait datasets from heterogeneous sources to be harmonized and merged, thus facilitating data compilation for any particular research focus.
@@
 Many issues in trait-based research arise when compiling datasets from several sources.
-Data may differ in taxonmic nomenclature and resolution (e.g.
+Data may differ in taxonomic nomenclature and resolution (e.g.
 reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the methodology applied in measurements.
 All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
-However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to acheive sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
+However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
 
@@
 Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
-These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
+These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data.
 <!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
@@
 <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
-It is likely that data will become increasingly available, but a lack of data standardisation will hamper the efficient re-use and synthesis of published datasets.
+It is likely that data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.
 
@@
 We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users.
-Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that is compatible with existing terminology standards for biodiversity data.
+Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that builds upon and is compatible with existing terminology standards for biodiversity data.
 Finally, we present an R package that assists the harmonisation of trait data from distributed sources.
@@
 Ontology | *Controlled vocabulary* that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
-defines a hierarchy of terms; thus a formal model of a the objects and their relationships in a domain of interest [@gruber95]
-Semantic web | The concept of connecting data on the internet via well-defined *data standards* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
+defines a hierarchy of terms; thus a formal model of the objects and their relationships in a domain of interest [@gruber95]
+Semantic web | An extension of the world wide web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
 specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
-Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being homogeneous across all data entries
+Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
 Data table | a two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
@@
 from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software;
-Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
+Relational database | Usage in this paper: Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
 Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
@@
 trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-File repository | a structure for data files, for archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
+File repository | a short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
 File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
@@
 Darwin Core Standard (Dwc) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
-Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* and primary data combined into a *relational database* via *identifier* columns. 
+Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns.
 Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures. 
@@
 
-Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier17)
+Table: Glossary of terms from the biodiversity data-management context as they are used in this paper (draws from @garnier17)
 
@@
 A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
-Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g. Figshare.com). 
-Furthermore, these services warrant long-term accessibility, offer DOIs for referencing, and support Public Domain dedication or Creative Commons licenses.
+Today, the online publication of such data is greatly facilitated by  *file hosting services* (e.g.
+Figshare.com), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
+These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management.
 However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
@@
 .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
-Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds for *metadata* documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management.
+Most importantly, these platforms enable public hosting of data with very low thresholds for *metadata* documentation and data standardisation. 
 As a result, datasets greatly vary in structure and terminology.
@@
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-Therefore, in an attempt to collate a list of existing, distributed and unstandardised datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list).
+Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list).
 We invite data owners and users to add further trait datasets to this spreadsheet.
@@
 
-As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
-With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g.
+As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with standardisation. 
+Concerning organism traits, with the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g.
 www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
@@
 A major challenge in trait-data standardisation is the lack of widely accepted and unambiguous trait definitions.
-Previous standard definitions of trait *concepts* range from listings of selected definitions in *glossaries*, over well-defined methodological *handbooks* and and comprehensive *thesauri*, to relational definitions of trait concepts in *ontologies*.
+Previous standard definitions of trait *concepts* range from listings of selected definitions in *glossaries*, over well-defined methodological *handbooks* and comprehensive *thesauri*, to relational definitions of trait concepts in *ontologies*.
 While glossaries may be seen as specific for a study context, the initiatives behind method handbooks, thesauri and ontologies are primarily concerned with consensus building on trait definitions in a wider community.
 
-Assigning an unambiguous methodological protocol to a trait is key for standardising the physical process of measuring.
+Assigning an unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring.
 Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].
@@
 A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
-Expanding on this, ontologies link the defined terms to each other by formally defining the relationships between them, with the objective of enabling a computational interpretation of data. 
+Expanding on this, ontologies link the defined terms by formally defining the relationships between them, with the objective of enabling a computational interpretation of data. 
 Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)* within own datasets.
@@
 
-Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for the collaborative development [@walls12]. 
-For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.
-<!-- Mention also NCBO BioPortal?
-https://bioportal.bioontology.org/ -->
+Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for their collaborative development [@walls12]. 
+For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/), provide centralised hosting for advanced trait ontologies and offer webservices for computational access.
 
 To conclude, there is already a suite of globally available thesauri and ontologies for traits that emerged from standardisation efforts of methodologies and community consensus processes.
-Albeit definitions in some domains are better covered than others.
+However, definitions in some domains are better covered than others.
 Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
@@
 To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
-For example the accessibility of ontologies will increase if advanced webservices, and particularly open *Application Programming Interfaces (APIs)*, are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+For example, the accessibility of ontologies will increase if open *Application Programming Interfaces (APIs)* are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
 Software then can assist researchers in linking own data to globally defined concepts.
@@
 
-These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
 For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies these terms.
@@
 These proposed standards are responses to a demand from biodiversity data managers for more structured input from the research community. 
-Hovewer, hardly any of the aforementioned trait datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
+However, hardly any of the aforementioned trait datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
 One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
@@
 -->
-The other reason is lack of awareness for the need for trait data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
+The other reason is lack of awareness for the need for trait data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
 
@@
 A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
-Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken.
+Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
+a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
 
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Martin Gossner, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts, the trait data standard vocabulary and the pre-releases of the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Kristin Bohn, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts, the trait data standard vocabulary and the pre-releases of the R-package.
 
+We thank the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
+infrastructure; Christiane Fischer and Jule Mangels for giving
+support through the central office, andMarkus Fischer, Eduard Linsenmair,
+Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
+Ernst-Detlef Schulze, Wolfgang W. Weisser and the late Elisabeth Kalko
+for their role in setting up the Biodiversity Exploratories project. 
 The work has been partly funded by the DFG Priority Program
@@
 BbgNatSchG).-->
-We thank <!--the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
-infrastructure;--> Christiane Fischer and Jule Mangels for giving
-support through the central office, Jens Niesschulze for establishing the central data base, and Markus Fischer, Eduard Linsenmair,
-Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
-Ernst-Detlef Schulze, Wolfgang W. Weisser and the late Elisabeth Kalko
-for their role in setting up the Biodiversity Exploratories project.
 
@@
 
-FDS, MJ, GP, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard; NKS authored the example thesaurus of traits; FDS developed the R package 'traitdataform'.
+FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP; NKS authored the example list of traits; FDS developed the R package 'traitdataform'; CP and FDS curated the living spreadsheet; AG and DF implemented the vocabulary in the GFBio terminology service.
 All authors contributed critically to the structure and content of the manuscript and gave final approval for publication.

### 2018-05-28 manuscript/manuscript.Rmd
update manuscript: comments MJ, NS and GLP

@@
 - name: Malte Jochum \footnotemark[4]
-- name: Gaetane Le Provost \footnotemark[3]
+- name: Gaëtane Le Provost \footnotemark[3]
 - name: Andreas Ostrowski \footnotemark[6]
@@
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[7]{Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
 \footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
+\footnotetext[7]{Chair for Terrestrial Ecology, Department of Ecology and Ecosystemmanagement,
+Technische Universität München, Freising, Germany}
 
@@
 Additionally, we provide an R-package that enables the transformation of any tabular dataset into the proposed format.
-This also allows trait datasets from heterogeneous sources to be harmonized and merged, thus facilitating data compilation for any particular research focus.
+This also allows trait datasets from heterogeneous sources to be harmonised and merged, thus facilitating data compilation for any particular research focus.
 5.
@@
 Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
-Because trait-based approaches to ecological and evolutionary research questions allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
+Because trait-based approaches allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
 Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10].
-With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
+With this mechanistic depth, the trait framework relates losses of ecosystem functions to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
@@
 TRY, www.try-db.org).
-Furthermore, these portals map heterogeneous data into a common scheme, and importantly, also offer access control and data usage policies.
+Furthermore, these portals map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies.
 As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers.
 Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
-These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data.
+These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
 <!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
 <!-- 4. distributed trait data must be integrated --> 
-While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly.
+While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, the number of decentralised trait datasets is increasing rapidly.
 This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15].
@@
 We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users.
-Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that builds upon and is compatible with existing terminology standards for biodiversity data.
+Based on these considerations, we propose a minimal structure and vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data.
 Finally, we present an R package that assists the harmonisation of trait data from distributed sources.
-With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.
+With this easy-to-use terminology and toolset, we hope to  convince trait-data providers and trait-data users about the general importance of trait-data standardisation and lay out the roadmap towards an accessible ecological trait data standard.
 
@@
 ----- | ---------
-Term | a word that describes a particular *concept* as part of the specialised vocabulary of a field
+Term | A word that describes a particular *concept* as part of the specialised vocabulary of a field
 Concept | An idea, notion or object that is made explicit in an information context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
@@
 specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
-Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
-Data table | a two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
-Database | a suite of data compiled from multiple *datasets*, i.e.
+Dataset | A set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
+Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
+Database | A suite of data compiled from multiple *datasets*, i.e.
 from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software;
 Relational database | Usage in this paper: Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
+Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
 Identifier (ID) | A unique label that relates entries within and across *datasets*; is used to connect *data tables* into a *relational database*; can be user-specific or, as a *URI*, point to a globally valid *ontology* or *thesaurus*
 Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; an example is 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
-Webservice | An exchange protocoll to access *online databases* directly and programmatically, i.e. by calls from a software tool
-Application Programming Interface (API) |  a set of clearly defined methods of communication between  software components, e.g.
+Webservice | An exchange protocol to access *online databases* directly and programmatically, i.e. by calls from a software tool
+Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
 client software and the *webservice* of an online databases; APIs are usually documented on the website of a database provider
@@
 trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-File repository | a short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
+File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
 File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
@@
 
-Table: Glossary of terms from the biodiversity data-management context as they are used in this paper (draws from @garnier17)
+Table: Glossary of terms from the biodiversity data-management context as they are used in this paper; draws from @garnier17.
 
@@
 
-In this section, we review four types of initiatives that are of relevance for trait data standardisation (see Glossary in Table 1 for italicised terms):
+In this section, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):
 
@@
 Most importantly, these platforms enable public hosting of data with very low thresholds for *metadata* documentation and data standardisation. 
-As a result, datasets greatly vary in structure and terminology.
+<!-- As a result, datasets greatly vary in structure and terminology. -->
 
-For trait-data, there are typical issues arising from the variability of data structures. 
-For instance, the column descriptions andterminology applied to taxa and traits are project specific, and rarely chosen to allow translation into larger database initiatives.
+For trait data, there are typical issues arising from the variability of data structures. 
+For instance, the column descriptions and terminology applied to taxa and traits are project specific, and rarely chosen to allow translation into larger database initiatives.
 Furthermore, metadata varies in its detail, e.g.
 for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
-In terms of structure, trait data usually are reported in a species$\times$trait wide-table format.
+In terms of structure, trait data usually are reported in a species$\times$traits wide-table format.
 In this format, each row contains a species (or taxon) for which multiple traits are reported in columns.
@@
 In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09].<!--[@wright04; ] -->
+@klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Appendix A table A1].<!--[@wright04; ] -->
 While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
@@
 As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with standardisation. 
-Concerning organism traits, with the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g.
+Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g.
 www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
 For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16].
-The resulting data are published as versioned data tables which are updated as new data sources are integrated.
+The resulting data are published as versioned data tables which are updated as new data sources become available.
 
@@
 the KEGG orthology, @kanehisa12].
-For animals, a single unified platform and harmonizing scheme for animal trait data is still lacking.
+For animals, a single unified platform and harmonising scheme for animal trait data is still lacking.
 The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17].
@@
 Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)* within own datasets.
-For example, a measurement of seed size could be linked to the Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
+For example, a measurement of seed size could be linked to the Planteome Trait Ontology (TO) definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
 Ontologies define terms based on other well-defined terms from published ontologies.
-The Planteome Trait Ontology definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
+The TO definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
+Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12].
 This way, each trait definition may link to a broader or narrower term.
 For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies.
-By providing this interlinkage of trait ontologies, a web of definitions is spun across the Internet which allows researchers and search engines to relate independent trait measurements with each other and connect it to the wider *semantic web* of online data [@berners-lee01; @page08].
+By providing this interlinkage of trait ontologies, a machine-readable web of definitions is spun across the Internet which allows researchers and search engines to relate independent trait measurements with each other and connect it to the wider *semantic web* of online data [@berners-lee01; @page08].
 The distinction of thesauri and ontologies is not truely binary.
@@
 Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for their collaborative development [@walls12]. 
-For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/), provide centralised hosting for advanced trait ontologies and offer webservices for computational access.
+For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/), provide centralised hosting for advanced trait ontologies and offer webservices for computational access.
 
@@
 
-In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach to providing an integrated structure for trait data.
+In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach of an integrated structure for trait data.
 The framework employs established terms provided by the DwC, relates trait definitions to trait ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
@@
 However, hardly any of the aforementioned trait datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
-One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
+One reason for this is most certainly complexity: the data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
 In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
@@
 -->
-The other reason is lack of awareness for the need for trait data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
+The other reason is lack of awareness for the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
 
@@
 database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.
-Low-threshold repositories offer the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, but data harmonisation of these distributed data sets is currently laborious.
-The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
+Low-threshold repositories offer the hosting of scientific primary data attracting a wealth of heterogeneous trait datasets, but data harmonisation of these distributed data sets is currently laborious.
+The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
 We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in the visibility and interconnectedness of published data.
@@
 The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
-Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.
+Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the semantic web.
 By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable.
@@
 We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible.
-The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17] each entry describes a trait value (i.e.
-quality) observed on an an individual or population (i.e.
+The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
+quality) observed on an individual or population (i.e.
 entity), of a biological taxon.
-It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data (Fig.
-2 *b*), rather than a species$\times$trait matrix (Fig.
+When applying the vocabulary, it is implicit to use a two-dimensional observation long-table format for the data (Fig.
+2 *b*), rather than a species$\times$traits matrix (Fig.
 2 *a*).
@@
 
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A.
+In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
+"Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A.
 and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
 In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-Wherever referring to individual terms of the vocabulary in publications or metadata,this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
+Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
 via the API; **in preparation!**).
@@
 The core terms describe minimal trait data according to the Entity-Quality model. 
-Beyond these core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information).
+Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information).
 These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
@@
 Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below). 
-The scope of the vocabulary may not yet cover all aspects morphological and evolutionary perspectives. 
+The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
 Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other Extensions of the Darwin Core.
@@
 2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
-3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
+3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`). 
 For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
@@
 Wherever possible, the column `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
-If no globally available trait definition exists as a reference, trait-datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits.
+If no globally available trait definition exists as a reference, trait datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits.
 Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g.
@@
 To access data from public databases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix.
-However, none of these packages provide the option to harmonization trait data into a unified scheme.
+The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
+However, none of these packages provide the option to harmonisation trait data into a unified scheme.
 To close this gap, we developed the R package 'traitdataform', which assists the production of data compliant with the trait data standard proposed above. 
@@
 To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions. 
-However, these distributed data are heterogeneous in form and description and initiatives to harmonize and compile these data require significant amounts of funding and personel. 
-To support the long-term rewards of standardization efforts, incentives should be sought to mitigate the cost of readying trait data for the 'semantic web' of biodiversity data and knowledge.
+However, these distributed data are heterogeneous in form and description and initiatives to harmonise and compile these data require significant amounts of funding and personel. 
+To support the long-term rewards of standardisation efforts, incentives should be sought to mitigate the cost of readying trait data for the 'semantic web' of biodiversity data and knowledge.
 This can be software tools or supporting infrastructures.
@@
 We also encourage the advancement of trait thesauri into more interrelated and complete ontologies. 
-The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
+The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
 This requires significant effort, but it returns  great scientific benefit by enabling synthesis on our general understanding of biodiversity and ecosystem function.
@@
 
-FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP; NKS authored the example list of traits; FDS developed the R package 'traitdataform'; CP and FDS curated the living spreadsheet; AG and DF implemented the vocabulary in the GFBio terminology service.
+FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP (forming the first tier of the author list); NKS authored the example list of traits; FDS developed the R package 'traitdataform'; CP and FDS curated the living spreadsheet.
+All contributing authors appear in alphabetical order in a second tier of the author list.
+AG and DF implemented the vocabulary in the GFBio terminology service.
 All authors contributed critically to the structure and content of the manuscript and gave final approval for publication.

### 2018-05-28 manuscript/manuscript.Rmd
update manuscript: comments of MG

@@
 
-1.	Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions.
+1.
+Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions.
 Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
@@
 
-Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
+Functional traits are phenotypic  (i.e.
+morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
 Because trait-based approaches allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
 Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10].
-With this mechanistic depth, the trait framework relates losses of ecosystem functions to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
+The trait framework relates losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15].
+This offers the mechanistic background to relate biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
@@
 Data may differ in taxonomic nomenclature and resolution (e.g.
-reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the methodology applied in measurements.
+reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements.
+These differences are not always documented in the metadata accompanying a dataset.
 All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
@@
 <!-- 3. trait databases have been developed, with standard structure and terms --> 
-To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
+To this end, the number of available trait datasets is increasing rapidly. 
+In the past, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
 PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
-These databases are then published as static data tables (i.e.
+<!--These databases are then published as static data tables (i.e.
 product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g.
 TRY, www.try-db.org).
-Furthermore, these portals map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies.
+Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies.
 As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers.
@@
 <!-- 4. distributed trait data must be integrated --> 
-While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, the number of decentralised trait datasets is increasing rapidly.
-This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15].
-As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
+Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15].
+As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
 <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
-It is likely that data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.
+It is likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.
 
@@
 
-In the field of comparative biology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been published in regular journal articles or books.
+In the field of comparative biology, morphological traits related to plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values for plants or ecological parameters of animals (e.g.
+reproductive traits, feeding biology, dispersal or body size) have been measured for decades, and have been published in regular journal articles or books.
 With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
@@
 For trait data, there are typical issues arising from the variability of data structures. 
-For instance, the column descriptions and terminology applied to taxa and traits are project specific, and rarely chosen to allow translation into larger database initiatives.
+For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific, and rarely chosen to allow translation into larger database initiatives.
 Furthermore, metadata varies in its detail, e.g.
@@
 <!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
-For example, by aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other trait data platform.
+For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait data platform.
 The *online database* enables selective data download and user permission and rights management.
@@
 Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12].
-Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13].
-Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g.
+Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13].
+Databases are also used to interpret OTUs in terms of their functional role [e.g.
 the KEGG orthology, @kanehisa12].
@@
 The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17].
-Nonetheless, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
+Nonetheless, initiatives for particular groups of animals, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), already exist.
 
@@
 
-Assigning an unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring.
+Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
+Assigning a more detailled and unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring.
 Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].
-Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
 
@@
 
-While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g.
-for taxa.
+While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
 To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.
@@
 More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e.
-trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e.
+trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e.
 traits) observed on a biological entity (i.e.
@@
 reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}
+a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire species or higher taxonomic group.}
 
@@
 It appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns.
+In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for a better mechanistic understanding of the relationship between drivers, communities and functions and providing new insights on global biodiversity patterns.
+Moreover it might be also a step towards a more predictive ecology as a broader set of available traits might enable more hypothesis based trait-based approaches. 
 In terms of data science, machine-readable, ontology-based data ease the application of big-data mining and machine-learning techniques.

### 2018-05-28 manuscript/manuscript.Rmd
finalize manuscript: ontologies, spell-checking, author list, microtypography

@@
 - name: Anton Güntsch \footnotemark[5]
+- name: Birgitta König-Ries\footnotemark[6]
 - name: Pete Manning \footnotemark[3]
-- name: Birgitta König-Ries\footnotemark[6]
 - name: Nadja K. Simons \footnotemark[7]
@@
 fontsize: 11pt
-geometry: margin=1.3in
+geometry: margin=1.4in
 output:
@@
 \footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
-\footnotetext[7]{Chair for Terrestrial Ecology, Department of Ecology and Ecosystemmanagement,
-Technische Universität München, Freising, Germany}
+\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
 
+\newpage
 \linenumbers
@@
 
-# Introduction
+\newpage
 
+\tiny
+
+Term | Definition 
+----- | ---------
+Term | A word that describes a particular *concept* as part of the specialised vocabulary of a field
+Concept | An idea, notion or object that is made explicit in an information context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
+Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a praticular context, while no unlisted entries are accepted
+Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*
+Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
+trait data), that is recognised by a large proportion of members of the application context
+Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte13] 
+Ontology | *Controlled vocabulary* that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
+defines a hierarchy of terms; thus a formal model of the objects and their relationships in a domain of interest [@gruber95]
+Semantic web | An extension of the world wide web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
+specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
+Dataset | A set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
+Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
+Database | A suite of data compiled from multiple *datasets*, i.e.
+from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software;
+Relational database | Usage in this paper: Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
+Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
+Identifier (ID) | A unique label that relates entries within and across *datasets*; is used to connect *data tables* into a *relational database*; can be user-specific or, as a *URI*, point to a globally valid *ontology* or *thesaurus*
+Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; an example is 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
+Webservice | An exchange protocol to access *online databases* directly and programmatically, i.e. by calls from a software tool
+Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
+client software and the *webservice* of an online databases; APIs are usually documented on the website of a database provider
+Online Portal | A website designed as a platform for the exchange of information, e.g.
+trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
+File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
+File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
+Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
+Darwin Core Standard (Dwc) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
+Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns.
+Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures. 
+Occurrence | A single observation instance of a taxon, i.e. an organism at a particular place at a particular time (http://rs.tdwg.org/dwc/terms/Organism)
+
+Table: Glossary of terms from the biodiversity data-management context as they are used in this paper; draws from @garnier17.
+
+<!--
+Measurement |  
+Aggregated Value |
+Value | 
+Fact | 
+-->
+
+\normalsize
+
+\newpage
 <!-- CP:  I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
@@
 
+# Introduction
 <!-- 1. Traits are important in ecology, evolution and conservation -->
@@
 <!-- 4. distributed trait data must be integrated --> 
-Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15].
+Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
 As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
@@
 -->
-\newpage
-
-\tiny
-
-Term | Definition 
------ | ---------
-Term | A word that describes a particular *concept* as part of the specialised vocabulary of a field
-Concept | An idea, notion or object that is made explicit in an information context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
-Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a praticular context, while no unlisted entries are accepted
-Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
-Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
-trait data), that is recognised by a large proportion of members of the application context
-Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte13] 
-Ontology | *Controlled vocabulary* that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
-defines a hierarchy of terms; thus a formal model of the objects and their relationships in a domain of interest [@gruber95]
-Semantic web | An extension of the world wide web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
-specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
-Dataset | A set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
-Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
-Database | A suite of data compiled from multiple *datasets*, i.e.
-from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software;
-Relational database | Usage in this paper: Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
-Identifier (ID) | A unique label that relates entries within and across *datasets*; is used to connect *data tables* into a *relational database*; can be user-specific or, as a *URI*, point to a globally valid *ontology* or *thesaurus*
-Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; an example is 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
-Webservice | An exchange protocol to access *online databases* directly and programmatically, i.e. by calls from a software tool
-Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
-client software and the *webservice* of an online databases; APIs are usually documented on the website of a database provider
-Online Portal | A website designed as a platform for the exchange of information, e.g.
-trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
-File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
-Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
-Darwin Core Standard (Dwc) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
-Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns.
-Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures. 
-Occurrence | A single observation instance of a taxon, i.e. an organism at a particular place at a particular time (http://rs.tdwg.org/dwc/terms/Organism)
-
-Table: Glossary of terms from the biodiversity data-management context as they are used in this paper; draws from @garnier17.
-
-<!--
-Measurement |  
-Aggregated Value |
-Value | 
-Fact | 
--->
-
-\normalsize
-
-\newpage
 
@@
 As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with standardisation. 
-Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g.
+Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g.
 www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
@@
 Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
-Assigning a more detailled and unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring.
-Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].
+Assigning a more detailed and unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring.
+Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that pre-date or ignore this standard [@kattge11a].
 
@@
 By providing this interlinkage of trait ontologies, a machine-readable web of definitions is spun across the Internet which allows researchers and search engines to relate independent trait measurements with each other and connect it to the wider *semantic web* of online data [@berners-lee01; @page08].
-The distinction of thesauri and ontologies is not truely binary.
+The distinction of thesauri and ontologies is not truly binary.
 Rather they mark idealised ends of a spectrum.
@@
 The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
-
 <!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
@@
 -->
-
 Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for their collaborative development [@walls12]. 
@@
 Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
+TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preferred for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
 The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11]. 
@@
 the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
-BExIS, http://bexis2.uni-jena.de/; Diversity Workbench, www.diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance.
+KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/; ) are likely to gain importance.
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
@@
 The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
-We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in the visibility and interconnectedness of published data.
+We propose the development of  tools and vocabularies that impose low thresholds and offer high pay-off in the visibility and interconnectedness of published data.
 
@@
 Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
-In literature data, summarized traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
+In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
 In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
-Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external terminology services (see above for resources).
+Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via URIs, which point to external terminology services (see above for resources).
 Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
@@
 In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
-"Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/".
+"Schneider, F.D., Jochum, M.,  Le Provost, G., Penone, C., Ostrowski, A.
+and Simons, N.K., 2017 Ecological Traitdata Standard v0.7, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.7/".
 In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
@@
 
-The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.
+The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
 The core terms describe minimal trait data according to the Entity-Quality model. 
@@
 These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence") that expand and refine terms of the Darwin Core Extensions (see below).
-This information may simply be added as extra columns to the core dataset.
-Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below). 
+For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence") that expand and refine terms of the Darwin Core Extensions (see below) which may simply be added as extra columns to the core dataset.
+Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below). 
 The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
-Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other Extensions of the Darwin Core.
+Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC.
 Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
@@
 For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.
-By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
+By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practised in TRY; @kattge11].
 This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.
@@
 Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-The data standard provides three extensions of the vocabulary that should be used to describe this information (Fig. 2 *d*):
+The data standard provides three extensions of the vocabulary that should be used to describe this information \mbox{(Fig. 2*d*)}:
 
-- The `Taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.
-- The `MeasurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.
+(http://ecologicaltraitdata.github.io/ETS/#extension-taxon)
+- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
 It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
@@
 literature source or expert reference).
-- The `Occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
+(https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact)
+- The `Occurrence` extension contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
 This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
+(https://ecologicaltraitdata.github.io/ETS/#extension-occurrence)
 
@@
 
-### Specification of traits
+### Specification of Metadata
 
-Wherever possible, the column `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
-If no globally available trait definition exists as a reference, trait datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits.
-Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g.
+Wherever possible, the column `traitID` should point to a publicly available, unambiguous trait definition in a published ontology. 
+If no globally available trait definition exists as an external reference, trait datasets should always be accompanied by a dataset-specific list of traits as part of the metadata or as an accompanying data table.
+Such a controlled vocabulary would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format of measured values or reported facts (e.g.
 units or legit factor levels).
-By providing a minimal vocabulary for trait thesauri (see https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset.
-The thesaurus might accompany the core data in a separate file (e.g.
-within a Darwin Core Archive), or be published on any other stable internet address (ideally providing stable URIs).
-<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
-See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards (https://ecologicaltraitdata.github.io/ETS/thesauri.html).
+Ideally, this definition refers to or refines terms from published trait ontologies.
+By providing a minimal vocabulary for trait lists (see https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the unambiguous definition of traits for trait datasets.
+This vocabulary might also prove useful for the future publication of trait ontologies. 
+<!--See our online reference for recommendations how to prepare a traitlist for  (https://ecologicaltraitdata.github.io/ETS/thesauri.html).
 We also provide an example thesaurus on Github (http://ecologicaltraitdata.github.io/TraitDataList/), which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project [@fischer10].
-
-### Specification of Metadata terms
+-->
 
@@
 
-The following example harmonises traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:
+The R-package further supports the mapping of trait names to a list of trait definitions and identifiers ( this lookup table is cast into an own object class called 'thesaurus'). 
+The following example harmonises traits based on a minimal list, referencing trait names with globally valid URIs provided by the BETSI thesaurus of soil invertebrate traits:
 
@@
 3).
-Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`).
-Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies.
+Other ways of defining a 'thesaurus' object are documented in the package vignette and function documentation (`?as.thesaurus`).
+Future iterations of the R package will aim at automatising the generation of thesaurus objects from globally available ontologies.
 The package functions form a tool-chain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing trait data as simple as:
+A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonising trait data as simple as:
 
@@
 To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions. 
-However, these distributed data are heterogeneous in form and description and initiatives to harmonise and compile these data require significant amounts of funding and personel. 
+However, these distributed data are heterogeneous in form and description and initiatives to harmonise and compile these data require significant amounts of funding and personnel. 
 To support the long-term rewards of standardisation efforts, incentives should be sought to mitigate the cost of readying trait data for the 'semantic web' of biodiversity data and knowledge.
@@
 
-We thank the managers of the three Exploratories, Kirsten Reichel-Jung, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project
-infrastructure; Christiane Fischer and Jule Mangels for giving
-support through the central office, andMarkus Fischer, Eduard Linsenmair,
+We thank the managers of the three Exploratories, Kirsten Reichel-Jung, Katrin Lorenzen, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project infrastructure;
+Christiane Fischer and Jule Mangels for giving
+support through the central office, and Markus Fischer, Eduard Linsenmair,
 Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
@@
 
+The Appendix A contains a static excerpt of the living spreadsheet on existing trait datasets and databases, which can be found at https://goo.gl/QxzfHy.
+
 The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+
+The development website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform.
+<!--An example of a published trait thesaurus for the Biodiversity Exploratories (www.biodiversity-exploratories.de) can be found at https://ecologicaltraitdata.github.io/TraitDataList/.-->
 
@@
 
-The development website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform.
-
-An example of a published trait thesaurus for the Biodiversity Exploratories (www.biodiversity-exploratories.de) can be found at https://ecologicaltraitdata.github.io/TraitDataList/.
-
-The living spreadsheet on existing trait datasets and databases can be found at https://goo.gl/QxzfHy.
-
 # References
 
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{5pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}
+
 \footnotesize

### 2018-05-28 manuscript/appendix_A.Rmd
fix appendix A author order

@@
 - name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
+- name: Malte Jochum \footnotemark[4]
+- name: Gaëtane Le Provost \footnotemark[3]
+- name: Andreas Ostrowski \footnotemark[6]
+- name: Caterina Penone \footnotemark[4]
 - name: David Fichtmüller \footnotemark[5]
+- name: Martin Gossner \footnotemark[8]
 - name: Anton Güntsch \footnotemark[5]
-- name: Malte Jochum \footnotemark[4]
 - name: Birgitta König-Ries\footnotemark[6]
-- name: Gaetane Le Provost \footnotemark[3]
 - name: Pete Manning \footnotemark[3]
-- name: Andreas Ostrowski \footnotemark[5]
-- name: Caterina Penone \footnotemark[4]
 - name: Nadja K. Simons \footnotemark[7]
@@
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[7]{Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}
+\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
+\footnotetext[7]{Chair for Terrestrial Ecology, Department of Ecology and Ecosystemmanagement,
+Technische Universität München, Freising, Germany}
 
 \newpage
+
+\renewcommand{\thetable}{A\arabic{table}}
 

### 2018-05-29 manuscript/appendix_A.Rmd
fix author names and substitut DOI to data standard v0.8

@@
 - name: David Fichtmüller \footnotemark[5]
-- name: Martin Gossner \footnotemark[8]
+- name: Martin M. Gossner \footnotemark[8]
 - name: Anton Güntsch \footnotemark[5]
@@
 \footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
-\footnotetext[7]{Chair for Terrestrial Ecology, Department of Ecology and Ecosystemmanagement,
-Technische Universität München, Freising, Germany}
+\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
 

### 2018-05-29 manuscript/manuscript.Rmd
fix author names and substitut DOI to data standard v0.8

@@
 "Schneider, F.D., Jochum, M.,  Le Provost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2017 Ecological Traitdata Standard v0.7, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.7/".
+and Simons, N.K., 2018 Ecological Traitdata Standard v0.8, DOI:  10.5281/zenodo.1255287, URL: https://ecologicaltraitdata.github.io/ETS/v0.8/".
 In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
@@
 Ideally, this definition refers to or refines terms from published trait ontologies.
-By providing a minimal vocabulary for trait lists (see https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the unambiguous definition of traits for trait datasets.
+By providing a minimal vocabulary for trait lists (see https://ecologicaltraitdata.github.io/ETS/#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
 This vocabulary might also prove useful for the future publication of trait ontologies. 

### 2018-09-10 manuscript/manuscript.Rmd
revise MS for comments of TRY and OpenTraits

@@
 2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
-3.
-In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a general and simple vocabulary as well as a simple data structure for storing and sharing ecological trait data.
+3. In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a standardised vocabulary to be used for storing and sharing ecological trait data. 
 4.
-Additionally, we provide an R-package that enables the transformation of any tabular dataset into the proposed format.
-This also allows trait datasets from heterogeneous sources to be harmonised and merged, thus facilitating data compilation for any particular research focus.
+This vocabulary allows trait datasets from heterogeneous sources to be merged more easily into databases, and facilitates the creation of data interfaces between software tools for trait-data handling and analysis.
 5.
-With these decentralised tools for trait-data harmonisation, we intend to facilitate the exchange and analysis of trait data within ecological research and enable global syntheses of traits across a wide range of taxa and ecosystems.
+By enabling decentralised trait-data standardisation, we intend to facilitate the exchange and analysis of trait data for a broader ecological research community and enable global syntheses of traits across a wide range of taxa and ecosystems.
 
@@
 
-Many issues in trait-based research arise when compiling datasets from several sources.
+Many issues in trait-based research arise when compiling datasets from multiple sources.
 Data may differ in taxonomic nomenclature and resolution (e.g.
@@
 TRY, www.try-db.org).
-Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies.
+Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, may also offer access control and data usage policies.
 As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers.
@@
 <!-- 4. distributed trait data must be integrated -->
+
 Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
@@
 We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users.
-Based on these considerations, we propose a minimal structure and vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data.
-Finally, we present an R package that assists the harmonisation of trait data from distributed sources.
-With this easy-to-use terminology and toolset, we hope to  convince trait-data providers and trait-data users about the general importance of trait-data standardisation and lay out the roadmap towards an accessible ecological trait data standard.
+Based on these considerations, we propose a versatile vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data.
+Since this standard vocabulary relies on the adoption of a broad research community we discuss incentives for its use and lay out mechanisms for a community-based future development towards an accessible ecological trait data standard.
+<!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
 
@@
 FDS: see Garnier et al. but may contain: 
-
 -->
@@
 Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g.
-Figshare.com), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
+Figshare.com, datadryad.org, Researchgate.net, or Zenodo.org), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
 These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management.
@@
 .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
-Most importantly, these platforms enable public hosting of data with very low thresholds for *metadata* documentation and data standardisation. 
-<!-- As a result, datasets greatly vary in structure and terminology. -->
+Most importantly, these platforms enable public hosting of data with very low thresholds regarding *metadata* documentation and data standardisation.
+As a result, published datasets greatly vary in structure and nomenclature.
 
@@
 A global overview of existing trait data for all taxa and trait types is difficult to obtain.
-Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list).
+Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for an excerpt of this list).
 We invite data owners and users to add further trait datasets to this spreadsheet.
@@
 Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
-Just as the individual datasets described above, many of these databases are published as static data tables on low-threshold file hosting platforms and are updated irregularly.
+Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms and may be updated irregularly.
 
-As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with standardisation. 
+As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
 Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g.
 www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
-For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16].
-The resulting data are published as versioned data tables which are updated as new data sources become available.
+For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16].
+The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).
 
@@
 <!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
-For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait data platform.
+For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform.
 The *online database* enables selective data download and user permission and rights management.
 As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+
 Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12].
@@
 The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17].
-Nonetheless, initiatives for particular groups of animals, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), already exist.
+Nonetheless, initiatives for particular groups of animals do exist, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
 
-Regarding open access, few of these centralised databases comply with the criteria demanded by journals and funding agencies for primary data publication.
 The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship. 
-With the proactive turn towards open access data (as stated in the Bouchot Declaration; http://www.bouchoutdeclaration.org/), it may be necessary to find other incentives for data submission.
+However, the upload of data to these centralised databases does currently not fulfill the criteria demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions.
+With data being publicly available, the role of database initiatives in standardising data and making them more accessible is paramount.
 
@@
 These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies these terms.
 These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
@@
 An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12].
+For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies the terms described therein.
 <!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
@@
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data publications and standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+One goal of such initiatives is to facilitate data usage in publications and provide standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+Another solution for data-users to access trait data in a structured way is offered by FENNEC [@ankenbrand18], an online tool or self-hosted service to extract trait information from multiple sources for a target species community.
 <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->
@@
 database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.
-Low-threshold repositories offer the hosting of scientific primary data attracting a wealth of heterogeneous trait datasets, but data harmonisation of these distributed data sets is currently laborious.
+On the other end, public repositories on file-hosting services offer the publication of scientific primary data according to current funding and journal criteria, attracting a wealth of heterogeneous trait datasets.
+Due to the low thresholds concerning data standardisation, compilation of these distributed data sets is currently laborious.
 The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
@@
 The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
-Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the semantic web.
+Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases:
+
+1.
+in  two-dimensional datasets, e.g.
+for labelling project-specific data, for local use and exchange with collaborators, or for publication of standardised primary data on open-access data repositories,
+2. as a consensus vocabulary when compiling data from distributed sources into databases, e.g. to map standardised columns and refer to taxa and trait definitions in a uniform way, and
+3. as an exchange format of data between web services and software tools, e.g. when providing database queries from a webservice or defining input and output formats of software packages.
+
 By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable.
@@
 In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We paid particular consideration to the work of @kattge11, @kattge11a, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
 Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.
-Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. 
-The vocabulary proposed is intended to form the foundations of a standard nomenclature that can be expanded and corrected by the wider community of researchers using trait-based approaches in ecology.
+<!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
+The vocabulary proposed is intended to form the foundations of a standard exchange format that can be expanded and revised by the wider community of researchers using trait-based approaches in ecology.
 <!--
@@
 
-## How to apply the standard vocabulary
-
-We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible.
-The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
-quality) observed on an individual or population (i.e.
-entity), of a biological taxon.
-When applying the vocabulary, it is implicit to use a two-dimensional observation long-table format for the data (Fig.
-2 *b*), rather than a species$\times$traits matrix (Fig.
-2 *a*).
-As the long-table format draws from a defined set of columns, merging datasets is easier.
-Long-table datasets also purport multiple advantages for data manipulation [e.g.
-filtering, sub-setting and aggregating data, @wickham14].
-
-Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary (Fig. 2 *c* & *d*).
-For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset.
-Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
-In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
-In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
-Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via URIs, which point to external terminology services (see above for resources).
-Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
-As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16.
-The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
-The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
-
-\end{figure}
-
-For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
-UTF-8 or ASCII).
-
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
-"Schneider, F.D., Jochum, M.,  Le Provost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2018 Ecological Traitdata Standard v0.8, DOI:  10.5281/zenodo.1255287, URL: https://ecologicaltraitdata.github.io/ETS/v0.8/".
-In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
-via the API; **in preparation!**).
-Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.
-
 ## Terms of the standard vocabulary
@@
 The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
-The core terms describe minimal trait data according to the Entity-Quality model. 
+The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
+quality) observed on an individual or population (i.e.
+entity), of a biological taxon.
 Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information).
-These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence") that expand and refine terms of the Darwin Core Extensions (see below) which may simply be added as extra columns to the core dataset.
+These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
+For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see below for best practice when publishing primary data).
 Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).
+
+Some of the terms expand and refine terms of the Darwin Core Extensions (http://rs.tdwg.org/dwc/terms/) or Glossary of EOL (http://eol.org/info/516).
+Wherever our glossary refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
+Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns.
+However, these should be clearly specified in the meta-information accompanying the dataset.
 The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
@@
 -->
-These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets (published separately or as part of a Darwin Core Archive), thus avoiding redundancy and duplication of content.
+These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content.
 A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
@@
 
-# Computational tools for producing compliant data
+## Best practice for primary data publication
 
-To access data from public databases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
-However, none of these packages provide the option to harmonisation trait data into a unified scheme.
-To close this gap, we developed the R package 'traitdataform', which assists the production of data compliant with the trait data standard proposed above. 
-There are two major use cases for the package:
+We focus this section on the first of the three use-cases for the ecological trait-data standard vocabulary: The publication of project-specific primary data.
+We assume that data managers and software developers who apply the vocabulary in use-case 2 and 3 are more fluent with the use of exchange formats and standard definitions and have established own best-practice principles.
 
-1. preparing trait datasets for publication on public hosting services and project databases, and
-2. automating the harmonisation of trait datasets from different sources by moulding them into a unified format.
+When publishing primary data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation.
+Thus, ideally, data are published as they have been recorded, i.e.
+in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard.
+This is true for the core terms, i.e.
+the trait value (the quality) and taxon name (entity), but also for any co-variates reported along with it.
+In the case of the core terms, data providers should use the '\*Std'-terms for standardised entries.
+For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g.
+elevation above sea-level.
+If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g.
+by appending '\*Orig' to the column name.
+Also, additional co-variates, e.g.
+on environmental growth conditions of the specimen, should be provided in user-defined columns.
+Any user-defined columns must be adequately defined in the accompanying meta-information.
 
-A comprehensive documentation of the package can be found on its Github repository (https://github.com/EcologicalTraitData/traitdataform) and the documentation website (http://EcologicalTraitData.github.io/traitdataform/).
-The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
+When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of reduced complexity. 
+For spreadsheets, it is recommended to use a observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). 
+As the long-table format draws from a defined set of columns, merging datasets is easier.
+Long-table datasets also purport multiple advantages for data manipulation [e.g.
+filtering, sub-setting and aggregating data, @wickham14].
 
-The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement long-table format (Fig.
-3).
-This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object.
-This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:
+Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary (Fig. 2 *c* & *d*).
+For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset.
+Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
+In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
+In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
+Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via URIs, which point to external terminology services (see above for resources).
+Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
+As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16.
+The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
+The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.
 
-```
-library("traitdataform")
-dataset <- as.traitdata(read.csv("path/to/data.csv"), 
-             traits = c("body_length", "antenna_length", 
-                        "metafemur_length"),
-             units = "mm",
-             taxa = "name_correct", 
-             keep = c(locationID = "location")
-             )
-
-```
-
-\begin{figure}[h]
+\begin{figure}[t]
 \centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
-\caption{Process chart of the functions provided within the R package 'traitdataform' to apply the standard vocabulary to any trait-data table.}
+\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
+\caption{Formats used for trait datasets.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
 
@@
 
-The parameter 'traits' lists column names that contain trait values.
-The column containing taxon names is given in parameter 'taxa'.
-Note that the parameter 'keep' specifies and renames any data that should be maintained in the output.
-The parameter 'units' is used to specify the input units of measurement.
-In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data  (Fig.
-3).
-<!-- BKR: In the example below, this looks like a manual matching.
-What is done automatically here and which information does the user have to provide?
-FDS: taxon is matched automatically, traits not.
-lack of trait APIs and webservices.
--->
-<!-- Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e.
-a thesaurus of traits.
-The same function harmonises trait values by converting values into target units and unified factor levels.
-<!-- BKR: Have you thought about linking to unit ontologies here?
-AO: This paper is a good reference and gives an idea of which ontology could be used;
-http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
-FDS: might be referred to in the section about thesauri -->
+For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
+UTF-8 or ASCII).
 
-The R-package further supports the mapping of trait names to a list of trait definitions and identifiers ( this lookup table is cast into an own object class called 'thesaurus'). 
-The following example harmonises traits based on a minimal list, referencing trait names with globally valid URIs provided by the BETSI thesaurus of soil invertebrate traits:
+In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
+"Schneider, F.D., Jochum, M.,  Le Provost, G., Penone, C., Ostrowski, A.
+and Simons, N.K., 2018 Ecological Traitdata Standard v0.8, DOI:  10.5281/zenodo.1255287, URL: https://ecologicaltraitdata.github.io/ETS/v0.8/".
+In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
+Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
+via the API; **in preparation!**).
 
-```
-traitlist <- as.thesaurus(
-   body_length = as.trait("body_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length"), 
-   antenna_length = as.trait("antenna_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Antenna_length"),
-   metafemur_length = as.trait("metafemur_length", 
-     expectedUnit = "mm", valueType = "numeric", 
-     identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Femur_length")
-)
+# Discussion
 
-datasetStd <- standardize.traits(dataset, thesaurus = traitlist)
-```
+To serve the demand for standardisation and harmonisation of ecological trait data that arises from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
+In the following paragraphs we want to discuss the incentives and impediments for individual researchers, and highlight the long-term pay-off and benefit for the community in adopting a standard terminology for trait-based ecological research.
 
-The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`  (Fig.
-3).
-Other ways of defining a 'thesaurus' object are documented in the package vignette and function documentation (`?as.thesaurus`).
-Future iterations of the R package will aim at automatising the generation of thesaurus objects from globally available ontologies.
-The package functions form a tool-chain where each function can be piped as an input into the next.
-A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonising trait data as simple as:
+<!-- community effort and consensus -->
+While it is clear, that a single vocabulary standard is not able to fullfill the needs of all research fields working with traits, we believe that it captures the a common core concept of trait data. 
+Consensus building on how traits are to be used and evaluated is currently under way in several fields with their taxonomic focus and project-specific questions.
+While these community discussions on trait definitions and measurement practice is leading to a better quality of data, naturally, the heterogeneity in trait-data quality and detail across fields will remain.
+Nonetheless, we provide a starting point and invite these communities to develop a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research, and to discuss and define terms and extensions to fit their specific needs.
+The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio terminology service.
 
-```
-datasetStd <- standardize(read.csv("path/to/data.csv"),
-            thesaurus = traitlist,
-            taxa = "name_correct", 
-            units = "mm"
-            )
-```
+<!-- investment and pay-off, incentives -->
+For individual researchers and small research projects, the investment of time and resources in trait-data standardisation before publication seems not to be yielding a direct benefit.
+Certainly, data released according to consensus standards, especially if published under open-access licenses, will be more easily re-used in compilations and synthesis studies. 
+A well-defined vocabulary for metadata will also ensure that authorship and license model are appropriately documented along the data life-cycle.
+However, for these incentives to take positive effect, data publications and data citations must become a valid scientific contribution in the community recognition and professional evaluation of individual researchers [@costello09; @roche15].
 
-Datasets that have been produced by these functions can easily be appended using the function `rbind()` of R base, while maintaining any available metadata information as separate column entries. 
-To merge datasets with additional information on the occurrence or measurement level, secondary data tables can be added as columns of the core dataset according to a unique identifier using the function `merge()`.
-This enables an easy handling of data sources that originate in a relational database format.
+<!-- Data quality issues --> 
+On the community level, shifting the task of standardisation from the data user side to the data owner side yields great gain in accurracy and reduces the risk of misinterpretation.
+For instance, length measurement results depend very much on the methodology and landmarks used and often systematic biases could be corrected for when converting it to a consensus definition.
+The detailled knowledge might not be delivered along with the data and a post-publication standardisation is inhibited.
+On the  other hand, plausibility checks and evaluation of statistical methods, e.g.
+for aggregating trait values to the species level, can only be done in comparison across a wide array of datasets.
+Currently, these "big data" volumes are only available in centralised databases.
+To establish a best-practice of data aggregation, an exploration and evaluation of different methods should be subject to a community discussion. 
+This is only possible with large quantities of distributed data being available in a harmonised way. Our ecological trait-data standard vocabulary facilitates such a community-driven comparison.
 
-Since the intention of the package is also to simplify the harmonisation of published trait data, the package offers direct access to trait datasets that have been released in the Public Domain or under Creative Commons licenses.
-<!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
+A wide adoption of our trait-data standard vocabulary would allow to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for a better mechanistic understanding of the relationship between drivers, communities and functions and providing new insights on global biodiversity patterns.
+Moreover it might be also a step towards a more predictive ecology as a broader set of available traits might enable new hypothesis-driven approaches. 
+However, to enable a global synthesis across trait datasets, there remains a need to advance the development of trait thesauri into more interrelated and complete ontologies.
+
+<!-- toolchain and semantic web -->
+To support the long-term rewards of data standardisation efforts, incentives should be sought to mitigate the cost of data providers for readying trait data for the 'semantic web' of biodiversity data.
+This can be software tools or supporting infrastructures.
+The R-package 'traitdataform' [@schneider18] is assisting the conversion of own data into compliant formats. 
+Similarly, the use and analysis of trait-data can be much facilitated by software packages for different tasks. 
+For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
+The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
+The package 'NatDB' [@pearse] collates distributed trait data and harmonizes them into a unified format.
+
+Our standard vocabulary is an important building block for a unified mode of data exchange between webservices and software packages and thus facilitates the development of a software toolchain for the trait-data lifecycle.
+Having well-defined terms is also a key precondition for developing exchange formats between large database initiatives and biodiversity data archives. 
+Even further downstream, machine-readable, ontology-based data will ease the application of big-data mining and machine-learning techniques. 
+Without clearly defined terms and concepts, computational handling of large amounts of trait-data will be impossible.
 
@@
 
-<!--The practical importance of these frameworks for data providers and data users is often not obvious and not appropriately considered when storing or uploading data to low-threshold  file-publishing services.
+To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions.
+These distributed data are heterogeneous in form and description and initiatives to harmonise, compile and analyse these data require significant amounts of funding and personnel.
+By using a constrained vocabulary with globally accessible definitions of terms,  distributed trait data can be accessed more easily by other researchers and harmonised into aggregated datasets.
 
-The investment is only paying out in the long run. Citations of trait datasets are higher when they are easier to access and relate to other data due to standard structures and vocabularies. 
-Along with the push towards open access publication of primary data, the scientific community must come up with ways to incentivize and require data standardization.
-
-More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.-->
-To serve the demand for simple ways to standardise and harmonise ecological trait data, we propose a versatile vocabulary for simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
-With the R-package 'traitdataform', we also present a toolbox in R to transfer and harmonise data into this scheme.
-
-It appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.
-In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for a better mechanistic understanding of the relationship between drivers, communities and functions and providing new insights on global biodiversity patterns.
-Moreover it might be also a step towards a more predictive ecology as a broader set of available traits might enable more hypothesis based trait-based approaches. 
-In terms of data science, machine-readable, ontology-based data ease the application of big-data mining and machine-learning techniques.
-
-To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions. 
-However, these distributed data are heterogeneous in form and description and initiatives to harmonise and compile these data require significant amounts of funding and personnel. 
-To support the long-term rewards of standardisation efforts, incentives should be sought to mitigate the cost of readying trait data for the 'semantic web' of biodiversity data and knowledge.
-This can be software tools or supporting infrastructures.
-The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories. 
-By using a constrained vocabulary with globally accessible definitions of terms,  distributed trait data can be accessed more easily by other researchers and harmonised into aggregated datasets.
-Also, it will ease the exchange of data between databases and facilitate the development computational methods and software tools that access and handle the data, based on the standard vocabulary. 
-We also encourage the advancement of trait thesauri into more interrelated and complete ontologies. 
-The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits.
+The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation and publication of data.
 This requires significant effort, but it returns  great scientific benefit by enabling synthesis on our general understanding of biodiversity and ecosystem function.
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Kristin Bohn, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts, the trait data standard vocabulary and the pre-releases of the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Kristin Bohn, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts and pre-print, the trait data standard vocabulary and the pre-releases of the R-package.
+We are grateful to Brian Enquist, Rachel Galagher and Brian Maitner for organising the OpenTraits workshop in New Orleans, USA, in August 2018.
 

### 2018-09-13 manuscript/manuscript.Rmd
update manuscript and biblio for internal review

@@
 
+<!-- 
+The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:
+
+Point 1: set the context for and purpose of the work;
+
+Point 2: indicate the approach and methods;
+
+Point 3: outline the main results;
+
+Point 4: identify the conclusions and the wider implications.
+-->
+
 1.
@@
 2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
-3. In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a standardised vocabulary to be used for storing and sharing ecological trait data. 
+3.
+In order to facilitate the standardisation and harmonisation of distributed trait datasets on the data-provider side, we propose a standardised vocabulary to be used for storing and sharing ecological trait data.
+We discuss impediments and incentives for a wide adoption of such a standard by data-providers.
 4.
-This vocabulary allows trait datasets from heterogeneous sources to be merged more easily into databases, and facilitates the creation of data interfaces between software tools for trait-data handling and analysis.
-5.
-By enabling decentralised trait-data standardisation, we intend to facilitate the exchange and analysis of trait data for a broader ecological research community and enable global syntheses of traits across a wide range of taxa and ecosystems.
+The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases, and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
+By aiding decentralised trait-data standardisation, our data standard facilitates the use of trait data for a broader ecological research community and enables global syntheses across a wide range of taxa and ecosystems.
 
@@
 
-Functional traits are phenotypic  (i.e.
-morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07].
-Because trait-based approaches allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
-Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10].
-The trait framework relates losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15].
-This offers the mechanistic background to relate biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
+Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
+Recent years have seen a proliferation of trait-based research in a wide range of fields. 
+Traits bridge the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and relate losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
+This offers the mechanistic background for linking biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
  <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
@@
 And Coux et al. 2016 Linking species functional roles to their network roles
-?
--->
+? -->
 <!--
@@
 @reich97].
-<!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ?
-I’m not sure that the literature cited is always appropriate here ?
--->
+<!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].-->
 <!--
@@
 -->
-<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions.
-Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->
 
-<!-- 2. assemble trait data into databases -->
+<!-- 2. trait databases have been developed, with standard structure and terms --> 
+However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
+To enable such functional ecological research, trait data have been harmonized and compiled into centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
+PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
+The heterogeneity of data from different research contexts <!-- i.e.
+differences in taxonomic nomenclature and resolution (e.g.
+reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements--> render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
+Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
+These include software to access trait data from the internet [e.g.
+@chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
 
-Many issues in trait-based research arise when compiling datasets from multiple sources.
-Data may differ in taxonomic nomenclature and resolution (e.g.
-reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements.
-These differences are not always documented in the metadata accompanying a dataset.
-All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
-However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
+<!-- 4. distributed trait data must be integrated --> 
+Meanwhile, open science reaches the mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
+As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services (e.g.
+Figshare.com, datadryad.org, Researchgate.net, or Zenodo.org), where no data standards are forced upon the uploaded material [@wilkinson16].
+It is thus likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient reuse and synthesis of published datasets.
 
-<!-- 3. trait databases have been developed, with standard structure and terms --> 
-To this end, the number of available trait datasets is increasing rapidly. 
-In the past, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
-PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
-<!--These databases are then published as static data tables (i.e.
-product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g.
-TRY, www.try-db.org).
-Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, may also offer access control and data usage policies.
-As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers.
-Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
-These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
-<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
-<!-- 4. distributed trait data must be integrated -->
-
-Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
-As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material.
-<!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
-It is likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.
-
-In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation.
-We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users.
+In this paper, we review existing trait databases and online portals, as well as initiatives for data standardisation from a pragmatic view of data providers and data users.
+We discuss the current efforts for making trait data visible, accessible and reuseable in downstream data analysis, as demanded by the FAIR principles for scientific data [@wilkinson16].
+We show how the current deficit in standardisation hampers the implementation of interoperability and reuse of trait data.
 Based on these considerations, we propose a versatile vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data. 
@@
 <!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
-
 <!-- FDS: Why 'ecological' trait data?
@@
 
-<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists
+# Initiatives for trait-data standardisation
 
-FDS: see Garnier et al. but may contain: 
--->
-
-# A review of initiatives for trait-data standardisation
-
-In this section, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):
+We review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):
 
@@
 Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].
+As it stands, the decentralisation and the lack of data standardisation of data in online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be reused and combined into broad synthesis analysis.
 
@@
 
-A global overview of existing trait data for all taxa and trait types is difficult to obtain.
+<!--A global overview of existing trait data for all taxa and trait types is difficult to obtain.
 Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for an excerpt of this list).
 We invite data owners and users to add further trait datasets to this spreadsheet.
-As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.
+-->
+<!-- FDS: this might be a task for community  building and decentralised registries, as discussed on the OpenTraits platform, e.g.
+FENNEC or MAD package do that, or a registry for earning badges of data quality would achieve that as a by-product.
+I think we should not aim for solving this registry problem here.
+-->
 
@@
 In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Appendix A table A1].<!--[@wright04; ] -->
+@klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Supplementary Material A, table A1].<!--[@wright04; ] -->
 While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
@@
 Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g.
-www.markmybird.org), data on body measurements are likely to grow exponentially in the near future.
+www.markmybird.org), data on body measurements are likely to skyrocket in the near future.
 For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16].
@@
 The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17].
-Nonetheless, initiatives for particular groups of animals do exist, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).
+Nonetheless, initiatives for particular groups of animals do exist: for instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), or the Coral Trait Database compiles ecological data on coral observations [@madin16].
 
-The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship. 
-However, the upload of data to these centralised databases does currently not fulfill the criteria demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions.
-With data being publicly available, the role of database initiatives in standardising data and making them more accessible is paramount.
+The role of database initiatives in standardising data and making them more accessible is paramount. 
+The trait-data platforms incentivise data submissions by offering increased data visibility and usage, while providing data-use policies that secure author attribution and, potentially, co-authorship.
+However, the upload of data to these centralised databases does often not fulfill all of the criteria that are increasingly demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions.
 
@@
 the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
-KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/; ) are likely to gain importance.
+KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to facilitate data usage in publications and provide standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.
+One goal of such initiatives is to provide standardisation of data and facilitate data usage, for instance by mapping unambiguous terminologies and ontologies for biodiversity data, and clarify conditions of data re-use.
 Another solution for data-users to access trait data in a structured way is offered by FENNEC [@ankenbrand18], an online tool or self-hosted service to extract trait information from multiple sources for a target species community.
@@
 
-## Conclusion of review
+## Summing-up
 
@@
 ontologies and data standards) and platforms for data management (e.g.
-database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.
+database and data management platforms) provide great visibility, accessability and improved interoperability of datasets, but raise relatively high thresholds for data and metadata preparation.
 On the other end, public repositories on file-hosting services offer the publication of scientific primary data according to current funding and journal criteria, attracting a wealth of heterogeneous trait datasets.
 Due to the low thresholds concerning data standardisation, compilation of these distributed data sets is currently laborious.
-The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
-We propose the development of tools and vocabularies that impose low thresholds and offer high pay-off in the visibility and interconnectedness of published data.
+The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by raising awareness for data standardisation, accessibility and re-useability on the side of data providers.
+By establishing this as a best practice in trait-based research, trait-data will eventually fullfill the FAIR guiding principles for scientific data [@wilkinson16].
+In the following sections, we propose the development of vocabularies that lower the thresholds and offer high pay-off in the visibility and reuse of published data.
 
@@
 As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
+The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
 The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
@@
 It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.
+By adopting the standard vocabulary, trait-data will fulfill the FAIR principles for scientific data [@wilkinson16].
+We describe best practice principles and give examples for the application of the vocabulary in Supplementary Material B.
 
@@
 We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
-Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.
 <!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
-The vocabulary proposed is intended to form the foundations of a standard exchange format that can be expanded and revised by the wider community of researchers using trait-based approaches in ecology.
 <!--
@@
 
-## Terms of the standard vocabulary
-
-The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
 The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
@@
 entity), of a biological taxon.
-Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information).
-These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see below for best practice when publishing primary data).
+<!--Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information).
+These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, -->
+We offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see sections below for extensions and best practice when publishing primary data).
 Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).
 
-Some of the terms expand and refine terms of the Darwin Core Extensions (http://rs.tdwg.org/dwc/terms/) or Glossary of EOL (http://eol.org/info/516).
-Wherever our glossary refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
+Some of the terms expand and refine terms of the Darwin Core and its Extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516).
+Wherever our vocabular refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
 Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns.
@@
 Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC. 
+The vocabulary proposed is intended to form the base of a standard exchange format.
 Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
 
-### Specification of core terms
+## Specification of core terms
 
@@
 
-### Extensions for additional data layers
+## Extensions for additional data layers
 
@@
 
-Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
-<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).
+<!--Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
+ If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).
 Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level.
@@
 
-### Specification of Metadata
+## Specification of Metadata
 
-Wherever possible, the column `traitID` should point to a publicly available, unambiguous trait definition in a published ontology. 
+Wherever possible, the column `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
 If no globally available trait definition exists as an external reference, trait datasets should always be accompanied by a dataset-specific list of traits as part of the metadata or as an accompanying data table.
@@
 in a separate metadata file, possibly applying the ecological metadata language, EML).
-In cases where individual data from different sources are compiled into a trait database, these information must be provided at the measurement level.
+In cases where individual data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain the reference to the original data provider and conditions of re-use of these data.
 This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
@@
 
-When publishing primary data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation.
-Thus, ideally, data are published as they have been recorded, i.e.
+Data publications of traits should apply the FAIR principles of scientific data management and aim at findability, accessibility, interoperability and reusability [@wilkinson16].
+By drawing column names and applying definitions of the ecological trait-data standard vocabulary, and providing reference to this standard in the metadata, authors can ensure that these principles are followed.
+
+Above all, when publishing primary data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation.
+Thus, data should be published as they have been recorded, i.e.
 in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard.
 This is true for the core terms, i.e.
-the trait value (the quality) and taxon name (entity), but also for any co-variates reported along with it.
-In the case of the core terms, data providers should use the '\*Std'-terms for standardised entries.
+the trait value (the quality) and taxon name (the entity), but also for any co-variates reported along with it.
+For the core terms, data providers should use the '\*Std'-terms for standardised entries.
 For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g.
@@
 
-When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of reduced complexity. 
+When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of simplicity. 
 For spreadsheets, it is recommended to use a observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). 
@@
 In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
-In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
+In larger compilations, a 'datasetID' allow to trace data origin to the primary source and keep track of authorship and original reference. 
 Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via URIs, which point to external terminology services (see above for resources).
@@
 UTF-8 or ASCII).
+The metadata should contain unambiguous information on authorship and conditions for re-use of the data, ideally by specifying a data license  (e.g.
+Creative Commons licenses) or public domain dedication.
 
@@
 via the API; **in preparation!**).
+To enable reuse in downstream analysis, the metadata should further specify a license, author and bibliographic citation that applies to the dataset.
+
+Supplementary Material B contains three worked examples of how to apply the vocabulary to own trait-datasets.
 
@@
 <!-- community effort and consensus -->
-While it is clear, that a single vocabulary standard is not able to fullfill the needs of all research fields working with traits, we believe that it captures the a common core concept of trait data. 
+While it is clear, that a single vocabulary standard is not able to fullfill the needs of all research fields working with traits, we believe that our proposal captures a common core concept of trait data.
 Consensus building on how traits are to be used and evaluated is currently under way in several fields with their taxonomic focus and project-specific questions.
-While these community discussions on trait definitions and measurement practice is leading to a better quality of data, naturally, the heterogeneity in trait-data quality and detail across fields will remain.
+While these community discussions on trait definitions and measurement practice are leading to a better quality of data, naturally, the heterogeneity in trait-data quality and detail across fields will remain.
 Nonetheless, we provide a starting point and invite these communities to develop a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research, and to discuss and define terms and extensions to fit their specific needs.
@@
 <!-- Data quality issues --> 
-On the community level, shifting the task of standardisation from the data user side to the data owner side yields great gain in accurracy and reduces the risk of misinterpretation.
+On the community level, shifting the task of standardisation from the data-user side to the data-owner side yields great gain in accurracy and reduces the risk of misinterpretation.
 For instance, length measurement results depend very much on the methodology and landmarks used and often systematic biases could be corrected for when converting it to a consensus definition.
@@
 This can be software tools or supporting infrastructures.
-The R-package 'traitdataform' [@schneider18] is assisting the conversion of own data into compliant formats. 
+The R-package 'traitdataform' [@schneider18a] is assisting the conversion of own data into compliant formats. 
 Similarly, the use and analysis of trait-data can be much facilitated by software packages for different tasks. 
@@
 The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
-The package 'NatDB' [@pearse] collates distributed trait data and harmonizes them into a unified format.
+The package 'NatDB' [@pearse] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data.
 
@@
 Having well-defined terms is also a key precondition for developing exchange formats between large database initiatives and biodiversity data archives. 
-Even further downstream, machine-readable, ontology-based data will ease the application of big-data mining and machine-learning techniques. 
-Without clearly defined terms and concepts, computational handling of large amounts of trait-data will be impossible.
+Even further downstream, readying the primary data for the semantic web via references to ontologies and data standards will ease the application of automatised big-data mining and machine-learning techniques.
+Without clearly defined terms and concepts, handling of large amounts of trait-data by computational assistance systems for scientific analysis ('e-Science') will be massively delayed [@wilkinson16].
 
@@
 
-To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions.
+To date, a rich distributed body of independently published trait datasets exists, with different focus on particular organism groups, ecosystem types or regions.
 These distributed data are heterogeneous in form and description and initiatives to harmonise, compile and analyse these data require significant amounts of funding and personnel.
 By using a constrained vocabulary with globally accessible definitions of terms,  distributed trait data can be accessed more easily by other researchers and harmonised into aggregated datasets. 
-
 The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation and publication of data.
-This requires significant effort, but it returns  great scientific benefit by enabling synthesis on our general understanding of biodiversity and ecosystem function.
+This requires significant effort, but it returns  great scientific benefit by enabling data-heavy synthesis for a general understanding of biodiversity and ecosystem functioning.
 
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Matthias Biber, Kristin Bohn, Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts and pre-print, the trait data standard vocabulary and the pre-releases of the R-package.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to <!--Matthias Biber, Kristin Bohn,--> Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary<!-- and the pre-releases of the R-package-->.
 We are grateful to Brian Enquist, Rachel Galagher and Brian Maitner for organising the OpenTraits workshop in New Orleans, USA, in August 2018.
@@
 
-FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP (forming the first tier of the author list); NKS authored the example list of traits; FDS developed the R package 'traitdataform'; CP and FDS curated the living spreadsheet.
+FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP (forming the first tier of the author list);<!--NKS authored the example list of traits; FDS developed the R package 'traitdataform';--> CP and FDS curated the living spreadsheet.
 All contributing authors appear in alphabetical order in a second tier of the author list.
@@
 
-The Appendix A contains a static excerpt of the living spreadsheet on existing trait datasets and databases, which can be found at https://goo.gl/QxzfHy.
+Supplementary Material A - static excerpt of a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use.
+We invite data owners and users to add further trait datasets to this spreadsheet.
+
+Supplementary Material B - contains three worked examples of how to apply the standard vocabulary to 1.) simple two-dimensional data using only core terms, 2.) advanced use with extensions and identifiers within one spreadsheet, 3.) compiled data from multiple sources using metadata terms, 4.) a relational database structure with multiple spreadsheets.
 
 The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+Any future development of the vocabulary is coordinated via https://github.com/EcologicalTraitData/ETS/.
 
-The development website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform.
+<!--The development website for the R-package 'traitdataform' is https://github.com/EcologicalTraitData/traitdataform.
 <!--An example of a published trait thesaurus for the Biodiversity Exploratories (www.biodiversity-exploratories.de) can be found at https://ecologicaltraitdata.github.io/TraitDataList/.-->
-
-Any future development of the vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.
 

### 2018-09-18 bexis_handbook/bexis_handbook_traits.Rmd
update BExIS handbook and example

@@
 title: "BExIS Manual for Traitdata"
-author: "Florian D. Schneider (florian.dirk.schneider@gmail.com)"
-date: "23 April 2018 (v0.3)"
+author: "Florian D.
+Schneider^[Copyright: Biodiversity Exploratories Information System (BExIS), This document is free to share for internal use in the Biodiversity Exploratories project.
+Contributors: Malte Jochum, Gaëtane LeProvost, Caterina Penone, Andreas Ostrowski, Nadja K.
+Simons]"
+date: "17 September 2018 (v0.4)"
 output:  
@@
 
-## Contributors
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
 
-Malte Jochum, Gaetane LeProvost, Caterina Penone, Andreas Ostrowski, Nadja K. Simons
+\renewcommand{\thefootnote}{\arabic{footnote}}
+
+\vspace{3em}
 
@@
 To facilitate the data handling by BExIS and other downstream services (e.g.
-GFBio or TRY) we defined a standard vocabulary for publishing trait datasets.
+GFBio or TRY) we defined a standard vocabulary for publishing trait datasets [@schneider18].
 
@@
 
-The vocabulary proposed here is designed for two-dimensional spreadsheet data.
-These simple table formats are what a typical research project on a particular organism group and locality produces.
+The trait-data standard is a list of terms, and is not enforcing a particular structure for trait-data.
+However, for the use in BExIS, it is required to bring the data into a simple spreadsheet format, where each row reports a single trait value.
+For use with the pre-defined BExIS template for trait data, the spreadsheet has also to contain all terms of the data standard as columns (except the section for 'trait lists').
+An empty template for the use with MS Excel or LibreOffice can be downloaded after creating the Metadata object in BExIS (see below steps 1 to 9) and clicking on 'Select' and 'Create template'.
+
 The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported taxon, the measurement or fact (e.g.
@@
 The vocabulary also contains terms for describing metadata accompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain.
-For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend to use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
+For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend to use relational database structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a].
 ]
 
-We assume that data are stored in an observation table, where each row is the reported measurement or fact for a single observation^[Another common format for trait data are species--trait matrices, where each row is a species and multiple trait values are stored in columns.
+We assume that data are stored in an observation table, where each row is the reported measurement or fact for a single trait observation^[Another common format for trait data are species--traits matrices, where each row is a species and multiple trait values are stored in columns.
 These have to be converted into observation tables when applying the trait vocabulary.
-The R-package 'traitdataform' (see below) facilitates this conversion.].
+The R-package 'traitdataform' (see below) has been designed to facilitates this conversion.].
 At minimum, the following columns are required (Fig.
@@
 
-![Elements of the original data set (species $\times$ trait matrix) are transferred into a observation table format with additional columns for trait and taxon standardization.
+![Elements of the original data set (species $\times$ traits matrix) are transferred into an observation table format with additional columns for trait and taxon standardization.
 Further information on the level of taxa, ameasurements, occurrences (i.e.
@@
 
-To make data compatible with a consensus standard of taxon names or a look-up tables of traits, it is highly recommended to map the dataset-specific names provided for taxa (`scientificName`) and traits (`traitName`) onto globally valid terminologies by adding uniform ressource identifiers ([URIs](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier)) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by human-readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
+To make data compatible with a consensus standard of taxon names or look-up tables of traits, it is highly recommended to map the dataset-specific names provided for taxa (`scientificName`) and traits (`traitName`) onto globally valid terminologies by adding uniform ressource identifiers ([URIs](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier)) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by human-readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g.
 GBIF Backbone Terminology; see Fig.
-1)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'].
+1)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.].
 
 By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
-This redundancy of data allows for continuity on the authors and data owners side while also ensuring quality checks and comparability for the data user.
+This redundancy of data allows for continuity on the author and data owner side while also ensuring quality checks and comparability for the data user.
 
@@
 Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the taxon, to the individual or specimen, or to the reported fact, measurement or sampling event.
-Here, the assignment of a unique identifier per single taxon (`taxonID`), measurement event (`measurementID`), or individual occurrence (`occurrenceID`), defines the structure that these data would take in a relational database context (Fig.
+Here, the assignment of a unique identifier per single taxon (`taxonID`), measurement event (`measurementID`), or individual occurrence (`occurrenceID`) defines the structure that these data would take in a relational database context (Fig.
 1).
@@
 
-- The `taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides terms to describe the hierarchical taxonomy information and taxon rank of the measurement. 
-- The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+- The `taxon` extension^[https://ecologicaltraitdata.github.io/ETS/#extension-taxon] provides terms to describe the hierarchical taxonomy information and taxon rank of the measurement. 
+- The `measurementOrFact` extension^[https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact] provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical aggregation.
 It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
@@
 literature source or expert reference).
-- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
-This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
-- The `BiodiversityExploratories` extension  (http://ecologicaltraitdata.github.io/ETS/#extension-biodiversity-exploratories)  provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`.
+- The `occurrence` extension^[https://ecologicaltraitdata.github.io/ETS/#extension-occurrence] contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
+This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important information to analyse trait variation due to differences in space and time.
+- The `BiodiversityExploratories` extension^[https://ecologicaltraitdata.github.io/ETS/#extension-biodiversity-exploratories] provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`.
 
@@
 
-1. preparing of own trait datasets for publication on low-threshold file-servers (e.g. Figshare or Researchgate) and project databases (like BExIS), and
+1. preparation of own trait datasets for publication on general-purpose data-repositories (e.g. Figshare or Researchgate) and project-specific databases (like BExIS), and
 2.
-automatizing the harmonisation of trait datasets from different sources by moulding them into a unified format^[The package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
+automated harmonization of trait datasets from different sources by moulding them into a unified format^[The package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses.
 See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.
@@
 
-Installation of the package is done via the 'devtools' package directly from Github^[The package is in beta stage, that is it is not yet complete in its desired functionality and undergoes heavy revisions.
-If you have difficulties with the package or the provided documentation, please report these on https://github.com/EcologicalTraitData/traitdataform/issues]:
+Installation of the package is done via the 'devtools' package directly from Github^[The package is in beta stage!
+It is not yet complete in its desired functionality and undergoes heavy revisions.
+If you have difficulties with the package or the provided documentation, please report these on https://github.com/EcologicalTraitData/traitdataform/issues.]:
 
@@
 This function maps column names into terms provided in the standard trait vocabulary and adds metadata as attributes to the output object.
-The function also will mould a species-trait-matrix into an observation table format.
-If data are measured on a single individual the function adds an occurrenceID to keep the reference context of individual measurements.
+The function will also mould a species$\times$ traits-matrix into an observation table format.
+If data are measured on a single individual, the function adds a field 'occurrenceID' to keep the reference context of individual measurements.
 This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:
@@
 Note that the parameter 'keep' specifies and renames any data that should be maintained in the output.
-The parameter 'units' is used to specify the input units of measurement.
+The parameter 'units' is used to specify the input units of measurements.
 
@@
 Also note the trait thesaurus provided for the Exploratories (https://ecologicaltraitdata.github.io/TraitDataList/), which will be included in the package in due time.].
-The same function harmonises trait values by converting values into target units and unified factor levels.
+The same function harmonises trait values by converting values into target units and unified factor levels^[Factor level harmonisation requires some manual mapping using the function `mutate.traitdata()`, but will be added to the `standardize.traits()` function soon.
+Please check the manual files for `?standardize.traits` for further detail.].
 
@@
             taxa = "name_correct", 
-            units = "mm",
-            bexis = TRUE
+            units = "mm"
             )
 
+```
+
+As a particular use case, the package includes a function `standardize.exploratories()` for data assembled within the Biodiversity Exploratories project.
+
+```
+datasetStd <- standardize.exploratories(datasetStd, plots = "locationID", fillall = TRUE)
+```
+
+ The function maps plot IDs to a universal geolocation and habitat information and adds these as columns to the core dataset.
+This is only relevant for traits that have been measured from specimens collected on the sites of the Exploratories.
+The parameter `fillall = TRUE` (default) creates an output with the complete list of columns (even if they remain empty) for an upload with the BExIS trait template.
+It should be switched to `FALSE` if the file is to be uploaded elsewhere.
+
+The final step is to save and upload the file in a clean '.txt' format.
+
+```
 write.table(datasetStd, file = "filename.txt", 
@@
             sep = "\t", 
+            quotes = FALSE,
             na = "NA", 
@@
 
-As a particular use case, the package includes a function `standardize.exploratories()` for data assembled within the Biodiversity Exploratories project.
-The function maps plot IDs to universal geolocation and habitat information and adds these as columns to the core dataset.
-This is only relevant for traits that have been measured from specimens collected on the sites of the Exploratories.
+\newpage
 
@@
 You should use the predefined template "traitTemplate" when creating a new trait dataset on BExIS. Please pay special attention to the following points when providing the Metadata information:
+
+**On page 1:** If the data have been uploaded in an unstandardized  format before, you should refer to all authors and owners as reported in the original file.
 
@@
 "\t") and decimal separator (usually ".").
-Structural missing values for empty columns created by the R package are filled with "NA", thus enter "NA" for missing data.
+The  Structural missing values for empty columns created by the R package are filled with "NA", thus enter "NA" for missing data.
 Known missing values (i.e.
@@
 The section on 'matrix data' can be ignored since the data have been transferred into observation table format.  
-The variables (i.e.
-column names) are already provided by the template.
-You may add non-standard columns if necessary.
+The 'quote' is the character that encloses character strings in your data table.
+By default, R's `write.table()` wraps all columns in quotes, regardless if they are character strings, factor levels or numbers.
+BExIS has trouble getting the factor levels right, if they are wrapped in quotes.
+So you must  make sure your file does not apply quotes on those fields (by specifying `quote = FALSE`).
+The variables (i.e. column names) are already provided by the template. You may add non-standard columns if necessary. But they must be placed in exactly the correct order.
 
@@
 
-After creating a Metadata object, you can append/upload a dataset to your BExIS entry. Please make sure that the order of columns is correctly aligned with the definitions of the template.
+**On page 8:** it is recommended to comment on how the dataset was obtained from a related dataset, i.e.
+by specifying "standardized copy of BExIS ID 21247, applying Schneider et al.
+2018, Ecological Trait-data Standard Vocabulary, v0.8, URL: https://ecologicaltraitdata.github.io/ETS/v0.8/ , DOI: 10.5281/zenodo.1255287"
+
+**page 9** just validates your input and asks you to go back, if there are any inconsistencies.
+
+After creating a Metadata object, you can append/upload a dataset to your BExIS entry by clicking on 'Select' and 'Upload observations from file'.
+Please make sure that the order of columns is correctly aligned with the definitions of the template.
 

### 2018-09-18 bexis_handbook/example01.Rmd
update BExIS handbook and example

@@
       in_header: header.tex
-fontsize: 10pt
+fontsize: 7pt
 geometry: margin=1.5in
@@
 
-```{r, eval = FALSE}
+```
 
-# load library
+# load library:
+
 library(traitdataform)
 
-# load non-standardized dataset from BExIS 
+# load non-standardized dataset from BExIS:
 # (requires access credentials and access rights)
@@
                   units = c("mg", "mm", # which units do apply
-                            "mm", "mm2", "mg/mm2"), 
+                            "mm", "mm2", "mg mm-2"), 
                   keep = c(locationID = "plot") # which other cols to keep
@@
 # standardize taxonomy:
-# applies a matching of taxonomic names to GBIF backbone taxonomy
+# (applies a matching of taxonomic names to GBIF backbone taxonomy)
 
@@
 
-traitlist <- as.thesaurus(data.frame(
-  trait = c(
-    "weight", 
-    "wing_length",
-    "wing_width",
-    "wing_area",
-    "wing_loading"
-  ),
-  traitDescription = c("dry weight of the individual", 
-                       "the distance between base and apex (xi)", 
-                       "the distance between tornus and costa (yi)", 
-                       "wing_width x wing_length", 
-                       "ratio of wing_area to body_mass"
-                       ),
-  valueType = c("numeric"),
-  expectedUnit = c("mg", "mm", "mm", "mm2", "mg/mm2")
+traitlist <- as.thesaurus(
+  weight = as.trait("body_weight",
+       expectedUnit = "mg", valueType = "numeric",
+       identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_weight"),
+  wing_length = as.trait("wing_length",
+       expectedUnit = "mm", valueType = "numeric",
+       identifier = ""),
+  wing_width = as.trait("wing_width",
+       expectedUnit = "mm", valueType = "numeric",
+       identifier = ""),
+  wing_area = as.trait("wing_area",
+       expectedUnit = "mm2", valueType = "numeric",
+       identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Wing_surface"),
+  wing_loading = as.trait("wing_loading",
+       expectedUnit = "mg mm-2", valueType = "numeric",
+       identifier = "")
 )
-)
+
+# This example does not specify identifiers, i.e. URI references to definitions
+# on the internet, for all traits. It would be advisable to provide such links,
+# but published ontologies do not cover all possible definitions of traits to
+# date. The metadata file should therefore include a definition of the traits in
+# a clear and unambiguous format.
 
@@
 
-dataset1Std <- standardize.exploratories(dataset1Std)
+dataset1Std <- standardize.exploratories(dataset1Std, plots = "locationID")
 
 # save to txt file, ready for BExIS upload: 
-# omitting row names, use tab as field delimiter, 
-# point as decimal delimiter, NA as missing value
+# (omitting row names, use tab as field delimiter, 
+# point as decimal delimiter, NA as missing value)
 
@@
             na = "NA", 
+            quote = FALSE,
             dec = "."

### 2018-10-10 manuscript/manuscript.Rmd
MS revision: revise review chapter

@@
 output:
-  pdf_document:
+  word_document:
     fig_caption: yes
-    fig_crop: no
-    keep_tex: yes
-    template: pretty.template
+    fig_width: 6
   html_document:
@@
     theme: united
-  word_document:
+  pdf_document:
     fig_caption: yes
-    fig_width: 6
+    fig_crop: no
+    keep_tex: yes
+    template: pretty.template
 bibliography: div05.bib
 ---
+
+<!--
+Title Page should include:
+
+    A concise and informative title. Do not include the authorities for taxonomic names.
+    A list of all authors' names with names and addresses of Institutions.
+    The name, address and e-mail address of the correspondence author.
+    A running headline of not more than 45 characters.
+-->
 
@@
 1.
-Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions.
-Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
-2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
+Trait-based approaches are widespread throughout ecological research, as they offer great potential to achieve a general understandnig of a wide range of ecological and evolutionary mechanisms.
+Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
+2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
 3.
-In order to facilitate the standardisation and harmonisation of distributed trait datasets on the data-provider side, we propose a standardised vocabulary to be used for storing and sharing ecological trait data.
-We discuss impediments and incentives for a wide adoption of such a standard by data-providers.
+In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data.
+We discuss potential incentives and challenges for the wide adoption of such a standard by data-providers.
 4.
-The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases, and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
-By aiding decentralised trait-data standardisation, our data standard facilitates the use of trait data for a broader ecological research community and enables global syntheses across a wide range of taxa and ecosystems.
+The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
+By aiding decentralised trait-data standardisation, our data standard may facilitate the use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.
 
@@
 
-functional ecology, species traits, semantic web, ontologies, data standard
+functional ecology, species traits, semantic web, ontologies, data standard, ecoinformatics
 
@@
 ----- | ---------
-Term | A word that describes a particular *concept* as part of the specialised vocabulary of a field
-Concept | An idea, notion or object that is made explicit in an information context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
-Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a praticular context, while no unlisted entries are accepted
+Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
+Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/)
+Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted
 Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*
@@
 trait data), that is recognised by a large proportion of members of the application context
-Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte13] 
-Ontology | *Controlled vocabulary* that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g.
-defines a hierarchy of terms; thus a formal model of the objects and their relationships in a domain of interest [@gruber95]
-Semantic web | An extension of the world wide web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
+Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g.
+may define a hierarchy of broader or narrower terms
+Ontology | *Controlled vocabulary* that defines *concepts* in relation to each other in a formal language, e.g.
+provides cross-references and semantic meaning; a semantic model of the objects and their relationships in a domain of interest [@gruber95]; commonly published in OWL format for machine readability.
+Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
 specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
-Dataset | A set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
+Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
 Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
 Database | A suite of data compiled from multiple *datasets*, i.e.
-from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software;
-Relational database | Usage in this paper: Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
-Identifier (ID) | A unique label that relates entries within and across *datasets*; is used to connect *data tables* into a *relational database*; can be user-specific or, as a *URI*, point to a globally valid *ontology* or *thesaurus*
-Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; an example is 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
-Webservice | An exchange protocol to access *online databases* directly and programmatically, i.e. by calls from a software tool
+from multiple study contexts or observation types; may take the form of a single two-dimensional *data tables*, but mostly is organised as multiple data tables linked via *identifiers* into *relational databases*; may be built using a specific database software or language (e.g.
+SQL)
+Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
+Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
+Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *relational database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*
+Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
+Web Service | A service to access *online databases* directly and computationally, i.e.
+by calls from a software tool; returns content of the database according to the request in different formats and *data standards*
 Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
-client software and the *webservice* of an online databases; APIs are usually documented on the website of a database provider
+client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider
 Online Portal | A website designed as a platform for the exchange of information, e.g.
 trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers)
-File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
+File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via persistent identifiers, e.g.
+DOIs (Digital Object Identifiers)
+File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
 Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
-Darwin Core Standard (Dwc) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
-Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns.
-Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures. 
-Occurrence | A single observation instance of a taxon, i.e. an organism at a particular place at a particular time (http://rs.tdwg.org/dwc/terms/Organism)
+Darwin Core Standard (DwC) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
+Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified e.g.
+using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns [@gbif17,  http://tools.gbif.org/dwca-assistant/]
+Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits @moretti17
+Occurrence | The observation context of a single entity of a taxon, i.e.
+an individual observed at a particular place and time (http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context
 
@@
 Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
-Recent years have seen a proliferation of trait-based research in a wide range of fields. 
-Traits bridge the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and relate losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
-This offers the mechanistic background for linking biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15].
+Recent years have seen a proliferation of trait-based research in a wide range of fields and have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
+This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
+Species traits have been suggested as indivator variable for monitoring  ecosystem health at the individual level [@kissling18].  
  <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
 Using traits is also a promising means of bypassing taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
+the fact that most species are yet undescribed and little is known of their interactions with other organisms and their environment.
 This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].
@@
 <!-- 2. trait databases have been developed, with standard structure and terms --> 
-However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.
-To enable such functional ecological research, trait data have been harmonized and compiled into centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g.
+Despite the importance of functional trait data, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time.
+However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
+<!-- i.e.
+differences in taxonomic nomenclature and resolution (e.g.
+reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements-->
+To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centred around particular research questions [e.g.
 PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
-The heterogeneity of data from different research contexts <!-- i.e.
-differences in taxonomic nomenclature and resolution (e.g.
-reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements--> render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
-Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed.
+Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
+Besides initiatives aiming at assembling data, tools to enable the compatibility of data across databases are being developed.
 These include software to access trait data from the internet [e.g.
-@chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
+@chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of standard terms [@walls12; @garnier17].
 
 <!-- 4. distributed trait data must be integrated --> 
-Meanwhile, open science reaches the mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
+Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
 As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services (e.g.
-Figshare.com, datadryad.org, Researchgate.net, or Zenodo.org), where no data standards are forced upon the uploaded material [@wilkinson16].
+Figshare.com, see Table 1), where no data standards areenforced upon the uploaded material [@wilkinson16].
 It is thus likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient reuse and synthesis of published datasets.
 
-In this paper, we review existing trait databases and online portals, as well as initiatives for data standardisation from a pragmatic view of data providers and data users.
-We discuss the current efforts for making trait data visible, accessible and reuseable in downstream data analysis, as demanded by the FAIR principles for scientific data [@wilkinson16].
-We show how the current deficit in standardisation hampers the implementation of interoperability and reuse of trait data.
-Based on these considerations, we propose a versatile vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data. 
-Since this standard vocabulary relies on the adoption of a broad research community we discuss incentives for its use and lay out mechanisms for a community-based future development towards an accessible ecological trait data standard.
+In this paper, we review existing trait databases and online portals, and initiatives for data standardisation from the pragmatic view of data providers, data curators and data users, as well as database managers.
+We discuss current efforts to making trait data visible, accessible, interoperable and re-useable in downstream data analysis, as demanded by the FAIR guiding principles for scientific data [@wilkinson16].
+Furthermore, we show how the current deficit in the standardisation of primary data hampers the implementation of interoperability and reuse of trait data. 
+Based on these considerations, we propose a versatile vocabulary for describing trait datasets that builds upon, and is compatible with , existing terminology standards for biodiversity data. 
+Since this standard vocabulary relies on the adoption of a broad research community, we discuss incentives for its use and lay out mechanisms for a future consensus-building and community development towards an accessible ecological trait data standard.
 <!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
@@
 
-We review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):
+The need for standardising trait-data arises from the prospective gain of compiling heterogeneous trait-datasets for data synthesis. 
+Often, the scientific scope and focus differs between data providers measuring and assessing the trait data in the first place and data users who re-use published data for a broader synthesis application.
+Data curators are taking up the task of providing compiled and harmonised datasets or databases.
+Data managers are concerned with the development of complex digital infrastructures for handling and analysing large amounts of data.
+These idealised roles of researchers cooperatively are dealing with trait-data standardisation throughout the data life-cycle.
+In this chapter, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):
 
-1. Initiatives that provide *trait datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature. 
-2. Initiatives that aim to harmonise trait data from the literature or from direct measurement into *trait databases* and make those data widely available. 
+1. Initiatives that provide trait *datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature. 
+2. Initiatives that aim to harmonise trait data from the literature or from direct measurements into trait *databases* and make those data widely available. 
 3.
-Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits, and provides standard *terminologies* in the form of *thesauri* and *ontologies*.
+Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits and provide standard *terminologies* in the form of *thesauri* and *ontologies*.
 4. Initiatives that aim to leverage *relational database* structures and *semantic web* technology to link trait data to a wider set of biodiversity data.
 
-We discuss these initiatives separately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
-We show how the degree of trait-data standardisation in existing datasets spans this entire spectrum and which tools and standards are applied to achieve harmonisation of data from multiple, distributed sources.
-The objective of this review is to raise awareness for the generic structure of trait data and aid researchers to share and publish own datasets in an appropriate form.
+We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
+We show how the degree of trait-data standardisation in existing datasets is highly variable and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources.
+The objective of this review is to raise awareness of the generic structure of trait data and aid researchers in how to share and publish their own datasets in an appropriate form.
 
@@
 
-In the field of comparative biology, morphological traits related to plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values for plants or ecological parameters of animals (e.g.
-reproductive traits, feeding biology, dispersal or body size) have been measured for decades, and have been published in regular journal articles or books.
-With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
+In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g.
+reproductive traits, feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books.
+With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
+A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, see e.g., @gossner15; see Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
+
 Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g.
-Figshare.com, datadryad.org, Researchgate.net, or Zenodo.org), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
-These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management.
-However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g.
+Figshare.com, see Table 1), which warrant long-term accessibility and citeability via DOIs, and govern data sharing via licensing.
+These platforms offer publicly accessible *repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management.
+However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g.
 .xlsx, .docx) or binary (e.g.
-.pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
-Most importantly, these platforms enable public hosting of data with very low thresholds regarding *metadata* documentation and data standardisation.
-As a result, published datasets greatly vary in structure and nomenclature.
+.pdf) data formats which make extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
+Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation.
+As a result, published datasets greatly vary in structure, nomenclature and metadata quality.
 
-For trait data, there are typical issues arising from the variability of data structures. 
-For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific, and rarely chosen to allow translation into larger database initiatives.
-Furthermore, metadata varies in its detail, e.g.
+For trait data, there are common issues arising from the variability of data structures. 
+For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives.
+Furthermore, metadata varies in its level of detail, e.g.
 for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
 In terms of structure, trait data usually are reported in a species$\times$traits wide-table format.
-In this format, each row contains a species (or taxon) for which multiple traits are reported in columns.
-Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
-Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].
-As it stands, the decentralisation and the lack of data standardisation of data in online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be reused and combined into broad synthesis analysis.
+In this intuitive *data table*, each row contains a species (or taxon) for which multiple traits are reported in columns.
+<!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).
+-->
+Similarly, when reporting raw data, researchers place observations of individual organisms (i.e.
+the *occurrences*) in rows with multiple trait measurements applied to the same individual across multiple columns.
+Co-variates on the taxon, the individual specimen (e.g.
+sex or lifestage) or context of observation (e.g.
+time and place of sampling) would be placed in additional columns, and would further expand the two-dimensional data table.
+The resolution or scope of these co-variates varies greatly depending on the research question and observation context.
+Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14].
+A task that is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).
 
 <!--To remedy these problems, data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/) or GFBio [the German Federation on for Biological Data; gfbio.org; @diepenbroek14], offer means for small projects to meet the high scientific criteria of data management.-->
-
 <!--A global overview of existing trait data for all taxa and trait types is difficult to obtain.
@@
 In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g.
-@klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Supplementary Material A, table A1].<!--[@wright04; ] -->
-While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
+@kleyer08; @oliveira17, see Supplementary Material A, Table A1].
+While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these *databases* apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
 -->
-Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
-Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms and may be updated irregularly.
+Documentation of metadata and methodology differs in the level of detail, depending on the research focus. 
+Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms, but they may be updated irregularly.
 
 As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
-Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g.
-www.markmybird.org), data on body measurements are likely to skyrocket in the near future.
-For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16].
+The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis17; @mantle12; @strobel18] and citizen science initiatives (e.g.
+www.markmybird.org),
+For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16].
 The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).
@@
 <!-- database backend, selective download, user access management --> 
-More specialised trait-database platforms have been created to cover certain trait types (e.g.
-floral traits, seed traits, root traits or wood density traits), interaction types (e.g.
-pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g.
-location or climatic data).
-Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
+Specialised *online portals* have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
 <!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
@@
 As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+For animal trait data, however, a single unified portal and harmonising scheme is still lacking.
+The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].
 
-Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12].
+Nonetheless, initiatives for particular groups of animals do exist: examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), or the Coral Trait Database [@madin16, see Supplementary Material A, Table A1].
+The role of online portals and database initiatives in standardising data and making them more accessible is paramount. 
+Trait-data portals incentivise data submissions by offering increased data visibility and usage, while providing data-use policies that secure author attribution and, potentially, co-authorship of associated articles.
+However, maintaining centralised database infrastructures is costly and requires long-term funding [@bach12].
+
+## Thesauri and Ontologies for traits
+
+<!--Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12].
 Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13].
@@
 the KEGG orthology, @kanehisa12].
-For animals, a single unified platform and harmonising scheme for animal trait data is still lacking.
-The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17].
-Nonetheless, initiatives for particular groups of animals do exist: for instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), or the Coral Trait Database compiles ecological data on coral observations [@madin16].
-
-The role of database initiatives in standardising data and making them more accessible is paramount. 
-The trait-data platforms incentivise data submissions by offering increased data visibility and usage, while providing data-use policies that secure author attribution and, potentially, co-authorship.
-However, the upload of data to these centralised databases does often not fulfill all of the criteria that are increasingly demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions.
-
-<!-- AO: In a sense of data availability to the public : yes.
-But do we discuss it here?
-We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs.
-non-open data, or?
-bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
-FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->
 
-## Thesauri and Ontologies for traits
+A major challenge in trait-data standardisation is the lack of widely accepted and unambiguous trait definitions [@williams06; @kissling18].
+Previous standard definitions of trait *concepts* range from listings of selected definitions in *terminologies*, over well-defined *method handbooks* and comprehensive *thesauri*, to formalized definitions of trait concepts in *ontologies*.
+The initiatives behind method handbooks, thesauri and ontologies are primarily concerned with building community consensus for trait definitions.
 
-A major challenge in trait-data standardisation is the lack of widely accepted and unambiguous trait definitions.
-Previous standard definitions of trait *concepts* range from listings of selected definitions in *glossaries*, over well-defined methodological *handbooks* and comprehensive *thesauri*, to relational definitions of trait concepts in *ontologies*.
-While glossaries may be seen as specific for a study context, the initiatives behind method handbooks, thesauri and ontologies are primarily concerned with consensus building on trait definitions in a wider community.
-
-Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
-Assigning a more detailed and unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring.
-Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that pre-date or ignore this standard [@kattge11a].
+Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13; @kissling18] aiming for a list of functional indicators for ecosystem health.
+Assigning a more detailed and unambiguous methodological protocol for a trait, including the units to use or the ordinal or factor levels to be assigned, is essential for standardising its measurement process.
+Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but are of limited use in harmonising trait data that pre-date or ignore this standard [@kattge11a].
 
 A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
-Expanding on this, ontologies link the defined terms by formally defining the relationships between them, with the objective of enabling a computational interpretation of data. 
-Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)* within own datasets.
-For example, a measurement of seed size could be linked to the Planteome Trait Ontology (TO) definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.
+In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms.
+For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
+Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)*.
+For example, a measurement of fruit mass could be linked to the definition of the term within the TOP Thesaurus of plant traits [@garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
+
+Ontologies link defined terms by formally describing the relationships between them and providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14].
 Ontologies define terms based on other well-defined terms from published ontologies.
-The TO definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
+The Planteome Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
 Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12].
-This way, each trait definition may link to a broader or narrower term.
-For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'.
-By providing this interlinkage of trait ontologies, a machine-readable web of definitions is spun across the Internet which allows researchers and search engines to relate independent trait measurements with each other and connect it to the wider *semantic web* of online data [@berners-lee01; @page08].
-The distinction of thesauri and ontologies is not truly binary.
-Rather they mark idealised ends of a spectrum.
-While thesauri may contain defined relations between terms within the standard, ontologies relate most terms to other defined concepts, and also link those to other standards.
+By providing ontologies in a formalized file format, like OWL or RDF, a machine-readable web of definitions is spun across the Internet allowing researchers and search engines to relate independent trait measurements with each other and connect them to the wider *semantic web* of online data [@berners-lee01; @page08; @williams06; @walls12].
 
@@
 -->
-Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for their collaborative development [@walls12]. 
-For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/), provide centralised hosting for advanced trait ontologies and offer webservices for computational access.
+Meta-ontology initiatives, like OBO Foundry [@smith07, http://www.obofoundry.org/], Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/), or the GFBio Terminology service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for advanced trait ontologies and offer harmonized web services computational access.
 
-To conclude, there is already a suite of globally available thesauri and ontologies for traits that emerged from standardisation efforts of methodologies and community consensus processes.
-However, definitions in some domains are better covered than others.
-Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
-Most importantly, while these defined vocabularies are widely used in biodiversity data management, distributed data repositories of smaller project contexts hardly make use of them. 
-A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
-To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
-For example, the accessibility of ontologies will increase if open *Application Programming Interfaces (APIs)* are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
-Software then can assist researchers in linking own data to globally defined concepts.
+To conclude, there is already a suite of globally available thesauri and ontologies for traits available.
+However, definitions in some domains are better covered than others and the interlinkage and accesibility of ontologies can be much improved to fulfil semantic web standards.
+While these defined vocabularies are widely used in biodiversity data management, distributed trait-data of smaller projects published in general-purpose file servers make little use of them.
 
-## Trait-data structures for the semantic web
+## Trait-data structures
 
-While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
-To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.
+Whiel trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
+To make sense of trait data in the context of more general databases, a consensus definition of trait data is therefore necessary.
 
-Trait data have been defined by @garnier17 to follow an entity-quality model (EQ), where a trait observation is 'an entity having a quality'. 
+Trait data have been defined by @garnier17 as following an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
 More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e.
@@
 traits) observed on a biological entity (i.e.
-an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e.
+an individual specimen, or its parts) that can be assigned to a biological taxon (i.e.
 a species or higher-level taxon).
-We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
-1*a*), by aggregating replicated measurements on multiple entities (Fig.
-1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g.
-databases, Fig.
-1*c*).
-Qualitative facts are assignments of an entity to a categorical level, e.g.
-of a behavioural or life-history trait (Fig.
-1*d*).
-The entity or observation (i.e.
-the *occurrence*) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
+The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
 a genus).
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Types of ecological trait data assume different entities or reported quantities.
-a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire species or higher taxonomic group.}
+These relationships between a trait measurement and an individual organism as an entity of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+These terminology frameworks specify terms and classes to describe the general structure of biodiversity databases.
+The terms may be used for  defining columns in data tables that contain measurement values, units and factor levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
+The terminologies provided by these standards are quite universal and would even cover most use cases of trait data.
+A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].  
+For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) refers to DwC terms for defining its database content.
 
-\end{figure}
-<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”.
-Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->
-
-These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
-The terminologies provided by these standards are quite universal and even cover most use cases of trait data.
-An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12].
-For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies the terms described therein.
 <!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter.
@@
 
-Specifically designed for plant traits, @kattge11 proposed a generic database structure that covers most potential use cases of trait-based ecology.
-This data structure is built around a central data table that contains observations, i.e.
-a single event of measurement on the same individual plant specimen at the same point in time.
-This structure emphasises the fact that multiple trait data are measured on the same individual organisms and used to analyse correlations between these multiple traits.
-Identifiers link the measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables.
+Specifically designed for plant traits, @kattge11a proposed a generic database structure that covers most potential use cases of trait-based ecology.
+This data structure is built around a central data table that contains observations of individual plants (entities) linked to several measurements of traits (qualities) via identifiers.
 The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
-This structure can be implemented in any relational database management system.
+The structure of @kattge11a can be mapped to the  Extensible Observation Ontology [OBOE, @madin07; @schildhauer16], which formalizes observations andd measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
 
 In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach of an integrated structure for trait data.
-The framework employs established terms provided by the DwC, relates trait definitions to trait ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
-Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preferred for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11]. 
+The framework employs established terms provided by the DwC, relates trait definitions to ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
+Additional layers of information cover bibliographic references, multimedia archives and ecological interactions. 
+TraitBank invites data submissions to the EOL database in a structured *Darwin Core Archive* [DwC-A, @robertson09a].
+
+All of these structures make  use of stable URIs to refer to taxon concepts.
 The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review [@franz16].
-Initiatives that aim at providing a stable reference for taxa are for instance the EOL Catalogue of Life [http://www.catalogueoflife.org/, @roskov18], the GBIF Backbone Taxonomy [@gbifsecretariat17], or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).
+Initiatives that aim at providing a stable reference for taxon concepts are for instance the EOL Catalogue of Life [http://www.catalogueoflife.org/, @roskov18], the GBIF Backbone Taxonomy [@gbifsecretariat17], or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).
 
-These proposed standards are responses to a demand from biodiversity data managers for more structured input from the research community. 
-However, hardly any of the aforementioned trait datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
-One reason for this is most certainly complexity: the data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
+## Closing gaps in the trait-data lifecycle
+
+Initiatives for standardisation (e.g.
+ontologies and data standards) and platforms for data management (e.g.
+database and data management platforms) provide great visibility, accessability and improved interoperability of datasets, but raise relatively high thresholds for data and metadata preparation.
+This contrasts with the growing numer of openly accessible trait datasets in public file-servers, which hardly any of the aforementioned trait datasets or databases for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
+As it stands, re-using these data in larger compilations is error-prone and inefficient and the potential for a broad synthesis is diminished.
+In sum, we attest a gap in the trait-data structures developed for data curators and managers and the unstructured data input by data providers.
+
+One likely reason for this is complexity: the proposed data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
 In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
@@
 -->
-The other reason is lack of awareness for the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
+The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management.
+In addition, complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
 
@@
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-One goal of such initiatives is to provide standardisation of data and facilitate data usage, for instance by mapping unambiguous terminologies and ontologies for biodiversity data, and clarify conditions of data re-use.
-Another solution for data-users to access trait data in a structured way is offered by FENNEC [@ankenbrand18], an online tool or self-hosted service to extract trait information from multiple sources for a target species community.
-<!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->
+The goal of such initiatives is to facilitate data re-use by providing standardisation of data, for instance by mapping unambiguous terminologies and ontologies for biodiversity data, and clarify conditions of data re-use.
 
-## Summing-up
+Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains. For example,
+- r packages
+FENNEC [@ankenbrand18], an online tool or self-hosted service that can extract trait information from multiple sources for a target species community.
 
-Initiatives for standardisation (e.g.
-ontologies and data standards) and platforms for data management (e.g.
-database and data management platforms) provide great visibility, accessability and improved interoperability of datasets, but raise relatively high thresholds for data and metadata preparation.
-On the other end, public repositories on file-hosting services offer the publication of scientific primary data according to current funding and journal criteria, attracting a wealth of heterogeneous trait datasets.
-Due to the low thresholds concerning data standardisation, compilation of these distributed data sets is currently laborious.
-The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by raising awareness for data standardisation, accessibility and re-useability on the side of data providers.
-By establishing this as a best practice in trait-based research, trait-data will eventually fullfill the FAIR guiding principles for scientific data [@wilkinson16].
-In the following sections, we propose the development of vocabularies that lower the thresholds and offer high pay-off in the visibility and reuse of published data.
+<!-- To remediate this, incentives for data standardisation before upload to file hosting services should be created.
+This could be implementations of data quality standards for trait-data as requirements for journal publications.
+Data publications should also be
+The recognition of data publications as a fully valid scientific contribution becomes also more likely if data are published according to the FAIR guiding principles for scientific data (data should be findable, accessible, interoperable and reusable; @).
+In this sense, trait-datasets will be findable if they are provided with explicit metadata and keywords, labelling them as trait-data for search engines (such as the Google Dataset Search).
+Accessibility and reusability are warranted by releasing datasets under public domain dedication or open licenses (such as Creative Commons 0 or Attribution license; @).
+Interoperability and
+-->
+
+Despite centralised database efforts, no unified definition of traits as a data type with a well-defined exchange format or data structure has emerged across the multiple sub-disciplines of trait-based research.
+
+However, unlike for other data types like for instance genetic data or movement data (@@), no unified format for trait data has successfully been established for application in all use-cases.
+
+Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
+A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
+To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
+For example, the accessibility of ontologies will increase if open *Application Programming Interfaces (APIs)* are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+Software then can assist researchers in linking own data to globally defined concepts.
+
+Therefore we must better integrate these distributed data into the global biodiversity data-management life-cycle by making data providers more aware of data visibility, accessibility, interoperability and re-useability.
+By establishing this as a best-practice standard in trait-based research, trait-data will eventually fulfil the FAIR guiding principles for scientific data [@wilkinson16].
+
+In the following chapter, we propose the development of vocabularies that lower the thresholds and offer high pay-off in the visibility and reuse of published data.
 
@@
 entity), of a biological taxon.
+
+We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
+1*a*), by aggregating replicated measurements on multiple entities (Fig.
+1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g.
+databases, Fig.
+1*c*).
+This encompasses a wide range of data types, including numeric, integer, intervals or ratios, or even values from an ordinal scale.
+Qualitative facts are assignments of an entity to a categorical level, e.g.
+of a behavioural or life-history trait (Fig.
+1*d*).
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
+\caption{Types of ecological trait data assume different entities or reported quantities.
+a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
+reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
+a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire species or higher taxonomic group.}
+
+\end{figure}
+<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”.
+Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->
+
 <!--Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information).
@@
 We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model.
-In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g.
-in a separate metadata file, possibly applying the ecological metadata language, EML).
+In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g.
+in a separate metadata file, possibly applying the ecological metadata language, EML, @knb11].
 In cases where individual data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain the reference to the original data provider and conditions of re-use of these data.
@@
 
-M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
+MMG obtained funding from Swiss National Science Foundation (SNF 310030E-173542/1);
+MJ was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
 

### 2018-10-11 manuscript/manuscript.Rmd
revise manuscript: update vocabulary description, discussion, bibliography

@@
 author:
-- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
-- name: Malte Jochum \footnotemark[4]
-- name: Gaëtane Le Provost \footnotemark[3]
-- name: Andreas Ostrowski \footnotemark[6]
-- name: Caterina Penone \footnotemark[4]
+- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
 - name: David Fichtmüller \footnotemark[5]
@@
 - name: Anton Güntsch \footnotemark[5]
+- name: Malte Jochum \footnotemark[4]
 - name: Birgitta König-Ries\footnotemark[6]
-- name: Pete Manning \footnotemark[3]
+- name: Gaëtane Le Provost \footnotemark[2]
+- name: Pete Manning \footnotemark[2]
+- name: Andreas Ostrowski \footnotemark[6]
+- name: Caterina Penone \footnotemark[4]
 - name: Nadja K. Simons \footnotemark[7]
@@
 \footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
-\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
 \footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
+\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
 \footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
-\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
 \footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
@@
 As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
-The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis17; @mantle12; @strobel18] and citizen science initiatives (e.g.
+The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g.
 www.markmybird.org),
@@
 
-Whiel trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
-To make sense of trait data in the context of more general databases, a consensus definition of trait data is therefore necessary.
+While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
 
-Trait data have been defined by @garnier17 as following an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
+To make sense of trait data in the context of more general databases, a consensus definition of trait data has been provided by  @garnier17: 
+The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
 More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e.
@@
 
-Initiatives for standardisation (e.g.
-ontologies and data standards) and platforms for data management (e.g.
-database and data management platforms) provide great visibility, accessability and improved interoperability of datasets, but raise relatively high thresholds for data and metadata preparation.
-This contrasts with the growing numer of openly accessible trait datasets in public file-servers, which hardly any of the aforementioned trait datasets or databases for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
+In sum, we attest a gap in the trait-data structures developed for data curators and managers and the data input by data providers.
+Hardly any of the aforementioned trait datasets or databases for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
 As it stands, re-using these data in larger compilations is error-prone and inefficient and the potential for a broad synthesis is diminished.
-In sum, we attest a gap in the trait-data structures developed for data curators and managers and the unstructured data input by data providers.
 
@@
 In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
-<!--If the traits are assembled for a specific research question with just few ancillary co-factors, it may be safer to append these as columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road.
-This way, if datasets of different sources are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.
--->
 The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management.
@@
 
-Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains. For example,
-- r packages
+Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains to facilitate the use and analysis of trait-data. 
+For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
+The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
+The package 'NatDB' [@pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. 
 FENNEC [@ankenbrand18], an online tool or self-hosted service that can extract trait information from multiple sources for a target species community.
@@
 
-Despite centralised database efforts, no unified definition of traits as a data type with a well-defined exchange format or data structure has emerged across the multiple sub-disciplines of trait-based research.
-
-However, unlike for other data types like for instance genetic data or movement data (@@), no unified format for trait data has successfully been established for application in all use-cases.
-
 Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards. 
@@
 To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
-For example, the accessibility of ontologies will increase if open *Application Programming Interfaces (APIs)* are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+For example, the accessibility of ontologies will increase if open APIs are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
 Software then can assist researchers in linking own data to globally defined concepts.
 
-Therefore we must better integrate these distributed data into the global biodiversity data-management life-cycle by making data providers more aware of data visibility, accessibility, interoperability and re-useability.
+To bring all these tools for datamanagement together, a common vocabulary and definition of consensus concepts and exchance formats is required that connects to all steps of the data life-cycle.
+Despite centralised data management efforts, however, no unified definition of traits as a data type with a well-defined exchange format or data structure has emerged across the multiple sub-disciplines of trait-based research.
+In the following chapter, we propose the a unified vocabulary for trait-data that can form the basis of such an exchange format.
+The simplicity of this standard terminology will lower the thresholds and offer high pay-off in the visibility and reuse of published data.
 By establishing this as a best-practice standard in trait-based research, trait-data will eventually fulfil the FAIR guiding principles for scientific data [@wilkinson16].
-
-In the following chapter, we propose the development of vocabularies that lower the thresholds and offer high pay-off in the visibility and reuse of published data.
 
@@
 As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
-The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
-The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
+The standard vocabulary is accessible at https://terminologies.gfbio.org/terms/ets/pages/.
+The aim of this vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement details. 
 Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases: 
+1.
+by data providers: for publication of standardised primary data on open-access data repositories, or for labelling project-specific data for local use and exchange with collaborators, e.g.
+in two-dimensional data tables or project databases
+2.
+by data users and curators: as a consensus vocabulary when compiling data from distributed sources into databases, e.g.
+to map standardised columns and refer to taxa and trait definitions in a uniform way, and
+3.
+by data managers: as a data exchange format of between databases, web services and software tools, e.g.
+when providing database queries via a web service or defining input and output formats of software packages.
 
-1.
-in  two-dimensional datasets, e.g.
-for labelling project-specific data, for local use and exchange with collaborators, or for publication of standardised primary data on open-access data repositories,
-2. as a consensus vocabulary when compiling data from distributed sources into databases, e.g. to map standardised columns and refer to taxa and trait definitions in a uniform way, and
-3. as an exchange format of data between web services and software tools, e.g. when providing database queries from a webservice or defining input and output formats of software packages.
-
-By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable.
-It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.
-By adopting the standard vocabulary, trait-data will fulfill the FAIR principles for scientific data [@wilkinson16].
+By using this standard vocabulary, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable.
+It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives, data synthesis and ensure long-term accessibility.
+By adopting the standard vocabulary, trait-data will support the FAIR principles for scientific data [@wilkinson16].
 We describe best practice principles and give examples for the application of the vocabulary in Supplementary Material B.
 
-In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
-We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
-<!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
+In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users and curators), and biodiversity informatics researchers (data managers).
+We paid particular attention to existing data standards [particularly @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+
 <!--
@@
 
-The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
-quality) observed on an individual or population (i.e.
+A trait dataset typically contains data according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
+quality) observerd on an individual or population (i.e.
 entity), of a biological taxon.
-
-We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
+The trait values may be quantitative measurements or qualitative facts. 
+Quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
 1*a*), by aggregating replicated measurements on multiple entities (Fig.
@@
 1*d*).
+Beyond these core observations, further information might be available that further specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation and date information).
+These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
+
 \begin{figure}[t]
@@
 \includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Types of ecological trait data assume different entities or reported quantities.
-a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire species or higher taxonomic group.}
+\caption{Types of ecological trait data assume different entities or reported qualities: a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
+reproductive rates, life spans); b) aggregated trait values are reported as means taken on multiple measures of members of a taxon; c) quantitative traits may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
+a sex) as the entity of description; d) qualitative traits are categorical, ordinal or binary descriptors of the entire species or higher taxonomic level (also called 'facts').}
 
@@
 
-<!--Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information).
-These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, -->
-We offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see sections below for extensions and best practice when publishing primary data).
-Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).
-
-Some of the terms expand and refine terms of the Darwin Core and its Extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516).
+Our vocabulary thus provides core terms to reflect the minimal Entity-Quality structure of trait data. 
+We further offer three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", and "Occurrence", "Taxon"; see section on extensions below). 
+All terms may be applied to describe columns of a data table, or to label content of a database (see Box 1 / Supplementary Material B for best practice principles and examples for publishing primary data).
+Additional terms are provided for metadata and for providing trait concepts (see section on metadata below).  
+Some of the terms expand and refine terms of DwC and its extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516).
 Wherever our vocabular refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
-Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns.
-However, these should be clearly specified in the meta-information accompanying the dataset.
-The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
-Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC. 
-The vocabulary proposed is intended to form the base of a standard exchange format.
-Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.
+The vocabulary can be mapped onto other existing ontologies (e.g.
+@madin07) or database structures (e.g.
+@kattge11a).
+Aspects not covered by the terminology may draw from terms provided by DwC and it's extensions or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).
 
@@
 
-To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum  (Fig.
-2 *b*):
+For trait data according to the Entity-Quality model, the following content is required at minimum (Fig. 2 *a&b*):
 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
-2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
-3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`). 
-For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
+2. a descriptive trait name (`traitName`) that links to a standardised definition (i.e. a concept); 
+3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`) that links to an accepted taxon concept.
+
+<!--### Relate to global taxon and trait definitions -->
+The trait name and taxon names as would ideally use unambiguous terms assigning both to clearly-defined concepts. 
 However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
-<!--### Relate to global taxon and trait definitions -->
 In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
-This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively).
-For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.
-By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practised in TRY; @kattge11].
-This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.
+This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively, Fig.
+2 *c*).
+For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  `https://www.gbif.org/species/3117424`; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  `http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass`.
+<!-- units and factor levels-->
+Similarly, standardisation of units can be achieved by relying on SI convention or by relating units to unambigous concepts via URIs provided by ontologies [@steinberg16; @keil18]. 
+For categorial, ordinal or binary traits, the factor levels should conform to expected levels as defined in the trait concept, or be unambiguously defined in the metadata of the dataset (see section metadata below).
+The vocabulary complements the user-defined values in dataset-specific units and factor levels with standardised entries (`traitValueStd` and `traitUnitStd`). 
+<!-- double record -->
+By suggesting for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
+This redundancy of data allows for continuity for data providers while also enabling quality checks and comparability for data curators.
 
@@
 
-Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event.
-The data standard provides three extensions of the vocabulary that should be used to describe this information \mbox{(Fig. 2*d*)}:
+Beyond measurement units or higher taxon information, further information might complement the core data which are related to the individual specimen, the reported fact, measurement or sampling event.
+We propose three extensions of the vocabulary that should be used to describe this information (Fig. 2*d*):
 
 - The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.
-(http://ecologicaltraitdata.github.io/ETS/#extension-taxon)
-- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation.
+(https://terminologies.gfbio.org/terms/ets/pages/#extension-taxon)
+- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature reference for the reported value, the method of measurement or statistical method of aggregation.
 It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
 variation in measurement method) or aggregated values (e.g.
-statistical method applied), as well as the source of reported facts (e.g.
+statistical method), as well as the source of reported facts (e.g.
 literature source or expert reference).
-(https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact)
-- The `Occurrence` extension contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age.
-This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
-(https://ecologicaltraitdata.github.io/ETS/#extension-occurrence)
+(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)
+- The `Occurrence` extension contains vocabulary to describe information on the observation context of individual specimens, such as sex, life stage or age.
+This also includes the method of sampling and preservation, as well as the date and geographical location, which provide an important resource to analyse trait variation due to differences in space and time.
+(https://terminologies.gfbio.org/terms/ets/pages/#extension-occurrence)
 
@@
 These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content.
-A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+A unique identifier links to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
 Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
-a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
+a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
+
+Identifiers also can be used to provide a structure within a two-dimensional data table, e.g.
+to link rows of measurements on the same individual (using `occurrenceID`) or observations at the same location (using `locationID`).
+The  values of multivariate measurements can be labelled using `measurementID`.
 
@@
 
-Wherever possible, the column `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
-If no globally available trait definition exists as an external reference, trait datasets should always be accompanied by a dataset-specific list of traits as part of the metadata or as an accompanying data table.
+Higher-level information about structure, authorship and data ownership and terms of use should be considered when sharing and working with trait datasets.
+Our metadata vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) provides terms to label these information. 
+In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g.
+in a separate file, possibly applying the ecological metadata language, EML, @knb11].
+Whenever data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain reference to the original data provider and conditions of re-use of these data.
+This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
+
+Wherever possible, the field `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
+If such an external reference does not exist, trait data should always be accompanied by a dataset-specific list of trait concepts. 
 Such a controlled vocabulary would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format of measured values or reported facts (e.g.
@@
 -->
-
-Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
-We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model.
-In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g.
-in a separate metadata file, possibly applying the ecological metadata language, EML, @knb11].
-In cases where individual data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain the reference to the original data provider and conditions of re-use of these data.
-This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
-
 <!--Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit.
@@
 
+In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
+"Schneider, F.D., Jochum, M., Le Provost, G., Penone, C., Ostrowski, A.
+and Simons, N.K., 2018 Ecological Traitdata Standard v0.9, DOI:  10.5281/zenodo.1441132, URL: https://terminologies.gfbio.org/terms/ets/pages/".
+Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their URIs,.
+The definitions of terms are hosted on the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
+via the API; https://terminologies.gfbio.org/api/terminologies/).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
+\caption{Formats used for trait datasets.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
+
+\end{figure}
+
+<!--
 ## Best practice for primary data publication
 
-We focus this section on the first of the three use-cases for the ecological trait-data standard vocabulary: The publication of project-specific primary data.
+In this section we focus on the publication of project-specific primary data by data providers (use case #1, see main document).
 We assume that data managers and software developers who apply the vocabulary in use-case 2 and 3 are more fluent with the use of exchange formats and standard definitions and have established own best-practice principles.
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
-
-\end{figure}
-
 For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
@@
 Supplementary Material B contains three worked examples of how to apply the vocabulary to own trait-datasets. 
+-->
 
@@
 
-To serve the demand for standardisation and harmonisation of ecological trait data that arises from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of simple, two-dimensional datasets as well as for the exchange and handling of trait data in the context of a 'semantic web'.
-In the following paragraphs we want to discuss the incentives and impediments for individual researchers, and highlight the long-term pay-off and benefit for the community in adopting a standard terminology for trait-based ecological research.
+To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, as well as for the exchange and handling of trait data in the context of a 'semantic web'.
 
 <!-- community effort and consensus -->
-While it is clear, that a single vocabulary standard is not able to fullfill the needs of all research fields working with traits, we believe that our proposal captures a common core concept of trait data.
-Consensus building on how traits are to be used and evaluated is currently under way in several fields with their taxonomic focus and project-specific questions.
-While these community discussions on trait definitions and measurement practice are leading to a better quality of data, naturally, the heterogeneity in trait-data quality and detail across fields will remain.
-Nonetheless, we provide a starting point and invite these communities to develop a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research, and to discuss and define terms and extensions to fit their specific needs.
-The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio terminology service.
+While it is clear, that a single vocabulary standard is not able to fulfil the needs of all research fields working with traits, we believe that our proposal captures a common core concept of trait data.
+Consensus building on how traits are to be used and evaluated is currently under way in several fields with their taxonomic focus and project-specific questions [@kissling18; @garnier17; @pey14; @moretti17].
+While community discussions on trait definitions and measurement practices are leading to a better quality of data, naturally, the heterogeneity in trait-data quality and detail across fields will remain.
+With our proposal of an Ecological Trait-data Standard vocabulary, we provide a starting point for the development of a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research.
+Therefore, we invite researchers to contribute to future versions of the standard vocabulary and develop own applications and ontologies that interact with it.
+The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service (https://terminologies.gfbio.org/terms/ets/pages/).
 
-<!-- investment and pay-off, incentives -->
-For individual researchers and small research projects, the investment of time and resources in trait-data standardisation before publication seems not to be yielding a direct benefit.
-Certainly, data released according to consensus standards, especially if published under open-access licenses, will be more easily re-used in compilations and synthesis studies. 
-A well-defined vocabulary for metadata will also ensure that authorship and license model are appropriately documented along the data life-cycle.
-However, for these incentives to take positive effect, data publications and data citations must become a valid scientific contribution in the community recognition and professional evaluation of individual researchers [@costello09; @roche15].
+<!-- ETS investment and pay-off for data providers, incentives -->
+For individual researchers and small research projects, the investment of time and resources in trait-data standardisation before publication seems currently not to be yielding a direct benefit.
+However, data released according to consensus standards, especially if published under open-access licenses, will be more easily re-used in compilations and synthesis studies. 
+A well-defined vocabulary for metadata will also ensure that authorship and terms of use are appropriately documented along the data life-cycle.
+However, for these incentives to take effect, data publications and data citations must become viewed as a valid scientific contribution to the community, and  recognized in the professional evaluation of individual researchers [@costello09; @roche15].
+To mitigate the cost of data standardisation efforts, supporting tools and infrastructures should be developed to mitigate the cost of data providers for readying trait data for the 'semantic web' of biodiversity data.
 
-<!-- Data quality issues --> 
-On the community level, shifting the task of standardisation from the data-user side to the data-owner side yields great gain in accurracy and reduces the risk of misinterpretation.
-For instance, length measurement results depend very much on the methodology and landmarks used and often systematic biases could be corrected for when converting it to a consensus definition.
-The detailled knowledge might not be delivered along with the data and a post-publication standardisation is inhibited.
+<!-- ETS solving Data compilation issues --> 
+At the community level, shifting the task of standardisation from the data-user side to the data-owner side yields great gain in accuracy and reduces the risk of misinterpretation.
+For instance, measurement results depend very much on the precise methodology used and often systematic biases could be corrected for when providing an undambiguous definition.
 On the other hand, plausibility checks and evaluation of statistical methods, e.g.
@@
 Currently, these "big data" volumes are only available in centralised databases.
-To establish a best-practice of data aggregation, an exploration and evaluation of different methods should be subject to a community discussion. 
+However, to establish a best-practice of data aggregation, an exploration and evaluation of different methods for quality assessment and quality control should be subject to a community discussion. 
 This is only possible with large quantities of distributed data being available in a harmonised way. Our ecological trait-data standard vocabulary facilitates such a community-driven comparison.
 
-A wide adoption of our trait-data standard vocabulary would allow to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for a better mechanistic understanding of the relationship between drivers, communities and functions and providing new insights on global biodiversity patterns.
-Moreover it might be also a step towards a more predictive ecology as a broader set of available traits might enable new hypothesis-driven approaches. 
-However, to enable a global synthesis across trait datasets, there remains a need to advance the development of trait thesauri into more interrelated and complete ontologies.
-
-<!-- toolchain and semantic web -->
-To support the long-term rewards of data standardisation efforts, incentives should be sought to mitigate the cost of data providers for readying trait data for the 'semantic web' of biodiversity data.
-This can be software tools or supporting infrastructures.
-The R-package 'traitdataform' [@schneider18a] is assisting the conversion of own data into compliant formats. 
-Similarly, the use and analysis of trait-data can be much facilitated by software packages for different tasks. 
-For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
-The package 'NatDB' [@pearse] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data.
-
-Our standard vocabulary is an important building block for a unified mode of data exchange between webservices and software packages and thus facilitates the development of a software toolchain for the trait-data lifecycle.
+<!-- ETS to build toolchain and semantic web -->
+Our standard vocabulary represents an important building block for a unified mode of data exchange between web services and software packages and thus facilitates the development of a software toolchain for the trait-data lifecycle.
 Having well-defined terms is also a key precondition for developing exchange formats between large database initiatives and biodiversity data archives. 
 Even further downstream, readying the primary data for the semantic web via references to ontologies and data standards will ease the application of automatised big-data mining and machine-learning techniques.
-Without clearly defined terms and concepts, handling of large amounts of trait-data by computational assistance systems for scientific analysis ('e-Science') will be massively delayed [@wilkinson16].
+Without clearly defined terms and concepts, handling of large amounts of trait-data by computational assistance systems for scientific analysis ('e-Science') will be massively hampered [@wilkinson16].
 
@@
 
-To date, a rich distributed body of independently published trait datasets exists, with different focus on particular organism groups, ecosystem types or regions.
-These distributed data are heterogeneous in form and description and initiatives to harmonise, compile and analyse these data require significant amounts of funding and personnel.
-By using a constrained vocabulary with globally accessible definitions of terms,  distributed trait data can be accessed more easily by other researchers and harmonised into aggregated datasets. 
+To date, there is a rich distributed body of independently published trait datasets, with different focus on particular organism groups, ecosystem types or regions.
+These distributed data are heterogeneous in form and description and initiatives to harmonise, compile and analyse these data require significant amounts of funding.
+
+By using a standard vocabulary with globally accessible definitions of terms,  would allow distributed trait data to be more easily re-used and harmonised into aggregated datasets. 
 The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation and publication of data.
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to <!--Matthias Biber, Kristin Bohn,--> Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Catrin Westphal, Markus Fischer and Jens Kattge for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary<!-- and the pre-releases of the R-package-->.
-We are grateful to Brian Enquist, Rachel Galagher and Brian Maitner for organising the OpenTraits workshop in New Orleans, USA, in August 2018.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to <!--Matthias Biber, Kristin Bohn,--> Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary<!-- and the pre-releases of the R-package-->.
+We are grateful to the organisers and participants of the Open Traits workshop in New Orleans, USA, in August 2018.
 
@@
 
-FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP (forming the first tier of the author list);<!--NKS authored the example list of traits; FDS developed the R package 'traitdataform';--> CP and FDS curated the living spreadsheet.
-All contributing authors appear in alphabetical order in a second tier of the author list.
-AG and DF implemented the vocabulary in the GFBio terminology service.
-All authors contributed critically to the structure and content of the manuscript and gave final approval for publication.
+FDS, AO, CP, and NKS conceived the idea and developed the vocabulary for the trait data standard with significant contributions of MJ and GLP;<!--NKS authored the example list of traits; FDS developed the R package 'traitdataform';--> CP and FDS curated the living spreadsheet; AG and DF implemented the vocabulary in the GFBio terminology service; all authors contributed critically to the structure and content of the manuscript and gave final approval for publication.
 
-# Online Resources
+# Data Accessibility
 

### 2018-10-19 manuscript/appendix_B.Rmd
init and final appendix_B.Rmd with examples, along with template

---
title: "Best practice examples for primary data publication"
subtitle: "Towards an Ecological Trait-data Standard - Supplementary Material B"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7]
fontsize: 11pt
geometry: margin=1.2in
documentclass: scrartcl
csl: amnat.csl
bibliography: div05.bib
nocite: | 
  @gossner15a, @gossner15
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
---

<!---->

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}

\newpage

This tutorial provides best-practice principles when standardising traits with terms of the Ecological Trait-data Standard (ETS, https://terminologies.gfbio.org/terms/ets/pages/).

We focus this document on the first of the three use-cases for the ETS vocabulary: The publication of project-specific primary data.  
To comply with the FAIR principles of scientific data management and aim at findability, accessibility, interoperability and reusability [@wilkinson16], trait data should apply the following principles when preparing data for publication:

- **Label your dataset!** data should be published with appropriate metadata and provide  a descriptive title and keywords to be findable by search engines. Keywords should include the general label "trait-data", and more specific labels such as "functional traits" or "morphometric traits", as well as the taxa covered by the dataset.

- **State terms of use!** To enable re-use, the metadata must contain the name of the author or owner of the data, the suggested citation for the data, as well as the license model or terms of use of the data. Ideally, these statements are provided by the data-hoster in a machine-readable way.

- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma- or tab-separated text files  ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g. UTF-8 or ASCII). For internationalisation, use points as decimal delimiters and set date-formats to ISO 8601, i.e. "YYYY-MM-DD".

- **Use standard terms!** Draw column or field names from the ETS vocabulary wherever possible. If co-variates or any ancillary data are not covered by the vocabulary, apply terms of the Darwin Core Standard (DwC).

- **Use URI references!** Wherever possible, the concepts used to describe the dataset should be referenced to globally accessible Uniform Resource Identifiers (URIs). These references allow data users and software tools to relate the dataset to well-defined concepts of taxa or traits, or measurement units, etc. The ETS-terms 'traitID' and 'taxonID' are defined to label fields with URIs as values.

- **Provide definitions!** Metadata must reference the data standards, taxonomic concepts and trait ontologies applied in the dataset as well as any user-defined terms and trait concepts. To provide own trait-concepts, use the terms provided within the ETS section "Terms for traitlists".

- **Publish raw data!** When publishing trait-data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation. Thus, data should be published as they have been recorded, i.e. in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard. This is true for the core terms, i.e. the trait value (the quality) and observation entity, but also for any co-variates reported along with it.

- **Keep original values!** For the core terms, data providers should complement user-defined entries (such as project specific taxon abbreviations) with standardised values by applying the '\*Std'-terms. For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g. elevation above sea-level. These should adhere to accepted standards wherever possible. If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g. by appending '\*Orig' to the column name. Also, additional co-variates, e.g. on environmental growth conditions of the specimen, should be provided in user-defined columns. Any user-defined columns must be adequately defined in the accompanying meta-information.

- **Report Uncertainty!** If publishing aggregate data (e.g. averages for multiple measurements), report the $n$ and dispersion of the aggregate value. This allows to weigh data from multiple sources when producing compilations. Any sources of bias or noise should be reported along with the data (e.g. observer ID, conditions of measurement etc.). Use terms of the ETS extension "Measurement or Fact".

- **Use Identifiers!** Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary (Fig. 2 *c* & *d*).
For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset. 
Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric coordinates could be linked via a 'measurementID'. 
In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
In larger compilations, a 'datasetID' allow to trace data origin to the primary source and keep track of authorship and original reference. 
Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via URIs, which point to external terminology services.

<!--
When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of simplicity. 
For spreadsheets, it is recommended to use a observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). 
As the long-table format draws from a defined set of columns, merging datasets is easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14]. 
-->

\newpage

# Example 1 - Apply standard terms for matrices

If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a observation-level wide-table format, with a column of values for each trait recorded and a row for each species (or taxon) for which data were available.
The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).

This wide-table format is straightforward for maintaining and managing new data for the data author. 
In terms of analysis, it allows to easily correlate traits with each other and analyse trade-offs between traits.
However, wide-tables need to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g. '999'). Additional information  (e.g. on variation of means or literature source) may be stored in secondary columns or accompanying datasheets. 
When applying the data standard to wide-table data, the columns for taxa or individuals may be labelled using the terms of the ETS. 
Taxon concepts should be referenced to globally available concepts using URIs, e.g. by referring to GBIF Taxonomy. Taxa authors and fine-grained higher taxonomy must then be consistent with the linked concept, or may be omitted as redundant and to avoid confusion.

Remarks should be provided in a human-readable form within the table and provided with further detail in the metadata, if necessary.

For wide tables, it is essential that the multiple trait columns are labelled in unambiguous ways and be linked to trait concepts provided in the metadata or a secondary spreadsheet.

\newpage

```{r, echo = FALSE, warning = FALSE, message=FALSE}
options(knitr.table.format = "latex")
library(kableExtra)

library(traitdataform)
pulldata('arthropodtraits')

caption = "Example 1 before standardisation; excerpt of original data from Gossner et al. (2015) \\textit{Scientific Data} 2:150013; data DOI: \\href{https://doi.org/10.5061/dryad.53ds2}{10.5061/dryad.53ds2} containing eight traits of arthropods from an European grassland; details provided in data publication \\href{https://www.nature.com/articles/sdata201513/tables/2}{Table 2}."

kable(arthropodtraits[1:45,c(1:8,17)], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

```

\newpage

```{r, echo = FALSE, warning = FALSE, message=FALSE}

arthropodtraits_head <- head(arthropodtraits, 45) 
names(arthropodtraits_head)[4] <- "scientificName"
names(arthropodtraits_head)[1:3] <- tolower(names(arthropodtraits_head)[1:3] )
arthropodtraits_head$taxonID <- get_gbif_taxonomy(levels(droplevels(arthropodtraits_head$scientificName)))$taxonID
names(arthropodtraits_head)[17] <- "measurementRemarks"
levels(arthropodtraits_head$measurementRemarks) <- c("","non-grassland species")

select_cols <- c(4,1,3,18,6:8,17)

caption = "Example 1 after standardisation. Column names mapped to ETS terms; taxa are mapped to taxon concepts in published terminologies via URIs. Trait columns are not standardised and respective definitions and units must be provided in the accompanying metadata."

kable(arthropodtraits_head[,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

```

\newpage

# Example 2 - provide structure in long-table

In observation wide-tables, each row is centered around a single physical occurrence of a species, e.g. phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns (see Table B1). 
Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients. Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism. 
This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data but rarely found for reported qualitative facts obtained from the literature.

A more effective method that allows for high flexibility and easier data merging is the storage of traitdata in long-table formats [@wickham14; @parr16]. 
This data format corresponds to the Entity-Quality Model: each row contains a single measurement or fact of a specific trait, referenced to a single specimen  (i.e., one occurrence)  which is assigned to a specific taxon  (see Table B2). 
Identifiers can provide structure and link measurements that are correlated by assigning a unique ID for each single individual. 
Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier. 
Long-table datasets purport multiple advantages for data handling (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. 
As long-table data share a larger set of columns, merging of datasets is in most cases easier for the end-user.

However, a disadvantage of the long-table data structure is the mixing of units or even factor levels in the 'traitValue' column. The mixed content of this column prohibits filtering data based on their values. Data-splitting techniques or a database system are tools to disentangle the data.

\newpage

```{r,  echo = FALSE, warning = FALSE, message=FALSE}

# 21228 life history traits of moths

# 21247 morphological traits of moths

moths <- read.service(21247, dec = ",", user = "fdschneider", pswd = "Ro9G3s8w")

attr(moths, 'thesaurus') <-  traitdataform:::as.thesaurus(
  weight = traitdataform:::as.trait("body_weight",
                                    expectedUnit = "mg", valueType = "numeric",
                                    identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_weight"),
  wing_length = traitdataform:::as.trait("wing_length",
                                         expectedUnit = "mm", valueType = "numeric",
                                         identifier = ""),
  wing_width = traitdataform:::as.trait("wing_width",
                                        expectedUnit = "mm", valueType = "numeric",
                                        identifier = ""),
  wing_area = traitdataform:::as.trait("wing_area",
                                       expectedUnit = "mm2", valueType = "numeric",
                                       identifier = "http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Wing_surface"),
  wing_loading = traitdataform:::as.trait("wing_loading",
                                          expectedUnit = "mg mm-2", valueType = "numeric",
                                          identifier = "")
)

attr(moths, 'taxa') <- "species"
attr(moths, 'keep') <-  c(locationID = "plot")
attr(moths, 'individuals') <- "label"
attr(moths, 'units') <- c(weight = "mg", wing_length = "mm", wing_width = "mm", wing_area = "mm2", wing_loading = "mg mm-2")

caption <- "Example 2 before standardisation. Data are stored in observation wide-table format, i.e. one row contains a single specimen with multiple trait measurements (in columns). Units and specific information on measurements must be reported in the metadata of the dataset. Data originate from Mangels et al. (unpublished)."
select_cols <- c("label","species","plot","weight","wing_length","wing_width","wing_area","wing_loading")

kable(moths[1:45,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

```

\blandscape

```{r, echo = FALSE, warning = FALSE, message=FALSE}

mothsStd <- standardize(moths)
mothsStd <- mothsStd[order(as.numeric(substr(mothsStd$occurrenceID,2,10))),]

caption <- "Example 2 after standardisation into measurement long-table format. Each row contains a single measurement linked to an observation of an individual specimen, i.e. an occurrence, via identifiers."
select_cols <- c("scientificName","traitName","traitValue","traitUnit","traitID","scientificNameStd","taxonID","occurrenceID", "measurementID", "locationID")

kable(mothsStd[1:35,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

```

\elandscape

# Example 3 - apply ETS extensions in trait-data table

If data contain covariate information, the two-dimensional data table can cover these by adding columns defined by the extensions of the ETS. 
The three extensions complement the core terms according to different layers of metadata information: 1) the "taxon" extension includes terms to describe more details on the taxon concept applied, 2) the "occurrence" extension provides terms to describe the indivdual specimen and its observation context, and 3) the "Measurement or Fact" extension adds information on the methodology or context of the measurement of the trait value.

Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain. As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16. 
The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
Although we cannot cover the structure of relational databases in this tutorial, the standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.

\blandscape

```{r, echo = FALSE, warning = FALSE, message=FALSE}
pulldata('heteroptera_raw')

heteroptera_raw$Center_Sampling_region <- gsub("°", " ", heteroptera_raw$Center_Sampling_region)
select_cols <- c(1:7,9,11,12:15)

# https://figshare.com/collections/Morphometric_measures_of_Heteroptera_sampled_in_grasslands_across_three_regions_of_Germany/3307611/1
caption = "Example 3 before standardisation; Excerpt of data from Gossner et al. (2015b) Ecology, 96, 1154–1154. doi: \\href{https://doi.org/10.1890/14-2159.1}{10.1890/14-2159.1} on measurements of traits in Heteroptera."

kable(heteroptera_raw[1:35,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

```

```{r, echo = FALSE, warning = FALSE, message=FALSE}
pulldata('heteroptera_raw')

heteroptera_long <- as.traitdata(heteroptera_raw) 
heteroptera_long <- standardize.taxonomy(heteroptera_long)
heteroptera_long <- heteroptera_long[order(as.numeric(heteroptera_long$occurrenceID)),]
heteroptera_long$basisOfRecordDescription <- as.factor(heteroptera_long$basisOfRecordDescription)

heteroptera_long$decimalLatitude <-  gsub("°", " ", substr(heteroptera_long$verbatimLocality,0,5))
heteroptera_long$decimalLongitute <-  gsub("°", " ", substr(heteroptera_long$verbatimLocality,10,14))

heteroptera_long$decimalLatitude <- round(as.numeric(measurements::conv_unit(heteroptera_long$decimalLatitude, from = 'deg_dec_min', to = 'dec_deg')),5)
heteroptera_long$decimalLongitute <- round(as.numeric(measurements::conv_unit(heteroptera_long$decimalLongitute, from = 'deg_dec_min', to = 'dec_deg')), 5)
heteroptera_long$basisOfRecord <- as.factor("PreservedSpecimen")

caption = "Example 3 after standardisation; application of terms of the ETS extensions structure the two-dimensional data table. Excerpt of data from Gossner et al. (2015b) Ecology, 96, 1154–1154. doi: \\href{https://doi.org/10.1890/14-2159.1}{10.1890/14-2159.1}."

select_cols <- c(1,2,3,4,8,6,15,18,22,16,20,21)
kable(heteroptera_long[heteroptera_long$traitName %in% c("Body_length", "Body_width", "Body_height", "Thorax_width"),select_cols][30:65,], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

```

\elandscape

# References

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}

\footnotesize

### 2018-10-19 manuscript/appendix_A.Rmd
adjust appendix_A.Rmd to new template and author list

@@
 ---
-title: "Towards an Ecological Trait-data Standard - Appendix A"
+title: "List of published trait datasets and databases"
+subtitle: "Towards an Ecological Trait-data Standard - Supplementary Material A"
 author:
-- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
-- name: Malte Jochum \footnotemark[4]
-- name: Gaëtane Le Provost \footnotemark[3]
-- name: Andreas Ostrowski \footnotemark[6]
-- name: Caterina Penone \footnotemark[4]
+- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
 - name: David Fichtmüller \footnotemark[5]
@@
 geometry: margin=0.5in
+documentclass: scrartcl
 classoption: landscape
@@
 \footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
-\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
-\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
+\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
+\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
 \footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
-\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
-\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
 \footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
@@
 library(kableExtra)
+
 gg <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTfQOIHjJv5x_YHgq14DR7VLTvycfsl2Njj2Ip4p2a5kca8AX5t5cmaial6Bg-MX9pBDOJaJfnlR7Qj/pub?output=csv", encoding = "UTF-8")
@@
 
-order <- c(14, 15,30,  # all organisms
+order <- c(14, 15,  # all organisms
            1,4,5,13,18,20,21, # plants
-           11,16,25,33, # all animals
+           11,16,25,30,33, # all animals
            3,17,31, ## birds
@@
            26,8,27, # other vertebrates
-           32,2, 6,9,23,10,22,29,# invertebrates
-           24,28, # fungi
-           19 # other
+           32,2, 6,9,23,10,22,29,19, # invertebrates
+           24,28 # fungi
+           # other
            )

### 2018-10-19 manuscript/manuscript.Rmd
finalize manuscript for last internal circuit, update fig1.pnd

@@
 geometry: margin=1.4in
+documentclass: scrartcl
 output:
@@
 
-\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
-\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
+\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com, c/o Birgitta König-Ries
+Friedrich-Schiller-Universität Jena, Institut f. Informatik, Ernst-Abbe-Platz 2, 07743 Jena
+}
+\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
 \footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
@@
 
+running title: Towards an Ecological Trait-data Standard
+
 \newpage
 \linenumbers
+\renewcommand\linenumberfont{\normalfont\tiny}
 
@@
 For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
-The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix.
-The package 'NatDB' [@pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. 
+The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits wide-table.
+<!--The package 'NatDB' [@pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. -->
 FENNEC [@ankenbrand18], an online tool or self-hosted service that can extract trait information from multiple sources for a target species community.
@@
 and Simons, N.K., 2018 Ecological Traitdata Standard v0.9, DOI:  10.5281/zenodo.1441132, URL: https://terminologies.gfbio.org/terms/ets/pages/".
-Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their URIs,.
+Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their URIs. 
 The definitions of terms are hosted on the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
@@
 \caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
 Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
@@
 \end{figure}
-
-<!--
-## Best practice for primary data publication
-
-In this section we focus on the publication of project-specific primary data by data providers (use case #1, see main document).
-We assume that data managers and software developers who apply the vocabulary in use-case 2 and 3 are more fluent with the use of exchange formats and standard definitions and have established own best-practice principles.
-
-Data publications of traits should apply the FAIR principles of scientific data management and aim at findability, accessibility, interoperability and reusability [@wilkinson16].
-By drawing column names and applying definitions of the ecological trait-data standard vocabulary, and providing reference to this standard in the metadata, authors can ensure that these principles are followed.
-
-Above all, when publishing primary data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation.
-Thus, data should be published as they have been recorded, i.e.
-in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard.
-This is true for the core terms, i.e.
-the trait value (the quality) and taxon name (the entity), but also for any co-variates reported along with it.
-For the core terms, data providers should use the '\*Std'-terms for standardised entries.
-For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g.
-elevation above sea-level.
-If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g.
-by appending '\*Orig' to the column name.
-Also, additional co-variates, e.g.
-on environmental growth conditions of the specimen, should be provided in user-defined columns.
-Any user-defined columns must be adequately defined in the accompanying meta-information.
-
-When applying the vocabulary in a primary data publication, most authors will chose a two-dimensional spreadsheet rather than relational databases for reasons of simplicity. 
-For spreadsheets, it is recommended to use a observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). 
-As the long-table format draws from a defined set of columns, merging datasets is easier.
-Long-table datasets also purport multiple advantages for data manipulation [e.g.
-filtering, sub-setting and aggregating data, @wickham14].
-
-Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary (Fig. 2 *c* & *d*).
-For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset.
-Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
-In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
-In larger compilations, a 'datasetID' allow to trace data origin to the primary source and keep track of authorship and original reference. 
-Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via URIs, which point to external terminology services (see above for resources).
-Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
-As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16.
-The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
-The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.
-
-For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
-UTF-8 or ASCII).
-The metadata should contain unambiguous information on authorship and conditions for re-use of the data, ideally by specifying a data license  (e.g.
-Creative Commons licenses) or public domain dedication.
-
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
-"Schneider, F.D., Jochum, M.,  Le Provost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2018 Ecological Traitdata Standard v0.8, DOI:  10.5281/zenodo.1255287, URL: https://ecologicaltraitdata.github.io/ETS/v0.8/".
-In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
-Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
-via the API; **in preparation!**).
-To enable reuse in downstream analysis, the metadata should further specify a license, author and bibliographic citation that applies to the dataset.
-
-Supplementary Material B contains three worked examples of how to apply the vocabulary to own trait-datasets. 
--->
 
@@
 
+The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+Any future development of the vocabulary is coordinated via https://github.com/EcologicalTraitData/ETS/.
+
 Supplementary Material A - static excerpt of a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use.
@@
 
-Supplementary Material B - contains three worked examples of how to apply the standard vocabulary to 1.) simple two-dimensional data using only core terms, 2.) advanced use with extensions and identifiers within one spreadsheet, 3.) compiled data from multiple sources using metadata terms, 4.) a relational database structure with multiple spreadsheets.
-
-The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
-Any future development of the vocabulary is coordinated via https://github.com/EcologicalTraitData/ETS/.
+Supplementary Material B - contains three worked examples of how to  1.) apply the core ETS vocabulary to species$\times$traits wide-trables, 2.) convert data to long-table format using identifiers 3.) apply ETS extensions and identifiers to include co-variate information on different layers of the dataset.
 

### 2018-10-31 manuscript/manuscript.Rmd
final comments on review, update bibliography

@@
 - name: Nadja K. Simons \footnotemark[7]
-csl: amnat.csl
+csl: mee.csl
 fontsize: 11pt
@@
 In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data.
-We discuss potential incentives and challenges for the wide adoption of such a standard by data-providers.
+We discuss potential incentives and challenges for the wide adoption of such a standard by data providers.
 4.
 The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
-By aiding decentralised trait-data standardisation, our data standard may facilitate the use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.
+By aiding decentralised trait-data standardisation, our data standard may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.
 
@@
 ----- | ---------
+Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
+client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
+Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/).
 Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
-Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/)
-Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted
-Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*
-Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
-trait data), that is recognised by a large proportion of members of the application context
+Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted.
+Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g.
+using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
+Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/).
+Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries.
+File repository | A storage or archiving of datasets on file-hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org; online repositories make data available for public access, provide *metadata*, state conditions of re-use, and (not always) facilitate citations via persistent identifiers, e.g.
+DOIs (Digital Object Identifiers).
+Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *relational database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  .
+Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
+Occurrence | The observation context of a single entity of a taxon, i.e.
+data on an individual observed at a particular place and time.
+(http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context.
+Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally.
+Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
+Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs).
+Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
+specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
+Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
 Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g.
-may define a hierarchy of broader or narrower terms
-Ontology | *Controlled vocabulary* that defines *concepts* in relation to each other in a formal language, e.g.
-provides cross-references and semantic meaning; a semantic model of the objects and their relationships in a domain of interest [@gruber95]; commonly published in OWL format for machine readability.
-Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
-specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
-Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
-Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
-Database | A suite of data compiled from multiple *datasets*, i.e.
-from multiple study contexts or observation types; may take the form of a single two-dimensional *data tables*, but mostly is organised as multiple data tables linked via *identifiers* into *relational databases*; may be built using a specific database software or language (e.g.
-SQL)
-Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
-Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally
-Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *relational database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*
-Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
+may define a hierarchy of broader or narrower terms.
+Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
 Web Service | A service to access *online databases* directly and computationally, i.e.
-by calls from a software tool; returns content of the database according to the request in different formats and *data standards*
-Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
-client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider
-Online Portal | A website designed as a platform for the exchange of information, e.g.
-trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
-File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via persistent identifiers, e.g.
-DOIs (Digital Object Identifiers)
-File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
-Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
-Darwin Core Standard (DwC) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
-Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified e.g.
-using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns [@gbif17,  http://tools.gbif.org/dwca-assistant/]
-Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits @moretti17
-Occurrence | The observation context of a single entity of a taxon, i.e.
-an individual observed at a particular place and time (http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context
+by calls from a software tool; returns content of the database according to the request in different formats and *data standards*.
 
@@
 
+<!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time.-->
+<!--Database | A suite of data compiled from multiple *datasets*, i.e.
+from multiple study contexts or observation types; may take the form of a single two-dimensional *data tables*, but mostly is organised as multiple data tables linked via *identifiers* into *relational databases*; may be built using a specific database software or language (e.g.
+SQL).-->
+<!--Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
+trait data), that is recognised by a large proportion of members of the application context.-->
+<!--File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org.-->
+<!--Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits [@moretti17].-->
+<!--Online portal | A website designed as a platform for the exchange of information, e.g.
+trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access.-->
 <!--
@@
 Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
-Recent years have seen a proliferation of trait-based research in a wide range of fields and have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
+Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
 This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
-Species traits have been suggested as indivator variable for monitoring  ecosystem health at the individual level [@kissling18].  
+Species traits have been suggested as indicator variables for monitoring  ecosystem health at the individual level, like for instance changes in average body sizes in a population of fish [@kissling18].
  <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
@@
 reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements-->
-To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centred around particular research questions [e.g.
+To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g.
 PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
@@
 These include software to access trait data from the internet [e.g.
-@chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of standard terms [@walls12; @garnier17].
+@chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of consensus terms [@walls12; @garnier17].
 
@@
 Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
-As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services (e.g.
-Figshare.com, see Table 1), where no data standards areenforced upon the uploaded material [@wilkinson16].
+As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services, where no data standards are enforced upon the uploaded material [@wilkinson16].
 It is thus likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient reuse and synthesis of published datasets.
@@
 In this paper, we review existing trait databases and online portals, and initiatives for data standardisation from the pragmatic view of data providers, data curators and data users, as well as database managers.
-We discuss current efforts to making trait data visible, accessible, interoperable and re-useable in downstream data analysis, as demanded by the FAIR guiding principles for scientific data [@wilkinson16].
+We discuss current efforts to make trait data visible, accessible, interoperable and re-useable in downstream data analysis, as demanded by the FAIR guiding principles for scientific data [@wilkinson16].
 Furthermore, we show how the current deficit in the standardisation of primary data hampers the implementation of interoperability and reuse of trait data. 
-Based on these considerations, we propose a versatile vocabulary for describing trait datasets that builds upon, and is compatible with , existing terminology standards for biodiversity data. 
-Since this standard vocabulary relies on the adoption of a broad research community, we discuss incentives for its use and lay out mechanisms for a future consensus-building and community development towards an accessible ecological trait data standard.
+Based on these considerations, we propose a versatile vocabulary for describing trait datasets that builds upon, and is compatible with, existing terminology standards for biodiversity data. 
+Since this standard vocabulary relies on the adoption by a broad research community, we discuss incentives for its use and lay out mechanisms for future consensus-building and community development towards an accessible ecological trait data standard.
 <!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
@@
 Often, the scientific scope and focus differs between data providers measuring and assessing the trait data in the first place and data users who re-use published data for a broader synthesis application.
-Data curators are taking up the task of providing compiled and harmonised datasets or databases.
+Furthermore, data curators and data managers are taking up the task of providing compiled and harmonised data and prepare them for future use and long-term preservation.
 Data managers are concerned with the development of complex digital infrastructures for handling and analysing large amounts of data.
-These idealised roles of researchers cooperatively are dealing with trait-data standardisation throughout the data life-cycle.
+These are idealised roles of researchers that are dealing with trait-data standardisation throughout the data life-cycle.
 In this chapter, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):
@@
 1. Initiatives that provide trait *datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature.
-2. Initiatives that aim to harmonise trait data from the literature or from direct measurements into trait *databases* and make those data widely available. 
-3.
-Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits and provide standard *terminologies* in the form of *thesauri* and *ontologies*.
-4. Initiatives that aim to leverage *relational database* structures and *semantic web* technology to link trait data to a wider set of biodiversity data.
+2. Initiatives that aim to harmonise trait data from the literature or from direct measurements into database infrastructures and make those data widely available on the Internet. 
+3. Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits and provide standard *terminologies*.
+4. Initiatives that aim to combine data (1 & 2) and terminologies (3) to link trait data to a wider set of biodiversity data.
 
 We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
-We show how the degree of trait-data standardisation in existing datasets is highly variable and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources.
+We show how the degree of trait-data standardisation in existing datasets is highly variable, and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources.
 The objective of this review is to raise awareness of the generic structure of trait data and aid researchers in how to share and publish their own datasets in an appropriate form.
@@
 In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g.
-reproductive traits, feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books.
+feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books.
 With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
@@
 
-Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g.
-Figshare.com, see Table 1), which warrant long-term accessibility and citeability via DOIs, and govern data sharing via licensing.
-These platforms offer publicly accessible *repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management.
+Today, the online publication of such data is greatly facilitated by file hosting services (e.g.
+Figshare, Zenodo, Researchgate, Data Dryad), which warrant long-term accessibility, and citeability via DOIs, and govern data sharing via license statements.
+These platforms offer the hosting of publicly accessible *file repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management.
 However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g.
 .xlsx, .docx) or binary (e.g.
-.pdf) data formats which make extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.
+.pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk.
 Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation.
@@
 For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives.
-Furthermore, metadata varies in its level of detail, e.g.
+Furthermore, the metadata vary in their level of detail, e.g.
 for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
 In terms of structure, trait data usually are reported in a species$\times$traits wide-table format.
-In this intuitive *data table*, each row contains a species (or taxon) for which multiple traits are reported in columns.
+In this intuitive data table, each row represents a species (or taxon) for which multiple traits are reported in columns.
 <!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).
@@
 
-## Database initiatives
+## Data aggregation initiatives
 
@@
 @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
-While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these *databases* apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
+While mostly concerned with issues of heterogeneity in units or levels of categorical variables, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other datasets.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
 -->
 Documentation of metadata and methodology differs in the level of detail, depending on the research focus. 
-Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms, but they may be updated irregularly.
+Trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiveristy informatics.  
+Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.
 
-As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
+As they deal with much larger amounts of data, initiatives that compile data from natural history museum collections are traditionally more concerned with standardisation. 
 The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g.
-www.markmybird.org),
+www.markmybird.org).
 For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16].
@@
 <!-- database backend, selective download, user access management --> 
-Specialised *online portals* have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
+Specialised online portals have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
 <!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
 For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform.
-The *online database* enables selective data download and user permission and rights management.
+The online portal enables selective data download and user permission and rights management.
 As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
-For animal trait data, however, a single unified portal and harmonising scheme is still lacking.
+For animal trait data, however, a single unified platform and harmonising scheme is still lacking.
 The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].
 
-Nonetheless, initiatives for particular groups of animals do exist: examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), or the Coral Trait Database [@madin16, see Supplementary Material A, Table A1].
+Nonetheless, initiatives for particular groups of animals do exist: examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), the Coral Trait Database [@madin16], or the Global Ants Database [@parr17, see Supplementary Material A, Table A1].
 The role of online portals and database initiatives in standardising data and making them more accessible is paramount. 
@@
 
-## Thesauri and Ontologies for traits
+## Terminology standards for traits
 
@@
 A major challenge in trait-data standardisation is the lack of widely accepted and unambiguous trait definitions [@williams06; @kissling18].
-Previous standard definitions of trait *concepts* range from listings of selected definitions in *terminologies*, over well-defined *method handbooks* and comprehensive *thesauri*, to formalized definitions of trait concepts in *ontologies*.
+Previous standard definitions of trait *concepts* range from listings of selected definitions in vocabularies, over well-defined method handbooks and comprehensive *thesauri*, to formalized definitions of trait concepts in *ontologies*.
 The initiatives behind method handbooks, thesauri and ontologies are primarily concerned with building community consensus for trait definitions.
@@
 
-A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
+A *thesaurus* provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
 In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms.
@@
 For example, a measurement of fruit mass could be linked to the definition of the term within the TOP Thesaurus of plant traits [@garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
-
-Ontologies link defined terms by formally describing the relationships between them and providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14].
+*Ontologies* link defined terms by formally describing the relationships between them and providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18].
 Ontologies define terms based on other well-defined terms from published ontologies.
-The Planteome Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
+The Plant Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
 Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12].
@@
 
-Comprehensive trait thesauri have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
+Comprehensive trait thesauri have been developed in the TOP Thesaurus of Plant Traits, which is employed in the TRY database [@garnier17], and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
 Ontologies of trait definitions have been developed for plants [e.g.
 the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
-the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13].
+the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13].
 The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
@@
 -->
-Meta-ontology initiatives, like OBO Foundry [@smith07, http://www.obofoundry.org/], Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/), or the GFBio Terminology service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for advanced trait ontologies and offer harmonized web services computational access.
 
 To conclude, there is already a suite of globally available thesauri and ontologies for traits available.
-However, definitions in some domains are better covered than others and the interlinkage and accesibility of ontologies can be much improved to fulfil semantic web standards.
-While these defined vocabularies are widely used in biodiversity data management, distributed trait-data of smaller projects published in general-purpose file servers make little use of them.
+However, definitions in some domains are better covered than others [@kissling18] and the interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
+To this end, the OBO Foundry is providing a development platform for (biological) ontologies and offers review and quality control [@smith07, http://www.obofoundry.org/].
+While defined vocabularies are increasingly used in biodiversity data management, distributed trait-data of smaller projects published in general-purpose file servers rarely refer to standard terminologies.
+Finding and applying the most suited and highest quality ontology from the range of available ontologies is not an easy task for ecological researchers. 
+To mitigate this effort, meta-ontology initiatives, like Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/, @whetzel11), or the GFBio Terminology service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for trait ontologies, structured browsing, and harmonized web services for computational access.
 
@@
 
-While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
+While trait thesauri and trait ontologies typically define concepts of measurements and observations for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
 
@@
 
-These relationships between a trait measurement and an individual organism as an entity of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These terminology frameworks specify terms and classes to describe the general structure of biodiversity databases.
-The terms may be used for  defining columns in data tables that contain measurement values, units and factor levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
+These relationships between a trait measurement and an individual organism as an entity of a particular taxon have been formalised in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+These frameworks specify terms and classes to describe the general structure of biodiversity databases.
+The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
 The terminologies provided by these standards are quite universal and would even cover most use cases of trait data.
 A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
-For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) refers to DwC terms for defining its database content.
+For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. 
+Both DwC and ABCD are ratified standards of the Biodiversity Information Standards TDWG (Taxonomic Databases Working Group; www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.
 
@@
 The structure of @kattge11a can be mapped to the  Extensible Observation Ontology [OBOE, @madin07; @schildhauer16], which formalizes observations andd measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
+Besides OBOE, @kissling18 discuss further ontologies that formalize the structure of observation data and attest that for the use with trait data the various ontologies are difficult to integrate.
 
-In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach of an integrated structure for trait data.
+The  Encyclopedia of Life (EOL) project has proposed TraitBank [@parr16] as a standard structure for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach of an integrated structure for trait data.
+It is the suggested format for upload of data to.
 The framework employs established terms provided by the DwC, relates trait definitions to ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
@@
 The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review [@franz16].
-Initiatives that aim at providing a stable reference for taxon concepts are for instance the EOL Catalogue of Life [http://www.catalogueoflife.org/, @roskov18], the GBIF Backbone Taxonomy [@gbifsecretariat17], or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).
+Initiatives that aim at providing a stable reference while tracking the changing taxon concepts are for instance the Catalogue of Life [http://www.catalogueoflife.org/, @roskov18] or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).
+The GBIF Backbone Taxonomy [@gbifsecretariat17] collects and bundles these terminologies into a single reference framework.
 
@@
 
-In sum, we attest a gap in the trait-data structures developed for data curators and managers and the data input by data providers.
-Hardly any of the aforementioned trait datasets or databases for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
-As it stands, re-using these data in larger compilations is error-prone and inefficient and the potential for a broad synthesis is diminished.
+In sum, we attest a gap between the trait-data structures developed for data curators and managers and the data input produced by data providers.
+Hardly any of the aforementioned standalone or aggregated trait datasets for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
+As it stands, re-using these data in larger compilations or integrating them into structured database initiatives is error-prone and labour-intensive and the potential for a broad synthesis is diminished.
 
-One likely reason for this is complexity: the proposed data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
+One likely reason for this lack of standardisation is the complexity of the task: the proposed data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
 In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
@@
 
-By filling this gap, data-brokering services [e.g.
-the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g.
+By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14;e.g.
+Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g.
 KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
@@
 Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains to facilitate the use and analysis of trait-data. 
-For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB.
+For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data directly from their source, including Birdlife, EOL TraitBank or BetyDB.
 The package 'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits wide-table.
 <!--The package 'NatDB' [@pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. -->
-FENNEC [@ankenbrand18], an online tool or self-hosted service that can extract trait information from multiple sources for a target species community.
+FENNEC [@ankenbrand18] is an online tool or self-hosted service capable of extracting trait information from multiple sources for a target species community.
+Such decentralised tools
 
@@
 Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards. 
-A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
-To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
-For example, the accessibility of ontologies will increase if open APIs are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
-Software then can assist researchers in linking own data to globally defined concepts.
+A more widespread implementation of ontologies would advance the possibilities to integrate datasets and reduce noise and uncertainty when aggregating data.  
+First, groups of trait researchers must take up the task of developing consensus definitions into semantically defined ontologies that are useful for their use case.
+Platforms like OBO Foundry can help structuring this process.
+Second, the reference to ontologies and thesauri must be incentivised and facilitated for individual data providers by the development of tools for  matching concepts from the available ontologies to their data.
+Third, frameworks for providing trait data in a unambiguous and machine-readable structure must be simplified to match the limited resources of small and intermediate research projects.
+This can be achieved by extending documentation or providing tools for the application of existing ontology frameworks and database structures, or by defining easy-to-use exchange formats that enable the interoperability of data at minimal effort.
+<!--
+This requires the availability open *Application Programming Interfaces (APIs)* for ontologies provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools.
+-->
+<!--To bring all these tools for data management together, a common vocabulary and definition of consensus concepts and exchange formats is required that connects to all steps of the data life-cycle.-->
 
-To bring all these tools for datamanagement together, a common vocabulary and definition of consensus concepts and exchance formats is required that connects to all steps of the data life-cycle.
-Despite centralised data management efforts, however, no unified definition of traits as a data type with a well-defined exchange format or data structure has emerged across the multiple sub-disciplines of trait-based research.
-In the following chapter, we propose the a unified vocabulary for trait-data that can form the basis of such an exchange format.
+Despite centralised data management efforts, however, no unified and widely-applicable definition of a data type for traits has emerged across the multiple sub-disciplines of trait-based research. 
+In the following chapter, we propose a unified vocabulary for trait-data that can serve as a minimal consensus exchange format.
 The simplicity of this standard terminology will lower the thresholds and offer high pay-off in the visibility and reuse of published data.

### 2018-10-31 manuscript/manuscript.Rmd
finalize manuscript for pre-print 2.0

@@
 
+\renewcommand{\baselinestretch}{2.0}
+
 <!-- 
@@
 
-\newpage
-
-\tiny
-
-Term | Definition 
------ | ---------
-Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
-client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
-Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/).
-Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
-Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted.
-Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g.
-using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/).
-Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries.
-File repository | A storage or archiving of datasets on file-hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org; online repositories make data available for public access, provide *metadata*, state conditions of re-use, and (not always) facilitate citations via persistent identifiers, e.g.
-DOIs (Digital Object Identifiers).
-Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *relational database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  .
-Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
-Occurrence | The observation context of a single entity of a taxon, i.e.
-data on an individual observed at a particular place and time.
-(http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context.
-Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally.
-Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
-Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs).
-Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
-specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
-Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
-Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g.
-may define a hierarchy of broader or narrower terms.
-Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
-Web Service | A service to access *online databases* directly and computationally, i.e.
-by calls from a software tool; returns content of the database according to the request in different formats and *data standards*.
-
-Table: Glossary of terms from the biodiversity data-management context as they are used in this paper; draws from @garnier17.
-
-<!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time.-->
-<!--Database | A suite of data compiled from multiple *datasets*, i.e.
-from multiple study contexts or observation types; may take the form of a single two-dimensional *data tables*, but mostly is organised as multiple data tables linked via *identifiers* into *relational databases*; may be built using a specific database software or language (e.g.
-SQL).-->
-<!--Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
-trait data), that is recognised by a large proportion of members of the application context.-->
-<!--File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org.-->
-<!--Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits [@moretti17].-->
-<!--Online portal | A website designed as a platform for the exchange of information, e.g.
-trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access.-->
-<!--
-Measurement |  
-Aggregated Value |
-Value | 
-Fact | 
--->
-
-\normalsize
-
-\newpage
 <!-- CP:  I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.
@@
 -->
+
+\newpage
 
@@
 Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->
+
+\newpage
+\scriptsize
+\singlespacing
+
+Term | Definition 
+----- | ---------
+Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
+client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
+Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/).
+Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
+Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted.
+Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g.
+using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
+Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/).
+Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries.
+Database | A structured colletion of data, usually organised as multiple data tables linked via *identifiers* into relational databases; usually constructed using a specific database management system, i.e.
+a software to provide a (offline or online) user interface.
+File repository | A storage or archiving of datasets on file-hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org; online repositories make data available for public access, provide *metadata*, state conditions of re-use, and (not always) facilitate citations via persistent identifiers, e.g.
+DOIs (Digital Object Identifiers).
+Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  .
+Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
+Occurrence | The observation context of a single entity of a taxon, i.e.
+data on an individual observed at a particular place and time.
+(http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context.
+Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
+Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
+specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
+Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
+Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g.
+may define a hierarchy of broader or narrower terms.
+Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
+Web Service | A service to access *databases* directly and computationally, i.e.
+by calls from a software tool; returns content of the database according to the request in different formats and data standards.
+
+Table: Glossary of terms from the biodiversity data-management context as they are used in this paper; draws from @garnier17.
+
+\normalsize
+\newpage
+
+\doublespacing
+
+<!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time.-->
+<!--Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
+trait data), that is recognised by a large proportion of members of the application context.-->
+<!--File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org.-->
+<!--Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits [@moretti17].-->
+<!--Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally.-->
+<!--Online portal | A website designed as a platform for the exchange of information, e.g.
+trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access.-->
+<!--Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs).-->
+<!--
+Measurement |  
+Aggregated Value |
+Value | 
+Fact | 
+-->
 
@@
 Finding and applying the most suited and highest quality ontology from the range of available ontologies is not an easy task for ecological researchers. 
-To mitigate this effort, meta-ontology initiatives, like Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/, @whetzel11), or the GFBio Terminology service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for trait ontologies, structured browsing, and harmonized web services for computational access.
+To mitigate this effort, meta-ontology initiatives, like Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/, @whetzel11), or the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for trait ontologies, structured browsing, and harmonized web services for computational access.
 
@@
 The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
-More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e.
+More specifically, a trait dataset contains information on quantitative measurement* or qualitative facts (i.e.
 trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e.
@@
 
-As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
-The standard vocabulary is accessible at https://terminologies.gfbio.org/terms/ets/pages/.
-The aim of this vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement details. 
-Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases: 
+As a response to the challenges outlined above, we propose a versatile standard vocabulary for trait-based ecological research. 
+The Ecological Trait-data Standard vocabulary is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different degrees of measurement details.
+Rather than prescribing a data structure, the vocabulary is intended as a more inclusive terminology that can be used in three major use cases:
+
 1.
@@
 2.
-by data users and curators: as a consensus vocabulary when compiling data from distributed sources into databases, e.g.
+by data users and curators: as a consensus vocabulary when compiling data from distributed sources into aggregate datasets, e.g.
 to map standardised columns and refer to taxa and trait definitions in a uniform way, and
 3.
-by data managers: as a data exchange format of between databases, web services and software tools, e.g.
+by data managers: as a data exchange format of between online resources, web services and software tools, e.g.
 when providing database queries via a web service or defining input and output formats of software packages.
@@
 1*c*).
-This encompasses a wide range of data types, including numeric, integer, intervals or ratios, or even values from an ordinal scale.
-Qualitative facts are assignments of an entity to a categorical level, e.g.
-of a behavioural or life-history trait (Fig.
-1*d*).
+This encompasses a wide range of numeric data types, including continuous, binary, integer, intervals or ratios, as well as ordinal values.
+Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
 Beyond these core observations, further information might be available that further specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation and date information).
@@
 Our vocabulary thus provides core terms to reflect the minimal Entity-Quality structure of trait data. 
-We further offer three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", and "Occurrence", "Taxon"; see section on extensions below). 
+We further offer three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", "Occurrence", and "Taxon"; see section on extensions below). 
 All terms may be applied to describe columns of a data table, or to label content of a database (see Box 1 / Supplementary Material B for best practice principles and examples for publishing primary data).
@@
 Some of the terms expand and refine terms of DwC and its extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516).
-Wherever our vocabular refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
+Wherever our vocabulary refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
 The vocabulary can be mapped onto other existing ontologies (e.g.
@@
 <!--### Relate to global taxon and trait definitions -->
-The trait name and taxon names as would ideally use unambiguous terms assigning both to clearly-defined concepts. 
+The trait name and taxon names would ideally use unambiguous terms assigning both to clearly-defined concepts. 
 However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
@@
 2 *c*).
-For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  `https://www.gbif.org/species/3117424`; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  `http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass`.
+For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to Flora Phenotype Ontology would be  'http://purl.obolibrary.org/obo/FLOPO_0005265'.
+
 <!-- units and factor levels-->
@@
 - The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.
-(https://terminologies.gfbio.org/terms/ets/pages/#extension-taxon)
+<!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-taxon)-->
 - The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature reference for the reported value, the method of measurement or statistical method of aggregation.
@@
 literature source or expert reference).
-(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)
+<!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)-->
 - The `Occurrence` extension contains vocabulary to describe information on the observation context of individual specimens, such as sex, life stage or age.
 This also includes the method of sampling and preservation, as well as the date and geographical location, which provide an important resource to analyse trait variation due to differences in space and time.
-(https://terminologies.gfbio.org/terms/ets/pages/#extension-occurrence)
+<!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-occurrence)-->
 
@@
 
-Identifiers also can be used to provide a structure within a two-dimensional data table, e.g.
+Identifiers can also be used to provide a structure within a two-dimensional data table, e.g.
 to link rows of measurements on the same individual (using `occurrenceID`) or observations at the same location (using `locationID`).
@@
 
-Higher-level information about structure, authorship and data ownership and terms of use should be considered when sharing and working with trait datasets.
-Our metadata vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) provides terms to label these information. 
+Higher-level information about structure, provenance of data, authorship and data ownership and terms of use should be considered when sharing and working with trait datasets [@michener06; @kissling18].
+Our metadata vocabulary ( https://terminologies.gfbio.org/terms/ets/pages/#metadata-vocabulary) provides terms to label these information. 
 In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g.
@@
 Ideally, this definition refers to or refines terms from published trait ontologies.
-By providing a minimal vocabulary for trait lists (see https://ecologicaltraitdata.github.io/ETS/#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
+By providing a minimal vocabulary for trait lists (see  https://terminologies.gfbio.org/terms/ets/pages//#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
 This vocabulary might also prove useful for the future publication of trait ontologies. 
@@
 \caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (occurrence) or the origin or confidence on the reported measurement or fact.}
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e.
+occurrence) or the origin or confidence on the reported measurement or fact.}
 
@@
 However, for these incentives to take effect, data publications and data citations must become viewed as a valid scientific contribution to the community, and  recognized in the professional evaluation of individual researchers [@costello09; @roche15].
-To mitigate the cost of data standardisation efforts, supporting tools and infrastructures should be developed to mitigate the cost of data providers for readying trait data for the 'semantic web' of biodiversity data.
+To mitigate the cost of data standardisation efforts, supporting tools and infrastructures should be developed for readying trait data for the 'semantic web' of biodiversity data.
 
@@
 To date, there is a rich distributed body of independently published trait datasets, with different focus on particular organism groups, ecosystem types or regions.
-These distributed data are heterogeneous in form and description and initiatives to harmonise, compile and analyse these data require significant amounts of funding.
+These distributed data are heterogeneous in form and description, and initiatives to harmonise, compile and analyse these data require significant amounts of funding.
 
-By using a standard vocabulary with globally accessible definitions of terms,  would allow distributed trait data to be more easily re-used and harmonised into aggregated datasets. 
+Using a standard vocabulary with globally accessible definitions of terms,  would allow distributed trait data to be more easily re-used and harmonised into aggregated datasets. 
 The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation and publication of data.
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to <!--Matthias Biber, Kristin Bohn,--> Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary<!-- and the pre-releases of the R-package-->.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to <!--Matthias Biber, Kristin Bohn,--> Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary<!-- and the pre-releases of the R-package-->.
 We are grateful to the organisers and participants of the Open Traits workshop in New Orleans, USA, in August 2018.
 
-We thank the managers of the three Exploratories, Kirsten Reichel-Jung, Katrin Lorenzen, Juliane Vogt, Miriam Teuscher, and all former managers for their work in maintaining the plot and project infrastructure;
-Christiane Fischer and Jule Mangels for giving
-support through the central office, and Markus Fischer, Eduard Linsenmair,
-Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot,
-Ernst-Detlef Schulze, Wolfgang W. Weisser and the late Elisabeth Kalko
-for their role in setting up the Biodiversity Exploratories project. 
+We thank the scientific coordinators, local managers and the data manager of the Biodiversity Exploratories program for their work, and Markus Fischer, Eduard Linsenmair, Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot, Ernst-Detlef Schulze, Wolfgang W.
+Weisser and the late Elisabeth Kalko for their role in setting up the Biodiversity Exploratories program.
 The work has been partly funded by the DFG Priority Program
-1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno.). <!--Field work
-permits were issued by the responsible state environmental offices of
-Baden-Württemberg, Thüringen, and Brandenburg (according to § 72
-BbgNatSchG).-->
-
+1374 "Infrastructure-Biodiversity-Exploratories" (DFG-Refno. Po362/18-3, MA7144/1-1, WE3081/21-1, KO2209/12-2);
 MMG obtained funding from Swiss National Science Foundation (SNF 310030E-173542/1);
@@
 
-The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is  https://terminologies.gfbio.org/terms/ets/pages/, stable DOI representing all versions: 10.5281/zenodo.1041732.
 Any future development of the vocabulary is coordinated via https://github.com/EcologicalTraitData/ETS/.

### 2018-11-01 manuscript/manuscript.Rmd
final edits on manuscript, fine tuning conclusion

@@
 
-By using this standard vocabulary, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable.
+All terms may be applied to describe columns of a data table, or to label content of a database (Fig.
+2; see Supplementary Material B for best practice principles and examples for publishing primary data).
+By applying standard terms, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable.
 It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives, data synthesis and ensure long-term accessibility.
-By adopting the standard vocabulary, trait-data will support the FAIR principles for scientific data [@wilkinson16].
-We describe best practice principles and give examples for the application of the vocabulary in Supplementary Material B.
-
-In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users and curators), and biodiversity informatics researchers (data managers).
-We paid particular attention to existing data standards [particularly @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.
+By adopting the standard vocabulary, trait-data will eventually support the FAIR principles for scientific data [@wilkinson16].
 
@@
 These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
+Our vocabulary thus offers three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", "Occurrence", and "Taxon"; see section on extensions below).
+Additional terms are provided for metadata and for defining trait concepts (see section on metadata below). 
+Aspects not covered by the vocabulary may draw from terms provided by DwC and it's extensions or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).
 
@@
 
-Our vocabulary thus provides core terms to reflect the minimal Entity-Quality structure of trait data. 
-We further offer three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", "Occurrence", and "Taxon"; see section on extensions below). 
-All terms may be applied to describe columns of a data table, or to label content of a database (see Box 1 / Supplementary Material B for best practice principles and examples for publishing primary data).
-Additional terms are provided for metadata and for providing trait concepts (see section on metadata below).
-Some of the terms expand and refine terms of DwC and its extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516).
-Wherever our vocabulary refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
-The vocabulary can be mapped onto other existing ontologies (e.g.
-@madin07) or database structures (e.g.
-@kattge11a).
-Aspects not covered by the terminology may draw from terms provided by DwC and it's extensions or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).
+## Building community consensus
+
+In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users and curators), and biodiversity informatics researchers (data managers).
+The aim was to develop a simple, easy-to-use template for standalone trait data publications or data compilations, to facilitate their integration into databases.
+Earlier proposals for trait-data standards [e.g.
+@kattge11a; @parr16] have been designed for more complex database structures and from a data manager perspective only, which may be the reason why they have so far not been adopted for primary data publications.
+We paid particular attention to these existing data standards [particularly @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our vocabulary with major trait databases and existing concepts for biodiversity data management.
+Nonetheless, we are aware of the diverse use-cases of trait data that might not yet be covered by the current version of the vocabulary.
+The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard, as a key ressource for trait-data standardisation.
+We invite communities of all trait-based research fields to discuss, revise and submit terms and extensions of the vocabulary (coordinated via Github Issues at https://github.com/EcologicalTraitData/ETS/issues).
+The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service (https://terminologies.gfbio.org/terms/ets/pages/).
 
@@
 <!-- community effort and consensus -->
-While it is clear, that a single vocabulary standard is not able to fulfil the needs of all research fields working with traits, we believe that our proposal captures a common core concept of trait data.
+While it is clear, that a single vocabulary standard is not able to fulfil the needs of all research fields working with traits, we believe that our proposal captures the common core concept of trait data in a single resource terminology.
 Consensus building on how traits are to be used and evaluated is currently under way in several fields with their taxonomic focus and project-specific questions [@kissling18; @garnier17; @pey14; @moretti17].
@@
 Therefore, we invite researchers to contribute to future versions of the standard vocabulary and develop own applications and ontologies that interact with it.
-The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service (https://terminologies.gfbio.org/terms/ets/pages/).
 
@@
 To date, there is a rich distributed body of independently published trait datasets, with different focus on particular organism groups, ecosystem types or regions.
-These distributed data are heterogeneous in form and description, and initiatives to harmonise, compile and analyse these data require significant amounts of funding.
+These distributed data are heterogeneous in form and description, hampering endeavours to harmonise, compile and analyse these data.
 
 Using a standard vocabulary with globally accessible definitions of terms,  would allow distributed trait data to be more easily re-used and harmonised into aggregated datasets. 
-The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation and publication of data.
+The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation before the publication of data.
 This requires significant effort, but it returns great scientific benefit by enabling data-heavy synthesis for a general understanding of biodiversity and ecosystem functioning.
@@
 
-Supplementary Material B - contains three worked examples of how to  1.) apply the core ETS vocabulary to species$\times$traits wide-trables, 2.) convert data to long-table format using identifiers 3.) apply ETS extensions and identifiers to include co-variate information on different layers of the dataset.
+Supplementary Material B - contains best-practice guidelines and three worked examples of how to  1.) apply the core ETS vocabulary to species$\times$traits wide-trables, 2.) convert data to long-table format using identifiers 3.) apply ETS extensions and identifiers to include co-variate information on different layers of the dataset.
 

### 2018-11-01 manuscript/appendix_B.Rmd
final edits on appendix B

@@
 - **State terms of use!** To enable re-use, the metadata must contain the name of the author or owner of the data, the suggested citation for the data, as well as the license model or terms of use of the data.
-Ideally, these statements are provided by the data-hoster in a machine-readable way.
+Recommended are copyright wavers and licenses that permit re-use without seeking authorization from the rights holder, such as Creative Commons (CC) licenses, which also can warrant attribution and citation (CC by) [@kissling18].
+Ideally, these statements are provided by the data-hoster in a machine-readable way, by using standardised fields of metadata or hyperlink references.
 
@@
 
-kable(mothsStd[1:35,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
+kable(mothsStd[1:28,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
       booktabs = T, caption = caption) %>%
@@
 
-kable(heteroptera_raw[1:35,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
+kable(heteroptera_raw[1:28,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
       booktabs = T, caption = caption) %>%
@@
 select_cols <- c(1,2,3,4,8,6,15,18,22,16,20,21)
-kable(heteroptera_long[heteroptera_long$traitName %in% c("Body_length", "Body_width", "Body_height", "Thorax_width"),select_cols][30:65,], longtable=TRUE, escape = TRUE, row.names = FALSE,
+kable(heteroptera_long[heteroptera_long$traitName %in% c("Body_length", "Body_width", "Body_height", "Thorax_width"),select_cols][30:58,], longtable=TRUE, escape = TRUE, row.names = FALSE,
       booktabs = T, caption = caption) %>%

### 2018-11-01 manuscript/coverletter.Rmd
add coverletter

---
output:
  rmarkdown::tufte_handout:
    keep_tex: true
    fig_width: 1
  pdf_document: default
  word_document: default
  html_document:
    fig_caption: yes
    fig_width: 1
    highlight: null
    theme: united
---

\vspace{3.1cm}
\setlength{\fboxsep}{3pt}
\setlength{\fboxrule}{0pt}
\fbox{
  \parbox[b]{8cm}{
  \small
Methods in Ecology and Evolution \\
The British Ecological Society\\
Charles Darwin House\\
12, Roger Street\\
London, UK\\
WC1N 2JU\\
  }
}

\vspace{4.2cm}

\noindent Dear editors,   \marginpar{01 November 2018}
\vspace{0.3cm}

We would like to submit our manuscript "Towards an Ecological Trait-data Standard" for consideration as a review article at Methods in Ecology and Evolution. The manuscript discusses the need for, and potential of, a better standardisation of trait data on the data-provider side.

Trait-based approaches are on the rise in many fields, ranging from local-scale functional ecology to global ecological synthesis. These works rely on the compilation of trait-data from heterogeneous sources. At the same time a wealth of often unused data is available in open-access repositories. Presently trait-data harmonization is time-consuming and error-prone due to an insufficient implementation of unified data standards in primary data publications. As it stands, the wealth of data is not easy to integrate in the semantic web of biodiversity data and downstream data synthesis is inhibited. We propose a simple and versatile vocabulary for primary data publications of trait-data to close this gap: the Ecological Trait-data Standard. We also discuss incentives and provide best-practice guidelines to facilitate the standardisation of trait-data before data publication.

We presented a pre-print version of this paper and our standard terminology earlier this year (https://doi.org/10.1101/328302).This received a lot of attention for a pre-print (1000+ reads on biorxiv.org and Researchgate) and initiated interactions with a broad community of researchers (in particular, with the OpenTraits.org community and the TRY plant trait database). Community feedback on this pre-print has been integrated in the current version of the standard (v0.9), and of this manuscript.

We consider Methods in Ecology and Evolution to be a perfect venue to disseminate this work, due to the general and international audience of the journal, many of whom work with trait data. By providing a primer on standardisation and interoperability of trait-data we feel that our paper makes an important contribution to both open science and trait-based ecology and evolution.

While we feel the paper is of importance, and an excellent fit to MEE, its format differs from those which are standard at the journal. The manuscript is structured into two chapters that we consider strongly interrelated: first, a review of the current state of implementation and the rationale for trait-data standardisation, and second, a proposal for a standard terminology to remediate the deficiencies in current standardisation practice for trait data publications. Due to this structure, it comes with a rather high word count of slightly below 10.000 words (including 2.470 for References and 250 for figure and table captions). Although these two chapters could in principle go into separate "Review" and "Practical Tools" papers, we would like to argue for keeping those two chapters in one single paper to have them well aligned, and to strengthen the impact of the work.

As reviewers, we suggest

-	Josh Madin (University of Hawaii), who knows all aspects of trait-data standardisation, including authoring and analysing trait data, as well compilation and management of large databases.,
-	Ramona Walls (University of Arizona), who developed seminal papers on the topic of biodiversity data management and ontologies, with particular focus on trait-data, 
-	Joseph Tobias (Imperial College London), who works on trait evolution and employs large compilations of trait data, and
-	Gavin Thomas (Sheffield University), who works on trait data measurements on birds.

We would like to oppose to inviting reviews from the lead authors of the TRY plant trait database (namely Jens Kattge, Gerhard Bönisch) who opposed to our vocabulary in a comment on biorxiv.org (https://www.biorxiv.org/content/early/2018/05/31/328302#comment-4010197361) highlighting competing interests with their centralised approach to data standardisation. Although we had a friendly exchange and took their concern into account and highlighted the complementarity of the two approaches when revising the manuscript, they might still perceive our proposal as competitive.

We look forward to hearing from you.

\vspace{1.5mm}

\noindent On behalf of all authors

\vspace{1.5mm}

\includegraphics[width=5.2cm]{FSchneider.eps}

\noindent Florian D. Schneider

### 2019-03-21 manuscript/manuscript.Rmd
edits on manuscript, part I

@@
 - name: Caterina Penone \footnotemark[4]
-- name: Nadja K. Simons \footnotemark[7]
+- name: Nadja K. Simons \footnotemark[7] \footnotemark[9]
 csl: mee.csl
@@
     template: pretty.template
+  word_document:
+    fig_caption: yes
+    fig_width: 6
   html_document:
@@
 \footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
+\footnotetext[9]{Ecological Networks, Technische Universität Darmstadt, Schnittspahnstraße 3, 64287 Darmstadt, Germany}
 
@@
 1.
-Trait-based approaches are widespread throughout ecological research, as they offer great potential to achieve a general understandnig of a wide range of ecological and evolutionary mechanisms.
+Trait-based approaches are widespread throughout ecological research as they offer great potential to achieve a general understanding of a wide range of ecological and evolutionary mechanisms.
 Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
@@
 4.
-The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
-By aiding decentralised trait-data standardisation, our data standard may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.
+The use of a standard vocabulary allows trait datasets from heterogeneous sources to be aggregated more easily into compilations and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
+By aiding decentralised trait-data standardisation, our vocabulary may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.
 
@@
 
-functional ecology, species traits, semantic web, ontologies, data standard, ecoinformatics
+functional ecology, species traits, semantic web, ontologies, data standardisation, ecoinformatics
 
@@
 Using traits is also a promising means of bypassing taxonomic impediment, i.e.
-the fact that most species are yet undescribed and little is known of their interactions with other organisms and their environment.
-This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].
+the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment.
+This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime01; @moretti17; @villeger17].
 
@@
 reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements-->
-To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g.
-PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
+To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g., PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
 Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
 Besides initiatives aiming at assembling data, tools to enable the compatibility of data across databases are being developed.
-These include software to access trait data from the internet [e.g.
-@chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of consensus terms [@walls12; @garnier17].
+These include software to access trait data from the internet [e.g., @chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of consensus terms [@walls12; @garnier17].
 
@@
 DOIs (Digital Object Identifiers).
-Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  .
-Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
-Occurrence | The observation context of a single entity of a taxon, i.e.
-data on an individual observed at a particular place and time.
-(http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context.
+Identifier (ID) | A unique label that relates data entries to information within and across *datasets* or external items of information; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*.
+Metadata | Data documentation of the higher-level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
+Occurrence | The observation context of a single individual, i.e. the existance of an organism at a particular place and time; Sometimes used as synonym of 'observation' in data management context.
 Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
@@
 Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
-Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g.
-may define a hierarchy of broader or narrower terms.
+Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g., may define a hierarchy of broader or narrower terms.
 Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
@@
 
-We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17].
+We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Characteristics [TOP; @garnier17].
 We show how the degree of trait-data standardisation in existing datasets is highly variable, and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources.
@@
 
-In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g.
-feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books.
+In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g., feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books.
 With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
-A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, see e.g., @gossner15; see Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
+A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, e.g., @gossner15 and Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
 
-Today, the online publication of such data is greatly facilitated by file hosting services (e.g.
-Figshare, Zenodo, Researchgate, Data Dryad), which warrant long-term accessibility, and citeability via DOIs, and govern data sharing via license statements.
+Today, the online publication of such data is greatly facilitated by file hosting services (e.g., Figshare, Zenodo, Researchgate, Data Dryad), which warrant long-term accessibility, and citeability via DOIs, and govern data sharing via license statements.
 These platforms offer the hosting of publicly accessible *file repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management.
-However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g.
+<!--However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g.
 .xlsx, .docx) or binary (e.g.
 .pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk.
+-->
 Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation.
-As a result, published datasets greatly vary in structure, nomenclature and metadata quality.
+Thus, although open for download, the trait datasets on data repositories might be stored in variable tabular structures and labelled following self-defined terms which makes extraction and further use unnecessarily tedious.
+<!--As a result, published datasets greatly vary in structure, nomenclature and metadata quality.
+-->
 
@@
 For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives.
-Furthermore, the metadata vary in their level of detail, e.g.
-for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
+Furthermore, the metadata vary in their level of detail, e.g., for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
 In terms of structure, trait data usually are reported in a species$\times$traits wide-table format.
-In this intuitive data table, each row represents a species (or taxon) for which multiple traits are reported in columns.
-<!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).
+In this intuitive data table, each row represents a species (or taxon) for which multiple traits are reported in columns.<!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).
 -->
-Similarly, when reporting raw data, researchers place observations of individual organisms (i.e.
-the *occurrences*) in rows with multiple trait measurements applied to the same individual across multiple columns.
-Co-variates on the taxon, the individual specimen (e.g.
-sex or lifestage) or context of observation (e.g.
-time and place of sampling) would be placed in additional columns, and would further expand the two-dimensional data table.
+Similarly, when reporting raw data, researchers place observations on individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
+Co-variates on the taxon, the individual specimen (e.g., sex or lifestage) or context of observation (e.g., time and place of sampling) would be placed in additional columns and would further expand the two-dimensional data table.
 The resolution or scope of these co-variates varies greatly depending on the research question and observation context.
@@
 <!-- community database backend, but static publication --> 
-In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g.
-@kleyer08; @oliveira17, see Supplementary Material A, Table A1].
-While mostly concerned with issues of heterogeneity in units or levels of categorical variables, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other datasets.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary.
--->
-Documentation of metadata and methodology differs in the level of detail, depending on the research focus. 
-Trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiveristy informatics.  
+In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g., @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
+While these initiatives address issues of heterogeneity in units or categorical variables, or aim for high taxonomic or geographic coverage, few of these compilations apply a standardised terminology for taxa or trait definitions.
+Additionally, documentation of metadata and methodology in the original datasets often differs in the level of detail, depending on the research focus, and those information are difficult to maintain in the compilation.
+Particularly, in the process of data aggregation, rich metadata content might be lost or the reference to the original dataset might be obscured, as only aggregated values are reported (e.g.
+means or medians).
+Such trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiversity informatics.
 Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.
@@
 As they deal with much larger amounts of data, initiatives that compile data from natural history museum collections are traditionally more concerned with standardisation. 
-The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g.
-www.markmybird.org).
+The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g., www.markmybird.org).
 For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16].
@@
 Specialised online portals have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
-<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
 For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform.
-The online portal enables selective data download and user permission and rights management.
-As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+The online portal enables selective data download and user permission and right management.<!--As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+-->
 For animal trait data, however, a single unified platform and harmonising scheme is still lacking.
-The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].
+<!--The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].-->
 
@@
 A *thesaurus* provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
+To provide a logic structure for trait terms, Garnier et al.
+(2017) suggest the Entity-Quality model (EQ), where a trait is defined as ‘an entity having a quality’ (for instance for trait ‘femur length’, ‘femur’ is the entity and ‘length’ the quality, Fig.
+1).
 In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms.
@@
 Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)*.
-For example, a measurement of fruit mass could be linked to the definition of the term within the TOP Thesaurus of plant traits [@garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
-*Ontologies* link defined terms by formally describing the relationships between them and providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18].
-Ontologies define terms based on other well-defined terms from published ontologies.
+For example, a measurement of fruit mass could be linked to the definition of the term within the Thesaurus of Plant Characteristics [TOP, @garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
+In addition to defining terms for human interpretation, *ontologies* define terms by their relationship to other defined terms, thereby providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18].
  The Plant Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
 Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12].
-By providing ontologies in a formalized file format, like OWL or RDF, a machine-readable web of definitions is spun across the Internet allowing researchers and search engines to relate independent trait measurements with each other and connect them to the wider *semantic web* of online data [@berners-lee01; @page08; @williams06; @walls12].
+By providing ontologies in a formalized syntax, like Web Ontology Language (OWL), a machine-readable web of definitions is spun across the Internet allowing researchers and search engines to relate independent trait measurements with each other and connect them to the wider *semantic web* of online data [@berners-lee01; @page08; @williams06; @walls12].
 
-Comprehensive trait thesauri have been developed in the TOP Thesaurus of Plant Traits, which is employed in the TRY database [@garnier17], and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
-Ontologies of trait definitions have been developed for plants [e.g.
-the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g.
-the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13].
+Comprehensive trait thesauri have been developed in TOP [which is employed in the TRY database, @garnier17] and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
+Ontologies of trait definitions have been developed for plants [e.g., the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g., the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13].
 The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
@@
 
-To conclude, there is already a suite of globally available thesauri and ontologies for traits available.
+To conclude, there is already a suite of globally available thesauri and ontologies for traits.
 However, definitions in some domains are better covered than others [@kissling18] and the interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
@@
 
-To make sense of trait data in the context of more general databases, a consensus definition of trait data has been provided by  @garnier17: 
+<!--To make sense of trait data in the context of more general databases, a consensus definition of trait data has been provided by  @garnier17: 
 The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
@@
 The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
-a genus).
+a genus).-->
+In terms of structure, a trait dataset contains measurement values or qualitative facts reporting on traits observed from one or multiple organisms or occurrences, i.e.
+observa-tions of an individual at a specific place and time, that are assigned to a scientific taxon.
+Information is thus centered around an observed numerical or categorical value, being further specified in what has been measured (i.e., a trait) and on what instance of a sci-entific taxon it has been observed.
+Additional information would describe the situation of the measurement, like methods of data aggregation, technical tools or observer identi-ty, and context of the occurrence, like environmental factors at the time and place of sampling or growth conditions.
+Depending on the research question, these co-variates are used to explain noise or drivers in trait values.
 
-These relationships between a trait measurement and an individual organism as an entity of a particular taxon have been formalised in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the general structure of biodiversity databases.
+Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+These frameworks specify terms and classes to describe the describe the relationships between any kind of measurement and occurrence of organisms of a particular taxon.
 The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
-The terminologies provided by these standards are quite universal and would even cover most use cases of trait data.
+The terminologies provided by these standards are quite universal and cover most use cases of trait data.
 A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
-For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. 
+<!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
 Both DwC and ABCD are ratified standards of the Biodiversity Information Standards TDWG (Taxonomic Databases Working Group; www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.
@@
 
-Specifically designed for plant traits, @kattge11a proposed a generic database structure that covers most potential use cases of trait-based ecology.
-This data structure is built around a central data table that contains observations of individual plants (entities) linked to several measurements of traits (qualities) via identifiers.
+Based on these terms, the Extensible Observation Ontology [OBOE, @madin07; @schildhauer16] formalizes observations andd measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
+By applying this scheme for plant traits, @kattge11a propose a generic database structure that covers most potential use cases of trait-based ecology.
+This data structure is built around a central data table that contains observations of individual plants linked to several measurements of traits via identifiers.
 The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
-The structure of @kattge11a can be mapped to the  Extensible Observation Ontology [OBOE, @madin07; @schildhauer16], which formalizes observations andd measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
-Besides OBOE, @kissling18 discuss further ontologies that formalize the structure of observation data and attest that for the use with trait data the various ontologies are difficult to integrate.
+@kissling18 discuss different ontologies (including OBOE) that formalize the structure of observation data and attest that for the use cases of trait data these ontologies are still difficult to integrate.
 
-The  Encyclopedia of Life (EOL) project has proposed TraitBank [@parr16] as a standard structure for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach of an integrated structure for trait data.
-It is the suggested format for upload of data to.
-The framework employs established terms provided by the DwC, relates trait definitions to ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives].
+While the abovementioned standards provide terms and concept definitions, and the logic relationships of those, they do not prescribe explicit structure for trait-data. 
+In this regard, the Encyclopedia of Life (EOL) project has proposed TraitBank [@parr16] as a standard structure for major physiological and life-history traits of all kingdoms of life.
+It is to date the most general approach of an integrated structure for trait data.
+The framework employs established terms provided by the DwC and the DwC MeasurementOrFact extension [@parr16]. 
 Additional layers of information cover bibliographic references, multimedia archives and ecological interactions. 
@@
 
-## Closing gaps in the trait-data lifecycle
+## Closing gaps to improve trait-data re-use
 
-In sum, we attest a gap between the trait-data structures developed for data curators and managers and the data input produced by data providers.
+In sum, we attest to a gap between the trait-data structures developed for data curators and managers and the data input produced by data providers.
 Hardly any of the aforementioned standalone or aggregated trait datasets for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
@@
 The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management.
-In addition, complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.
+In addition, complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects, especially since funders often do not require these efforts to be included into proposals.
 
-By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14;e.g.
-Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g.
-KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
+By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14;e.g.,  Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g., KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-The goal of such initiatives is to facilitate data re-use by providing standardisation of data, for instance by mapping unambiguous terminologies and ontologies for biodiversity data, and clarify conditions of data re-use.
+The goal of such initiatives is to facilitate data re-use by providing standardisation of data, for instance by mapping to unambiguous terminologies and ontologies for biodiversity data, and clarifying conditions of data re-use.
 
@@
 The package 'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits wide-table.
-<!--The package 'NatDB' [@pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. -->
+<!--The package 'MAD' [Making a Database, @pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. -->
 FENNEC [@ankenbrand18] is an online tool or self-hosted service capable of extracting trait information from multiple sources for a target species community.
-Such decentralised tools
-
 <!-- To remediate this, incentives for data standardisation before upload to file hosting services should be created.
@@
 Second, the reference to ontologies and thesauri must be incentivised and facilitated for individual data providers by the development of tools for  matching concepts from the available ontologies to their data.
-Third, frameworks for providing trait data in a unambiguous and machine-readable structure must be simplified to match the limited resources of small and intermediate research projects.
-This can be achieved by extending documentation or providing tools for the application of existing ontology frameworks and database structures, or by defining easy-to-use exchange formats that enable the interoperability of data at minimal effort.
+Third, frameworks for providing trait data in an unambiguous and machine-readable structure must be simplified to match the limited resources of small and intermediate research projects.
+This can be achieved by extending documentation or providing tools for the application of existing ontology frameworks and database structures, or by defining easy-to-use standard vocabularies that enable the interoperability of data at minimal effort.
 <!--
@@
 
-Despite centralised data management efforts, however, no unified and widely-applicable definition of a data type for traits has emerged across the multiple sub-disciplines of trait-based research. 
-In the following chapter, we propose a unified vocabulary for trait-data that can serve as a minimal consensus exchange format.
+However, no unified and widely-adopted terminology for for primary trait-data publications has emerged across the multiple sub-disciplines of trait-based research. 
+In the following chapter, we propose a unified vocabulary for trait-data that can serve as a minimal consensus for describing and labelling trait data.
 The simplicity of this standard terminology will lower the thresholds and offer high pay-off in the visibility and reuse of published data.
-By establishing this as a best-practice standard in trait-based research, trait-data will eventually fulfil the FAIR guiding principles for scientific data [@wilkinson16].
+By establishing this as a "best-practice" in trait-based research, trait-data will eventually fulfil the FAIR guiding principles for scientific data [@wilkinson16].
 
@@
 
-# An ecological trait-data standard vocabulary
+# Introducing the Ecological Trait-data Standard Vocabulary
 
 As a response to the challenges outlined above, we propose a versatile standard vocabulary for trait-based ecological research. 
-The Ecological Trait-data Standard vocabulary is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different degrees of measurement details.
-Rather than prescribing a data structure, the vocabulary is intended as a more inclusive terminology that can be used in three major use cases:
+The Ecological Trait-data Standard vocabulary (ETS) is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different degrees of measurement details.
+Rather than prescribing a data structure or exchange format, the vocabulary is intended as a more inclusive terminology that can be used in three major use cases:
 
@@
 3.
-by data managers: as a data exchange format of between online resources, web services and software tools, e.g.
-when providing database queries via a web service or defining input and output formats of software packages.
+by data managers: in developing data exchange formats between online resources, web services and software tools, e.g., when providing database queries via a web service or defining input and output formats of software packages.
 
@@
 It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives, data synthesis and ensure long-term accessibility.
-By adopting the standard vocabulary, trait-data will eventually support the FAIR principles for scientific data [@wilkinson16].
 

### 2019-03-22 manuscript/manuscript.Rmd
edits on manuscript, part II

@@
 ---
-title: "Towards an Ecological Trait-data Standard"
+title: "Towards an Ecological Trait-data Standard Vocabulary"
 author:
@@
 <!-- 1. Traits are important in ecology, evolution and conservation -->
-
 Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
@@
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
-Using traits is also a promising means of bypassing taxonomic impediment, i.e.
-the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment.
+Using traits is also a promising means of bypassing taxonomic impediment, i.e., the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment.
 This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime01; @moretti17; @villeger17].
@@
 Furthermore, we show how the current deficit in the standardisation of primary data hampers the implementation of interoperability and reuse of trait data. 
-Based on these considerations, we propose a versatile vocabulary for describing trait datasets that builds upon, and is compatible with, existing terminology standards for biodiversity data. 
-Since this standard vocabulary relies on the adoption by a broad research community, we discuss incentives for its use and lay out mechanisms for future consensus-building and community development towards an accessible ecological trait data standard.
+Based on these considerations, we propose a versatile vocabulary for describing ecological trait datasets, which builds upon, and is compatible with, existing terminology standards for biodiversity data, in particular the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+Since a standard vocabulary relies on the adoption by a broad research community, we discuss incentives for its use and lay out mechanisms for future consensus-building and community development towards an accessible and easy-to-use ecological trait-data standard vocabulary.
 <!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
@@
 
-Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the Darwin Core Standard for biodiversity data [DwC; @wieczorek12].
+Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the *Darwin Core Standard* for biodiversity data [DwC; @wieczorek12].
 These frameworks specify terms and classes to describe the describe the relationships between any kind of measurement and occurrence of organisms of a particular taxon.
@@
 As a response to the challenges outlined above, we propose a versatile standard vocabulary for trait-based ecological research. 
-The Ecological Trait-data Standard vocabulary (ETS) is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different degrees of measurement details.
+The Ecological Trait-data Standard Vocabulary (ETS) is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different degrees of measurement details.
 Rather than prescribing a data structure or exchange format, the vocabulary is intended as a more inclusive terminology that can be used in three major use cases:
@@
 by data providers: for publication of standardised primary data on open-access data repositories, or for labelling project-specific data for local use and exchange with collaborators, e.g.
-in two-dimensional data tables or project databases
+in two-dimensional data tables or project databases,
 2.
@@
 
-All terms may be applied to describe columns of a data table, or to label content of a database (Fig.
-2; see Supplementary Material B for best practice principles and examples for publishing primary data).
-By applying standard terms, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable.
+All terms may be applied to describe columns of a data table (Fig. 2; see Supplementary Material B for best practice principles and examples for publishing primary data). 
+By applying these standard terms, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable.
 It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives, data synthesis and ensure long-term accessibility.
+
+The definitions of terms are hosted on the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/], providing permanent and redirectable individual URIs and URLs for each term.
+The service can be accessed programmatically (i.e., via the API; https://terminologies.gfbio.org/api/terminologies/).
 
@@
 
-A trait dataset typically contains data according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e.
-quality) observerd on an individual or population (i.e.
-entity), of a biological taxon.
-The trait values may be quantitative measurements or qualitative facts. 
+A trait dataset typically contains multiple data entries, where each entry describes a trait value observed on an instance of a scientific taxon. 
+The item on which the value has been observed can be very variable as well, ranging from an occurrence of an individual at a specific place and time in its natural environment or a preserved specimen in a collection (Fig.
+1*a*), a group of individuals of a specific taxon (Fig.
+1*b*), or an entire population of a species (Fig.
+1*c-d*).
+The reported trait values may be quantitative measurements or qualitative facts. 
 Quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
@@
 1*c*).
-This encompasses a wide range of numeric data types, including continuous, binary, integer, intervals or ratios, as well as ordinal values.
-Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
-Beyond these core observations, further information might be available that further specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation and date information).
+This encompasses a wide range of numeric data types, including continuous, binary, integer, intervals or ratios, as well as categorical (ordinal or nominal) values.
+Qualitative facts are assignments of categorical information, often on entire taxa, e.g., of a behavioural or life-history trait (Fig. 1*d*).
+
+Beyond these core observations, further information might be available that specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation and date information).
 These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-Our vocabulary thus offers three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", "Occurrence", and "Taxon"; see section on extensions below).
-Additional terms are provided for metadata and for defining trait concepts (see section on metadata below). 
-Aspects not covered by the vocabulary may draw from terms provided by DwC and it's extensions or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).
+Our vocabulary thus offers three extensions to contain ancillary information on different layers of the core data, in analogy to DwC extensions ("Taxon", "Measurement or Fact", and "Occurrence"; see section on extensions below).
+Additional terms are provided for dealing with metadata, particularly in trait-data compilations, and for defining trait concepts (see section on metadata below). 
+Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular Dublin Core and DwC and it's extensions), or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).
 
@@
 In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users and curators), and biodiversity informatics researchers (data managers).
-The aim was to develop a simple, easy-to-use template for standalone trait data publications or data compilations, to facilitate their integration into databases.
-Earlier proposals for trait-data standards [e.g.
-@kattge11a; @parr16] have been designed for more complex database structures and from a data manager perspective only, which may be the reason why they have so far not been adopted for primary data publications.
-We paid particular attention to these existing data standards [particularly @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our vocabulary with major trait databases and existing concepts for biodiversity data management.
+The aim was to develop a simple, easy-to-use template for standalone trait data publications or data compilations, to facilitate their re-use for synthesis and integration into larger database structures.
+Earlier proposals for trait-data standards [e.g.,  @kattge11a; @parr16] have been designed for more complex database structures or from a data manager perspective, which may be the reason why they have so far not been adopted for primary data publications.
+We paid particular attention to these existing data standards [e.g., @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our vocabulary with major trait databases and existing concepts for biodiversity data management.
+
 Nonetheless, we are aware of the diverse use-cases of trait data that might not yet be covered by the current version of the vocabulary.
-The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard, as a key ressource for trait-data standardisation.
+The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation.
+For future development of the vocabulary, we will engage with a broader communty of trait researchers, in particular via the Open Traits Network (http:/opentraits.org), and ensure compatibilty with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
 We invite communities of all trait-based research fields to discuss, revise and submit terms and extensions of the vocabulary (coordinated via Github Issues at https://github.com/EcologicalTraitData/ETS/issues).
-The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service (https://terminologies.gfbio.org/terms/ets/pages/).
+The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service.
 
@@
 
-For trait data according to the Entity-Quality model, the following content is required at minimum (Fig. 2 *a&b*):
+To qualify trait data, the following content is required at minimum (Fig. 2 *a&b*):
 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
-2. a descriptive trait name (`traitName`) that links to a standardised definition (i.e. a concept); 
-3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`) that links to an accepted taxon concept.
+2. a descriptive trait name (`traitName`) that links the observation to a standardised definition (i.e. a concept); 
+3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`) that links the observation to an accepted taxon concept.
 
 <!--### Relate to global taxon and trait definitions -->
-The trait name and taxon names would ideally use unambiguous terms assigning both to clearly-defined concepts. 
-However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
-In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
-This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively, Fig.
+The `traitName` and `scientificName` would use unambiguous terms assigning both to clearly-defined concepts.
+Eventually, disambiguation can be warranted by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`).
+For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to Flora Phenotype Ontology would be  'http://purl.obolibrary.org/obo/FLOPO_0005265'.
+However, to ensure compatibility with project-specific databases or analytical code, it might be necessary to keep user-specific identifiers for those terms, for which we are suggesting the use of `verbatimScientificName`  and `verbatimTraitName` (Fig.
 2 *c*).
-For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to Flora Phenotype Ontology would be  'http://purl.obolibrary.org/obo/FLOPO_0005265'.
+<!-- double record -->
+By allowing user-side entries along with consensus terms we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards [as also practiced in TRY,  @kattge11a; @kattge11].
+The redundancy of labelling allows for continuity for data providers while also enabling quality checks and comparability for data curators.
 
@@
 Similarly, standardisation of units can be achieved by relying on SI convention or by relating units to unambigous concepts via URIs provided by ontologies [@steinberg16; @keil18]. 
-For categorial, ordinal or binary traits, the factor levels should conform to expected levels as defined in the trait concept, or be unambiguously defined in the metadata of the dataset (see section metadata below).
-The vocabulary complements the user-defined values in dataset-specific units and factor levels with standardised entries (`traitValueStd` and `traitUnitStd`). 
-<!-- double record -->
-By suggesting for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11].
-This redundancy of data allows for continuity for data providers while also enabling quality checks and comparability for data curators.
+For categorical, ordinal or binary traits, the categories should conform to expected levels as defined in the trait concept or be unambiguously defined in the metadata of the dataset (see section metadata below).
+The vocabulary offers terms for keeping the user-defined values in dataset-specific units and factor levels along with standardised entries (`verbatimTraitValue` and `verbatimTraitUnit`, Fig. 2 *c*).
 
@@
 Beyond measurement units or higher taxon information, further information might complement the core data which are related to the individual specimen, the reported fact, measurement or sampling event.
-We propose three extensions of the vocabulary that should be used to describe this information (Fig. 2*d*):
+We propose three extensions of the vocabulary that should be used to describe this information (Fig. 2*d*), in line with the existing DwC extension structure:
 
@@
 These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content.
-A unique identifier links to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
-Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g.
-a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
+A unique identifier links to these other datasheets, encoding single measurements or reported facts (`measurementID`), individual occurrences of a species (`occurrenceID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g., a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
 
-Identifiers can also be used to provide a structure within a two-dimensional data table, e.g.
-to link rows of measurements on the same individual (using `occurrenceID`) or observations at the same location (using `locationID`).
-The  values of multivariate measurements can be labelled using `measurementID`.
+Identifiers can also be used to provide a structure within a two-dimensional data table, e.g., to link rows of measurements on the same individual (using `occurrenceID`) or observations at the same location (using `locationID`).
+The  values of multivariate measurements can be linked using `measurementID`.
+The terms of the extensions draw from terms of the DwC extensions of particular relevance for trait data.
+See the documentation of the ETS for further detail on the use of extensions.
 
@@
 
-Higher-level information about structure, provenance of data, authorship and data ownership and terms of use should be considered when sharing and working with trait datasets [@michener06; @kissling18].
-Our metadata vocabulary ( https://terminologies.gfbio.org/terms/ets/pages/#metadata-vocabulary) provides terms to label these information. 
-In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g.
-in a separate file, possibly applying the ecological metadata language, EML, @knb11].
-Whenever data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain reference to the original data provider and conditions of re-use of these data.
-This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.
+Higher-level information about structure, provenance of data, authorship and data ownership, as well as terms of use should be considered when sharing and working with trait datasets [@michener06; @kissling18].
+In the case of primary measurement data, this information usually applies to the entire trait dataset, and would be stored along with the published data as metadata [e.g., in a separate XML file, ideally applying Ecological Metadata Language, EML, @knb11].
+Whenever data from different sources are compiled into a single dataset, these information would become part of the data table, as each data entry would have to maintain reference to the original data provider and conditions of re-use of these data.
+This can be achieved by appending the metadata terms as columns to the core dataset, or by linking to a secondary data table via an unambiguous `datasetID` (e.g., a URI pointing to the source DOI) and a descriptive `datasetName` (e.g., a  descriptive name for the source).
+Our metadata vocabulary ( https://terminologies.gfbio.org/terms/ets/pages/#metadata-vocabulary) provides terms for a minimal set of information that should be provided along with trait data. 
+The suggested terms originate from Dublin Core Metadata Initiative (DCMI), and are widely compatible with terms provided by the DataCite Metadata Schema [@datacite19].
+The terms can be extended and complemented by using terms from these resources.
 
-Wherever possible, the field `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
-If such an external reference does not exist, trait data should always be accompanied by a dataset-specific list of trait concepts. 
-Such a controlled vocabulary would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format of measured values or reported facts (e.g.
-units or legit factor levels).
+As mentioned above, wherever possible, the field `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
+If no suitable reference exists, trait data should always be accompanied by a dataset-specific listing of trait concepts. 
+Such a controlled vocabulary would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format of measured values or reported facts (e.g., units or legit factor levels).
 Ideally, this definition refers to or refines terms from published trait ontologies.
-By providing a minimal vocabulary for trait lists (see  https://terminologies.gfbio.org/terms/ets/pages//#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
+By providing a minimal vocabulary for trait lists (see  https://terminologies.gfbio.org/terms/ets/pages/#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
 This vocabulary might also prove useful for the future publication of trait ontologies. 
@@
 
-In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g.
-"Schneider, F.D., Jochum, M., Le Provost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2018 Ecological Traitdata Standard v0.9, DOI:  10.5281/zenodo.1441132, URL: https://terminologies.gfbio.org/terms/ets/pages/".
-Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their URIs. 
-The definitions of terms are hosted on the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e.
-via the API; https://terminologies.gfbio.org/api/terminologies/).
+In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset , e.g., by entering "Schneider, F.D., Jochum, M., Le Provost, G., Penone, C., Ostrowski, A.
+and Simons, N.K., 2019, Ecological Traitdata Standard Vocabulary v0.10, DOI:  10.5281/zenodo.1441132, URL: https://terminologies.gfbio.org/terms/ets/pages/" in the metadata field `conformsTo`.
+Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their individual URIs.
 
@@
 
+<!-- continue here -->
+
 # Discussion
 
-To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, as well as for the exchange and handling of trait data in the context of a 'semantic web'.
+To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, that will be of use for the exchange and handling of trait data in the context of the semantic web.
 
 <!-- community effort and consensus -->
-While it is clear, that a single vocabulary standard is not able to fulfil the needs of all research fields working with traits, we believe that our proposal captures the common core concept of trait data in a single resource terminology.
-Consensus building on how traits are to be used and evaluated is currently under way in several fields with their taxonomic focus and project-specific questions [@kissling18; @garnier17; @pey14; @moretti17].
-While community discussions on trait definitions and measurement practices are leading to a better quality of data, naturally, the heterogeneity in trait-data quality and detail across fields will remain.
-With our proposal of an Ecological Trait-data Standard vocabulary, we provide a starting point for the development of a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research.
-Therefore, we invite researchers to contribute to future versions of the standard vocabulary and develop own applications and ontologies that interact with it.
+Consensus building on how traits are to be used and evaluated is currently under way in several fields of ecological research with their taxonomic focus and project-specific questions [@kissling18; @garnier17; @pey14; @moretti17].
+Such community discussions on trait definitions and measurement practices are leading to a better quality of data, naturally.
+However, they still require a stronger linkage into the global biodiversity data initiatives.
+With our proposal of an Ecological Trait-data Standard Vocabulary (ETS), we aim to capture the common core concept of trait data in a single resource terminology, and provide a starting point for the development of a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research.
+To enable the ETS to capture the different approaches in trait-based research aross fields, we invite researchers to contribute to future versions of the standard vocabulary and develop their own applications and ontologies that interact with it.
+Development will also aim at linking the initiative to the joint efforts for biodiversity data terminologies, in particular within Biodiversity Information Standards (TDWG).
 
 <!-- ETS investment and pay-off for data providers, incentives -->
-For individual researchers and small research projects, the investment of time and resources in trait-data standardisation before publication seems currently not to be yielding a direct benefit.
-However, data released according to consensus standards, especially if published under open-access licenses, will be more easily re-used in compilations and synthesis studies. 
-A well-defined vocabulary for metadata will also ensure that authorship and terms of use are appropriately documented along the data life-cycle.
+Data released according to consensus standards, especially if published under open-access licenses, are more easily re-used in compilations and synthesis studies.
+By providing the ETS, an easy-to-use vocabulary for trait-based research, the investment of time and resources in trait-data standardisation before publication will be mitigated for individual researchers and small research projects.
+A well-defined minimial vocabulary for metadata will also ensure that authorship and terms of use are appropriately documented along the data life-cycle. 
 However, for these incentives to take effect, data publications and data citations must become viewed as a valid scientific contribution to the community, and recognized in the professional evaluation of individual researchers [@costello09; @roche15].
-To mitigate the cost of data standardisation efforts, supporting tools and infrastructures should be developed for readying trait data for the 'semantic web' of biodiversity data.
 
@@
 At the community level, shifting the task of standardisation from the data-user side to the data-owner side yields great gain in accuracy and reduces the risk of misinterpretation.
-For instance, measurement results depend very much on the precise methodology used and often systematic biases could be corrected for when providing an undambiguous definition.
-On the other hand, plausibility checks and evaluation of statistical methods, e.g.
-for aggregating trait values to the species level, can only be done in comparison across a wide array of datasets.
+For instance, measurement results depend very much on the precise methodology used and often systematic biases could be corrected for when providing an unambiguous definition.
+On the other hand, plausibility checks and evaluation of statistical methods, e.g., for aggregating trait values to the species level, can only be done in comparison across a wide array of datasets.
 Currently, these "big data" volumes are only available in centralised databases.
 However, to establish a best-practice of data aggregation, an exploration and evaluation of different methods for quality assessment and quality control should be subject to a community discussion. 
-This is only possible with large quantities of distributed data being available in a harmonised way. Our ecological trait-data standard vocabulary facilitates such a community-driven comparison.
+This is only possible with large quantities of distributed data being available in a harmonised way. The ETS facilitates such a community-driven comparison.
 
 <!-- ETS to build toolchain and semantic web -->
-Our standard vocabulary represents an important building block for a unified mode of data exchange between web services and software packages and thus facilitates the development of a software toolchain for the trait-data lifecycle.
+The ETS represents an important building block for a unified mode to ease data exchange between web services and software packages and thus facilitates the development of a software toolchain for the trait-data lifecycle.
 Having well-defined terms is also a key precondition for developing exchange formats between large database initiatives and biodiversity data archives. 
@@
 
-To date, there is a rich distributed body of independently published trait datasets, with different focus on particular organism groups, ecosystem types or regions.
+To date, there is a rich distributed body of independently published trait datasets, each with a specific focus on particular organism groups, ecosystem types or regions.
 These distributed data are heterogeneous in form and description, hampering endeavours to harmonise, compile and analyse these data.
 
-Using a standard vocabulary with globally accessible definitions of terms,  would allow distributed trait data to be more easily re-used and harmonised into aggregated datasets. 
+Using a standard vocabulary with globally accessible definitions of terms would allow distributed trait data to be more easily re-used and harmonised into aggregated datasets. 
 The biggest challenge in future standardisation of trait data may be consensus building for standard terms, the establishment of incentives and the development of tools for a user-side standardisation before the publication of data.
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to <!--Matthias Biber, Kristin Bohn,--> Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary<!-- and the pre-releases of the R-package-->.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary.
 We are grateful to the organisers and participants of the Open Traits workshop in New Orleans, USA, in August 2018.
@@
 
-Supplementary Material A - static excerpt of a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use.
-We invite data owners and users to add further trait datasets to this spreadsheet.
+Supplementary Material A - an exemplary list of published trait datasets and data compilations, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use.
 

### 2019-03-25 manuscript/manuscript.Rmd
update manuscript, finalize revision

@@
 fontsize: 11pt
-geometry: margin=1.4in
+geometry: margin=1.35in
 documentclass: scrartcl
@@
 \footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
-\footnotetext[9]{Ecological Networks, Technische Universität Darmstadt, Schnittspahnstraße 3, 64287 Darmstadt, Germany}
+\footnotetext[9]{Ecological Networks, Department of Biology, Technische Universität Darmstadt, Darmstadt, Germany}
 
@@
 Using traits is also a promising means of bypassing taxonomic impediment, i.e., the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment.
-This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime01; @moretti17; @villeger17].
+This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime06; @moretti17; @villeger17].
 
@@
 Metadata | Data documentation of the higher-level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
+In the context of trait-data, such additional information can move to the body of the primary data table when data are compiled from different sources.
 Occurrence | The observation context of a single individual, i.e. the existance of an organism at a particular place and time; Sometimes used as synonym of 'observation' in data management context.
@@
 
-We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Characteristics [TOP; @garnier17].
+We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant characteristics [TOP; @garnier17].
 We show how the degree of trait-data standardisation in existing datasets is highly variable, and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources.
@@
 -->
+
 Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation.
@@
 
-For trait data, there are common issues arising from the variability of data structures. 
-For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives.
-Furthermore, the metadata vary in their level of detail, e.g., for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].
+For trait data, there are common issues arising from the variability of data structures and metadata quality. 
 In terms of structure, trait data usually are reported in a species$\times$traits wide-table format.
@@
 The resolution or scope of these co-variates varies greatly depending on the research question and observation context.
+The column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives. 
 Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. 
-A task that is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).
+Furthermore, the metadata provided along with the primary data vary in their level of detail, e.g., for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
+While in some primary data publications information like geolocation or sampling date and time might be dataset-level information, thus qualifying as metadata, in other datasets they might be collected on a level of individual observations (see section on data compilations below).
+More importantly, clear statements on ownership and authorship, terms of use, or internationalisation (e.g., separators and delimiters), are often still neglected in primary trait-data publications. 
+The task of harmonizing trait data is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).
 
@@
 
-## Data aggregation initiatives
+## Data compilation initiatives
 
@@
 means or medians).
-Such trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiversity informatics.
+Such trait-data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiversity informatics.
 Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.
@@
 To provide a logic structure for trait terms, Garnier et al.
-(2017) suggest the Entity-Quality model (EQ), where a trait is defined as ‘an entity having a quality’ (for instance for trait ‘femur length’, ‘femur’ is the entity and ‘length’ the quality, Fig.
-1).
+(2017) suggest the Entity-Quality model (EQ), where a trait is defined as ‘an entity having a quality’ (for instance for trait ‘femur length’, ‘femur’ is the entity and ‘length’ the quality).
 In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms.
@@
 Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)*.
-For example, a measurement of fruit mass could be linked to the definition of the term within the Thesaurus of Plant Characteristics [TOP, @garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
+For example, a measurement of fruit mass could be linked to the definition of the term within the Thesaurus of Plant characteristics [TOP, @garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
 In addition to defining terms for human interpretation, *ontologies* define terms by their relationship to other defined terms, thereby providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18].
@@
 Comprehensive trait thesauri have been developed in TOP [which is employed in the TRY database, @garnier17] and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].
-Ontologies of trait definitions have been developed for plants [e.g., the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g., the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13].
+Ontologies of trait definitions have been developed for plants [e.g., the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], and for specific animal taxa [e.g., the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13].
+The UBERON ontology is an integrated cross-species anatomy ontology for all animals, which combines concepts from different existing ontologies, with wide application in biomedical or physiological research [@mungall12].
 The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
@@
 a genus).-->
-In terms of structure, a trait dataset contains measurement values or qualitative facts reporting on traits observed from one or multiple organisms or occurrences, i.e.
-observa-tions of an individual at a specific place and time, that are assigned to a scientific taxon.
+<!--In terms of structure, a trait dataset contains measurement values or qualitative facts reporting on traits observed from one or multiple organisms or occurrences, i.e., observations of an individual at a specific place and time, that are assigned to a scientific taxon (Fig.
+1).
 Information is thus centered around an observed numerical or categorical value, being further specified in what has been measured (i.e., a trait) and on what instance of a sci-entific taxon it has been observed.
-Additional information would describe the situation of the measurement, like methods of data aggregation, technical tools or observer identi-ty, and context of the occurrence, like environmental factors at the time and place of sampling or growth conditions.
+Additional information would describe the situation of the measurement, like methods of data aggregation, technical tools or observer identity, and context of the occurrence, like environmental factors at the time and place of sampling or growth conditions.
 Depending on the research question, these co-variates are used to explain noise or drivers in trait values.
+-->
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
+\caption{Types of ecological trait data assume different entities or reported qualities: a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
+reproductive rates, life spans); b) aggregated trait values are reported as means taken on multiple measures of organisms of a taxon; c) quantitative traits may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
+a sex) as the subject of description; d) qualitative traits are categorical, ordinal or binary descriptors of the entire species or higher taxonomic level (also called 'facts').}
+
+\end{figure}
+
+A trait dataset typically contains multiple data entries, where each entry describes a trait value observed on an instance of a scientific taxon. 
+The item on which the value has been observed can be very variable, ranging from an *occurrence* of an individual at a specific place and time in its natural environment or a preserved specimen in a collection (Fig.
+1*a*), a group of individuals of a specific taxon (Fig.
+1*b*), or an entire population of a species (Fig.
+1*c-d*).
+The reported trait values may be quantitative measurements or qualitative facts. 
+Quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
+1*a*), by aggregating replicated measurements on multiple entities (Fig.
+1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g.
+databases, Fig.
+1*c*).
+This encompasses a wide range of numeric data types, including continuous, binary, integer, intervals or ratios, as well as categorical (ordinal or nominal) values.
+Qualitative facts are assignments of categorical information, often on entire taxa, e.g., of a behavioural or life-history trait (Fig. 1*d*).
+
+Beyond these core observations, further information might be available that specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation as well as date and time of sampling).
+As such information may be useful for future analysis of the causal reasons of trait variation or to explain noise in measurement data, they should always be published along with the core data.
+In most cases, these information apply to the entire dataset, and thus would be included in the metadata acompanying a data publication [potentially applying Ecological Metadata Language, EML, @knb11 as a formal structure].
+In the case of trait-data and depending on the research scope, such information may also have been collected on a level of measurement, occurrence or taxon level.
+Geolocation or date and time would then not be provided as metadata, but as covariate information in additional columns of the primary dataset.
+When compiling datasets, it is a key task of data curators to deal with dataset-level information and incorporate them into the compiled data table to maintain the information for downstream analysis.
 
 Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the *Darwin Core Standard* for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe the describe the relationships between any kind of measurement and occurrence of organisms of a particular taxon.
-The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details.
-The terminologies provided by these standards are quite universal and cover most use cases of trait data.
-A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
+These frameworks specify terms and classes to describe any kind of measurement and occurrence of organisms.
+The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, information of time and date of observation, and methodological details.
+A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12]. The terminologies provided by these standards are quite universal and cover most use cases of trait data. 
 <!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
@@
 This data structure is built around a central data table that contains observations of individual plants linked to several measurements of traits via identifiers.
-The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
+The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment (which may have been extracted from metadata).
 @kissling18 discuss different ontologies (including OBOE) that formalize the structure of observation data and attest that for the use cases of trait data these ontologies are still difficult to integrate.
@@
 While the abovementioned standards provide terms and concept definitions, and the logic relationships of those, they do not prescribe explicit structure for trait-data. 
-In this regard, the Encyclopedia of Life (EOL) project has proposed TraitBank [@parr16] as a standard structure for major physiological and life-history traits of all kingdoms of life.
+In this regard, the Encyclopedia of Life (EOL) has proposed TraitBank [@parr16] as a standard structure for uploading data on physiological and life-history traits of all kingdoms of life.
 It is to date the most general approach of an integrated structure for trait data.
@@
 Additional layers of information cover bibliographic references, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured *Darwin Core Archive* [DwC-A, @robertson09a].
+TraitBank invites data submissions to the EOL database in a structured *Darwin Core Archive* [DwC-A, @robertson09a], including simple text files (csv) and a metadata description using EML (the correct specification of which can be validated before upload on https://tools.gbif.org/dwca-validator/).
 
@@
 One likely reason for this lack of standardisation is the complexity of the task: the proposed data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice.
-In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
+In the eyes of the data-provider, in most cases, any co-variates can be appended as extra columns to the dataset.
 The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management.
@@
 
-A trait dataset typically contains multiple data entries, where each entry describes a trait value observed on an instance of a scientific taxon. 
-The item on which the value has been observed can be very variable as well, ranging from an occurrence of an individual at a specific place and time in its natural environment or a preserved specimen in a collection (Fig.
-1*a*), a group of individuals of a specific taxon (Fig.
-1*b*), or an entire population of a species (Fig.
-1*c-d*).
-The reported trait values may be quantitative measurements or qualitative facts. 
-Quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig.
-1*a*), by aggregating replicated measurements on multiple entities (Fig.
-1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g.
-databases, Fig.
-1*c*).
-This encompasses a wide range of numeric data types, including continuous, binary, integer, intervals or ratios, as well as categorical (ordinal or nominal) values.
-Qualitative facts are assignments of categorical information, often on entire taxa, e.g., of a behavioural or life-history trait (Fig. 1*d*).
+Our vocabulary offers three extensions to contain additional information on the context of the observation along with the core data in analogy to DwC extensions ("Taxon", "Measurement or Fact", and "Occurrence"; see section on extensions below).
+Further terms are provided for dealing with typical dataset-level information on authorship and rights of re-use of the data, as well as for defining trait concepts (see section on metadata below). 
+Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular Dublin Core and DwC and it's extensions), or be added as user-defined columns (which should then be clearly specified in the metadata-information accompanying the dataset).
 
-Beyond these core observations, further information might be available that specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation and date information).
-These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.
-Our vocabulary thus offers three extensions to contain ancillary information on different layers of the core data, in analogy to DwC extensions ("Taxon", "Measurement or Fact", and "Occurrence"; see section on extensions below).
-Additional terms are provided for dealing with metadata, particularly in trait-data compilations, and for defining trait concepts (see section on metadata below). 
-Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular Dublin Core and DwC and it's extensions), or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Types of ecological trait data assume different entities or reported qualities: a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated trait values are reported as means taken on multiple measures of members of a taxon; c) quantitative traits may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the entity of description; d) qualitative traits are categorical, ordinal or binary descriptors of the entire species or higher taxonomic level (also called 'facts').}
-
-\end{figure}
 <!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”.
@@
 Nonetheless, we are aware of the diverse use-cases of trait data that might not yet be covered by the current version of the vocabulary.
-The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation.
+The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research.
 For future development of the vocabulary, we will engage with a broader communty of trait researchers, in particular via the Open Traits Network (http:/opentraits.org), and ensure compatibilty with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
+This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
 We invite communities of all trait-based research fields to discuss, revise and submit terms and extensions of the vocabulary (coordinated via Github Issues at https://github.com/EcologicalTraitData/ETS/issues).
@@
 
-To qualify trait data, the following content is required at minimum (Fig. 2 *a&b*):
+To qualify as trait data complying with the ETS, the following content is required at minimum (Fig. 2 *b*):
 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
@@
 For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to Flora Phenotype Ontology would be  'http://purl.obolibrary.org/obo/FLOPO_0005265'.
-However, to ensure compatibility with project-specific databases or analytical code, it might be necessary to keep user-specific identifiers for those terms, for which we are suggesting the use of `verbatimScientificName`  and `verbatimTraitName` (Fig.
+However, to ensure compatibility with project-specific databases or analytical code, it might be in the interest of the data author to keep user-specific identifiers for those terms, for which we are suggesting the use of `verbatimScientificName`  and `verbatimTraitName` (Fig.
 2 *c*).
@@
 Similarly, standardisation of units can be achieved by relying on SI convention or by relating units to unambigous concepts via URIs provided by ontologies [@steinberg16; @keil18]. 
-For categorical, ordinal or binary traits, the categories should conform to expected levels as defined in the trait concept or be unambiguously defined in the metadata of the dataset (see section metadata below).
+For categorical, ordinal or binary traits, the categories should conform to expected levels as defined in the trait concept or be unambiguously defined in the metadata of the dataset. 
 The vocabulary offers terms for keeping the user-defined values in dataset-specific units and factor levels along with standardised entries (`verbatimTraitValue` and `verbatimTraitUnit`, Fig. 2 *c*).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
+\caption{Formats used for trait datasets.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers.
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e.
+occurrence) or the origin or confidence on the reported measurement or fact.}
+
+\end{figure}
 
@@
 
-## Specification of Metadata
+## Specification of metadata
 
-Higher-level information about structure, provenance of data, authorship and data ownership, as well as terms of use should be considered when sharing and working with trait datasets [@michener06; @kissling18].
+Dataset-level information about structure, provenance of data, authorship and data ownership, as well as terms of use should be considered when sharing and working with trait datasets [@michener06; @kissling18].
 In the case of primary measurement data, this information usually applies to the entire trait dataset, and would be stored along with the published data as metadata [e.g., in a separate XML file, ideally applying Ecological Metadata Language, EML, @knb11].
@@
 Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their individual URIs.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
-\caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and c) relating it to a standard trait definition and accepted taxon name using unambiguous identifiers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e.
-occurrence) or the origin or confidence on the reported measurement or fact.}
-
-\end{figure}
-
-<!-- continue here -->
 
@@
 <!-- ETS to build toolchain and semantic web -->
+Without clearly defined terms and concepts, handling of large amounts of trait-data by computational assistance systems for scientific analysis ('e-Science') will be massively hampered [@wilkinson16].
 The ETS represents an important building block for a unified mode to ease data exchange between web services and software packages and thus facilitates the development of a software toolchain for the trait-data lifecycle.
@@
 Even further downstream, readying the primary data for the semantic web via references to ontologies and data standards will ease the application of automatised big-data mining and machine-learning techniques.
-Without clearly defined terms and concepts, handling of large amounts of trait-data by computational assistance systems for scientific analysis ('e-Science') will be massively hampered [@wilkinson16].
 

### 2019-03-25 manuscript/manuscript.Rmd
spell checking

@@
     template: pretty.template
-  word_document:
-    fig_caption: yes
-    fig_width: 6
   html_document:
@@
 1.
-Traits are important in ecology, evolution and conservation (i.e.
-It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
+Traits are important in ecology, evolution and conservation (i.e.,  It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).
 
@@
 When we do this we can answer new questions and find generalities.
-So there is a growing interest in assembling trait data from different sources and taxa (e.g.
-give the number of papers using TRY).
+So there is a growing interest in assembling trait data from different sources and taxa (e.g., give the number of papers using TRY).
 
@@
 <!-- GLP: 
- Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g.
-TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals).
-Say also something on the development of general protocols for measuring traits (e.g.
-Moretti et al.
+ Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g., TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals).
+Say also something on the development of general protocols for measuring traits (e.g., Moretti et al.
 2016; Freschet et al.
@@
 I wonder if you should speak about ‘Functional traits’here or just Traits ?
-As you are not only considered Functional traits in this paper (e.g.
-life history traits) ?
+As you are not only considered Functional traits in this paper (e.g., life history traits) ?
 Maybe remove the definition no ?
 -->
-<!-- PM: Thuis is quite complex- e.g.
-‘taxonomic impediment’etc.
+<!-- PM: Thuis is quite complex- e.g., ‘taxonomic impediment’etc.
 I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples.
@@
 <!-- 1. Traits are important in ecology, evolution and conservation -->
-Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
+Functional traits are phenotypic (i.e., morphological, physiological, behavioural) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
 Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
@@
 <!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04].
-Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e.
-the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
+Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e., the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms.
 Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [ @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13].
@@
 <!--
-Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g.
-@reich97].
+Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g., @reich97].
 <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].-->
@@
 However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
-<!-- i.e.
-differences in taxonomic nomenclature and resolution (e.g.
-reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements-->
-To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g., PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
+<!-- i.e., differences in taxonomic nomenclature and resolution (e.g., reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements-->
+To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centred around particular research questions [e.g., PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17].
 Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
@@
 ----- | ---------
-Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g.
-client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
+Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g., client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
 Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/).
 Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
-Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted.
-Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g.
-using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
+Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a particular context, while no unlisted entries are accepted.
+Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g., using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
 Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/).
 Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries.
-Database | A structured colletion of data, usually organised as multiple data tables linked via *identifiers* into relational databases; usually constructed using a specific database management system, i.e.
-a software to provide a (offline or online) user interface.
-File repository | A storage or archiving of datasets on file-hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org; online repositories make data available for public access, provide *metadata*, state conditions of re-use, and (not always) facilitate citations via persistent identifiers, e.g.
-DOIs (Digital Object Identifiers).
+Database | A structured collection of data, usually organised as multiple data tables linked via *identifiers* into relational databases; usually constructed using a specific database management system, i.e., a software to provide a (offline or online) user interface.
+File repository | A storage or archiving of datasets on file-hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org; online repositories make data available for public access, provide *metadata*, state conditions of re-use, and (not always) facilitate citations via persistent identifiers, e.g., DOIs (Digital Object Identifiers).
 Identifier (ID) | A unique label that relates data entries to information within and across *datasets* or external items of information; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*.
@@
 In the context of trait-data, such additional information can move to the body of the primary data table when data are compiled from different sources.
-Occurrence | The observation context of a single individual, i.e. the existance of an organism at a particular place and time; Sometimes used as synonym of 'observation' in data management context.
+Occurrence | The observation context of a single individual, i.e., the existence of an organism at a particular place and time; Sometimes used as synonym of 'observation' in data management context.
 Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
-Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e.
-specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
+Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e., specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
 Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
@@
 Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
-Web Service | A service to access *databases* directly and computationally, i.e.
-by calls from a software tool; returns content of the database according to the request in different formats and data standards.
+Web Service | A service to access *databases* directly and computationally, i.e., by calls from a software tool; returns content of the database according to the request in different formats and data standards.
 
@@
 
-<!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time.-->
-<!--Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g.
-trait data), that is recognised by a large proportion of members of the application context.-->
+<!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e., they are not altered or filtered across time.-->
+<!--Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g., trait data), that is recognised by a large proportion of members of the application context.-->
 <!--File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org.-->
@@
 <!--Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally.-->
-<!--Online portal | A website designed as a platform for the exchange of information, e.g.
-trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access.-->
+<!--Online portal | A website designed as a platform for the exchange of information, e.g., trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access.-->
 <!--Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs).-->
@@
 These platforms offer the hosting of publicly accessible *file repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management.
-<!--However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g.
-.xlsx, .docx) or binary (e.g.
-.pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk.
+<!--However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g., .xlsx, .docx) or binary (e.g., .pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk.
 -->
 
-Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation.
+Most importantly, these platforms enable public hosting of data with very low quality-thresholds regarding *metadata* documentation and data standardisation.
 Thus, although open for download, the trait datasets on data repositories might be stored in variable tabular structures and labelled following self-defined terms which makes extraction and further use unnecessarily tedious.
@@
 Similarly, when reporting raw data, researchers place observations on individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
-Co-variates on the taxon, the individual specimen (e.g., sex or lifestage) or context of observation (e.g., time and place of sampling) would be placed in additional columns and would further expand the two-dimensional data table.
+Co-variates on the taxon, the individual specimen (e.g., sex or life-stage) or context of observation (e.g., time and place of sampling) would be placed in additional columns and would further expand the two-dimensional data table.
 The resolution or scope of these co-variates varies greatly depending on the research question and observation context.
-The column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives. 
+The column descriptions and terminology applied to taxa and traits are mostly project-specific and rarely chosen to allow translation into larger database initiatives. 
 Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. 
@@
 -->
-<!-- FDS: this might be a task for community  building and decentralised registries, as discussed on the OpenTraits platform, e.g.
-FENNEC or MAD package do that, or a registry for earning badges of data quality would achieve that as a by-product.
+<!-- FDS: this might be a task for community  building and decentralised registries, as discussed on the OpenTraits platform, e.g., FENNEC or MAD package do that, or a registry for earning badges of data quality would achieve that as a by-product.
 I think we should not aim for solving this registry problem here.
@@
 While these initiatives address issues of heterogeneity in units or categorical variables, or aim for high taxonomic or geographic coverage, few of these compilations apply a standardised terminology for taxa or trait definitions.
-Additionally, documentation of metadata and methodology in the original datasets often differs in the level of detail, depending on the research focus, and those information are difficult to maintain in the compilation.
-Particularly, in the process of data aggregation, rich metadata content might be lost or the reference to the original dataset might be obscured, as only aggregated values are reported (e.g.
-means or medians).
+Additionally, documentation of metadata and methodology in the original datasets often differs in the level of detail, depending on the research focus.
+Particularly, in the process of data aggregation, rich metadata content might be lost or the reference to the original dataset might be obscured, as only aggregated values are reported (e.g., means or medians).
 Such trait-data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiversity informatics.
@@
 Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13].
-Databases are also used to interpret OTUs in terms of their functional role [e.g.
-the KEGG orthology, @kanehisa12].
+Databases are also used to interpret OTUs in terms of their functional role [e.g., the KEGG orthology, @kanehisa12].
 -->
@@
 The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
-<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g.
-units or legit factor levels) of measured values or reported facts.
+<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g., units or legit factor levels) of measured values or reported facts.
 It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits.
@@
 The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
-More specifically, a trait dataset contains information on quantitative measurement* or qualitative facts (i.e.
-trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e.
-traits) observed on a biological entity (i.e.
-an individual specimen, or its parts) that can be assigned to a biological taxon (i.e.
-a species or higher-level taxon).
-The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g.
-a genus).-->
+More specifically, a trait dataset contains information on quantitative measurement* or qualitative facts (i.e., trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e., traits) observed on a biological entity (i.e., an individual specimen, or its parts) that can be assigned to a biological taxon (i.e., a species or higher-level taxon).
+The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g., a genus).-->
 <!--In terms of structure, a trait dataset contains measurement values or qualitative facts reporting on traits observed from one or multiple organisms or occurrences, i.e., observations of an individual at a specific place and time, that are assigned to a scientific taxon (Fig.
@@
 \includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
-\caption{Types of ecological trait data assume different entities or reported qualities: a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g.
-reproductive rates, life spans); b) aggregated trait values are reported as means taken on multiple measures of organisms of a taxon; c) quantitative traits may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g.
-a sex) as the subject of description; d) qualitative traits are categorical, ordinal or binary descriptors of the entire species or higher taxonomic level (also called 'facts').}
+\caption{Types of ecological trait data assume different entities or reported qualities: a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g., reproductive rates, life spans); b) aggregated trait values are reported as means taken on multiple measures of organisms of a taxon; c) quantitative traits may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g., a sex) as the subject of description; d) qualitative traits are categorical, ordinal or binary descriptors of the entire species or higher taxonomic level (also called 'facts').}
 
@@
 1*a*), by aggregating replicated measurements on multiple entities (Fig.
-1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g.
-databases, Fig.
+1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g., databases, Fig.
 1*c*).
@@
 Beyond these core observations, further information might be available that specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation as well as date and time of sampling).
-As such information may be useful for future analysis of the causal reasons of trait variation or to explain noise in measurement data, they should always be published along with the core data.
-In most cases, these information apply to the entire dataset, and thus would be included in the metadata acompanying a data publication [potentially applying Ecological Metadata Language, EML, @knb11 as a formal structure].
-In the case of trait-data and depending on the research scope, such information may also have been collected on a level of measurement, occurrence or taxon level.
-Geolocation or date and time would then not be provided as metadata, but as covariate information in additional columns of the primary dataset.
-When compiling datasets, it is a key task of data curators to deal with dataset-level information and incorporate them into the compiled data table to maintain the information for downstream analysis.
+As such data may be useful for future analysis of the causal reasons of trait variation or to explain noise in measurement data, it should always be published along with the core data.
+In most cases, information on place and time apply to the entire dataset, and thus would be included in the metadata accompanying a data publication [potentially applying Ecological Metadata Language, EML, @knb11 as a formal structure].
+In the case of trait-data and depending on the research scope, the information may also have been collected on a level of measurement, occurrence or taxon level.
+Geolocation or date and time would then not be provided as metadata, but as covariate data in additional columns of the primary dataset.
+When compiling datasets, it is a key task of data curators to deal with dataset-level information and maintain it for downstream analysis by incorporating it into the compiled data table.
 
@@
 
-Based on these terms, the Extensible Observation Ontology [OBOE, @madin07; @schildhauer16] formalizes observations andd measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
+Based on these terms, the Extensible Observation Ontology [OBOE, @madin07; @schildhauer16] formalizes observations and measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
 By applying this scheme for plant traits, @kattge11a propose a generic database structure that covers most potential use cases of trait-based ecology.
@@
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
-The goal of such initiatives is to facilitate data re-use by providing standardisation of data, for instance by mapping to unambiguous terminologies and ontologies for biodiversity data, and clarifying conditions of data re-use.
+The goal of such initiatives is to facilitate data re-use by providing standardisation of data, for instance by mapping to unambiguous terminologies and ontologies for biodiversity data and clarifying conditions of data re-use.
 
 Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains to facilitate the use and analysis of trait-data. 
-For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data directly from their source, including Birdlife, EOL TraitBank or BetyDB.
+For instance, the R-package 'traits' [@chamberlain17] contains functions to extract trait data directly from their source, including Birdlife, EOL TraitBank or BetyDB.
 The package 'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits wide-table.
@@
 
-However, no unified and widely-adopted terminology for for primary trait-data publications has emerged across the multiple sub-disciplines of trait-based research. 
+However, no unified and widely-adopted terminology for primary trait-data publications has emerged across the multiple sub-disciplines of trait-based research. 
 In the following chapter, we propose a unified vocabulary for trait-data that can serve as a minimal consensus for describing and labelling trait data.
@@
 Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted.
-E.g.
-spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on.
+e.g., spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on.
 Your wexample hat already different variables (=traits) with different units as columns.
@@
 <!-- AO: You say it already 2 sentences before --><!--
-This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g.
-'999').
-It may store additional information (e.g.
-on variation of means or literature source) in secondary columns or accompanying datasheets.
+This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g., '999').
+It may store additional information (e.g., on variation of means or literature source) in secondary columns or accompanying datasheets.
 
 Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a].
-Also at this level of information, missing entries, e.g.
-about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14; @taugourdeau14] and using Bayesian approaches [@schrodt15].
+Also at this level of information, missing entries, e.g., about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14; @taugourdeau14] and using Bayesian approaches [@schrodt15].
 <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].
@@
 In these occurrence wide-tables (Fig.
-2*b*), each row is centered around a single physical occurrence of a species, e.g.
-phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns.
+2*b*), each row is centered around a single physical occurrence of a species, e.g., phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns.
 Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
@@
 Both wide-table formats report the different trait values stored in multiple columns.
-Each trait column represents one value that applies to the entity in question (i.e.
-to a species  or an occurrence).
+Each trait column represents one value that applies to the entity in question (i.e., to a species  or an occurrence).
 As the column names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.
@@
 But not necessarily computational more effective at all.
-E.g.
-I start to mix up numeric values with descriptive values.
+e.g., I start to mix up numeric values with descriptive values.
 I start to mix up different units.
-Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g.
-quality stuff.
+Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g., quality stuff.
 Ergo less effective from that perspective.
@@
 We can simply argue that a well established routine should now also executed on trait level!
-Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.
+Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g., traits to species.
 
@@
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.
-Long-table datasets purport multiple advantages for data manipulation (e.g.
-filtering, sub-setting and aggregating data), visualization (e.g.
-plot measured values by factor variable or taxon) and statistical modelling (e.g.
-ANOVA for testing difference of trait value by sex) [@wickham14].
+Long-table datasets purport multiple advantages for data manipulation (e.g., filtering, sub-setting and aggregating data), visualization (e.g., plot measured values by factor variable or taxon) and statistical modelling (e.g., ANOVA for testing difference of trait value by sex) [@wickham14].
 As long-table fromats draw from a defined set of columns,  merging of datasets is much easier.
@@
 Of course, you have less problems on the schema or structure level.
-But, on the other hand, the differences (e.g.
-different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows).
+But, on the other hand, the differences (e.g., different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows).
 Thus, at least with respect to integration, you just move the problem from one level to the other.
 To ease integration, one would need to agree on the entries in the features column
-AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
+AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g., for taxon, trait_names, units, more?
 We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
@@
 \caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) wide-tables report measurements on the specimen-level, i.e.
-one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to a trait definition and taxon via unambiguous identifiers.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species $\times$ traits matrix; b) wide-tables report measurements on the specimen-level, i.e., one physical occurrence of a taxon; c) a measurement long-table is a well defined and tidy data format, reporting one single measurement per row and relating it to a trait definition and taxon via unambiguous identifiers.
 Additional identifiers relate each row to other layers of information on d) the specimen (occurrence) or e) the origin or confidence on the reported measurement or fact.}
@@
 
-<!--When traits are assessed on a higher level of organisation, e.g.
-microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
+<!--When traits are assessed on a higher level of organisation, e.g., microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits.
 Here, location or measurement information are the primary focus of the investigation and must be contained in the .
@@
 1.
-by data providers: for publication of standardised primary data on open-access data repositories, or for labelling project-specific data for local use and exchange with collaborators, e.g.
-in two-dimensional data tables or project databases,
+by data providers: for publication of standardised primary data on open-access data repositories, or for labelling project-specific data for local use and exchange with collaborators, e.g., in two-dimensional data tables or project databases,
 2.
-by data users and curators: as a consensus vocabulary when compiling data from distributed sources into aggregate datasets, e.g.
-to map standardised columns and refer to taxa and trait definitions in a uniform way, and
+by data users and curators: as a consensus vocabulary when compiling data from distributed sources into aggregate datasets, e.g., to map standardised columns and refer to taxa and trait definitions in a uniform way, and
 3.
@@
 Further terms are provided for dealing with typical dataset-level information on authorship and rights of re-use of the data, as well as for defining trait concepts (see section on metadata below). 
-Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular Dublin Core and DwC and it's extensions), or be added as user-defined columns (which should then be clearly specified in the metadata-information accompanying the dataset).
+Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular Dublin Core and DwC and its extensions), or be added as user-defined columns (which should then be clearly specified in the metadata-information accompanying the dataset).
 
@@
 The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research.
-For future development of the vocabulary, we will engage with a broader communty of trait researchers, in particular via the Open Traits Network (http:/opentraits.org), and ensure compatibilty with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
+For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http:/opentraits.org), and ensure compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
 This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
@@
 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
-2. a descriptive trait name (`traitName`) that links the observation to a standardised definition (i.e. a concept); 
+2. a descriptive trait name (`traitName`) that links the observation to a standardised definition (i.e., a concept); 
 3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`) that links the observation to an accepted taxon concept.
@@
 <!-- units and factor levels-->
-Similarly, standardisation of units can be achieved by relying on SI convention or by relating units to unambigous concepts via URIs provided by ontologies [@steinberg16; @keil18]. 
+Similarly, standardisation of units can be achieved by relying on SI convention or by relating units to unambiguous concepts via URIs provided by ontologies [@steinberg16; @keil18]. 
 For categorical, ordinal or binary traits, the categories should conform to expected levels as defined in the trait concept or be unambiguously defined in the metadata of the dataset. 
@@
 a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e.
-occurrence) or the origin or confidence on the reported measurement or fact.}
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e., occurrence) or the origin or confidence on the reported measurement or fact.}
 
@@
 - The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature reference for the reported value, the method of measurement or statistical method of aggregation.
-It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g.
-variation in measurement method) or aggregated values (e.g.
-statistical method), as well as the source of reported facts (e.g.
-literature source or expert reference).
+It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g., variation in measurement method) or aggregated values (e.g., statistical method), as well as the source of reported facts (e.g., literature source or expert reference).
 <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)-->
@@
 In the case of primary measurement data, this information usually applies to the entire trait dataset, and would be stored along with the published data as metadata [e.g., in a separate XML file, ideally applying Ecological Metadata Language, EML, @knb11].
-Whenever data from different sources are compiled into a single dataset, these information would become part of the data table, as each data entry would have to maintain reference to the original data provider and conditions of re-use of these data.
+Whenever data from different sources are compiled into a single dataset, metadata information would become part of the resulting data table, as each data entry would have to maintain reference to the original data provider and conditions of re-use of these data.
 This can be achieved by appending the metadata terms as columns to the core dataset, or by linking to a secondary data table via an unambiguous `datasetID` (e.g., a URI pointing to the source DOI) and a descriptive `datasetName` (e.g., a  descriptive name for the source).
@@
 However, they still require a stronger linkage into the global biodiversity data initiatives.
-With our proposal of an Ecological Trait-data Standard Vocabulary (ETS), we aim to capture the common core concept of trait data in a single resource terminology, and provide a starting point for the development of a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research.
-To enable the ETS to capture the different approaches in trait-based research aross fields, we invite researchers to contribute to future versions of the standard vocabulary and develop their own applications and ontologies that interact with it.
+With our proposal of an Ecological Trait-data Standard Vocabulary (ETS), we aim to capture the common core concept of trait data in a single resource terminology and provide a starting point for the development of a joint language and terminology around traits as a cross-sectoral topic of ecological and evolutionary research.
+To enable the ETS to capture the different approaches in trait-based research across fields, we invite researchers to contribute to future versions of the standard vocabulary and develop their own applications and ontologies that interact with it.
 Development will also aim at linking the initiative to the joint efforts for biodiversity data terminologies, in particular within Biodiversity Information Standards (TDWG).
@@
 By providing the ETS, an easy-to-use vocabulary for trait-based research, the investment of time and resources in trait-data standardisation before publication will be mitigated for individual researchers and small research projects.
-A well-defined minimial vocabulary for metadata will also ensure that authorship and terms of use are appropriately documented along the data life-cycle. 
-However, for these incentives to take effect, data publications and data citations must become viewed as a valid scientific contribution to the community, and recognized in the professional evaluation of individual researchers [@costello09; @roche15].
+A well-defined minimal vocabulary for metadata will also ensure that authorship and terms of use are appropriately documented along the data life-cycle. 
+However, for these incentives to take effect, data publications and data citations must become viewed as a valid scientific contribution to the community and recognized in the professional evaluation of individual researchers [@costello09; @roche15].
 
@@
 Having well-defined terms is also a key precondition for developing exchange formats between large database initiatives and biodiversity data archives. 
-Even further downstream, readying the primary data for the semantic web via references to ontologies and data standards will ease the application of automatised big-data mining and machine-learning techniques.
+Even further downstream, readying the primary data for the semantic web via references to ontologies and data standards will ease the application of automatized big-data mining and machine-learning techniques.
 

### 2019-03-27 manuscript/appendix_A.Rmd
update Appendix A for resubmission

@@
 - name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
-- name: David Fichtmüller \footnotemark[5]
-- name: Martin M. Gossner \footnotemark[8]
-- name: Anton Güntsch \footnotemark[5]
-- name: Malte Jochum \footnotemark[4]
+- name: David Fichtmüller \footnotemark[3]
+- name: Martin M. Gossner \footnotemark[4]
+- name: Anton Güntsch \footnotemark[3]
+- name: Malte Jochum \footnotemark[5] \hspace{1mm} \footnotemark[8]
 - name: Birgitta König-Ries\footnotemark[6]
@@
 - name: Andreas Ostrowski \footnotemark[6]
-- name: Caterina Penone \footnotemark[4]
-- name: Nadja K. Simons \footnotemark[7]
+- name: Caterina Penone \footnotemark[5]
+- name: Nadja K. Simons \footnotemark[7] \hspace{1mm} \footnotemark[9]
 fontsize: 12pt
@@
 
-\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
+\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com,  \newline
+c/o Birgitta König-Ries, Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Ernst-Abbe-Platz 2, 07743 Jena, Germany
+}
 \footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
-\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
-\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
-\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[3]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[4]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
+\footnotetext[5]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[8]{German Centre for Integrative Biodiversity Research (iDiv) Halle‐Jena‐Leipzig, Leipzig, Germany \& Leipzig University, Institute of Biology, Leipzig, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
 \footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
-
+\footnotetext[9]{Ecological Networks, Department of Biology, Technische Universität Darmstadt, Darmstadt, Germany}
 \newpage
@@
   column_spec(2, width = "5em") %>%
-  column_spec(3, width = "16em") %>%
+  column_spec(3, width = "18em") %>%
   column_spec(4, width = "6em") %>%

### 2019-03-27 manuscript/manuscript.Rmd
fix references in main document, fix affiliations, update figures

@@
 ---
-title: "Towards an Ecological Trait-data Standard Vocabulary"
+title: "Towards an Ecological Trait-data Standard"
 author:
 - name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
-- name: David Fichtmüller \footnotemark[5]
-- name: Martin M. Gossner \footnotemark[8]
-- name: Anton Güntsch \footnotemark[5]
-- name: Malte Jochum \footnotemark[4]
+- name: David Fichtmüller \footnotemark[3]
+- name: Martin M. Gossner \footnotemark[4]
+- name: Anton Güntsch \footnotemark[3]
+- name: Malte Jochum \footnotemark[5] \hspace{1mm} \footnotemark[8]
 - name: Birgitta König-Ries\footnotemark[6]
@@
 - name: Andreas Ostrowski \footnotemark[6]
-- name: Caterina Penone \footnotemark[4]
-- name: Nadja K. Simons \footnotemark[7] \footnotemark[9]
+- name: Caterina Penone \footnotemark[5]
+- name: Nadja K. Simons \footnotemark[7] \hspace{1mm} \footnotemark[9]
 csl: mee.csl
@@
 
-\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com, c/o Birgitta König-Ries
-Friedrich-Schiller-Universität Jena, Institut f. Informatik, Ernst-Abbe-Platz 2, 07743 Jena
+\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com,  \newline
+c/o Birgitta König-Ries, Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Ernst-Abbe-Platz 2, 07743 Jena, Germany
 }
 \footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
-\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
-\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
-\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[3]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[4]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
+\footnotetext[5]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[8]{German Centre for Integrative Biodiversity Research (iDiv) Halle‐Jena‐Leipzig, Leipzig, Germany \& Leipzig University, Institute of Biology, Leipzig, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
@@
 ----- | ---------
-Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g., client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
-Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/).
-Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
+Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference<!-- (https://www.w3.org/TR/skos-reference/)-->.
 Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a particular context, while no unlisted entries are accepted.
-Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g., using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
-Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/).
+Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity; maintained by the Biodiversity Information Standards TDWG (http://rs.tdwg.org/dwc/).
 Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries.
@@
 Occurrence | The observation context of a single individual, i.e., the existence of an organism at a particular place and time; Sometimes used as synonym of 'observation' in data management context.
-Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
+Ontology | A semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
 Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e., specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
+Term | A word that names or labels a particular *concept* as part of the specialised vocabulary of a field. 
 Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
@@
 Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
-Web Service | A service to access *databases* directly and computationally, i.e., by calls from a software tool; returns content of the database according to the request in different formats and data standards.
 
@@
 \doublespacing
-
+<!--Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g., client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.-->
+<!--Darwin Core Archive (DwC-A) | A textfile archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g., using Ecological Metadata Language, EML) [@robertson09].-->
 <!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e., they are not altered or filtered across time.-->
@@
 <!--Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs).-->
+<!--Web Service | A service to access *databases* directly and computationally, i.e., by calls from a software tool; returns content of the database according to the request in different formats and data standards.-->
 <!--
@@
 
-In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g., feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books.
+In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g., feeding biology, dispersal, metabolic rate and body size) have been assessed for decades and have been published in regular journal articles or books.
 With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits.
@@
 The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g., www.markmybird.org).
-For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16].
-The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).
+For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections; the resulting data are published as versioned data tables which are updated as new data sources become available  [http://vertnet.org, @guralnick16].
 
@@
 <!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
-Both DwC and ABCD are ratified standards of the Biodiversity Information Standards TDWG (Taxonomic Databases Working Group; www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.
+Both DwC and ABCD are ratified standards of the Biodiversity Information Standards (TDWG, www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.
 
@@
 Additional layers of information cover bibliographic references, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured *Darwin Core Archive* [DwC-A, @robertson09a], including simple text files (csv) and a metadata description using EML (the correct specification of which can be validated before upload on https://tools.gbif.org/dwca-validator/).
+TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], which is a set of simple text files (csv) and a metadata description using EML (the correct specification of which can be validated before upload on https://tools.gbif.org/dwca-validator/).
 
@@
 
-By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14;e.g.,  Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g., KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
+By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14;e.g.,  Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g., KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, http://diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
@@
 Our vocabulary offers three extensions to contain additional information on the context of the observation along with the core data in analogy to DwC extensions ("Taxon", "Measurement or Fact", and "Occurrence"; see section on extensions below).
-Further terms are provided for dealing with typical dataset-level information on authorship and rights of re-use of the data, as well as for defining trait concepts (see section on metadata below). 
-Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular Dublin Core and DwC and its extensions), or be added as user-defined columns (which should then be clearly specified in the metadata-information accompanying the dataset).
+Further terms are provided for dealing with typical dataset-level information on authorship and rights of re-use of the data (based on terms of Dublin Core Metadata Initiative, DCMI), as well as for defining trait concepts (see section on metadata below).
+Aspects not covered by the vocabulary may draw from terms provided by  other existing terminologies (in particular DCMI and DwC and its extensions), or be added as user-defined columns (which should then be clearly specified in the metadata-information accompanying the dataset).
 
@@
 The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research.
-For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http:/opentraits.org), and ensure compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
+For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and ensure compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
 This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
@@
 This can be achieved by appending the metadata terms as columns to the core dataset, or by linking to a secondary data table via an unambiguous `datasetID` (e.g., a URI pointing to the source DOI) and a descriptive `datasetName` (e.g., a  descriptive name for the source).
-Our metadata vocabulary ( https://terminologies.gfbio.org/terms/ets/pages/#metadata-vocabulary) provides terms for a minimal set of information that should be provided along with trait data. 
+Our metadata vocabulary<!--( https://terminologies.gfbio.org/terms/ets/pages/#metadata-vocabulary)--> provides terms for a minimal set of information that should be provided along with trait data. 
 The suggested terms originate from Dublin Core Metadata Initiative (DCMI), and are widely compatible with terms provided by the DataCite Metadata Schema [@datacite19].
@@
 Ideally, this definition refers to or refines terms from published trait ontologies.
-By providing a minimal vocabulary for trait lists (see  https://terminologies.gfbio.org/terms/ets/pages/#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
+By providing a minimal vocabulary for trait lists<!--(see  https://terminologies.gfbio.org/terms/ets/pages/#terms-for-trait-definitions)-->, we hope to facilitate the unambiguous definition of traits for trait datasets.
 This vocabulary might also prove useful for the future publication of trait ontologies. 
@@
 
-We thank the scientific coordinators, local managers and the data manager of the Biodiversity Exploratories program for their work, and Markus Fischer, Eduard Linsenmair, Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot, Ernst-Detlef Schulze, Wolfgang W.
+We thank the past and present scientific coordinators, local managers and the data manager of the Biodiversity Exploratories program for their work, and Markus Fischer, Eduard Linsenmair, Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot, Ernst-Detlef Schulze, Wolfgang W.
 Weisser and the late Elisabeth Kalko for their role in setting up the Biodiversity Exploratories program.
@@
 
-The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is  https://terminologies.gfbio.org/terms/ets/pages/, stable DOI representing all versions: 10.5281/zenodo.1041732.
+The online reference for the Ecological Trait-data Standard Vocabulary described in this paper is  https://terminologies.gfbio.org/terms/ets/pages/, with a stable DOI representing all versions: 10.5281/zenodo.1041732.
 Any future development of the vocabulary is coordinated via https://github.com/EcologicalTraitData/ETS/.

### 2019-03-27 manuscript/appendix_B.Rmd
update Appendix B, update affiliations, fix Example 2

@@
 - name: Andreas Ostrowski \footnotemark[6]
-- name: Caterina Penone \footnotemark[4]
-- name: Nadja K. Simons \footnotemark[7]
+- name: Caterina Penone \footnotemark[5]
+- name: Nadja K. Simons \footnotemark[7] \hspace{1mm} \footnotemark[9]
 fontsize: 11pt
@@
 
-\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
+\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com,  \newline
+c/o Birgitta König-Ries, Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Ernst-Abbe-Platz 2, 07743 Jena, Germany
+}
 \footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
-\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
-\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
-\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[3]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
+\footnotetext[4]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
+\footnotetext[5]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
+\footnotetext[8]{German Centre for Integrative Biodiversity Research (iDiv) Halle‐Jena‐Leipzig, Leipzig, Germany \& Leipzig University, Institute of Biology, Leipzig, Germany}
 \footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
 \footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
+\footnotetext[9]{Ecological Networks, Department of Biology, Technische Universität Darmstadt, Darmstadt, Germany}
 
@@
 
-- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma- or tab-separated text files  ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
+- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats but rather in software-independent file formats (like .xlsx or .odt), or comma- or tab-separated text files ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
 UTF-8 or ASCII).
@@
 This is true for the core terms, i.e.
-the trait value (the quality) and observation entity, but also for any co-variates reported along with it.
+the trait value and the item of observation and its taxon assignment, but also for any co-variates reported along with it.
 
@@
 A more effective method that allows for high flexibility and easier data merging is the storage of traitdata in long-table formats [@wickham14; @parr16]. 
-This data format corresponds to the Entity-Quality Model: each row contains a single measurement or fact of a specific trait, referenced to a single specimen  (i.e., one occurrence)  which is assigned to a specific taxon  (see Table B2).
+This data format corresponds has a row-based structure: each row contains a single measurement or fact of a specific trait, referenced to a single specimen  (or one occurrence of an individual organism at a particular space and time) which is assigned to a specific taxon  (see Table B2).
 Identifiers can provide structure and link measurements that are correlated by assigning a unique ID for each single individual. 
@@
 ```{r,  echo = FALSE, warning = FALSE, message=FALSE}
+library(XML)
+library(RCurl)
+library(getPass)
 
+read.service <- function(datasetid, 
+                         user = NULL, pswd = NULL, 
+                         dec=".", 
+                         na.strings="NA", 
+                         fill=FALSE, sep="\t", 
+                         quote=if(identical(sep, "\n")) "" else "'\"",
+                         fileEncoding = "UTF-8")
+{
+
+  if(is.null(user)) user <- readline("user name: ") 
+  if(is.null(pswd)) pswd <- getPass::getPass("password: ", noblank = FALSE)
+
+  opts = curlOptions(encoding="CE_UTF8" ,ssl.verifypeer = FALSE, httpheader=c(Accept = "text/plain"))
+  params = c("datasetId"=datasetid, "username"=user, "password"=pswd)
+  data <- postForm(uri="https://www.bexis.uni-jena.de/WebServices/DataService.asmx/DownloadData", .params=params, .opts=opts, style="post")  
+  x <- xmlTreeParse(data, asText = T, trim=FALSE)
+  txt <- xmlValue(xmlRoot(x)[[1]])
+  txt <-gsub("#","?", txt)
+  f <- textConnection(toString(txt))   
+  d <- utils::read.table(file=f, sep=sep, header=T, dec=dec, na.strings=na.strings, fill=fill, quote=quote, fileEncoding = fileEncoding)
+  close(f)
+  d
+}
 # 21228 life history traits of moths
@@
 attr(moths, 'keep') <-  c(locationID = "plot")
-attr(moths, 'individuals') <- "label"
+attr(moths, 'occurrences') <- "label"
 attr(moths, 'units') <- c(weight = "mg", wing_length = "mm", wing_width = "mm", wing_area = "mm2", wing_loading = "mg mm-2")
+
+select_rows <- 1:12
 
@@
 
-kable(moths[1:45,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
-      booktabs = T, caption = caption) %>%
+kable(rbind(moths[select_rows, select_cols]), longtable=TRUE, escape = TRUE, row.names = FALSE,
+      booktabs = T, caption = caption, digits = c(NA,NA,NA,3, 2, 2,4,4)) %>%
 kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)
@@
 mothsStd <- mothsStd[order(as.numeric(substr(mothsStd$occurrenceID,2,10))),]
+names(mothsStd)[c(1,2,6,7)] <- c("verbatimScientificName", "verbatimTraitName", "scientificName", "traitName")
 
@@
 an occurrence, via identifiers."
-select_cols <- c("scientificName","traitName","traitValue","traitUnit","traitID","scientificNameStd","taxonID","occurrenceID", "measurementID", "locationID")
+select_cols <- c("scientificName","traitName","traitValue","traitUnit","traitID","verbatimTraitName","taxonID","occurrenceID", "measurementID", "locationID")
+select_rows <- which(mothsStd$traitName %in% c("body_weight", "wing_area", "wing_length") )
 
-kable(mothsStd[1:28,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
-      booktabs = T, caption = caption) %>%
+kable(mothsStd[select_rows[1:28],select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
+      booktabs = T, caption = caption, digits = c(4)) %>%
 kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

### 2019-03-28 manuscript/appendix_B.Rmd
update appendix B, add Ex 4, fix examples, fix layout issues

@@
 
-- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats but rather in software-independent file formats (like .xlsx or .odt), or comma- or tab-separated text files ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g.
-UTF-8 or ASCII).
-For internationalisation, use points as decimal delimiters and set date-formats to ISO 8601, i.e.
-"YYYY-MM-DD".
+- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats but rather in software-independent file formats (like .xlsx or .odt), or comma- or tab-separated text files ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g., UTF-8 or ASCII).
+For internationalisation, use points as decimal delimiters and set date-formats to ISO 8601, i.e., "YYYY-MM-DD".
 
@@
 - **Publish raw data!** When publishing trait-data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation.
-Thus, data should be published as they have been recorded, i.e.
-in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard.
-This is true for the core terms, i.e.
-the trait value and the item of observation and its taxon assignment, but also for any co-variates reported along with it.
+Thus, data should be published as they have been recorded, i.e., in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard.
+This is true for the core terms, i.e., the trait value and the item of observation and its taxon assignment, but also for any co-variates reported along with it.
 
-- **Keep original values!** For the core terms, data providers should complement user-defined entries (such as project specific taxon abbreviations) with standardised values by applying the '\*Std'-terms.
-For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g.
-elevation above sea-level.
+- **Keep original values!** For the core terms, data providers should provide standardised values conforming to consensus terms of the own research field.
+The ETS provides terms for maintaining user-defined entries (such as project specific taxon abbreviations) along with the standard terms (using the
+'verbatim\*' duplicates of the terms). 
+For co-variates, the extensions of the vocabulary provide column names for standardised values only, e.g., elevation above sea-level.
 These should adhere to accepted standards wherever possible.
-If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g.
-by appending '\*Orig' to the column name.
-Also, additional co-variates, e.g.
-on environmental growth conditions of the specimen, should be provided in user-defined columns.
+If any conversion was applied to obtain these data, users should keep the original values in user-defined columns, e.g., by appending 'verbatim\*' to the column name.
+Also, additional co-variates, e.g., on environmental growth conditions of the specimen, should be provided in user-defined columns.
 Any user-defined columns must be adequately defined in the accompanying meta-information.
 
-- **Report Uncertainty!** If publishing aggregate data (e.g.
-averages for multiple measurements), report the $n$ and dispersion of the aggregate value.
+- **Report Uncertainty!** If publishing aggregate data (e.g., averages for multiple measurements), report the $n$ and dispersion of the aggregate value.
 This allows to weigh data from multiple sources when producing compilations.
-Any sources of bias or noise should be reported along with the data (e.g.
-observer ID, conditions of measurement etc.).
+Any sources of bias or noise should be reported along with the data (e.g., observer ID, conditions of measurement etc.).
 Use terms of the ETS extension "Measurement or Fact".
@@
 Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric coordinates could be linked via a 'measurementID'. 
-In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
+In literature data, summarised traits are usually given at the taxon level instead of the individual organism (e.g., reported as means or factorials) and a 'taxonID' is the key identifier.
 In larger compilations, a 'datasetID' allow to trace data origin to the primary source and keep track of authorship and original reference. 
@@
 As the long-table format draws from a defined set of columns, merging datasets is easier.
-Long-table datasets also purport multiple advantages for data manipulation [e.g.
-filtering, sub-setting and aggregating data, @wickham14].
+Long-table datasets also purport multiple advantages for data manipulation [e.g., filtering, sub-setting and aggregating data, @wickham14].
 -->
@@
 In terms of analysis, it allows to easily correlate traits with each other and analyse trade-offs between traits.
-However, wide-tables need to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g.
-'999').
-Additional information  (e.g.
-on variation of means or literature source) may be stored in secondary columns or accompanying datasheets.
+However, wide-tables need to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g., '999').
+Additional information  (e.g., on variation of means or literature source) may be stored in secondary columns or accompanying datasheets.
 When applying the data standard to wide-table data, the columns for taxa or individuals may be labelled using the terms of the ETS. 
-Taxon concepts should be referenced to globally available concepts using URIs, e.g.
-by referring to GBIF Taxonomy.
+Taxon concepts should be referenced to globally available concepts using URIs, e.g., by referring to GBIF Taxonomy.
 Taxa authors and fine-grained higher taxonomy must then be consistent with the linked concept, or may be omitted as redundant and to avoid confusion.
@@
 
-In observation wide-tables, each row is centered around a single physical occurrence of a species, e.g.
-phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns (see Table B1).
+In observation wide-tables, each row is centered around a single physical occurrence of a species, e.g., phenotypic variation of traits within species would usually be recorded per observation where each replicate taken on a distinct individual specimen would be recorded in a row, with the respective trait values stored in columns (see Table B1).
 Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients.
@@
 Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier. 
-Long-table datasets purport multiple advantages for data handling (e.g.
-filtering, sub-setting and aggregating data), visualization (e.g.
-plot measured values by factor variable or taxon) and statistical modelling (e.g.
-ANOVA for testing difference of trait value by sex) [@wickham14].
+Long-table datasets purport multiple advantages for data handling (e.g., filtering, sub-setting and aggregating data), visualization (e.g., plot measured values by factor variable or taxon) and statistical modelling (e.g., ANOVA for testing difference of trait value by sex) [@wickham14].
 As long-table data share a larger set of columns, merging of datasets is in most cases easier for the end-user.
@@
 Data-splitting techniques or a database system are tools to disentangle the data.
+Note also, how the datasets become bulkier, as many information and labels are now repeated on all measurements as instances of the same individual organism (i.e., one occurrence).
 
@@
 # 21247 morphological traits of moths
+#https://www.bexis.uni-jena.de/Data/ShowXml.aspx?DatasetId=21247
 
@@
 caption <- "Example 2 before standardisation.
-Data are stored in observation wide-table format, i.e.
-one row contains a single specimen with multiple trait measurements (in columns).
+Data are stored in observation wide-table format, i.e., one row contains a single specimen with multiple trait measurements (in columns).
 Units and specific information on measurements must be reported in the metadata of the dataset.
-Data originate from Mangels et al.
-(unpublished)."
+Data extract from Mangels et al.
+(released within the data management system of the Biodiversity Exploratories, BExIS:21247)."
 select_cols <- c("label","species","plot","weight","wing_length","wing_width","wing_area","wing_loading")
@@
 caption <- "Example 2 after standardisation into measurement long-table format.
-Each row contains a single measurement linked to an observation of an individual specimen, i.e.
-an occurrence, via identifiers."
+Each row contains a single measurement linked to an observation of an individual specimen, i.e., an occurrence, via identifiers.
+Only three traits have been selected from the original data, and rows have been cut off to fit the page."
 select_cols <- c("scientificName","traitName","traitValue","traitUnit","traitID","verbatimTraitName","taxonID","occurrenceID", "measurementID", "locationID")
@@
 
-If data contain covariate information, the two-dimensional data table can cover these by adding columns defined by the extensions of the ETS. 
-The three extensions complement the core terms according to different layers of metadata information: 1) the "taxon" extension includes terms to describe more details on the taxon concept applied, 2) the "occurrence" extension provides terms to describe the indivdual specimen and its observation context, and 3) the "Measurement or Fact" extension adds information on the methodology or context of the measurement of the trait value.
+If data contain covariate information, the two-dimensional data table can cover these by adding more columns.
+The three extensions of the ETS complement the core terms according to different layers of metadata information: 1) the "taxon" extension includes terms to describe more details on the taxon concept applied, 2) the "Measurement or Fact" extension adds information on the methodology or context of the measurement of the trait value, and 3) the "occurrence" extension provides terms to describe the individual specimen and its observation context.
 
-Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
-As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16.
-The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
-Although we cannot cover the structure of relational databases in this tutorial, the standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.
+A data table for instance should report on the taxonomic resolution of the trait observation, especially if it is reported on something else than the species level (by setting `taxonRank`: genus).
+The additional terms for taxon hierarchies (`kingdom` to `genus`) would allow the data user to extract information on genera or families, but this would be redundant with an unambiguous taxon concept has being provided within `taxonID`.
+Further columns would provide detail on the measurement accuracy (`measurementResolution`) or the number of individuals on which an aggregated measure has been determined (by reporting `individualCount`: *n* and `aggregateMeasure`: 'TRUE').
+Data providers can easily contain information on sex or life stage of the specimen (using terms `sex` or `lifeStage`), or on the location of sampling (e.g., `verbatimLocality`, all 'occurrence' terms).
+
+If data are provided in a single two-dimensional data table that applies standard terms of the ETS, this may increase readability and facilitate re-use by synthesis researchers as they do not need to process information stored elsewhere.
+Data tables conforming to the same data standard can be merged with ease.
 
@@
 
-kable(heteroptera_raw[1:28,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
+select_rows <- which(heteroptera_raw$ID %in% 8:15)
+
+kable(heteroptera_raw[select_rows,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
       booktabs = T, caption = caption) %>%
@@
 ```
+
+\newpage
 
@@
 
+select_rows <- which(heteroptera_raw$ID %in% 8:15)
+
 select_cols <- c(1,2,3,4,8,6,15,18,22,16,20,21)
 kable(heteroptera_long[heteroptera_long$traitName %in% c("Body_length", "Body_width", "Body_height", "Thorax_width"),select_cols][30:58,], longtable=TRUE, escape = TRUE, row.names = FALSE,
+      booktabs = T, caption = caption) %>%
+kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)
+
+```
+
+\elandscape
+
+# Example 4 - apply ETS extensions in relational databse
+
+Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain.
+As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational database structures may be indicated.
+Relational databases help to avoid content duplication, as the core table will only maintain identifiers, that relate to secondary data tables containing multiple specifications on the identifier level.
+For instance, the core table may specify 'occurrenceID' for which specific information can be found in a lookup table. 
+The ETS may be applied to describe columns within the individual data tables of relational databases.
+Compatibility with other database structures is warranted, as ETS terms map to terms of the Darwin Core Standard used in other suggested database structures for trait data, like the generic structure for plant trait databases [@kattge11a] or the TraitBank structure [@parr16].
+
+For publishing relational databases, it is however important to provide sufficient information on how the individual data tables are related, e.g., in a machine readable metadata specification (e.g., in EML language).
+Further, it must be ensured that data tables are always contained in a single data publication and are not distributed independently of each other, as this would cause issues of broken links if data are updated or lost.
+Both requirements are met if data are released as a packaged archive, for instance following the Darwin Core Archive specification [@robertson09].
+
+\newpage
+
+```{r, echo = FALSE, warning = FALSE, message=FALSE}
+pulldata('heteroptera_raw')
+
+heteroptera_long <- as.traitdata(heteroptera_raw) 
+heteroptera_long <- standardize.taxonomy(heteroptera_long)
+heteroptera_long <- heteroptera_long[order(as.numeric(heteroptera_long$occurrenceID)),]
+heteroptera_long$basisOfRecordDescription <- as.factor(heteroptera_long$basisOfRecordDescription)
+
+heteroptera_long$decimalLatitude <-  gsub("°", " ", substr(heteroptera_long$verbatimLocality,0,5))
+heteroptera_long$decimalLongitute <-  gsub("°", " ", substr(heteroptera_long$verbatimLocality,10,14))
+
+heteroptera_long$decimalLatitude <- round(as.numeric(measurements::conv_unit(heteroptera_long$decimalLatitude, from = 'deg_dec_min', to = 'dec_deg')),5)
+heteroptera_long$decimalLongitute <- round(as.numeric(measurements::conv_unit(heteroptera_long$decimalLongitute, from = 'deg_dec_min', to = 'dec_deg')), 5)
+heteroptera_long$basisOfRecord <- as.factor("PreservedSpecimen")
+
+caption = "Example 4: original data of Example 3 after standardisation into relational database structure, providing taxon and occurrence level information in secondary data tables (Fig.
+8 and 9, respectively)."
+
+select_rows <- which(heteroptera_raw$ID %in% 8:15)
+
+select_cols <- c(1,2,3,4,6,8)
+kable(heteroptera_long[heteroptera_long$traitName %in% c("Body_length", "Body_width", "Body_height", "Thorax_width"),select_cols][30:60,], longtable=TRUE, escape = TRUE, row.names = FALSE,
+      booktabs = T, caption = caption) %>%
+kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)
+
+```
+
+\blandscape
+
+```{r, echo = FALSE, warning = FALSE, message=FALSE}
+taxa <- attributes(heteroptera_long)$taxonomy
+names(taxa)[c(1,3,4)]<- c("verbatimScientificName",  "scientificName", "scientificNameAuthorship") 
+select_cols <- c(14,3,4,5,7,8,9,10,11,12,13)
+select_rows <- taxa$taxonID %in% unique(heteroptera_long$taxonID[heteroptera_long$occurrenceID %in% 8:15])
+
+caption = "Example 4, data table on taxon level information after standardisation. Data extracted from GBIF backbone taxonomy."
+
+kable(taxa[select_rows,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
+      booktabs = T, caption = caption) %>%
+kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)
+
+```
+
+```{r, echo = FALSE, warning = FALSE, message=FALSE}
+
+caption = "Example 4, data table on occurrence level information after standardisation, accompanying Table 7."
+
+select_rows <- c(155, 177, 199, 221, 243, 265, 287, 309)
+select_cols <- c(8,18,22,16,17,20,21)
+kable(heteroptera_long[select_rows,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
       booktabs = T, caption = caption) %>%

### 2019-03-28 manuscript/manuscript.Rmd
final edits for resubmission of manuscript

@@
 4.
-The use of a standard vocabulary allows trait datasets from heterogeneous sources to be aggregated more easily into compilations and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
+The use of a standard vocabulary allows for trait datasets from heterogeneous sources to be aggregated more easily into compilations and facilitates the creation of interfaces between software tools for trait-data handling and analysis.
 By aiding decentralised trait-data standardisation, our vocabulary may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.
@@
 The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, information of time and date of observation, and methodological details.
-A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12]. The terminologies provided by these standards are quite universal and cover most use cases of trait data. 
+A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
+Of particular importance for trait data is the 'MeasurementOrFact' extension, which typically would be used in database management and bioinformatics to structure trait observations [@parr16].
 <!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
@@
 The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research.
-For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and ensure compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
+For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and ensure full compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
 This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
@@
 a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers.
-Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e., occurrence) or the origin or confidence on the reported measurement or fact.}
+Additional identifiers relate each row to other layers of information on d) the taxon resolution, the individual organism (i.e.
+occurrence), or the origin or confidence on the reported measurement or fact.}
 
@@
 <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)-->
-- The `Occurrence` extension contains vocabulary to describe information on the observation context of individual specimens, such as sex, life stage or age.
+- The `Occurrence` extension contains vocabulary to describe information on the observation context of individual organisms, such as sex, life stage or age.
 This also includes the method of sampling and preservation, as well as the date and geographical location, which provide an important resource to analyse trait variation due to differences in space and time.
@@
 These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content.
-A unique identifier links to these other datasheets, encoding single measurements or reported facts (`measurementID`), individual occurrences of a species (`occurrenceID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+A unique identifier links to these other datasheets, encoding single measurements or reported facts (`measurementID`), individual organisms of a species (`occurrenceID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).
+
+The concept of 'occurrence' is prone to cause confusion.
+By definition of DwC it is "An existence of an Organism at a particular place at a particular time", thus any individual observed twice would have two distinct 'occurrences'.
+In most ecological contexts, sampling of an individual is only performed once, resulting in any occurrence being semantically identical with the individual organism (i.e., the DwC term 'organism').
+Also, as 'occurrence' is strictly defined by a date-time event, it may mean the same as the common-sense concept of 'observation'. 
 Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g., a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
+As occurrence is the narrower term and the key concept for linking an individual organism to a place and time in DwC, and since it is indeed relevant to distinguish between multiple 'occurrences' of the same organism in some trait-based research applications, the ETS sticks to this terminology.
 
@@
 In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset , e.g., by entering "Schneider, F.D., Jochum, M., Le Provost, G., Penone, C., Ostrowski, A.
-and Simons, N.K., 2019, Ecological Traitdata Standard Vocabulary v0.10, DOI:  10.5281/zenodo.1441132, URL: https://terminologies.gfbio.org/terms/ets/pages/" in the metadata field `conformsTo`.
+and Simons, N.K., 2019, Ecological Traitdata Standard Vocabulary v0.10, DOI:  10.5281/zenodo.2605377, URL: https://terminologies.gfbio.org/terms/ets/pages/" in the metadata field `conformsTo`.
 Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their individual URIs.
@@
 
-To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, that will be of use for the exchange and handling of trait data in the context of the semantic web.
+To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, and for the exchange and handling of trait data in the context of the semantic web.
 
@@
 
-Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary.
+Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal and two anonymous reviewers for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary.
 We are grateful to the organisers and participants of the Open Traits workshop in New Orleans, USA, in August 2018.

### 2019-03-29 manuscript/manuscript.Rmd
update manuscript final go for clarity, reduction of references to match word limit

@@
 Functional traits are phenotypic (i.e., morphological, physiological, behavioural) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
-Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10].
+Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @bello10].
 This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
@@
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
-Using traits is also a promising means of bypassing taxonomic impediment, i.e., the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment.
-This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime06; @moretti17; @villeger17].
+Because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime06; @moretti17; @villeger17], their measurement is also a promising means of bypassing taxonomic impediment, i.e., the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment.
 
@@
 <!-- 2. trait databases have been developed, with standard structure and terms --> 
-Despite the importance of functional trait data, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time.
+Despite the importance of trait-based approaches, fully exploiting their potential relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time.
 However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
@@
 <!-- 4. distributed trait data must be integrated --> 
-Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15].
+Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [ @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @allison15a].
+<!--@emerson15; @swan12;>
 As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services, where no data standards are enforced upon the uploaded material [@wilkinson16].
@@
 
-In this paper, we review existing trait databases and online portals, and initiatives for data standardisation from the pragmatic view of data providers, data curators and data users, as well as database managers.
+In this paper, we review existing initiatives for trait-data collection and tandardisation from the pragmatic view of data providers, data curators and data users, as well as database managers.
 We discuss current efforts to make trait data visible, accessible, interoperable and re-useable in downstream data analysis, as demanded by the FAIR guiding principles for scientific data [@wilkinson16].
@@
 Identifier (ID) | A unique label that relates data entries to information within and across *datasets* or external items of information; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*.
-Metadata | Data documentation of the higher-level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
+Metadata | Data documentation of the higher-level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener06].
 In the context of trait-data, such additional information can move to the body of the primary data table when data are compiled from different sources.
@@
 1. Initiatives that provide trait *datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature.
-2. Initiatives that aim to harmonise trait data from the literature or from direct measurements into database infrastructures and make those data widely available on the Internet. 
+2.
+Initiatives that aim to harmonise trait data from the literature or from direct measurements into data compilations or database infrastructures and make those data widely available on the Internet.
 3. Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits and provide standard *terminologies*.
-4. Initiatives that aim to combine data (1 & 2) and terminologies (3) to link trait data to a wider set of biodiversity data.
+4. Initiatives that aim to combine data (1 & 2) and terminologies (3) into formalized structures for knowledge representation to link trait data to a wider set of biodiversity data.
 
@@
 The resolution or scope of these co-variates varies greatly depending on the research question and observation context.
-The column descriptions and terminology applied to taxa and traits are mostly project-specific and rarely chosen to allow translation into larger database initiatives. 
+The column descriptions and terminology applied to taxa and traits are mostly project-specific and rarely chosen to for compatibility with larger database initiatives. 
 Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. 
-Furthermore, the metadata provided along with the primary data vary in their level of detail, e.g., for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
-While in some primary data publications information like geolocation or sampling date and time might be dataset-level information, thus qualifying as metadata, in other datasets they might be collected on a level of individual observations (see section on data compilations below).
+Furthermore, metadata provided along with the primary data vary in their level of detail, e.g., for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
+While, in some datasets, information like geolocation or sampling date and time might be dataset-level information, thus qualifying as metadata, in other datasets they might be collected on a level of individual observations (see section on data compilations below).
 More importantly, clear statements on ownership and authorship, terms of use, or internationalisation (e.g., separators and delimiters), are often still neglected in primary trait-data publications. 
@@
 In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g., @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
-While these initiatives address issues of heterogeneity in units or categorical variables, or aim for high taxonomic or geographic coverage, few of these compilations apply a standardised terminology for taxa or trait definitions.
-Additionally, documentation of metadata and methodology in the original datasets often differs in the level of detail, depending on the research focus.
-Particularly, in the process of data aggregation, rich metadata content might be lost or the reference to the original dataset might be obscured, as only aggregated values are reported (e.g., means or medians).
-Such trait-data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiversity informatics.
+While these initiatives successfully address issues of heterogeneity in units or categorical variables, or achieve high taxonomic or geographic coverage, few of these compilations apply a standardised terminology for taxa or trait definitions.
+Additionally, in the process of data aggregation, rich metadata content might be lost, as the detail in the original files differs, while the reference to the original dataset becomes obscured, as only aggregated values are reported (e.g., means or medians).
+Such trait-data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict data-management sense.
 Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.
@@
 For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform.
-The online portal enables selective data download and user permission and right management.<!--As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
+The online portal enables selective data download and management of user permissions.<!--As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).
 -->
@@
 <!--The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].-->
-
-Nonetheless, initiatives for particular groups of animals do exist: examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), the Coral Trait Database [@madin16], or the Global Ants Database [@parr17, see Supplementary Material A, Table A1].
+Nonetheless, initiatives for particular groups of animals do exist.
+Examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), the Coral Trait Database [@madin16], or the Global Ants Database [@parr17, see Supplementary Material A, Table A1].
 The role of online portals and database initiatives in standardising data and making them more accessible is paramount. 
@@
 Previous standard definitions of trait *concepts* range from listings of selected definitions in vocabularies, over well-defined method handbooks and comprehensive *thesauri*, to formalized definitions of trait concepts in *ontologies*.
-The initiatives behind method handbooks, thesauri and ontologies are primarily concerned with building community consensus for trait definitions.
+The initiatives behind method handbooks, thesauri and ontologies are essential for building community consensus for trait definitions.
 
 Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13; @kissling18] aiming for a list of functional indicators for ecosystem health.
-Assigning a more detailed and unambiguous methodological protocol for a trait, including the units to use or the ordinal or factor levels to be assigned, is essential for standardising its measurement process.
+
+Assigning a detailed and unambiguous methodological protocol for a trait, including the units to use or the ordinal or factor levels to be assigned, is essential for standardising its measurement process.
 Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but are of limited use in harmonising trait data that pre-date or ignore this standard [@kattge11a].
@@
 To provide a logic structure for trait terms, Garnier et al.
-(2017) suggest the Entity-Quality model (EQ), where a trait is defined as ‘an entity having a quality’ (for instance for trait ‘femur length’, ‘femur’ is the entity and ‘length’ the quality).
+(2017) suggest the Entity-Quality model (EQ), where a trait is defined as 'an entity having a quality' (for instance for trait 'femur length', 'femur' is the entity and 'length' the quality).
 In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms.
@@
 For example, a measurement of fruit mass could be linked to the definition of the term within the Thesaurus of Plant characteristics [TOP, @garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
+
 In addition to defining terms for human interpretation, *ontologies* define terms by their relationship to other defined terms, thereby providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18].
- The Plant Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
-Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12].
+The Plant Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)".
+Thus, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12].
 By providing ontologies in a formalized syntax, like Web Ontology Language (OWL), a machine-readable web of definitions is spun across the Internet allowing researchers and search engines to relate independent trait measurements with each other and connect them to the wider *semantic web* of online data [@berners-lee01; @page08; @williams06; @walls12].
@@
 The UBERON ontology is an integrated cross-species anatomy ontology for all animals, which combines concepts from different existing ontologies, with wide application in biomedical or physiological research [@mungall12].
-The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
 <!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g., units or legit factor levels) of measured values or reported facts.
@@
 To conclude, there is already a suite of globally available thesauri and ontologies for traits. 
-However, definitions in some domains are better covered than others [@kissling18] and the interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
+However, definitions in some domains are better covered than others [@kissling18], and different curation strategies and measures for peer-review and community building are employed. 
 To this end, the OBO Foundry is providing a development platform for (biological) ontologies and offers review and quality control [@smith07, http://www.obofoundry.org/]. 
@@
 
-While trait thesauri and trait ontologies typically define concepts of measurements and observations for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
+While trait thesauri and trait ontologies typically define concepts of measurements and observations for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and labelled.
 
@@
 Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the *Darwin Core Standard* for biodiversity data [DwC; @wieczorek12].
-These frameworks specify terms and classes to describe any kind of measurement and occurrence of organisms.
-The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, information of time and date of observation, and methodological details.
+Both DwC and ABCD are ratified standards of the Biodiversity Information Standards (TDWG, www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.
+These terms may be used for defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, information of time and date of observation, and methodological details [@robertson09].
 A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
@@
 <!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
-Both DwC and ABCD are ratified standards of the Biodiversity Information Standards (TDWG, www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.
 
@@
 
-Based on these terms, the Extensible Observation Ontology [OBOE, @madin07; @schildhauer16] formalizes observations and measurements of traits into a machine-readable ontology, thus being easily integrated into larger database management systems.
+While the abovementioned standards provide terms and concept definitions, and the logic relationships of those, they do not prescribe explicit structure for trait-data. 
+Based on the terms of DwC, the Extensible Observation Ontology [OBOE, @madin07; @schildhauer16] formalizes observations and measurements into a machine-readable ontology, thus being easily integrated into larger database management systems.
 By applying this scheme for plant traits, @kattge11a propose a generic database structure that covers most potential use cases of trait-based ecology.
 This data structure is built around a central data table that contains observations of individual plants linked to several measurements of traits via identifiers.
-The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment (which may have been extracted from metadata).
+The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment.
 @kissling18 discuss different ontologies (including OBOE) that formalize the structure of observation data and attest that for the use cases of trait data these ontologies are still difficult to integrate.
 
-While the abovementioned standards provide terms and concept definitions, and the logic relationships of those, they do not prescribe explicit structure for trait-data. 
-In this regard, the Encyclopedia of Life (EOL) has proposed TraitBank [@parr16] as a standard structure for uploading data on physiological and life-history traits of all kingdoms of life.
+The Encyclopedia of Life (EOL) has proposed TraitBank [@parr16] as a standard structure for uploading data on physiological and life-history traits of all kingdoms of life.
 It is to date the most general approach of an integrated structure for trait data.
@@
 Additional layers of information cover bibliographic references, multimedia archives and ecological interactions. 
-TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], which is a set of simple text files (csv) and a metadata description using EML (the correct specification of which can be validated before upload on https://tools.gbif.org/dwca-validator/).
+TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A,  @gbif17], which is a set of simple text files (csv), a file to specify relationships between these text files (called meta.xml), and a file for metadata descriptions using EML [called, EML.xml, see @gbif17 for specifications, archives can be validated before upload on https://tools.gbif.org/dwca-validator/].
 
-All of these structures make  use of stable URIs to refer to taxon concepts.
+All of these structures suggest the use of stable URIs to refer to taxon concepts.
 The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review [@franz16].
 Initiatives that aim at providing a stable reference while tracking the changing taxon concepts are for instance the Catalogue of Life [http://www.catalogueoflife.org/, @roskov18] or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).
-The GBIF Backbone Taxonomy [@gbifsecretariat17] collects and bundles these terminologies into a single reference framework.
+The GBIF Backbone Taxonomy [@gbifsecretariat17] collects and bundles existing terminologies into a single reference framework.
 
@@
 In sum, we attest to a gap between the trait-data structures developed for data curators and managers and the data input produced by data providers.
-Hardly any of the aforementioned standalone or aggregated trait datasets for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards.
+Hardly any of the aforementioned standalone or aggregated trait datasets for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or data standards.
 As it stands, re-using these data in larger compilations or integrating them into structured database initiatives is error-prone and labour-intensive and the potential for a broad synthesis is diminished.
@@
 In the eyes of the data-provider, in most cases, any co-variates can be appended as extra columns to the dataset.
-The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management.
+The other reason is lack of awareness of the need for trait-data standardisation among data providers, who are not trained in the demands of biodiversity data-management.
 In addition, complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects, especially since funders often do not require these efforts to be included into proposals.
 
-By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14;e.g.,  Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g., KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, http://diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/) are likely to gain importance.
+By filling this gap, data-brokering services [the German Federation for Biological Data; gfbio.org, @diepenbroek14; e.g.,  Data Observation Network for Earth, DataONE, @michener11] or data management systems for scientific projects (e.g., KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, http://diversityworkbench.net; BExIS2, http://bexis2.uni-jena.de/) are likely to gain importance.
 These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures.
@@
 
-Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards. 
 A more widespread implementation of ontologies would advance the possibilities to integrate datasets and reduce noise and uncertainty when aggregating data. 
@@
 Third, frameworks for providing trait data in an unambiguous and machine-readable structure must be simplified to match the limited resources of small and intermediate research projects.
-This can be achieved by extending documentation or providing tools for the application of existing ontology frameworks and database structures, or by defining easy-to-use standard vocabularies that enable the interoperability of data at minimal effort.
+This can be achieved by extending documentation or providing tools for the application of existing ontology frameworks and database structures (e.g.
+data validator services), and by defining easy-to-use standard vocabularies that enable the interoperability of data at minimal effort.
 <!--
@@
 As a response to the challenges outlined above, we propose a versatile standard vocabulary for trait-based ecological research. 
-The Ecological Trait-data Standard Vocabulary (ETS) is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different degrees of measurement details.
+The Ecological Trait-data Standard Vocabulary (ETS) is accessible at https://terminologies.gfbio.org/terms/ets/pages/ and combines terms of DwC with newly defined terms to cover the variety of trait-based approaches and their different needs to report measurement detail.
 Rather than prescribing a data structure or exchange format, the vocabulary is intended as a more inclusive terminology that can be used in three major use cases:
@@
 
-All terms may be applied to describe columns of a data table (Fig. 2; see Supplementary Material B for best practice principles and examples for publishing primary data). 
-By applying these standard terms, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable.
-It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives, data synthesis and ensure long-term accessibility.
+All terms may be applied to describe columns of a data table (Fig. 2; see Supplementary Material B for best-practice principles and examples for publishing primary data). 
+By applying these standard terms, data providers can ensure that the description of trait measurements uploaded into public data repositories will be unambiguous.
+It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives and data synthesis, while warranting long-term accessibility.
 
@@
 Our vocabulary offers three extensions to contain additional information on the context of the observation along with the core data in analogy to DwC extensions ("Taxon", "Measurement or Fact", and "Occurrence"; see section on extensions below).
-Further terms are provided for dealing with typical dataset-level information on authorship and rights of re-use of the data (based on terms of Dublin Core Metadata Initiative, DCMI), as well as for defining trait concepts (see section on metadata below).
+Further terms are provided for dealing with typical dataset-level information on authorship and rights of re-use of the data (based on terms of Dublin Core Metadata Initiative, DCMI), as well as for defining own trait concepts (see section on metadata below).
 Aspects not covered by the vocabulary may draw from terms provided by other existing terminologies (in particular DCMI and DwC and its extensions), or be added as user-defined columns (which should then be clearly specified in the metadata-information accompanying the dataset).
@@
 
-In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users and curators), and biodiversity informatics researchers (data managers).
+In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
 The aim was to develop a simple, easy-to-use template for standalone trait data publications or data compilations, to facilitate their re-use for synthesis and integration into larger database structures.
-Earlier proposals for trait-data standards [e.g.,  @kattge11a; @parr16] have been designed for more complex database structures or from a data manager perspective, which may be the reason why they have so far not been adopted for primary data publications.
-We paid particular attention to these existing data standards [e.g., @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our vocabulary with major trait databases and existing concepts for biodiversity data management.
+Earlier proposals for trait-data standards [e.g.,  @kattge11a; @parr16] have been designed for relational database structures from a data manager perspective, which may be the reason why they have so far hardly been adopted for primary data publications.
+We paid particular attention to these existing data standards [e.g., @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to maximize compatibility.
 
@@
 The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research.
-For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and ensure full compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
+For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and work towards full compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org).
 This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
@@
 \centering
-\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
+\includegraphics[width=1\textwidth, angle=0]{figures/fig2.pdf}
 \caption{Formats used for trait datasets.
-a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers.
+a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well-defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers.
 Additional identifiers relate each row to other layers of information on d) the taxon resolution, the individual organism (i.e.
@@
 The concept of 'occurrence' is prone to cause confusion.
-By definition of DwC it is "An existence of an Organism at a particular place at a particular time", thus any individual observed twice would have two distinct 'occurrences'.
-In most ecological contexts, sampling of an individual is only performed once, resulting in any occurrence being semantically identical with the individual organism (i.e., the DwC term 'organism').
-Also, as 'occurrence' is strictly defined by a date-time event, it may mean the same as the common-sense concept of 'observation'. 
-Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g., a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
-As occurrence is the narrower term and the key concept for linking an individual organism to a place and time in DwC, and since it is indeed relevant to distinguish between multiple 'occurrences' of the same organism in some trait-based research applications, the ETS sticks to this terminology.
+By definition of DwC it is "An existence of an Organism at a particular place at a particular time".
+Thus any individual observed twice would have two distinct 'occurrences'.
+If sampling of an individual is only performed once, this results in any occurrence being semantically identical with the individual organism (i.e., the DwC term 'organism').
+Also, as 'occurrence' is strictly defined by a date-time event, it may be identical the common-sense concept of 'observation'. 
+Some data-types directly refer to existing global identifiers for occurrence IDs, e.g., a GBIF URI or a stable identifier references the precise specimen at a particular place and time from which the measurement was taken [@groom17; @guntsch17].
+As occurrence is the narrower term and the key concept for linking an individual organism to a location and sampling event in DwC, and since it is indeed relevant to distinguish between multiple 'occurrences' of the same organism in some trait-based research applications, the ETS sticks to this terminology.
 
@@
 
-We thank the past and present scientific coordinators, local managers and the data manager of the Biodiversity Exploratories program for their work, and Markus Fischer, Eduard Linsenmair, Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot, Ernst-Detlef Schulze, Wolfgang W.
+We thank the past and present scientific coordinators, local managers and data managers of the Biodiversity Exploratories program for their work, and Markus Fischer, Eduard Linsenmair, Dominik Hessenmöller, Daniel Prati, Ingo Schöning, François Buscot, Ernst-Detlef Schulze, Wolfgang W.
 Weisser and the late Elisabeth Kalko for their role in setting up the Biodiversity Exploratories program.

### 2019-03-29 manuscript/manuscript.Rmd
fix layout and line breaks

@@
 Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [ @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @allison15a].
-<!--@emerson15; @swan12;>
+<!--@emerson15; @swan12;-->
 As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services, where no data standards are enforced upon the uploaded material [@wilkinson16].
@@
 -->
-
 Most importantly, these platforms enable public hosting of data with very low quality-thresholds regarding *metadata* documentation and data standardisation.
